### 2013-04-01 sbreview.tex
Initial commit

%For arXiv, uncomment this block and comment the next block.
\documentclass{amsart}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}
\usepackage[round]{natbib}
\newcommand{\forarxiv}[1]{#1}
\newcommand{\notforarxiv}[1]{}

% % For Sys Bio, uncomment this block and comment the previous block.
% \documentclass[12pt,letterpaper]{article}
% \usepackage{fixltx2e}
% \usepackage{textcomp}
% \usepackage{fullpage}
% \usepackage{amsfonts}
% \usepackage{verbatim}
% \usepackage[english]{babel}
% \usepackage{pifont}
% \usepackage{color}
% \usepackage{setspace}
% \usepackage{lscape}
% \usepackage{indentfirst}
% \usepackage[normalem]{ulem}
% \usepackage{booktabs}
% %\usepackage{nag}
% \usepackage{natbib}
% %\usepackage{bibtex}
% \usepackage{float}
% \usepackage{latexsym}
% %\usepackage{hyperref}
% \usepackage{url}
% %\usepackage{html}
% \usepackage{hyperref}
% \usepackage{epsfig}
% \usepackage{graphicx}
% \usepackage{amssymb}
% \usepackage{amsmath}
% \usepackage{bm}
% \usepackage{array}
% %\usepackage{mhchem}
% \usepackage{ifthen}
% \usepackage{caption}
% \usepackage{hyperref}
% %\usepackage{xcolor}
% \usepackage{amsthm}
% \usepackage{amstext}
% \linespread{1.66}
% \raggedright
% \setlength{\parindent}{0.5in}
% \setcounter{secnumdepth}{0}
% \pagestyle{empty}
% \renewcommand{\section}[1]{%
% \bigskip
% \begin{center}
% \begin{Large}
% \normalfont\scshape #1
% \medskip
% \end{Large}
% \end{center}}
% \renewcommand{\subsection}[1]{%
% \bigskip
% \begin{center}
% \begin{large}
% \normalfont\itshape #1
% \end{large}
% \end{center}}
% \renewcommand{\subsubsection}[1]{%
% \vspace{2ex}
% \noindent
% \textit{#1.}---}
% \renewcommand{\tableofcontents}{}
% \bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
% \newcommand{\forarxiv}[1]{}
% \newcommand{\notforarxiv}[1]{#1}

% show keys for eqs, etc.
% \usepackage[notref,notcite]{showkeys}

% code
\newcommand{\program}{\textsf{program}}

% theorems, etc
\newtheorem{lem}{Lemma}
\newtheorem{cor}{Corollary}
\newtheorem{prop}{Proposition}
\newtheorem{thm}{Theorem}
\newtheorem{prob}{Problem}
\newtheorem{defn}{Definition}
\newtheorem{obs}{Observation}
\newtheorem{alg}{Algorithm}

\newcommand{\eat}[1]{}

\newcommand{\FIGmassTransport}{\
\begin{figure}[ht]
\begin{center}
  \forarxiv{\includegraphics[width=13cm]{mass_transport.pdf}}
\end{center}
\caption{\
  Caption goes here.
}
\label{FIGmassTransport}
\end{figure}
}
\newcommand{\refFIGmassTransport}{1}

\begin{document}

\notforarxiv{
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: PHYLOGENETICS AND THE HUMAN MICROBIOME
\bigskip
\medskip
\begin{center}

% Insert your title:
\noindent{\Large \bf Phylogenetics and the human microbiome}
\bigskip

\noindent {\normalsize \sc
Frederick A. Matsen IV$^1$}\\
\noindent {\small \it
$^1$
Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA, 91802, USA}\\
\end{center}
\medskip
\noindent{\bf Corresponding author:} Frederick A Matsen, Program in Computational Biology, Fred Hutchinson Cancer Research Center, Seattle, WA, 91802, USA; E-mail: matsen@fhcrc.org.\\
\vspace{1in}
}

\forarxiv{\
\title{Phylogenetics and the human microbiome}
\author{Frederick A. Matsen IV}
\date{\today}
\begin{abstract}
}
\notforarxiv{
\subsubsection{Abstract}
}

The human microbiome is the collection of cells that live inside and on humans.
The Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortium have advanced human microboime research recently by doing lots of sequencing, as well as bioinformatics and statistical development.
In this paper I review the impact of phylogenetics and tree-thinking on the methods used in the analysis of this data, as well as describing current challenges for phylogenetics coming from this type of work.

\forarxiv{
\end{abstract}
\maketitle

\section{Introduction}
}

\notforarxiv{
\noindent (Keywords: human microbiome; microbial ecology; zzz)\\
\vspace{1.5in}
}

\section{Introduction to the human microbiome}

% The HMP planned to comprehensively characterize the taxonomic
% composition of the microbiome by averaging 5,000 454 FLX 16S rRNA gene
% sequences from all 300 subjects, 18 body sites, and multiple time
% points. This design, combined with more than a 1,000-fold increase in
% sequencing throughput over the course of the HMP, forced the
% consortium to develop novel tools for processing large 16S rRNA gene
% datasets, tackling issues specific to 454 sequence data quality, and
% addressing novel biological questions that were previously
% inaccessible due to limited sample sizes.

Now we will embark on our journey.

There have been hundreds of papers concerning the human microbiome. It
is not our desire to survey all of these. Rather, our idea is to
highlight areas that may be of particular interest for the
phylogenetics community by potentially having wider applications or by
being areas for future exploration.

Please note that many of the methods and issues described in this paper will be equally useful for microbial ecology in general, but by narrowing our scope we were able to keep our paper within reasonable size limits.

Antibiotics.
\cite{dethlefsen2008pervasive,dethlefsen2011incomplete,jakobsson2010short,jernberg2007long}

\section{Tree-thinking in human microbiome research}

\subsection{phylogenetics and taxonomy}

Without a doubt, the greatest impact that phylogenetics has had on microbial ecology research is on changing our view of the taxonomic relationships between microorganisms.
In the absence of sequencing, microbiologist have to rely on microscopes and staining.
Lots of little things look alike.

The most stunning example of this is the discovery of the archaea by Woese.

However, there have been

There are projects that use phylogeny to revise taxonomy, short of PhyloCode.
GreenGenes.
The reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.

Much of this work has been done using the 16s ribosomal rRNA gene.

With the advent of high throughput sequencing, there has been a lot of interest concerning which variable region to use.

\subsection{Diversity estimates using phylogenetics}

Variants of phylogenetic diversity and their use.

\subsection{Community comparison using phylogenetics}

UniFrac distance and variants.

Figure: example principal coordinates plot.

\section{Phylogenetic inference as practiced by human microbiome researchers}

Very large data sets.

Some groups are not interested in accuracy per se.

Alignment.

Networks not considered.

Species delineation problem not explicitly considered. Clustering.

Chimeric sequences are a real problem, and similar to recombination.

\section{Phylogenetic challenges for human microbiome research}

\subsection{The challenge that lies ahead: functional genes and phylogenomics}

There has been a shift from interest in community dynamics to fine-scale interactions between host and immune system.

\cite{wylie2012sequence}
\cite{chen2012associating}

\section{Discussion}

What we can expect next.

\notforarxiv{
\newpage
\section{Figure Legends}
%\FIGmassTransport
\clearpage

\newpage
}

\bibliographystyle{plainnat}
\bibliography{sbreview}
\end{document}

### 2013-04-01 trash.tex
Initial commit

\subsection{Software pipelines}
\cite{caporaso2010qiime}
\cite{schloss2009introducing}
\cite{lynch2013axiome}

### 2013-04-01 sbreview.tex
something like an outline

@@
 The human microbiome is the collection of cells that live inside and on humans.
-The Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortium have advanced human microboime research recently by doing lots of sequencing, as well as bioinformatics and statistical development.
+The Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortia have advanced human microboime research by generating thousands of 16s surveys and terabases of metagenomic data, as well as funding bioinformatics and statistical development.
 In this paper I review the impact of phylogenetics and tree-thinking on the methods used in the analysis of this data, as well as describing current challenges for phylogenetics coming from this type of work.
@@
 
-% The HMP planned to comprehensively characterize the taxonomic
-% composition of the microbiome by averaging 5,000 454 FLX 16S rRNA gene
-% sequences from all 300 subjects, 18 body sites, and multiple time
-% points. This design, combined with more than a 1,000-fold increase in
-% sequencing throughput over the course of the HMP, forced the
-% consortium to develop novel tools for processing large 16S rRNA gene
-% datasets, tackling issues specific to 454 sequence data quality, and
-% addressing novel biological questions that were previously
-% inaccessible due to limited sample sizes.
+Starter paragraph about the scope of human microbiome research.
 
-Now we will embark on our journey.
-
-There have been hundreds of papers concerning the human microbiome. It
-is not our desire to survey all of these. Rather, our idea is to
-highlight areas that may be of particular interest for the
-phylogenetics community by potentially having wider applications or by
-being areas for future exploration.
-
+We will begin by introducing the results of recent research.
+There have been hundreds of papers concerning the human microbiome.
+It is not our desire to survey all of these.
+Rather, our idea is to highlight areas that may be of particular interest for the phylogenetics community by potentially having wider applications or by being areas for future exploration.
 Please note that many of the methods and issues described in this paper will be equally useful for microbial ecology in general, but by narrowing our scope we were able to keep our paper within reasonable size limits.
 
-Antibiotics.
+Major finding: dynamic in terms of representation but diverse in terms of function.
+
+Major finding: antibiotics.
 \cite{dethlefsen2008pervasive,dethlefsen2011incomplete,jakobsson2010short,jernberg2007long}
+
+Enterotypes discussion.
 
@@
 
-\subsection{phylogenetics and taxonomy}
+\subsection{Phylogenetics and taxonomy}
 
@@
 
-However, there have been
-
-There are projects that use phylogeny to revise taxonomy, short of PhyloCode.
+However, there have been a number projects that use phylogeny to revise taxonomy, short of PhyloCode.
 GreenGenes.
+tax2tree.
+Bergey's taxonomy is supposed to be.
+SILVA.
 The reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.
@@
 
-With the advent of high throughput sequencing, there has been a lot of interest concerning which variable region to use.
+Tree-based classification.
+
+Curation of HM databases using phylogenetics?
 
 \subsection{Diversity estimates using phylogenetics}
+
+Alpha diversity.
 
@@
 
+Beta diversity.
+
 UniFrac distance and variants.
@@
 Figure: example principal coordinates plot.
+
+\subsection{Inference of horizontal gene transfer}
 
@@
 
+ARB software.
+Phylogenetic placement.
+
 \section{Phylogenetic challenges for human microbiome research}
 
-\subsection{The challenge that lies ahead: functional genes and phylogenomics}
+Non-overlapping reads.
+Assemblies.
+
+Functional genes and phylogenomics.
+
+HGT.
 
@@
 What we can expect next.
+
+Field is still not yet mature.
 
@@
 \bibliography{sbreview}
+
 \end{document}

### 2013-04-02 sbreview.tex
tuning and tweaking

@@
 
+% Noah suggests making a table for
+%alpha diversity
+%beta diversity
+%PCA
+%taxonomic assignment/classification
+%clustering/binning
+
 \begin{document}
@@
 
-% Insert your title:
-\noindent{\Large \bf Phylogenetics and the human microbiome}
+\noindent{\Large \bf Phylogenetics and the human microbiome.}
 \bigskip
@@
 \forarxiv{\
-\title{Phylogenetics and the human microbiome}
+\title{Phylogenetics and the human microbiome.}
 \author{Frederick A. Matsen IV}
@@
 The human microbiome is the collection of cells that live inside and on humans.
-The Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortia have advanced human microboime research by generating thousands of 16s surveys and terabases of metagenomic data, as well as funding bioinformatics and statistical development.
+The Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortia have advanced human microboime research by generating thousands of 16s surveys and terabases of metagenomic data on the human microbiome, as well as funding bioinformatics and statistical development.
 In this paper I review the impact of phylogenetics and tree-thinking on the methods used in the analysis of this data, as well as describing current challenges for phylogenetics coming from this type of work.
@@
 \notforarxiv{
-\noindent (Keywords: human microbiome; microbial ecology; zzz)\\
+\noindent (Keywords: human microbiome; microbial ecology; phylogenetic methods; review)\\
 \vspace{1.5in}
@@
 We will begin by introducing the results of recent research.
-There have been hundreds of papers concerning the human microbiome.
-It is not our desire to survey all of these.
-Rather, our idea is to highlight areas that may be of particular interest for the phylogenetics community by potentially having wider applications or by being areas for future exploration.
-Please note that many of the methods and issues described in this paper will be equally useful for microbial ecology in general, but by narrowing our scope we were able to keep our paper within reasonable size limits.
+% There have been hundreds of papers concerning the human microbiome, and it is not our desire to survey all of these.
+% Rather, our idea is to highlight areas that may be of particular interest for the phylogenetics community by potentially having wider applications or by being areas for future exploration.
+% Please note that many of the methods and issues described in this paper will be equally useful for microbial ecology in general, but by narrowing our scope we were able to keep our paper within reasonable size limits.
 
-Major finding: dynamic in terms of representation but diverse in terms of function.
+The human microbiome has remarkable metabolic potential.
+cite qin paper, and follow up
 
-Major finding: antibiotics.
-\cite{dethlefsen2008pervasive,dethlefsen2011incomplete,jakobsson2010short,jernberg2007long}
+The microbiome is dynamic in terms of representation but apparently constant in terms of function.
+cites
 
-Enterotypes discussion.
+Human microbiome is spatially organized.
+cite segre,
+
+Inter-individual versus intra-individual variation.
+
+16s surveys have shown that antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts years.
+\cite{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}
+
+Substantial attention was given to whether microbial ``types'' exist.
+A high-profile paper from Per Bork's group
+http://www.nature.com/news/gut-microbial-enterotypes-become-less-clear-cut-1.10276
+
+Important bacteria that were previously unknown?
+
+obesity
+turnbaugh papers
+http://www.nature.com/nature/journal/v444/n7122/abs/nature05414.html
+http://europepmc.org/abstract/MED/18407065/reload=0;jsessionid=Ea4YLfXjd7Cfpx5Hvafq.2
+amish study
+http://precedings.nature.com/documents/4957/version/1
+
+Data products
+reference genomes
+HMP DACC.
+
+Terminology: survey versus metagenome.
+definition of a single sample.
 
 \section{Tree-thinking in human microbiome research}
+
+Phylogenetics is primarily used in this realm as a framework with which data can be organized.
+It is common for researchers to equate 16s phylogeny with taxonomy.
+Trait evolution methods are not common.
 
@@
 
-However, there have been a number projects that use phylogeny to revise taxonomy, short of PhyloCode.
+In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
+These attempts are less ambitious than the development of the PhyloCode, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed.
+Bergey's taxonomy is supposed to be.
 GreenGenes.
 tax2tree.
-Bergey's taxonomy is supposed to be.
 SILVA.
-The reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.
+%The reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.
 
 Much of this work has been done using the 16s ribosomal rRNA gene.
+Exceptions.
 
@@
 
-Alpha diversity.
+Because 16s surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
+For the most part, this literature adapts methods from the classical ecological literature.
 
-Variants of phylogenetic diversity and their use.
+Here we will focus on variants of phylogenetic diversity and their use.
+
+GEBA. Most wanted.
+
+Eisen darling paper on extractability.
 
@@
 
-Beta diversity.
+The level of similarity between samples or groups thereof are often called \emph{beta diversity}; it is the current standard to employ phylogenetic tree structure in this sort of comparison.
 
@@
 
+Trait evolution not considered.
+
 \section{Phylogenetic challenges for human microbiome research}
@@
 There has been a shift from interest in community dynamics to fine-scale interactions between host and immune system.
+
+Trait evolution methods are conspicuously absent.
 

### 2013-04-02 sbreview.tex
lots more citations

--- 2013-04-01 sbreview.tex
@@
 
+\cite{hugenholtz2002review}
+{{E}xploring prokaryotic diversity in the genomic era}
+
 Major finding: antibiotics.
@@
 Enterotypes discussion.
+
+Core microbiome
+\cite{turnbaugh2008core}
+
+HMP nature paper
+\cite{methe2012framework}
+
+% applications
+\cite{costello2009bacterial}
+Bacterial community variation in human body habitats across space and time
+
+\cite{ley2006microbial}
+Microbial ecology: human gut microbes associated with obesity
 
@@
 
+\cite{parks2011classifying}
+Classifying short genomic fragments from novel lineages using composition and homology
+
+\cite{mcdonald2011improved}
+An improved Greengenes taxonomy with explicit ranks for ecological and evolutionary analyses of bacteria and archaea
+
+\cite{phylopythia}
+McHardy, A. C.  and Martin, H. G.  and Tsirigos, A.  and Hugenholtz, P.  and Rigoutsos, I.
+
+\cite{munch2008statistical}
+Statistical assignment of DNA sequences using Bayesian phylogenetics
+
+\cite{munch2008fast}
+Fast phylogenetic DNA barcoding
+
+\cite{rosen2011nbc}
+NBC: the Naive Bayes Classification tool webserver for taxonomic classification of metagenomic reads
+
+\cite{segata2011metagenomic}
+Metagenomic biomarker discovery and explanation
+
+\cite{segata2012metagenomic}
+Metagenomic microbial community profiling using unique clade-specific marker genes
+
 Curation of HM databases using phylogenetics?
+
+\cite{forey2001phylocode}
+{The PhyloCode: description and commentary}
+
+\cite{matsen2011reconciling}
+Reconciling taxonomy and phylogenetic inference: formalism and algorithms for describing discord and inferring taxonomic roots
+
+\cite{carma}
+
+\cite{kreig1984bergey}
+Bergey's Manual of Systematic Bacteriology.
+
+\cite{dalevi2007automated}
+Automated group assignment in large phylogenetic trees using GRUNT: GRouping, Ungrouping, Naming Tool
+
+\cite{desantis2006greengenes}
+{Greengenes, a chimera-checked 16S rRNA gene database and workbench compatible with ARB}
+
+\cite{bazinet2012comparative}
+A comparative evaluation of sequence classification programs
+
+\cite{brady2011phymmbl}
+PhymmBL expanded: confidence scores, custom databases, parallelization and more
+
+\cite{brady2009phymm}
+Phymm and PhymmBL: metagenomic phylogenetic classification with interpolated Markov models
+
+\cite{ColeEaRDP2009}
+
+\cite{liu2008accurate}
+Accurate taxonomy assignments from 16S rRNA sequences produced by highly parallel pyrosequencers
+
+\cite{lan2012using}
+Using the RDP Classifier to Predict Taxonomic Novelty and Reduce the Search Space for Finding Novel Organisms
+
+\cite{macdonald2012rapid}
+Rapid identification of high-confidence taxonomic assignments for metagenomic data
+
+\cite{maidak1997rdp}
+The RDP (ribosomal database project)
 
@@
 
+\cite{hill1973diversity}
+Diversity and evenness: a unifying notation and its consequences
+
+\cite{odwyer2012phylogenetic}
+Phylogenetic Diversity Theory Sheds Light on the Structure of Microbial Communities
+
+cite kembel paper
+
+\cite{morlon2011spatial}
+Spatial patterns of phylogenetic diversity
+
 \subsection{Community comparison using phylogenetics}
@@
 
+\cite{chen2012associating}
+Associating microbiome composition with environment-- generalized unifrac
+
 UniFrac distance and variants.
+\cite{LozuponeKnightUniFrac05}
+\cite{LozuponeEaWeightedUnifrac07}
 
 Figure: example principal coordinates plot.
+
+\cite{BikEaMicrobiotaStomach06}
+
+hill numbers
+\cite{chao2010phylogenetic}
+
+\cite{evans2012phylogenetic}
+The phylogenetic Kantorovich-Rubinstein metric for environmental sequence samples
+
+\cite{kuczynski2010microbial}
+Microbial community resemblance methods differ in their ability to detect biologically relevant patterns
+
+\cite{jaccard1908nouvelles}
+Nouvelles recherches sur la distribution florale
+
+\cite{matsen2013edge}
+Edge principal components and squash clustering: using the special structure of phylogenetic placement data for sample comparison
+
+\cite{PurdomAnalyzingDataGraphs08}
+
+\cite{nemergut2011global}
+?? Global patterns in the biogeography of bacterial taxa
 
@@
 \section{Phylogenetic inference as practiced by human microbiome researchers}
+
+\cite{felsenstein1981evolutionary}
+{Evolutionary trees from DNA sequences: a maximum likelihood approach}
+
+\cite{eddy1998profile}
+Profile hidden Markov models.
+
+\cite{li2006cdhit}
+Li, W.  and Godzik, A.
+\cite{edgar2010usearch}
+
+\cite{mirarabsepp}
+{SEPP: SAT{\'e}-Enabled Phylogenetic Placement}
+
+\cite{monierEaLargeViruses08}
+{Taxonomic distribution of large DNA viruses in the sea}
+
+\cite{huson2007megan}
+
+\cite{mcnabb2004hsp65}
+{{A}ssessment of partial sequencing of the 65-kilodalton heat shock protein gene (hsp65) for routine identification of {M}ycobacterium species isolated from clinical sources}
+
+\cite{price2010fasttree}
+{FastTree 2--approximately maximum-likelihood trees for large alignments}
+
+\cite{navlakha2009finding}
+Finding biologically accurate clusterings in hierarchical tree decompositions using the variation of information
+
+\cite{nawrocki2009infernal}
+Infernal 1.0: inference of RNA alignments
+
+\cite{white2010alignment}
+Alignment and clustering of phylogenetic markers-implications for microbial diversity studies
+
+RNA copy number
+\cite{morganEaInVitroSimulatedMetagenome10}
 
@@
 
+\cite{wu2008simple}
+A simple, fast, and accurate method of phylogenomic inference
+
+\cite{matsen2010pplacer}
+
+\cite{ludwig2004arb}
+{{ARB}: a software environment for sequence data}
+\cite{berger2011performance}
+
+\cite{berger2011aligning}
+Aligning short reads to reference alignments and trees
+
+\cite{caporaso2010qiime}
+
+\cite{schloss2009introducing}
+Introducing mothur: open-source, platform-independen
+
+\cite{lynch2013axiome}
+AXIOME: automated exploration of microbial diversity
+
+\cite{wang2007naive}
+{Naive Bayesian classifier for rapid assignment of rRNA sequences into the new bacterial taxonomy}
+
+\cite{werner2011impact}
+Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
+
+\cite{stark2010mltreemap}
+{{MLTreeMap}-accurate Maximum Likelihood placement of environmental DNA sequences into taxonomic and functional reference phylogenies.}
+
+\cite{vonMeringEaQuantitative08}
+{Quantitative phylogenetic assessment of microbial communities in diverse environments}
+
+% alternative loci
+\cite{case2007rpob}
+
+% databases
+\cite{chen2010human}
+The Human Oral Microbiome Database: a web accessible resource for investigating oral microbe taxonomic and genomic information
+
+\cite{griffen2011core}
+CORE: a phylogenetically-curated 16S rDNA database of the core oral microbiome
+
+\cite{srinivasan2012bacterial}
+
 \section{Phylogenetic challenges for human microbiome research}
@@
 There has been a shift from interest in community dynamics to fine-scale interactions between host and immune system.
-
 \cite{wylie2012sequence}
 \cite{chen2012associating}
+
+% medical
+\cite{clarridge2004}
+{{I}mpact of 16{S} r{RNA} gene sequence analysis for identification of bacteria on clinical microbiology and infectious diseases}
+
+\cite{phillips2012microbiome}
+Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography
 
@@
 \end{document}
+
+\cite{hummelen2010deep}
+Deep sequencing of the vaginal microbiota of women with HIV
+
+\cite{ravel2011vaginal}
+Vaginal microbiome of reproductive-age women
+
+\cite{fredricks2005molecular}
+Molecular identification of bacteria associated with bacterial vaginosis

### 2013-04-02 sbreview.tex

@@
 We will begin by introducing the results of recent research.
-There have been hundreds of papers concerning the human microbiome.
-It is not our desire to survey all of these.
-Rather, our idea is to highlight areas that may be of particular interest for the phylogenetics community by potentially having wider applications or by being areas for future exploration.
-Please note that many of the methods and issues described in this paper will be equally useful for microbial ecology in general, but by narrowing our scope we were able to keep our paper within reasonable size limits.
+% There have been hundreds of papers concerning the human microbiome, and it is not our desire to survey all of these.
+% Rather, our idea is to highlight areas that may be of particular interest for the phylogenetics community by potentially having wider applications or by being areas for future exploration.
+% Please note that many of the methods and issues described in this paper will be equally useful for microbial ecology in general, but by narrowing our scope we were able to keep our paper within reasonable size limits.
 
-Major finding: dynamic in terms of representation but diverse in terms of function.
+The human microbiome has remarkable metabolic potential.
+cite qin paper, and follow up
+
+The microbiome is dynamic in terms of representation but apparently constant in terms of function.
+cites
+
+Human microbiome is spatially organized.
+cite segre,
+
+Inter-individual versus intra-individual variation.
+
+16s surveys have shown that antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts years.
+\cite{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}
+
+Substantial attention was given to whether microbial ``types'' exist.
+A high-profile paper from Per Bork's group
+http://www.nature.com/news/gut-microbial-enterotypes-become-less-clear-cut-1.10276
+
+Important bacteria that were previously unknown?
 
@@
 \cite{dethlefsen2008pervasive,dethlefsen2011incomplete,jakobsson2010short,jernberg2007long}
+obesity
+turnbaugh papers
+http://www.nature.com/nature/journal/v444/n7122/abs/nature05414.html
+http://europepmc.org/abstract/MED/18407065/reload=0;jsessionid=Ea4YLfXjd7Cfpx5Hvafq.2
+amish study
+http://precedings.nature.com/documents/4957/version/1
 
-Enterotypes discussion.
+Data products
+reference genomes
+HMP DACC.
 
@@
 Microbial ecology: human gut microbes associated with obesity
+Terminology: survey versus metagenome.
+definition of a single sample.
 
 \section{Tree-thinking in human microbiome research}
+
+Phylogenetics is primarily used in this realm as a framework with which data can be organized.
+It is common for researchers to equate 16s phylogeny with taxonomy.
+Trait evolution methods are not common.
 
@@
 There has been a shift from interest in community dynamics to fine-scale interactions between host and immune system.
+
+Trait evolution methods are conspicuously absent.
+
 \cite{wylie2012sequence}

### 2013-04-02 sbreview.tex
more!

@@
 \maketitle
-
-\section{Introduction}
 }
@@
 Starter paragraph about the scope of human microbiome research.
+Our idea is to highlight areas that may be of particular interest for the phylogenetics community by potentially having wider applications or by being areas for future exploration.
 
-We will begin by introducing the results of recent research.
-% There have been hundreds of papers concerning the human microbiome, and it is not our desire to survey all of these.
-% Rather, our idea is to highlight areas that may be of particular interest for the phylogenetics community by potentially having wider applications or by being areas for future exploration.
-% Please note that many of the methods and issues described in this paper will be equally useful for microbial ecology in general, but by narrowing our scope we were able to keep our paper within reasonable size limits.
+Terminology: survey versus metagenome.
+Definition of a single sample.
 
-The human microbiome has remarkable metabolic potential.
-cite qin paper, and follow up
+The human microbiome is.
+Virome.
+Has been opened up by high-throughput sequencing.
+Non-culturable organisms.
+
+This sequencing has shown that the microbiome contains a diverse collection of organisms, has remarkable metabolic potential.
+Number of species.
+\cite{qin2010human}
+% To understand the impact of gut microbes on human health and well-being it is crucial to assess their genetic potential.
+Here we describe the Illumina-based metagenomic sequencing, assembly and characterization of 3.3 million non-redundant microbial genes, derived from 576.7 gigabases of sequence, from faecal samples of 124 European individuals.
+The gene set, ~150 times larger than the human gene complement, contains an overwhelming majority of the prevalent (more frequent) microbial genes of the cohort and probably includes a large proportion of the prevalent human intestinal microbial genes.
+The genes are largely shared among individuals of the cohort.
+Over 99% of the genes are bacterial, indicating that the entire cohort harbours between 1,000 and 1,150 prevalent bacterial species and each individual at least 160 such species, which are also largely shared.
+We define and describe the minimal gut metagenome and the minimal gut bacterial genome in terms of functions present in all individuals and most bacteria, respectively.
 
 The microbiome is dynamic in terms of representation but apparently constant in terms of function.
-cites
+\cite{hmp2012structure}
 
-Human microbiome is spatially organized.
-cite segre,
+Human microbiome is spatially organized, as can be seen on skin \cite{grice2009topographical}.
 
-Inter-individual versus intra-individual variation.
+Inter-individual versus intra-individual variation \cite{hmp2012structure}
 
@@
 Substantial attention was given to whether microbial ``types'' exist.
-A high-profile paper from Per Bork's group
-http://www.nature.com/news/gut-microbial-enterotypes-become-less-clear-cut-1.10276
-
-Important bacteria that were previously unknown?
+A high-profile paper from Per Bork's group had enterotypes: Bacteroides, Prevotella or Ruminococcus.
+\cite{arumugam2011enterotypes}
+Appears that there is a real tendency to cluster around Bacteroides or Prevotella \cite{wu2011linking}
 
 \cite{hugenholtz2002review}
-{{E}xploring prokaryotic diversity in the genomic era}
+{Exploring prokaryotic diversity in the genomic era}
 
@@
 \cite{dethlefsen2008pervasive,dethlefsen2011incomplete,jakobsson2010short,jernberg2007long}
+
 obesity
@@
 
-% applications
 \cite{costello2009bacterial}
@@
 Microbial ecology: human gut microbes associated with obesity
-Terminology: survey versus metagenome.
-definition of a single sample.
 
@@
 
-GEBA. Most wanted.
+A phylogeny-driven genomic encyclopaedia of Bacteria and Archaea \cite{wu2009phylogeny}.
+
+Most wanted.
 

### 2013-05-01 sbreview.tex
minor changes

@@
 
-The human microbiome is the collection of cells that live inside and on humans.
+The human microbiome is the collection of microbes that live inside and on humans.
 The Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortia have advanced human microboime research by generating thousands of 16s surveys and terabases of metagenomic data on the human microbiome, as well as funding bioinformatics and statistical development.
-In this paper I review the impact of phylogenetics and tree-thinking on the methods used in the analysis of this data, as well as describing current challenges for phylogenetics coming from this type of work.
+In this talk I will review the impact of phylogenetics and tree-thinking on the methods used in the analysis of this data, as well as describing current challenges for phylogenetics coming from this type of work.
 

### 2013-06-05 sbreview.tex
some progress organizing

--- 2013-04-02 sbreview.tex
@@
+% morgan et al extraction
+% PyNAST
+
 %For arXiv, uncomment this block and comment the next block.
@@
 
-% % For Sys Bio, uncomment this block and comment the previous block.
-% \documentclass[12pt,letterpaper]{article}
-% \usepackage{fixltx2e}
-% \usepackage{textcomp}
-% \usepackage{fullpage}
-% \usepackage{amsfonts}
-% \usepackage{verbatim}
-% \usepackage[english]{babel}
-% \usepackage{pifont}
-% \usepackage{color}
-% \usepackage{setspace}
-% \usepackage{lscape}
-% \usepackage{indentfirst}
-% \usepackage[normalem]{ulem}
-% \usepackage{booktabs}
-% %\usepackage{nag}
-% \usepackage{natbib}
-% %\usepackage{bibtex}
-% \usepackage{float}
-% \usepackage{latexsym}
-% %\usepackage{hyperref}
-% \usepackage{url}
-% %\usepackage{html}
-% \usepackage{hyperref}
-% \usepackage{epsfig}
-% \usepackage{graphicx}
-% \usepackage{amssymb}
-% \usepackage{amsmath}
-% \usepackage{bm}
-% \usepackage{array}
-% %\usepackage{mhchem}
-% \usepackage{ifthen}
-% \usepackage{caption}
-% \usepackage{hyperref}
-% %\usepackage{xcolor}
-% \usepackage{amsthm}
-% \usepackage{amstext}
-% \linespread{1.66}
-% \raggedright
-% \setlength{\parindent}{0.5in}
-% \setcounter{secnumdepth}{0}
-% \pagestyle{empty}
-% \renewcommand{\section}[1]{%
-% \bigskip
-% \begin{center}
-% \begin{Large}
-% \normalfont\scshape #1
-% \medskip
-% \end{Large}
-% \end{center}}
-% \renewcommand{\subsection}[1]{%
-% \bigskip
-% \begin{center}
-% \begin{large}
-% \normalfont\itshape #1
-% \end{large}
-% \end{center}}
-% \renewcommand{\subsubsection}[1]{%
-% \vspace{2ex}
-% \noindent
-% \textit{#1.}---}
-% \renewcommand{\tableofcontents}{}
-% \bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
-% \newcommand{\forarxiv}[1]{}
-% \newcommand{\notforarxiv}[1]{#1}
-
 % show keys for eqs, etc.
@@
 \newcommand{\program}{\textsf{program}}
-
-% theorems, etc
-\newtheorem{lem}{Lemma}
-\newtheorem{cor}{Corollary}
-\newtheorem{prop}{Proposition}
-\newtheorem{thm}{Theorem}
-\newtheorem{prob}{Problem}
-\newtheorem{defn}{Definition}
-\newtheorem{obs}{Observation}
-\newtheorem{alg}{Algorithm}
 
@@
 
-The human microbiome is the collection of cells that live inside and on humans.
+The human microbiome is the collection of microbes that live inside of and on the outside of humans.
 The Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortia have advanced human microboime research by generating thousands of 16s surveys and terabases of metagenomic data on the human microbiome, as well as funding bioinformatics and statistical development.
@@
 
-Starter paragraph about the scope of human microbiome research.
-Our idea is to highlight areas that may be of particular interest for the phylogenetics community by potentially having wider applications or by being areas for future exploration.
+The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
+In this section I will review what is known ...
 
-Terminology: survey versus metagenome.
-Definition of a single sample.
+Archaea has been found in the gut.
 
-The human microbiome is.
-Virome.
-Has been opened up by high-throughput sequencing.
-Non-culturable organisms.
+Genetics.
 
-This sequencing has shown that the microbiome contains a diverse collection of organisms, has remarkable metabolic potential.
-Number of species.
+The microbiome is dynamic in terms of representation but apparently constant in terms of function.
+\cite{hmp2012structure}
+
+The microbiome has remarkable metabolic potential.
 \cite{qin2010human}
@@
 
-The microbiome is dynamic in terms of representation but apparently constant in terms of function.
-\cite{hmp2012structure}
-
+Core microbiome \cite{turnbaugh2008core}.
 Human microbiome is spatially organized, as can be seen on skin \cite{grice2009topographical}.
-
-Inter-individual versus intra-individual variation \cite{hmp2012structure}
+Bacterial community variation in human body habitats across space and time \cite{costello2009bacterial}.
+Inter-individual versus intra-individual variation \cite{hmp2012structure}.
 
@@
 \cite{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}
+
+Major finding: antibiotics.
+\cite{dethlefsen2008pervasive,dethlefsen2011incomplete,jakobsson2010short,jernberg2007long}
+
+Obesity
+
+Microbial ecology: human gut microbes associated with obesity \cite{ley2006microbial}
+
+Amish study \cite{zupancic2012analysis}
+"Neither BMI nor any metabolic syndrome trait was associated with a particular gut community.
+Network analysis identified twenty-two bacterial species and four OTUs that were either positively or inversely correlated with metabolic syndrome traits, suggesting that certain members of the gut microbiota may play a role in these metabolic derangements."
+
+pregancy \cite{koren2012host}
 
@@
 
+Although the focus of much of the work has been on bacteria and archaea, other groups of life have attracted more attention lately, such as the fungi.
+Viruses too
+febrile study \cite{wylie2012sequence}
+
+\section{Methods used to investigate the human microbiome}
+
+Terminology: survey versus metagenome.
+Definition of a single sample.
+Despite their importance, I will not describe experimental techniques such as gnotobiotic mice to stay within scope.
+
 \cite{hugenholtz2002review}
@@
 
-Major finding: antibiotics.
-\cite{dethlefsen2008pervasive,dethlefsen2011incomplete,jakobsson2010short,jernberg2007long}
+\subsection{Marker genes}
+History-- bacteria?
 
-obesity
-turnbaugh papers
-http://www.nature.com/nature/journal/v444/n7122/abs/nature05414.html
-http://europepmc.org/abstract/MED/18407065/reload=0;jsessionid=Ea4YLfXjd7Cfpx5Hvafq.2
-amish study
-http://precedings.nature.com/documents/4957/version/1
+Our knowledge of the microbial world is due to a large part derived from the work of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
+Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
+This gene was chosen because...
+They developed the first "universal" 16S PCR primers.
+The great accomplishment of this is learning that archaea, although morphologically similar to bacteria, form their own separate lineage.
 
-Data products
+Non-culturable organisms.
+
+While Woese and colleagues used primitive sequencing methods to gain information, modern researchers have the luxury of high throughput sequencing.
+This can be
+
+Despite the high-throughput and low cost of modern sequencing, inherent challenges remain for the use of population census by marker gene sequencing.
+Most fundamentally, various microbes have different DNA extraction efficiencies, even with hard core protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \cite{morganEaInVitroSimulatedMetagenome10}.
+Furthermore, primer bias.
+Current sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
+Different regions may be good for different groups of organisms.
+
+16S exists in multiple copies and there can be diversity within the copies.
+Kembel efforts to correct for it.
+
+% alternative loci
+\cite{case2007rpob}
+
+Noise and the rare biosphere?
+
+Knowledge of the impact of microbial communities has been
+
+\subsection{Metagenomes}
+
+\subsection{Data generated from the various studies}
 reference genomes
@@
 
-Core microbiome
-\cite{turnbaugh2008core}
-
 HMP nature paper
 \cite{methe2012framework}
-
-\cite{costello2009bacterial}
-Bacterial community variation in human body habitats across space and time
-
-\cite{ley2006microbial}
-Microbial ecology: human gut microbes associated with obesity
 
@@
 It is common for researchers to equate 16s phylogeny with taxonomy.
-Trait evolution methods are not common.
 
@@
 
+The reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.
+
 The most stunning example of this is the discovery of the archaea by Woese.
@@
 These attempts are less ambitious than the development of the PhyloCode, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed.
-Bergey's taxonomy is supposed to be.
+Bergey's Manual of Systematic Bacteriology \cite{kreig1984bergey}.
+
 GreenGenes.
+\cite{desantis2006greengenes}
+{Greengenes, a chimera-checked 16S rRNA gene database and workbench compatible with ARB}
+\cite{dalevi2007automated}
+Automated group assignment in large phylogenetic trees using GRUNT: GRouping, Ungrouping, Naming Tool
+
 tax2tree.
 SILVA.
-%The reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.
 
@@
 Tree-based classification.
-
-\cite{parks2011classifying}
-Classifying short genomic fragments from novel lineages using composition and homology
-
-\cite{mcdonald2011improved}
-An improved Greengenes taxonomy with explicit ranks for ecological and evolutionary analyses of bacteria and archaea
-
-\cite{phylopythia}
-McHardy, A. C.  and Martin, H. G.  and Tsirigos, A.  and Hugenholtz, P.  and Rigoutsos, I.
-
-\cite{munch2008statistical}
-Statistical assignment of DNA sequences using Bayesian phylogenetics
-
-\cite{munch2008fast}
-Fast phylogenetic DNA barcoding
-
-\cite{rosen2011nbc}
-NBC: the Naive Bayes Classification tool webserver for taxonomic classification of metagenomic reads
-
-\cite{segata2011metagenomic}
-Metagenomic biomarker discovery and explanation
-
-\cite{segata2012metagenomic}
-Metagenomic microbial community profiling using unique clade-specific marker genes
-
-Curation of HM databases using phylogenetics?
 
@@
 \cite{carma}
-
-\cite{kreig1984bergey}
-Bergey's Manual of Systematic Bacteriology.
-
-\cite{dalevi2007automated}
-Automated group assignment in large phylogenetic trees using GRUNT: GRouping, Ungrouping, Naming Tool
-
-\cite{desantis2006greengenes}
-{Greengenes, a chimera-checked 16S rRNA gene database and workbench compatible with ARB}
 
@@
 
+\cite{mcdonald2011improved}
+An improved Greengenes taxonomy with explicit ranks for ecological and evolutionary analyses of bacteria and archaea
+
+\cite{parks2011classifying}
+Classifying short genomic fragments from novel lineages using composition and homology
+
+\cite{phylopythia}
+McHardy, A. C.  and Martin, H. G.  and Tsirigos, A.  and Hugenholtz, P.  and Rigoutsos, I.
+
+\cite{munch2008statistical}
+Statistical assignment of DNA sequences using Bayesian phylogenetics
+
+\cite{munch2008fast}
+Fast phylogenetic DNA barcoding
+
+\cite{rosen2011nbc}
+NBC: the Naive Bayes Classification tool webserver for taxonomic classification of metagenomic reads
+
+\cite{segata2011metagenomic}
+Metagenomic biomarker discovery and explanation
+
+\cite{segata2012metagenomic}
+Metagenomic microbial community profiling using unique clade-specific marker genes
+
 \subsection{Diversity estimates using phylogenetics}
@@
 Most wanted.
-
-Eisen darling paper on extractability.
 
@@
 
-\subsection{Inference of horizontal gene transfer}
+\subsection{Genome-scale inquiries using phylogenetics}
+
+GEBA \cite{wu2009phylogeny}
+
+Sushi paper, network of gene transfer.
+
+Reconciliation, team phyldog.
 
 \section{Phylogenetic inference as practiced by human microbiome researchers}
+
+NJ
 
@@
 
-\cite{caporaso2010qiime}
-
-\cite{schloss2009introducing}
-Introducing mothur: open-source, platform-independen
-
-\cite{lynch2013axiome}
-AXIOME: automated exploration of microbial diversity
-
 \cite{wang2007naive}
@@
 
-% alternative loci
-\cite{case2007rpob}
+\subsection{Software}
+\cite{caporaso2010qiime}
+
+\cite{schloss2009introducing}
+Introducing mothur: open-source, platform-independen
+
+\cite{lynch2013axiome}
+AXIOME: automated exploration of microbial diversity
 
@@
 
+hooper2012interactions,
+
 Trait evolution methods are conspicuously absent.
@@
 
+\subsection{Notably absent}
+Trait evolution methods are not common.
+
 \section{Discussion}
@@
 What we can expect next.
+Interactions with the immune system:
+\cite{hooper2012interactions}
+
+Clinical applications, but MALDI-TOF
 
 Field is still not yet mature.
+Need to think about study design.
 

### 2013-06-05 sbreview.tex
cranking away...

@@
 The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
-In this section I will review what is known ...
-
-Archaea has been found in the gut.
-
-Genetics.
-
-The microbiome is dynamic in terms of representation but apparently constant in terms of function.
-\cite{hmp2012structure}
+In this section I will review what is known about them and their effects.
+More bacteria than human cells.
 
@@
 We define and describe the minimal gut metagenome and the minimal gut bacterial genome in terms of functions present in all individuals and most bacteria, respectively.
+The microbiome is dynamic in terms of representation but apparently constant in terms of function.
+\cite{hmp2012structure}
+Core microbiome \cite{turnbaugh2008core}.
 
-Core microbiome \cite{turnbaugh2008core}.
+Microbiome is an ecosystem.
+It is remarkably diverse, though less diverse than soil.
 Human microbiome is spatially organized, as can be seen on skin \cite{grice2009topographical}.
@@
 
-16s surveys have shown that antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts years.
+When we perturb the microbiome, sometimes it makes fundamental changes.
+Antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts years.
 \cite{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}
 
-Major finding: antibiotics.
-\cite{dethlefsen2008pervasive,dethlefsen2011incomplete,jakobsson2010short,jernberg2007long}
+Sometimes these changes lead to phenotypic changes in the host.
+There is strong evidence of interaction between the gut microbiome and obesity, but the story is yet to be fully explored.
+An intervention study has established human gut microbes associated with obesity \cite{ley2006microbial}.
+An obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \cite{turnbaugh2006obesity}.
+Pregnant human gut microbiome \cite{koren2012host} leads to metabolic disorder in mice.
 
-Obesity
-
-Microbial ecology: human gut microbes associated with obesity \cite{ley2006microbial}
-
+These findings have yet to find their way into a real understanding.
 Amish study \cite{zupancic2012analysis}
@@
 
-pregancy \cite{koren2012host}
+In order to really get an understanding, researchers are shifting from community dynamics to fine-scale interactions between host and microbe \cite{hooper2012interactions}.
+Some major findings there.
 
-Substantial attention was given to whether microbial ``types'' exist.
-A high-profile paper from Per Bork's group had enterotypes: Bacteroides, Prevotella or Ruminococcus.
-\cite{arumugam2011enterotypes}
-Appears that there is a real tendency to cluster around Bacteroides or Prevotella \cite{wu2011linking}
+So far we have been focusing on bacteria, though archaea also play a role.
+Archaea have also been found in the gut.
 
 Although the focus of much of the work has been on bacteria and archaea, other groups of life have attracted more attention lately, such as the fungi.
+
 Viruses too
@@
 
-There has been a shift from interest in community dynamics to fine-scale interactions between host and immune system.
-
-hooper2012interactions,
-
-Trait evolution methods are conspicuously absent.
-
-\cite{wylie2012sequence}
-\cite{chen2012associating}
-
-% medical
-\cite{clarridge2004}
-{{I}mpact of 16{S} r{RNA} gene sequence analysis for identification of bacteria on clinical microbiology and infectious diseases}
-
-\cite{phillips2012microbiome}
-Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography
-
 \subsection{Notably absent}
@@
 \section{Discussion}
+What we can expect next.
+As described above, we can expect much more development in terms of specifics.
+\cite{hooper2012interactions}
+Except to the extent that it is useful to look at gene trees to understand function, phylogenenetics doesn't really have a role.
 
-What we can expect next.
-Interactions with the immune system:
-\cite{hooper2012interactions}
+But in addition to deeper probing we can also expect wider coverage.
+Lots of data.
+Citizen microbiome
 
-Clinical applications, but MALDI-TOF
+Comparative studies of microbiomes.
+\cite{phillips2012microbiome}
+Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography
 
-Field is still not yet mature.
-Need to think about study design.
+Clinical applications,
+\cite{clarridge2004}
+{{I}mpact of 16{S} r{RNA} gene sequence analysis for identification of bacteria on clinical microbiology and infectious diseases}
+Understand cause and effect of diseases.
+Take an ecololgical view on atopic dermititis.
+
+For routine identification of a single microbe in culture, MALDI-TOF kicks ass and is cheap.
+
+The field has experienced a frenetic rate of expansion over the past xxx years, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
+Although the human microbiome has led to a lot of "hype," the microbes inside of us and research on them is here to stay.
+Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
 

### 2013-06-05 trash.tex
cranking away...

@@
-\subsection{Software pipelines}
-\cite{caporaso2010qiime}
-\cite{schloss2009introducing}
-\cite{lynch2013axiome}
+Substantial attention was given to whether microbial ``types'' exist.
+A high-profile paper from Per Bork's group had enterotypes: Bacteroides, Prevotella or Ruminococcus.
+\cite{arumugam2011enterotypes}
+Appears that there is a real tendency to cluster around Bacteroides or Prevotella \cite{wu2011linking}

### 2013-06-05 sbreview.tex
....

@@
-% morgan et al extraction
-% PyNAST
+% PI-CRUST
 
-%For arXiv, uncomment this block and comment the next block.
 \documentclass{amsart}
@@
 Definition of a single sample.
-Despite their importance, I will not describe experimental techniques such as gnotobiotic mice to stay within scope.
+There are a number of non-sequence-based experimental techniques that have been crucial for the development of the field, such as manipulation of gnotobiotic mice, but these are beyond the scope of this paper.
 
-\cite{hugenholtz2002review}
-{Exploring prokaryotic diversity in the genomic era}
+Many of these methods cross over with microbial ecology in general.
+Exploring prokaryotic diversity in the genomic era \cite{hugenholtz2002review}.
 
-\subsection{Marker genes}
+\subsection{Microbial community estimation using marker genes}
 History-- bacteria?
 
-Our knowledge of the microbial world is due to a large part derived from the work of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
+Our modern knowledge of the microbial world is due to a large part derived from the work of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
 Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
@@
 While Woese and colleagues used primitive sequencing methods to gain information, modern researchers have the luxury of high throughput sequencing.
-This can be
+This can be done with a high level of multiplexing, making an explicit trade-off between depth of sequencing for each specimen and the number of specimens able to be put on the sequencer at the same time.
+This leads to the question of how many sequences are needed to characterize the microbial diversity of a given environment.
+For simply separating two rather different samples, rather few samples are needed \cite{kuczynski2010microbial}.
 
 Despite the high-throughput and low cost of modern sequencing, inherent challenges remain for the use of population census by marker gene sequencing.
-Most fundamentally, various microbes have different DNA extraction efficiencies, even with hard core protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \cite{morganEaInVitroSimulatedMetagenome10}.
+Most fundamentally, various microbes have different DNA extraction efficiencies, even with hard core protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \cite{morgan2010metagenomic}.
 Furthermore, primer bias.
@@
 
-% alternative loci
-\cite{case2007rpob}
+Some groups have reported advantages to using alternate single-copy genes as markers, (e.g. \cite{case2007rpob},\cite{mcnabb2004hsp65}), however nothing has really come close to 16S.
 
@@
 
-Knowledge of the impact of microbial communities has been
+As we will see, 16S is used frequently as a way of measuring diversity.
+Those used to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
+However, \cite{zaneveld}
+This approach has recently been taken to its logical conclusion by trying to infer.
 
 \subsection{Metagenomes}
+
+As described above ``metagenome" means that data is sheared randomly.
+This type of data has a complication above that taken from 16S survey data, which is that the genetic region sequenced is unknown in addition to the organism it was taken from.
+
+Metagenomic data is typically used to infer functional information directly.
+
+One approach to using metagenomic data is simply to subset to genes that are found in a large proportion of micro-organisms.
+This is the approach taken by the team of Jonathan Eisen, starting first with AMPHORA \cite{wu2008amphora} and subsequently with PhyloSift (in preparation).
+
+In parallel, MLTreeMap.
+
+Because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another.
+One solution is assembly and then phylogenetic analysis.
+metAMOS pipeline \cite{treangen2011metamos}.
+
+A fascinating approach is that of \citet{segata2012metagenomic}, who start by compiling a database of clade-specific genes \citep{segata2011metagenomic}.
+Then they classify the origin of a given read as being the only clade that has this gene.
+They show that this has good specificity, however, sensitivity is limited to genomes that have been sequenced.
 
@@
 
-tax2tree.
-SILVA.
-
-Much of this work has been done using the 16s ribosomal rRNA gene.
-Exceptions.
-
-Tree-based classification.
-
 \cite{forey2001phylocode}
@@
 Reconciling taxonomy and phylogenetic inference: formalism and algorithms for describing discord and inferring taxonomic roots
+
+tax2tree.
+SILVA.
+
+Tree-based classification.
 
@@
 
-\cite{mcnabb2004hsp65}
-{{A}ssessment of partial sequencing of the 65-kilodalton heat shock protein gene (hsp65) for routine identification of {M}ycobacterium species isolated from clinical sources}
-
 \cite{price2010fasttree}
@@
 RNA copy number
-\cite{morganEaInVitroSimulatedMetagenome10}
+\cite{morgan2010metagenomic}
 
@@
 
-Trait evolution not considered.
-
 \section{Phylogenetic challenges for human microbiome research}
@@
 {{I}mpact of 16{S} r{RNA} gene sequence analysis for identification of bacteria on clinical microbiology and infectious diseases}
+For routine identification of a single microbe in culture, MALDI-TOF kicks ass and is cheap.
+Diagnostics involving mixed communities.
+Bioinformatics will get hardened into fixed workflows.
+
 Understand cause and effect of diseases.
@@
 
-For routine identification of a single microbe in culture, MALDI-TOF kicks ass and is cheap.
+Microbial outbreak tracking.
+Very common in RNA viruses.
+Already using similar methods by applying whole-genome sequencing.
 
 The field has experienced a frenetic rate of expansion over the past xxx years, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
-Although the human microbiome has led to a lot of "hype," the microbes inside of us and research on them is here to stay.
+Although the human microbiome has led to a lot of "hype," our microbes are here to stay and so is research on them.
 Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.

### 2013-06-11 sbreview.tex
-

@@
-% PI-CRUST
-
 \documentclass{amsart}
@@
 16S exists in multiple copies and there can be diversity within the copies.
-Kembel efforts to correct for it.
+\citet{kembel2012incorporating}
 
@@
 As we will see, 16S is used frequently as a way of measuring diversity.
-Those used to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
-However, \cite{zaneveld}
-This approach has recently been taken to its logical conclusion by trying to infer.
+Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
+However, \citep{zaneveld2010ribosomal} have shown that there is a correlation between pairwise distances of 16S genes and genetic repertoire.
+This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees (cite PI-CRUST).
 
@@
 
-Metagenomic data is typically used to infer functional information directly.
+Metagenomic data is typically used to infer information about metabolic capacity (cite Humann) rather than make statements of an evolutionary nature.
+One approach to using metagenomic data is simply to subset to genes that are found in a large proportion of micro-organisms.
+This is the approach taken by the teams of Christian von Mering \cite{von2007quantitative,stark2010mltreemap} and Jonathan Eisen \cite{wu2008amphora} (in preparation).
+Because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another, and even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
 
-One approach to using metagenomic data is simply to subset to genes that are found in a large proportion of micro-organisms.
-This is the approach taken by the team of Jonathan Eisen, starting first with AMPHORA \cite{wu2008amphora} and subsequently with PhyloSift (in preparation).
-
-In parallel, MLTreeMap.
-
-Because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another.
-One solution is assembly and then phylogenetic analysis.
-metAMOS pipeline \cite{treangen2011metamos}.
+An alternative approach is to use metagenomic reads directly for analysis.
+The authors of the metAMOS pipeline \cite{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
 
@@
 Phylogenetics is primarily used in this realm as a framework with which data can be organized.
-It is common for researchers to equate 16s phylogeny with taxonomy.
+It is common for researchers to equate 16S phylogeny with taxonomy.
 
@@
 In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
-These attempts are less ambitious than the development of the PhyloCode, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed.
-Bergey's Manual of Systematic Bacteriology \cite{kreig1984bergey}.
-
-GreenGenes.
-\cite{desantis2006greengenes}
-{Greengenes, a chimera-checked 16S rRNA gene database and workbench compatible with ARB}
-\cite{dalevi2007automated}
-Automated group assignment in large phylogenetic trees using GRUNT: GRouping, Ungrouping, Naming Tool
-
-\cite{forey2001phylocode}
-{The PhyloCode: description and commentary}
-
-\cite{matsen2011reconciling}
-Reconciling taxonomy and phylogenetic inference: formalism and algorithms for describing discord and inferring taxonomic roots
-
-tax2tree.
-SILVA.
+These attempts are less ambitious than the development of the PhyloCode \cite{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed.
+Bergey's Manual of Systematic Bacteriology has adopted 16S \cite{kreig1984bergey}.
+The GreenGenes taxonomy \cite{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \cite{dalevi2007automated} and more recently with their tax2tree tool \cite{mcdonald2011improved}.
+Rather than proposing changes to the taxonomy, our group \cite{matsen2011reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy.
 
 Tree-based classification.
-
-\cite{carma}
-
 \cite{bazinet2012comparative}
-A comparative evaluation of sequence classification programs
-
-\cite{brady2011phymmbl}
-PhymmBL expanded: confidence scores, custom databases, parallelization and more
-
-\cite{brady2009phymm}
-Phymm and PhymmBL: metagenomic phylogenetic classification with interpolated Markov models
-
-\cite{ColeEaRDP2009}
-
-\cite{liu2008accurate}
-Accurate taxonomy assignments from 16S rRNA sequences produced by highly parallel pyrosequencers
-
-\cite{lan2012using}
-Using the RDP Classifier to Predict Taxonomic Novelty and Reduce the Search Space for Finding Novel Organisms
-
-\cite{macdonald2012rapid}
-Rapid identification of high-confidence taxonomic assignments for metagenomic data
-
-\cite{maidak1997rdp}
-The RDP (ribosomal database project)
-
-\cite{mcdonald2011improved}
-An improved Greengenes taxonomy with explicit ranks for ecological and evolutionary analyses of bacteria and archaea
-
-\cite{parks2011classifying}
-Classifying short genomic fragments from novel lineages using composition and homology
-
-\cite{phylopythia}
-McHardy, A. C.  and Martin, H. G.  and Tsirigos, A.  and Hugenholtz, P.  and Rigoutsos, I.
-
-\cite{munch2008statistical}
-Statistical assignment of DNA sequences using Bayesian phylogenetics
-
-\cite{munch2008fast}
-Fast phylogenetic DNA barcoding
-
-\cite{rosen2011nbc}
-NBC: the Naive Bayes Classification tool webserver for taxonomic classification of metagenomic reads
-
-\cite{segata2011metagenomic}
-Metagenomic biomarker discovery and explanation
-
-\cite{segata2012metagenomic}
-Metagenomic microbial community profiling using unique clade-specific marker genes
 
@@
 cite kembel paper
-
-\cite{morlon2011spatial}
-Spatial patterns of phylogenetic diversity
 
@@
 
-\subsection{Notably absent}
-Trait evolution methods are not common.
-
 \section{Discussion}

### 2013-06-11 trash.tex
-

@@
 Appears that there is a real tendency to cluster around Bacteroides or Prevotella \cite{wu2011linking}
+
+\cite{ColeEaRDP2009}
+\cite{carma}
+\cite{bazinet2012comparative}
+A comparative evaluation of sequence classification programs
+\cite{brady2011phymmbl}
+PhymmBL expanded: confidence scores, custom databases, parallelization and more
+\cite{brady2009phymm}
+Phymm and PhymmBL: metagenomic phylogenetic classification with interpolated Markov models
+\cite{liu2008accurate}
+Accurate taxonomy assignments from 16S rRNA sequences produced by highly parallel pyrosequencers
+\cite{lan2012using}
+Using the RDP Classifier to Predict Taxonomic Novelty and Reduce the Search Space for Finding Novel Organisms
+\cite{macdonald2012rapid}
+Rapid identification of high-confidence taxonomic assignments for metagenomic data
+\cite{maidak1997rdp}
+The RDP (ribosomal database project)
+\cite{mcdonald2011improved}
+An improved Greengenes taxonomy with explicit ranks for ecological and evolutionary analyses of bacteria and archaea
+\cite{parks2011classifying}
+Classifying short genomic fragments from novel lineages using composition and homology
+\cite{phylopythia}
+McHardy, A. C.  and Martin, H. G.  and Tsirigos, A.  and Hugenholtz, P.  and Rigoutsos, I.
+\cite{munch2008statistical}
+Statistical assignment of DNA sequences using Bayesian phylogenetics
+\cite{munch2008fast}
+Fast phylogenetic DNA barcoding
+\cite{rosen2011nbc}
+NBC: the Naive Bayes Classification tool webserver for taxonomic classification of metagenomic reads
+\cite{segata2011metagenomic}
+Metagenomic biomarker discovery and explanation
+\cite{segata2012metagenomic}
+Metagenomic microbial community profiling using unique clade-specific marker genes
+
+\cite{morlon2011spatial}
+Spatial patterns of phylogenetic diversity

### 2013-09-12 sbreview.tex
fungi in the gut
more refs
abstract tweaks

@@
 
-The human microbiome is the collection of microbes that live inside of and on the outside of humans.
-The Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortia have advanced human microboime research by generating thousands of 16s surveys and terabases of metagenomic data on the human microbiome, as well as funding bioinformatics and statistical development.
-In this paper I review the impact of phylogenetics and tree-thinking on the methods used in the analysis of this data, as well as describing current challenges for phylogenetics coming from this type of work.
+The human microbiome is the collection of microbes that live inside and on the outside of humans.
+Recent work including the Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortia has advanced human microboime research by generating thousands of 16s surveys and terabases of metagenomic data on the human microbiome.
+In this paper I review the impact of phylogenetics and tree-thinking on the methods used in the analysis of human microbiome data, as well as describing current challenges for phylogenetics coming from this type of work.
 
@@
 
+more virome
+http://www.pnas.org/content/early/2013/07/03/1300833110.short
+review paper:http://genomebiology.com/2013/14/7/127
+
+mycobiome
+http://genomemedicine.com/content/5/7/63
+
 In order to really get an understanding, researchers are shifting from community dynamics to fine-scale interactions between host and microbe \cite{hooper2012interactions}.
@@
 Archaea have also been found in the gut.
+and fungi too
+http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0066019
 

### 2013-09-24 sbreview.tex
Back to work

@@
 
-\section{Introduction to the human microbiome}
+\section{Introduction}
 
+Why, as systematic biologists, should we be interested in what these microbial ecologists have done?
+System with a lot of data and a lot of questions.
+Indeed, just to do ecology you have to do sequencing.
+Many of the issues that are raised by high throughput sequencing of beasties that are more commonly of interest for phylogenetics folk are present, and worse, in microbial studies.
+Taxonomic issues.
+
+Lack of "obvious" ways of grouping beasties has led to innovation.
+It means something that they have developed their own way of using trees, although there is some overlap with previous literature.
+Questions of microbial genomics, translated into questions of function, have real consequences.
+
+Why human microbiome versus microbial ecology?
+Relevance to health.
+We know a lot more about these bugs.
+Can manipulate the systems-- feeding studies and gnotobiotic mice.
+
+Opportunities for SB audience to contribute to this field.
+
+\section{The human microbiome}
 The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
@@
 
+% \cite{hummelen2010deep} Deep sequencing of the vaginal microbiota of women with HIV
+% \cite{ravel2011vaginal} Vaginal microbiome of reproductive-age women
+% \cite{fredricks2005molecular} Molecular identification of bacteria associated with bacterial vaginosis
+
 \section{Tree-thinking in human microbiome research}
@@
 It is common for researchers to equate 16S phylogeny with taxonomy.
+UniFrac is interesting.
 
@@
 cite kembel paper
+
+Surprisingly little work on alpha diversity.
 
@@
 
+Usual challenges.
 Non-overlapping reads.
 Assemblies.
+
+However, the apparent success of 16s tree based comparisons leads to some interesting questions.
+
+16s tree has taken us remarkably far, but it's clearly limited.
+Continued research on the correlation between phylogeny and function.
+Can we correct the 16s tree somehow, or what sort of thing can we use that is a better predictor?
+
+There continues to be discussion about whether there is a tree of life.
+Examples.
+This can be a theoretical one.
+What is the correct representation of the genetic ancestry of an organism that allows us to best predict.
+
+Hubbell's neutral theory applied to human microbiome \cite{fierer2012animalcules,costello2012application}.
 
@@
 
+There are limitations to what we can learn.
+Genome doesn't tell all.
+Host side is sometimes missing, but experiments are good.
+
 The field has experienced a frenetic rate of expansion over the past xxx years, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
@@
 \end{document}
-
-\cite{hummelen2010deep}
-Deep sequencing of the vaginal microbiota of women with HIV
-
-\cite{ravel2011vaginal}
-Vaginal microbiome of reproductive-age women
-
-\cite{fredricks2005molecular}
-Molecular identification of bacteria associated with bacterial vaginosis

### 2013-09-24 sbreview.tex
Traits and evolution.

@@
 System with a lot of data and a lot of questions.
+Microbes can only be identified by their molecular sequence.
 Indeed, just to do ecology you have to do sequencing.
 Many of the issues that are raised by high throughput sequencing of beasties that are more commonly of interest for phylogenetics folk are present, and worse, in microbial studies.
+
 Taxonomic issues.
+Lack of "obvious" ways of grouping beasties.
+Lack of broad phenotypic characterization.
+These issues motivate sequence- and evolution-based methodology.
 
-Lack of "obvious" ways of grouping beasties has led to innovation.
+This has led to innovation in ways of using trees.
 It means something that they have developed their own way of using trees, although there is some overlap with previous literature.
-Questions of microbial genomics, translated into questions of function, have real consequences.
+Questions of microbial genomics, translated into questions of function, have important consequences for human health.
 
@@
 
-Opportunities for SB audience to contribute to this field.
+In this review I will describe human microbiome research from the perspective of a phylogenetics researcher.
+I will first briefly review the recent literature on the human microbiome, then describe novel ways in which human microbiome researchers have used trees and the challenges that human microbiome research poses for the phylogenetics community.
+I will finish with opportunities for SB audience to contribute to this field.
 
@@
 However, \citep{zaneveld2010ribosomal} have shown that there is a correlation between pairwise distances of 16S genes and genetic repertoire.
-This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees (cite PI-CRUST).
+This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} using either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
 
@@
 One approach to using metagenomic data is simply to subset to genes that are found in a large proportion of micro-organisms.
-This is the approach taken by the teams of Christian von Mering \cite{von2007quantitative,stark2010mltreemap} and Jonathan Eisen \cite{wu2008amphora} (in preparation).
+This is the approach taken by \citep{von2007quantitative,stark2010mltreemap} and Jonathan Eisen \citep{wu2008amphora} (in preparation).
 Because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another, and even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
@@
 
-\section{Phylogenetic challenges for human microbiome research}
+\section{Phylogenetic challenges and opportunities in human microbiome research}
 
@@
 
+There is a strange divide between trait evolution methods for macro- and micro-fauna.
+The results of simple methods are reasonable.
+Possible that improved methods, perhaps involving whole-genome evolutionary modeling, could shed light on the problem.
+
 There continues to be discussion about whether there is a tree of life.
@@
 What is the correct representation of the genetic ancestry of an organism that allows us to best predict.
+
+In any case, what about trait evolution models that take HGT into account?
+What about trait evolution models that are on KEGG?
 
@@
 Lots of data.
-Citizen microbiome
+Citizen microbiome projects.
 
@@
 
+Perhaps there is a role for trait evolution methods, or perhaps a simple method does prediction to within how good the data is.
+There is the possibility that we don't have to think about individual organisms at all, but rather can think about the collection of genes and their metabolic network as a meta-organism (cite Elbo).
+Question: to what extent is modeling cellular-level evolutionary history interesting?
+
 The field has experienced a frenetic rate of expansion over the past xxx years, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
@@
 Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
+
+\section{Acknowledgements}
+Aaron Darling, Connor McCoy,
+FHCRC folk.
+ATD grant.
 

### 2013-09-24 sbreview.tex
Still reorganizing and thinking.

@@
 Definition of a single sample.
-There are a number of non-sequence-based experimental techniques that have been crucial for the development of the field, such as manipulation of gnotobiotic mice, but these are beyond the scope of this paper.
 
@@
 
+Gnotobiotic mice.
+Human intervention.
+
 \subsection{Microbial community estimation using marker genes}
@@
 
-Our modern knowledge of the microbial world is due to a large part derived from the work of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
+Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
 Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
@@
 
-As we will see, 16S is used frequently as a way of measuring diversity.
-Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
-However, \citep{zaneveld2010ribosomal} have shown that there is a correlation between pairwise distances of 16S genes and genetic repertoire.
-This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} using either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
-
 \subsection{Metagenomes}
@@
 The authors of the metAMOS pipeline \cite{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
-
-A fascinating approach is that of \citet{segata2012metagenomic}, who start by compiling a database of clade-specific genes \citep{segata2011metagenomic}.
-Then they classify the origin of a given read as being the only clade that has this gene.
-They show that this has good specificity, however, sensitivity is limited to genomes that have been sequenced.
 
@@
 
-Without a doubt, the greatest impact that phylogenetics has had on microbial ecology research is on changing our view of the taxonomic relationships between microorganisms.
+Phylogenetics has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
+The most stunning example of this is the discovery of the archaea by Woese.
 In the absence of sequencing, microbiologist have to rely on microscopes and staining.
@@
 
-The reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.
-
-The most stunning example of this is the discovery of the archaea by Woese.
+However, the reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.
 
@@
 
+A fascinating approach is that of \citet{segata2012metagenomic}, who start by compiling a database of clade-specific genes \citep{segata2011metagenomic}.
+Then they classify the origin of a given read as being the only clade that has this gene.
+They show that this has good specificity, however, sensitivity is limited to genomes that have been sequenced.
+
 \subsection{Diversity estimates using phylogenetics}
@@
 Here we will focus on variants of phylogenetic diversity and their use.
-
-A phylogeny-driven genomic encyclopaedia of Bacteria and Archaea \cite{wu2009phylogeny}.
-
-Most wanted.
 
@@
 
-Surprisingly little work on alpha diversity.
+Despite this work, and the enthusiasm with which microbial ecologists have accepted between-community comparison, phylogenetic alpha diversity seems under-developed.
+We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \cite{mccoy2013abundance}.
+We have also looked at PD under rarefaction \cite{nipperess2013mean}.
 
@@
 
+\subsection{Phylogeny and function}
+
+As we will see below, 16S is frequently used as a way of measuring diversity.
+Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
+However, \citep{zaneveld2010ribosomal} have shown that there is a correlation between pairwise distances of 16S genes and genetic repertoire.
+This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} using either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
+
 \subsection{Genome-scale inquiries using phylogenetics}
 
-GEBA \cite{wu2009phylogeny}
+A phylogeny-driven genomic encyclopaedia of Bacteria and Archaea \cite{wu2009phylogeny}.
+
+Most wanted.
 
@@
 
-NJ
-
-\cite{felsenstein1981evolutionary}
-{Evolutionary trees from DNA sequences: a maximum likelihood approach}
-
-\cite{eddy1998profile}
-Profile hidden Markov models.
-
-\cite{li2006cdhit}
-Li, W.  and Godzik, A.
-\cite{edgar2010usearch}
-
-\cite{mirarabsepp}
-{SEPP: SAT{\'e}-Enabled Phylogenetic Placement}
-
-\cite{monierEaLargeViruses08}
-{Taxonomic distribution of large DNA viruses in the sea}
-
-\cite{huson2007megan}
-
-\cite{price2010fasttree}
-{FastTree 2--approximately maximum-likelihood trees for large alignments}
-
-\cite{navlakha2009finding}
-Finding biologically accurate clusterings in hierarchical tree decompositions using the variation of information
-
-\cite{nawrocki2009infernal}
-Infernal 1.0: inference of RNA alignments
-
-\cite{white2010alignment}
-Alignment and clustering of phylogenetic markers-implications for microbial diversity studies
-
-RNA copy number
-\cite{morgan2010metagenomic}
+\subsection{Alignment and tree inference}
 
@@
 Alignment.
+\cite{eddy1998profile}
+Profile hidden Markov models.
+
+\cite{nawrocki2009infernal}
+Infernal 1.0: inference of RNA alignments
+
+PYNAST
+
+\cite{felsenstein1981evolutionary}
+{Evolutionary trees from DNA sequences: a maximum likelihood approach}
+
+NJ
+\cite{price2010fasttree}
+{FastTree 2--approximately maximum-likelihood trees for large alignments}
+
+Species delineation problem not explicitly considered. Clustering.
+\cite{li2006cdhit}
+Li, W.  and Godzik, A.
+\cite{edgar2010usearch}
+\cite{navlakha2009finding} Finding biologically accurate clusterings in hierarchical tree decompositions using the variation of information
+\cite{white2010alignment} Alignment and clustering of phylogenetic markers-implications for microbial diversity studies
 
 Networks not considered.
-
-Species delineation problem not explicitly considered. Clustering.
 
@@
 
-\cite{wang2007naive}
-{Naive Bayesian classifier for rapid assignment of rRNA sequences into the new bacterial taxonomy}
+\cite{mirarabsepp}
+{SEPP: SAT{\'e}-Enabled Phylogenetic Placement}
 
-\cite{werner2011impact}
-Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
+\cite{monierEaLargeViruses08}
+{Taxonomic distribution of large DNA viruses in the sea}
 
@@
 {Quantitative phylogenetic assessment of microbial communities in diverse environments}
+
+\cite{huson2007megan}
+
+\cite{wang2007naive}
+{Naive Bayesian classifier for rapid assignment of rRNA sequences into the new bacterial taxonomy}
+
+\cite{werner2011impact}
+Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
 
@@
 Examples.
-This can be a theoretical one.
-What is the correct representation of the genetic ancestry of an organism that allows us to best predict.
+This can be a theoretical one, but here is a practical formalism: what is the correct representation of the genetic ancestry of an organism that allows us to best predict.
 
@@
 
+The field has already benefited significantly from phylogeny using 16S.
+Now we have more full-genome data.
+We also have more full-genome methods.
+What can we do with that?
+
 HGT.
+What about tree models of within-organism transfer?
+Could we do tree shape statistics?
 
@@
 What we can expect next.
-As described above, we can expect much more development in terms of specifics.
-\cite{hooper2012interactions}
-Except to the extent that it is useful to look at gene trees to understand function, phylogenenetics doesn't really have a role.
 
@@
 
+Perhaps there is a role for trait evolution methods, or perhaps a simple method does prediction to within how good the data is.
+There is the possibility that we don't have to think about individual organisms at all, but rather can think about the collection of genes and their metabolic network as a meta-organism (cite Elbo).
+Has limitation that you forget about cellular boundaries, which matter-- not a freely diffusing soup.
+Question: to what extent is modeling cellular-level evolutionary history interesting?
+As an extension microbes have to be close to one another-- requires population modeling.
+
 There are limitations to what we can learn.
@@
 Host side is sometimes missing, but experiments are good.
-
-Perhaps there is a role for trait evolution methods, or perhaps a simple method does prediction to within how good the data is.
-There is the possibility that we don't have to think about individual organisms at all, but rather can think about the collection of genes and their metabolic network as a meta-organism (cite Elbo).
-Question: to what extent is modeling cellular-level evolutionary history interesting?
+As described above, we can expect much more development in terms of specifics.
+\cite{hooper2012interactions}
 

### 2013-09-24 sbreview.tex
Tree-thinking pass.

@@
 \usepackage{url}
+\usepackage{palatino}
 \usepackage[round]{natbib}
@@
 Lack of broad phenotypic characterization.
+In the absence of sequencing, microbiologist have to rely on microscopes and staining.
+Lots of little things look alike.
 These issues motivate sequence- and evolution-based methodology.
@@
 
-\section{Methods used to investigate the human microbiome}
+\section{Methods used to investigate the human microbiome via sequencing.}
 
 Terminology: survey versus metagenome.
-Definition of a single sample.
-
-Many of these methods cross over with microbial ecology in general.
-Exploring prokaryotic diversity in the genomic era \cite{hugenholtz2002review}.
-
-Gnotobiotic mice.
-Human intervention.
+Many of these methods cross over with microbial ecology in general \citep{hugenholtz2002review} (cite this?).
+I will not be talking about manipulations of these systems, such as gnotobiotic mice and human intervention.
 
 \subsection{Microbial community estimation using marker genes}
-History-- bacteria?
 
@@
 They developed the first "universal" 16S PCR primers.
-The great accomplishment of this is learning that archaea, although morphologically similar to bacteria, form their own separate lineage.
 
@@
 
-Noise and the rare biosphere?
-
 \subsection{Metagenomes}
@@
 
-Metagenomic data is typically used to infer information about metabolic capacity (cite Humann) rather than make statements of an evolutionary nature.
+Some use metagenomic data as an expanded marker gene library.
+\cite{kembel2011phylogenetic}
+AMPHORA, Phylosift.
 One approach to using metagenomic data is simply to subset to genes that are found in a large proportion of micro-organisms.
@@
 
+If not used as an expanded marker gene panel then typically used to infer information about metabolic capacity (cite Humann) rather than make statements of an evolutionary nature.
+
 \subsection{Data generated from the various studies}
@@
 \cite{methe2012framework}
-
-% \cite{hummelen2010deep} Deep sequencing of the vaginal microbiota of women with HIV
-% \cite{ravel2011vaginal} Vaginal microbiome of reproductive-age women
-% \cite{fredricks2005molecular} Molecular identification of bacteria associated with bacterial vaginosis
 
@@
 Phylogenetics has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
-The most stunning example of this is the discovery of the archaea by Woese.
-In the absence of sequencing, microbiologist have to rely on microscopes and staining.
-Lots of little things look alike.
+The most stunning example of this is the discovery that archaea, although morphologically similar to bacteria, form their own separate lineage.
 
@@
 \subsection{Diversity estimates using phylogenetics}
-
 Because 16s surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
@@
 
+On expanded marker tree \cite{kembel2011phylogenetic}.
+
+hill numbers
+\cite{chao2010phylogenetic}
+
 \cite{odwyer2012phylogenetic}
 Phylogenetic Diversity Theory Sheds Light on the Structure of Microbial Communities
-
-cite kembel paper
 
@@
 
-The level of similarity between samples or groups thereof are often called \emph{beta diversity}; it is the current standard to employ phylogenetic tree structure in this sort of comparison.
+The level of similarity between samples or groups thereof are often called \emph{beta diversity};
+Traditional way: \cite{jaccard1908nouvelles} Nouvelles recherches sur la distribution florale
+
+UniFrac distance and variants has been very successful.
+\cite{LozuponeKnightUniFrac05}
+\cite{LozuponeEaWeightedUnifrac07}
+\cite{kuczynski2010microbial} Microbial community resemblance methods differ in their ability to detect biologically relevant patterns
+
+It recovers gradients.
+\cite{nemergut2011global}
+?? Global patterns in the biogeography of bacterial taxa
+
+Figure: example principal coordinates plot.
+
+This has precedents.
+\cite{evans2012phylogenetic}
+The phylogenetic Kantorovich-Rubinstein metric for environmental sequence samples
+
+\cite{BikEaMicrobiotaStomach06}
+\cite{PurdomAnalyzingDataGraphs08}
 
@@
 
-UniFrac distance and variants.
-\cite{LozuponeKnightUniFrac05}
-\cite{LozuponeEaWeightedUnifrac07}
-
-Figure: example principal coordinates plot.
-
-\cite{BikEaMicrobiotaStomach06}
-
-hill numbers
-\cite{chao2010phylogenetic}
-
-\cite{evans2012phylogenetic}
-The phylogenetic Kantorovich-Rubinstein metric for environmental sequence samples
-
-\cite{kuczynski2010microbial}
-Microbial community resemblance methods differ in their ability to detect biologically relevant patterns
-
-\cite{jaccard1908nouvelles}
-Nouvelles recherches sur la distribution florale
-
 \cite{matsen2013edge}
 Edge principal components and squash clustering: using the special structure of phylogenetic placement data for sample comparison
-
-\cite{PurdomAnalyzingDataGraphs08}
-
-\cite{nemergut2011global}
-?? Global patterns in the biogeography of bacterial taxa
 

### 2013-09-24 sbreview.tex
More rearrangin'

@@
 
-more virome
-http://www.pnas.org/content/early/2013/07/03/1300833110.short
-review paper:http://genomebiology.com/2013/14/7/127
-
-mycobiome
-http://genomemedicine.com/content/5/7/63
-
 In order to really get an understanding, researchers are shifting from community dynamics to fine-scale interactions between host and microbe \cite{hooper2012interactions}.
@@
 
-So far we have been focusing on bacteria, though archaea also play a role.
-Archaea have also been found in the gut.
-and fungi too
-http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0066019
-
-Although the focus of much of the work has been on bacteria and archaea, other groups of life have attracted more attention lately, such as the fungi.
-
-Viruses too
-febrile study \cite{wylie2012sequence}
+We will be focusing on bacterial, as they have formed the bulk of the research.
+However, archaea \cite{hoffmann2013archaea} and fungi play a role.
+Viruses too \citep{reyes2010viruses,minot2011human,wylie2012sequence,minot2013rapid}.
 
@@
 
-\subsection{Data generated from the various studies}
+\subsection{Whole genomes}
+Whole-genome sequencing used for microbial outbreak tracking \cite{snitkin2012tracking}.
+Now from culture, but next we will see single-cell sequencing.
+
+\subsection{Data availability and future}
 reference genomes
@@
 \cite{methe2012framework}
+
+Future will give deeper sequencing, which can mean massive parallelization.
+Citizen microbiome projects.
+
+Comparative studies of microbiomes.
+\cite{phillips2012microbiome}
+Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography
 
@@
 \subsection{Software}
-\cite{caporaso2010qiime}
 
-\cite{schloss2009introducing}
-Introducing mothur: open-source, platform-independen
+People do most phylogenetic inferences as part of a pipeline.
+The most popular ones are QIIME \cite{caporaso2010qiime} and mothur \cite{schloss2009introducing}.
+AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \cite{lynch2013axiome}.
 
-\cite{lynch2013axiome}
-AXIOME: automated exploration of microbial diversity
+mothur has ported in the clearcut \cite{evans2006relaxed,sheneman2006clearcut} program.
+QIIME wraps clearcut and \cite{price2010fasttree}.
 
-% databases
+Bayesian inference is very rare.
+
+Databases:
 \cite{chen2010human}
@@
 
-But in addition to deeper probing we can also expect wider coverage.
-Lots of data.
-Citizen microbiome projects.
-
-Comparative studies of microbiomes.
-\cite{phillips2012microbiome}
-Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography
-
 Clinical applications,
@@
 Take an ecololgical view on atopic dermititis.
-
-Microbial outbreak tracking.
-Very common in RNA viruses.
-Already using similar methods by applying whole-genome sequencing.
 

### 2013-09-25 sbreview.tex
Introduction roughed out.

@@
 
-Why, as systematic biologists, should we be interested in what these microbial ecologists have done?
-System with a lot of data and a lot of questions.
-Microbes can only be identified by their molecular sequence.
+The parameter regime and focus of microbiome research sits outside of the traditional setting for systematic biology, thus why should our community be interested in what these microbial ecologists and medical researchers have done?
+This system is data- and question-rich.
+It absolutely requires molecular methods, as microbes can be identified by their molecular sequence, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
 Indeed, just to do ecology you have to do sequencing.
 Many of the issues that are raised by high throughput sequencing of beasties that are more commonly of interest for phylogenetics folk are present, and worse, in microbial studies.
+For instance, the diversity of these organisms is great and organizing them into a taxonomy is a serious challenge, especially in the absence of obvious morphological features.
 
-Taxonomic issues.
-Lack of "obvious" ways of grouping beasties.
-Lack of broad phenotypic characterization.
-In the absence of sequencing, microbiologist have to rely on microscopes and staining.
-Lots of little things look alike.
-These issues motivate sequence- and evolution-based methodology.
+Because of challenges with species definitions, microbial ecology researchers have developed their own techniques for doing comparing samples which are explicitly phylogenetic.
+Although there is some overlap with previous literature, these techniques are novel and could be used in a wider setting and deserve wider consideration by the phylogenetics community.
 
-This has led to innovation in ways of using trees.
-It means something that they have developed their own way of using trees, although there is some overlap with previous literature.
-Questions of microbial genomics, translated into questions of function, have important consequences for human health.
-
-Why human microbiome versus microbial ecology?
-Relevance to health.
-We know a lot more about these bugs.
-Can manipulate the systems-- feeding studies and gnotobiotic mice.
+The human microbiome part of microbial ecology is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
+Additionally, due to more than a century of hospital lab medicine, our knowledge about human-associated microbes is much richer than about microbes in general.
+It is relatively routine to manipulate the human microbiome or models thereof via intervention studies and germ-free animals.
 
@@
 I will first briefly review the recent literature on the human microbiome, then describe novel ways in which human microbiome researchers have used trees and the challenges that human microbiome research poses for the phylogenetics community.
-I will finish with opportunities for SB audience to contribute to this field.
+I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
 
@@
 
-\section{Methods used to investigate the human microbiome via sequencing.}
+\section{Investigating the human microbiome via sequencing}
 
@@
 Many of these methods cross over with microbial ecology in general \citep{hugenholtz2002review} (cite this?).
-I will not be talking about manipulations of these systems, such as gnotobiotic mice and human intervention.
+% I will not be talking about manipulations of these systems, such as gnotobiotic mice and human intervention.
 
 \subsection{Microbial community estimation using marker genes}
+
+Definition. Goal.
 
@@
 For simply separating two rather different samples, rather few samples are needed \cite{kuczynski2010microbial}.
+This has led to desire for massive parallelization (cite).
 

### 2013-09-25 sbreview.tex
Intro to human microbiome fleshed out.

@@
 The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
-In this section I will review what is known about them and their effects.
-More bacteria than human cells.
+These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells.
+The microbiome has remarkable metabolic potential, with a set of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
+Much of our metabolic interaction with the outside world is mediated by our microbiome, as it has important roles in both vitamin
+% http://www.sciencedirect.com/science/article/pii/S095816691200119X
+and drug \citep{maurice2013xenobiotics} metabolism; our food and drug intake also impacts the diversity of microbes present.
+In this section I will briefly review what is known about the human microbiome and its effect on our health.
 
-The microbiome has remarkable metabolic potential.
-\cite{qin2010human}
-% To understand the impact of gut microbes on human health and well-being it is crucial to assess their genetic potential.
-Here we describe the Illumina-based metagenomic sequencing, assembly and characterization of 3.3 million non-redundant microbial genes, derived from 576.7 gigabases of sequence, from faecal samples of 124 European individuals.
-The gene set, ~150 times larger than the human gene complement, contains an overwhelming majority of the prevalent (more frequent) microbial genes of the cohort and probably includes a large proportion of the prevalent human intestinal microbial genes.
-The genes are largely shared among individuals of the cohort.
-Over 99% of the genes are bacterial, indicating that the entire cohort harbours between 1,000 and 1,150 prevalent bacterial species and each individual at least 160 such species, which are also largely shared.
-We define and describe the minimal gut metagenome and the minimal gut bacterial genome in terms of functions present in all individuals and most bacteria, respectively.
-The microbiome is dynamic in terms of representation but apparently constant in terms of function.
-\cite{hmp2012structure}
-Core microbiome \cite{turnbaugh2008core}.
+The human microbiome is an ecosystem.
+The microbiome is dynamic in terms of representation but apparently constant in terms of function \citep{hmp2012structure}.
+The ``core'' microbiome, or the microbiome shared between all humans, has been described \citep{turnbaugh2008core}.
+Human microbiome is spatially organized, as can be seen on skin \citep{grice2009topographical}, with substantial variation in human body habitats across space and time \citep{costello2009bacterial}.
+There is a substantial range of inter-individual versus intra-individual variation \citep{hmp2012structure}.
 
-Microbiome is an ecosystem.
-It is remarkably diverse, though less diverse than soil.
-Human microbiome is spatially organized, as can be seen on skin \cite{grice2009topographical}.
-Bacterial community variation in human body habitats across space and time \cite{costello2009bacterial}.
-Inter-individual versus intra-individual variation \cite{hmp2012structure}.
+Our actions can shift the composition of our microbiome.
+Changes in diet very quickly shift its composition, and there is a strong correlation between long-term diet and microbiome \citep{wu2011linking}.
+Antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts for years \citep{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}.
 
-When we perturb the microbiome, sometimes it makes fundamental changes.
-Antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts years.
-\cite{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}
+Considerable attention has been given to how the microbiome influences host phenotype.
+There is strong evidence of interaction between the gut microbiome and obesity, but the story is yet clear.
+An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
+An obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}.
+Pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}.
+However, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.
 
-Sometimes these changes lead to phenotypic changes in the host.
-There is strong evidence of interaction between the gut microbiome and obesity, but the story is yet to be fully explored.
-An intervention study has established human gut microbes associated with obesity \cite{ley2006microbial}.
-An obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \cite{turnbaugh2006obesity}.
-Pregnant human gut microbiome \cite{koren2012host} leads to metabolic disorder in mice.
+Bacteria have been the primary focus of human microbiome research, however other domains have been investigated.
+Archaea and fungi interact with bacterial residents \citep{hoffmann2013archaea}.
+Virus populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
 
-These findings have yet to find their way into a real understanding.
-Amish study \cite{zupancic2012analysis}
-"Neither BMI nor any metabolic syndrome trait was associated with a particular gut community.
-Network analysis identified twenty-two bacterial species and four OTUs that were either positively or inversely correlated with metabolic syndrome traits, suggesting that certain members of the gut microbiota may play a role in these metabolic derangements."
-
-In order to really get an understanding, researchers are shifting from community dynamics to fine-scale interactions between host and microbe \cite{hooper2012interactions}.
-Some major findings there.
-
-We will be focusing on bacterial, as they have formed the bulk of the research.
-However, archaea \cite{hoffmann2013archaea} and fungi play a role.
-Viruses too \citep{reyes2010viruses,minot2011human,wylie2012sequence,minot2013rapid}.
+In this paper we will be primarily be describing the human microbiome from the perspective of ecological dynamics.
+We will be ignoring recent research on fine-scale immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
+Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and with the human body, ecological methods will continue to have an important role.
 
@@
 This leads to the question of how many sequences are needed to characterize the microbial diversity of a given environment.
-For simply separating two rather different samples, rather few samples are needed \cite{kuczynski2010microbial}.
+For simply separating two rather different samples, rather few samples are needed \citep{kuczynski2010microbial}.
 This has led to desire for massive parallelization (cite).
@@
 Despite the high-throughput and low cost of modern sequencing, inherent challenges remain for the use of population census by marker gene sequencing.
-Most fundamentally, various microbes have different DNA extraction efficiencies, even with hard core protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \cite{morgan2010metagenomic}.
+Most fundamentally, various microbes have different DNA extraction efficiencies, even with hard core protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
 Furthermore, primer bias.
@@
 
-Some groups have reported advantages to using alternate single-copy genes as markers, (e.g. \cite{case2007rpob},\cite{mcnabb2004hsp65}), however nothing has really come close to 16S.
+Some groups have reported advantages to using alternate single-copy genes as markers, (e.g. \citep{case2007rpob},\citep{mcnabb2004hsp65}), however nothing has really come close to 16S.
 
@@
 Some use metagenomic data as an expanded marker gene library.
-\cite{kembel2011phylogenetic}
+\citep{kembel2011phylogenetic}
 AMPHORA, Phylosift.
@@
 An alternative approach is to use metagenomic reads directly for analysis.
-The authors of the metAMOS pipeline \cite{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
+The authors of the metAMOS pipeline \citep{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
 
@@
 \subsection{Whole genomes}
-Whole-genome sequencing used for microbial outbreak tracking \cite{snitkin2012tracking}.
+Whole-genome sequencing used for microbial outbreak tracking \citep{snitkin2012tracking}.
 Now from culture, but next we will see single-cell sequencing.
@@
 HMP nature paper
-\cite{methe2012framework}
+\citep{methe2012framework}
 
@@
 Comparative studies of microbiomes.
-\cite{phillips2012microbiome}
+\citep{phillips2012microbiome}
 Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography
@@
 In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
-These attempts are less ambitious than the development of the PhyloCode \cite{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed.
-Bergey's Manual of Systematic Bacteriology has adopted 16S \cite{kreig1984bergey}.
-The GreenGenes taxonomy \cite{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \cite{dalevi2007automated} and more recently with their tax2tree tool \cite{mcdonald2011improved}.
-Rather than proposing changes to the taxonomy, our group \cite{matsen2011reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy.
+These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed.
+Bergey's Manual of Systematic Bacteriology has adopted 16S \citep{kreig1984bergey}.
+The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
+Rather than proposing changes to the taxonomy, our group \citep{matsen2011reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy.
 
 Tree-based classification.
-\cite{bazinet2012comparative}
+\citep{bazinet2012comparative}
 
@@
 
-\cite{hill1973diversity}
+\citep{hill1973diversity}
 Diversity and evenness: a unifying notation and its consequences
 
-On expanded marker tree \cite{kembel2011phylogenetic}.
+On expanded marker tree \citep{kembel2011phylogenetic}.
 
 hill numbers
-\cite{chao2010phylogenetic}
+\citep{chao2010phylogenetic}
 
-\cite{odwyer2012phylogenetic}
+\citep{odwyer2012phylogenetic}
 Phylogenetic Diversity Theory Sheds Light on the Structure of Microbial Communities
@@
 Despite this work, and the enthusiasm with which microbial ecologists have accepted between-community comparison, phylogenetic alpha diversity seems under-developed.
-We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \cite{mccoy2013abundance}.
-We have also looked at PD under rarefaction \cite{nipperess2013mean}.
+We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}.
+We have also looked at PD under rarefaction \citep{nipperess2013mean}.
 
@@
 The level of similarity between samples or groups thereof are often called \emph{beta diversity};
-Traditional way: \cite{jaccard1908nouvelles} Nouvelles recherches sur la distribution florale
+Traditional way: \citep{jaccard1908nouvelles} Nouvelles recherches sur la distribution florale
 
 UniFrac distance and variants has been very successful.
-\cite{LozuponeKnightUniFrac05}
-\cite{LozuponeEaWeightedUnifrac07}
-\cite{kuczynski2010microbial} Microbial community resemblance methods differ in their ability to detect biologically relevant patterns
+\citep{LozuponeKnightUniFrac05}
+\citep{LozuponeEaWeightedUnifrac07}
+\citep{kuczynski2010microbial} Microbial community resemblance methods differ in their ability to detect biologically relevant patterns
 
 It recovers gradients.
-\cite{nemergut2011global}
+\citep{nemergut2011global}
 ?? Global patterns in the biogeography of bacterial taxa
@@
 This has precedents.
-\cite{evans2012phylogenetic}
+\citep{evans2012phylogenetic}
 The phylogenetic Kantorovich-Rubinstein metric for environmental sequence samples
 
-\cite{BikEaMicrobiotaStomach06}
-\cite{PurdomAnalyzingDataGraphs08}
+\citep{BikEaMicrobiotaStomach06}
+\citep{PurdomAnalyzingDataGraphs08}
 
-\cite{chen2012associating}
+\citep{chen2012associating}
 Associating microbiome composition with environment-- generalized unifrac
 
-\cite{matsen2013edge}
+\citep{matsen2013edge}
 Edge principal components and squash clustering: using the special structure of phylogenetic placement data for sample comparison
@@
 
-A phylogeny-driven genomic encyclopaedia of Bacteria and Archaea \cite{wu2009phylogeny}.
+A phylogeny-driven genomic encyclopaedia of Bacteria and Archaea \citep{wu2009phylogeny}.
 
@@
 Alignment.
-\cite{eddy1998profile}
+\citep{eddy1998profile}
 Profile hidden Markov models.
 
-\cite{nawrocki2009infernal}
+\citep{nawrocki2009infernal}
 Infernal 1.0: inference of RNA alignments
@@
 
-\cite{felsenstein1981evolutionary}
+\citep{felsenstein1981evolutionary}
 {Evolutionary trees from DNA sequences: a maximum likelihood approach}
@@
 NJ
-\cite{price2010fasttree}
+\citep{price2010fasttree}
 {FastTree 2--approximately maximum-likelihood trees for large alignments}
@@
 Species delineation problem not explicitly considered. Clustering.
-\cite{li2006cdhit}
+\citep{li2006cdhit}
 Li, W.  and Godzik, A.
-\cite{edgar2010usearch}
-\cite{navlakha2009finding} Finding biologically accurate clusterings in hierarchical tree decompositions using the variation of information
-\cite{white2010alignment} Alignment and clustering of phylogenetic markers-implications for microbial diversity studies
+\citep{edgar2010usearch}
+\citep{navlakha2009finding} Finding biologically accurate clusterings in hierarchical tree decompositions using the variation of information
+\citep{white2010alignment} Alignment and clustering of phylogenetic markers-implications for microbial diversity studies
 
@@
 
-\cite{wu2008simple}
+\citep{wu2008simple}
 A simple, fast, and accurate method of phylogenomic inference
 
-\cite{matsen2010pplacer}
+\citep{matsen2010pplacer}
 
-\cite{ludwig2004arb}
+\citep{ludwig2004arb}
 {{ARB}: a software environment for sequence data}
-\cite{berger2011performance}
+\citep{berger2011performance}
 
-\cite{berger2011aligning}
+\citep{berger2011aligning}
 Aligning short reads to reference alignments and trees
 
-\cite{mirarabsepp}
+\citep{mirarabsepp}
 {SEPP: SAT{\'e}-Enabled Phylogenetic Placement}
 
-\cite{monierEaLargeViruses08}
+\citep{monierEaLargeViruses08}
 {Taxonomic distribution of large DNA viruses in the sea}
 
-\cite{stark2010mltreemap}
+\citep{stark2010mltreemap}
 {{MLTreeMap}-accurate Maximum Likelihood placement of environmental DNA sequences into taxonomic and functional reference phylogenies.}
 
-\cite{vonMeringEaQuantitative08}
+\citep{vonMeringEaQuantitative08}
 {Quantitative phylogenetic assessment of microbial communities in diverse environments}
 
-\cite{huson2007megan}
+\citep{huson2007megan}
 
-\cite{wang2007naive}
+\citep{wang2007naive}
 {Naive Bayesian classifier for rapid assignment of rRNA sequences into the new bacterial taxonomy}
 
-\cite{werner2011impact}
+\citep{werner2011impact}
 Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
@@
 People do most phylogenetic inferences as part of a pipeline.
-The most popular ones are QIIME \cite{caporaso2010qiime} and mothur \cite{schloss2009introducing}.
-AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \cite{lynch2013axiome}.
+The most popular ones are QIIME \citep{caporaso2010qiime} and mothur \citep{schloss2009introducing}.
+AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
 
-mothur has ported in the clearcut \cite{evans2006relaxed,sheneman2006clearcut} program.
-QIIME wraps clearcut and \cite{price2010fasttree}.
+mothur has ported in the clearcut \citep{evans2006relaxed,sheneman2006clearcut} program.
+QIIME wraps clearcut and \citep{price2010fasttree}.
 
@@
 Databases:
-\cite{chen2010human}
+\citep{chen2010human}
 The Human Oral Microbiome Database: a web accessible resource for investigating oral microbe taxonomic and genomic information
 
-\cite{griffen2011core}
+\citep{griffen2011core}
 CORE: a phylogenetically-curated 16S rDNA database of the core oral microbiome
 
-\cite{srinivasan2012bacterial}
+\citep{srinivasan2012bacterial}
 
@@
 
-Hubbell's neutral theory applied to human microbiome \cite{fierer2012animalcules,costello2012application}.
+Hubbell's neutral theory applied to human microbiome \citep{fierer2012animalcules,costello2012application}.
 
@@
 Clinical applications,
-\cite{clarridge2004}
+\citep{clarridge2004}
 {{I}mpact of 16{S} r{RNA} gene sequence analysis for identification of bacteria on clinical microbiology and infectious diseases}
@@
 As described above, we can expect much more development in terms of specifics.
-\cite{hooper2012interactions}
+\citep{hooper2012interactions}
 

### 2013-09-25 sbreview.tex
Start of data section.

@@
 \newcommand{\refFIGmassTransport}{1}
+
+\hyphenation{Ge-nome Ge-nomes Me-ta-ge-nome Me-ta-ge-nomes Ma-cro-ev-o-lu-tion-ary}
 
@@
 Our actions can shift the composition of our microbiome.
-Changes in diet very quickly shift its composition, and there is a strong correlation between long-term diet and microbiome \citep{wu2011linking}.
+Changes in diet very quickly shift its composition, and there is a strong correlation between long-term diet and microbiome \citep{li2009human,wu2011linking}.
 Antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts for years \citep{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}.
@@
 
-Terminology: survey versus metagenome.
-Many of these methods cross over with microbial ecology in general \citep{hugenholtz2002review} (cite this?).
+% Many of these methods cross over with microbial ecology in general \citep{hugenholtz2002review} (cite this?).
 % I will not be talking about manipulations of these systems, such as gnotobiotic mice and human intervention.
 
-\subsection{Microbial community estimation using marker genes}
+It is now possible to assay microbial communities in high throughput using via sequencing.
+There are two ways of doing so.
+The first is to amplify a specific ``marker'' gene in the genome, and perform sequencing.
+The second is to randomly shear input DNA and/or RNA and then perform sequencing directly.
+Although there is some vagueness in the usage of these words, we will consistently refer to the former as a \textit{survey} and the second a \textit{metagenome}.
+
+\subsection{Microbial community estimation using marker gene surveys}
 

### 2013-09-25 sbreview.tex
Single gene survey roughed out.

@@
 \section{Investigating the human microbiome via sequencing}
-
-% Many of these methods cross over with microbial ecology in general \citep{hugenholtz2002review} (cite this?).
-% I will not be talking about manipulations of these systems, such as gnotobiotic mice and human intervention.
-
 It is now possible to assay microbial communities in high throughput using via sequencing.
@@
 \subsection{Microbial community estimation using marker gene surveys}
+Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
+Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
+This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
+Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which initiated surveys of all organisms regardless of whether they can be cultured.
 
-Definition. Goal.
-
-Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
-Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
-This gene was chosen because...
-They developed the first "universal" 16S PCR primers.
-
-Non-culturable organisms.
-
-While Woese and colleagues used primitive sequencing methods to gain information, modern researchers have the luxury of high throughput sequencing.
+Where Woese and colleagues labored over digestion and gel electrophoresis to infer sequences, modern researchers have the luxury of high throughput sequencing.
 This can be done with a high level of multiplexing, making an explicit trade-off between depth of sequencing for each specimen and the number of specimens able to be put on the sequencer at the same time.
+This has led to extensive parallelization, most recently by sequencing dozens of samples on the Illumina instrument \citep{degnan2011illumina,caporaso2012ultra}.
 This leads to the question of how many sequences are needed to characterize the microbial diversity of a given environment.
-For simply separating two rather different samples, rather few samples are needed \citep{kuczynski2010microbial}.
-This has led to desire for massive parallelization (cite).
+For simply separating two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.
 
@@
 Most fundamentally, various microbes have different DNA extraction efficiencies, even with hard core protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
-Furthermore, primer bias.
 Current sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
-Different regions may be good for different groups of organisms.
-
-16S exists in multiple copies and there can be diversity within the copies.
-\citet{kembel2012incorporating}
-
-Some groups have reported advantages to using alternate single-copy genes as markers, (e.g. \citep{case2007rpob},\citep{mcnabb2004hsp65}), however nothing has really come close to 16S.
+``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
+Worse, multiplex PCR is known to create chimeric sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
+Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb};
+recent work \citep{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been successful despite a moderate evolutionary signal \citep{klappenbach2000rrna}.
+Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
 
@@
 Phylogenetics has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
-The most stunning example of this is the discovery that archaea, although morphologically similar to bacteria, form their own separate lineage.
+The most stunning example of this is the discovery that archaea, although morphologically similar to bacteria, form their own separate lineage \citep{woese1977phylogenetic}.
 

### 2013-09-25 sbreview.tex
Roughed out the data section.

@@
 
+Lots of survey, metagenome, and whole-genome sequencing data has been generated by the Human Microbiome Project \citep{methe2012framework} and is available on its website\footnote{\url{http://www.hmpdacc.org/}}.
+
 \subsection{Microbial community estimation using marker gene surveys}
@@
 This type of data has a complication above that taken from 16S survey data, which is that the genetic region sequenced is unknown in addition to the organism it was taken from.
+However, because it does proceed through an amplification step, it does not have the same primer biases as a marker gene survey, although sequencing is known to have biases.
 
-Some use metagenomic data as an expanded marker gene library.
-\citep{kembel2011phylogenetic}
-AMPHORA, Phylosift.
-One approach to using metagenomic data is simply to subset to genes that are found in a large proportion of micro-organisms.
-This is the approach taken by \citep{von2007quantitative,stark2010mltreemap} and Jonathan Eisen \citep{wu2008amphora} (in preparation).
+It is possible to use metagenomic data as an expanded set of marker genes.
+That is, one can use 16S reads that appear in the metagenome as well as other ``core" genes present in a large proportion of micro-organisms that are expected to follow the same evolutionary path \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}.
 Because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another, and even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
 
-An alternative approach is to use metagenomic reads directly for analysis.
+Metagenomic data can be used to infer information about metabolic capacity \cite{greenblum2012metagenomic,abubucker2012metabolic}.
 The authors of the metAMOS pipeline \citep{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
-
-If not used as an expanded marker gene panel then typically used to infer information about metabolic capacity (cite Humann) rather than make statements of an evolutionary nature.
+The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \cite{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.
 
 \subsection{Whole genomes}
-Whole-genome sequencing used for microbial outbreak tracking \citep{snitkin2012tracking}.
-Now from culture, but next we will see single-cell sequencing.
-
-\subsection{Data availability and future}
-reference genomes
-HMP DACC.
-
-HMP nature paper
-\citep{methe2012framework}
-
-Future will give deeper sequencing, which can mean massive parallelization.
-Citizen microbiome projects.
-
-Comparative studies of microbiomes.
-\citep{phillips2012microbiome}
-Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography
+Whole-genome sequencing from culture is currently being used for microbial outbreak tracking \citep{koser2012rapid,snitkin2012tracking}.
+The Food and Drug Administration maintains GenomeTrakr, an openly accessible database of whole genome sequences from culture\footnote{\url{http://www.fda.gov/Food/FoodScienceResearch/WholeGenomeSequencingProgramWGS/}}.
+This data will become common for unculturable organisms as single-cell sequencing \citep[reviewed in][]{kalisky2011single} becomes common.
 
@@
 
+Future will give deeper sequencing, which can mean massive parallelization.
+Citizen microbiome projects.
+
+Comparative studies of microbiomes.
+\citep{phillips2012microbiome}
+Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography
+
 Clinical applications,

### 2013-09-26 sbreview.tex
Outta steam for the day.

@@
 
-Phylogenetics is primarily used in this realm as a framework with which data can be organized.
-It is common for researchers to equate 16S phylogeny with taxonomy.
-UniFrac is interesting.
+In this section I consider the ways in which phylogenetic thinking has impacted human microbiome research.
+What may be most interesting for this audience is the way in which phylogenetic trees are being used to actively revise taxonomy and used as a structure on which to perform sample comparison.
 
@@
 
-Phylogenetics has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
+Phylogenetic inference has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
 The most stunning example of this is the discovery that archaea, although morphologically similar to bacteria, form their own separate lineage \citep{woese1977phylogenetic}.
 
-However, the reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.
+In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
+These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed to preserve information associated with taxonomic classifications.
+Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process is somewhat opaque \citep{kreig1984bergey}.
+The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved} which uses a heuristic algorithm to optimize $F$-measure of precision and recall.
+Interestingly, this latter tool allows for polyphyletic taxonomic groups as allowance for either lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis.
+Our group \citep{matsen2011reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a previously described coloring problem.
 
-In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
-These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed.
-Bergey's Manual of Systematic Bacteriology has adopted 16S \citep{kreig1984bergey}.
-The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
-Rather than proposing changes to the taxonomy, our group \citep{matsen2011reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy.
+An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
+There is also SAP.
+However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple na\"ive Bayes classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
+
+Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
 
 Tree-based classification.
-\citep{bazinet2012comparative}
 
@@
 For the most part, this literature adapts methods from the classical ecological literature.
+
+One approach is to apply diversity metrics to classified reads.
 

### 2013-09-26 sbreview.tex
AM session. More ideas.

@@
 A phylogeny-driven genomic encyclopaedia of Bacteria and Archaea \citep{wu2009phylogeny}.
+However, for the most wanted \cite{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
+% http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294
 
-Most wanted.
-
-Sushi paper, network of gene transfer.
+Sushi paper \cite{hehemann2010transfer}
+network of gene transfer \cite{smillie2011ecology}
 
@@
 
-However, the apparent success of 16s tree based comparisons leads to some interesting questions.
+In this paper I have devoted considerable space to the ways in which microbial ecologists have used the 16s tree as a proxy structure for the evolutionary history of their favorite organisms.
+They have even shown that 16S distance recapitulates gene content divergence and used this correlation to predict gene functions.
+16s tree has taken us remarkably far, but it is not a complete representation of the evolutionary history of these microbes.
 
-16s tree has taken us remarkably far, but it's clearly limited.
-Continued research on the correlation between phylogeny and function.
-Can we correct the 16s tree somehow, or what sort of thing can we use that is a better predictor?
+Indeed, in the microbial phylogenetics community there continues to be discussion about whether there is a tree of life.
+Examples.
 
-There is a strange divide between trait evolution methods for macro- and micro-fauna.
+However, the apparent success of 16s tree based comparisons begs the question of if a better representation of the evolutionary history of the microbes would yield better predictions.
+That is, if the 16S tree is incomplete and useful, what could we replace it with that would also be incomplete but even more useful?
+This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of an organism that allows us to best predict?
+This representation could be simple.
+For example, one of the results of \citet{zaneveld2010ribosomal} is that 16S correlates better with gene repertoire in some taxonomic groups than others.
+If we were to equip the 16S tree with some measure of the strength of that correlation, would that allow for more accurate prediction?
+Would a species tree reconciled with a collection of gene trees allow for better prediction?
+It is quite possible that there is nothing better given the inherent noise of the data, but if we allow an arbitrary ``hidden'' object, what such object would perform best?
+
+What can we do with improved genome-level phylo reconstruction methods like Bastien's.
+
+Trait evolution models are notably absent for the most part, except for PICRUSt.
 The results of simple methods are reasonable.
 Possible that improved methods, perhaps involving whole-genome evolutionary modeling, could shed light on the problem.
-
-There continues to be discussion about whether there is a tree of life.
-Examples.
-This can be a theoretical one, but here is a practical formalism: what is the correct representation of the genetic ancestry of an organism that allows us to best predict.
-
 In any case, what about trait evolution models that take HGT into account?
 What about trait evolution models that are on KEGG?
+Continued research on the correlation between phylogeny and function.
+
+Trait evolution is used for macroorganisms to describe, say, environmental effects on recent evolution.
+However, evolution of microbes is all ancient?
+Or we can assume that all the genetic diversity needed is already out there, and it's just a question of who gets what?
 
 Hubbell's neutral theory applied to human microbiome \citep{fierer2012animalcules,costello2012application}.
+What about macroevolutionary modeling in general.
 
-Functional genes and phylogenomics.
+Preservation of diversity.
+"Overall these studies reveal that human microbiome data has been preserved in some coprolites, and these preserved human microbiomes match more closely to those from the rural communities than to those from cosmopolitan communities.
+These results suggest that the modern cosmopolitan lifestyle resulted in a dramatic change to the human gut microbiome."
 
-The field has already benefited significantly from phylogeny using 16S.
-Now we have more full-genome data.
-We also have more full-genome methods.
-What can we do with that?
+% note there is also http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0013963
+\citet{ochman2010evolutionary} found identical trees for primate evolution and microbiome.
+They used maximum parsimony such that each column was a bug and each entry took discrete states according to how much of that bug was present.
+What sort of models could be used for microbiome evolution?
+Questions of convergent evolution from nutrition versus history.
 
-HGT.
-What about tree models of within-organism transfer?
-Could we do tree shape statistics?
+Coevolutionary studies between hosts and microbes.
+
+What about diversity of organisms within a population.
+ILS?
+
+How could we do independent contrasts on a collection of reconciled gene trees?
+
+Gut microbiomes of Malawian twin pairs discordant for kwashiorkor \cite{smith2013gut}
 
@@
 {{I}mpact of 16{S} r{RNA} gene sequence analysis for identification of bacteria on clinical microbiology and infectious diseases}
+
 For routine identification of a single microbe in culture, MALDI-TOF kicks ass and is cheap.
 Diagnostics involving mixed communities.
+
 Bioinformatics will get hardened into fixed workflows.
@@
 Understand cause and effect of diseases.
-Take an ecololgical view on atopic dermititis.
+Take an ecological view on atopic dermititis.
 
-Perhaps there is a role for trait evolution methods, or perhaps a simple method does prediction to within how good the data is.
+Although we have raised the possibility of using organism-level modeling via phylogeny
 There is the possibility that we don't have to think about individual organisms at all, but rather can think about the collection of genes and their metabolic network as a meta-organism (cite Elbo).
@@
 
-There are limitations to what we can learn.
-Genome doesn't tell all.
+There are limitations to what we can learn using genetics because more intricate processes such as gene regulation may be at play, limiting what sequence-level phylogenetics can do.
+However, evolution still acts and there is nothing saying that we couldn't model evolution of phenotype in some other way.
+
+Future work will need to include specific interactions.
 Host side is sometimes missing, but experiments are good.
@@
 The field has experienced a frenetic rate of expansion over the past xxx years, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
-Although the human microbiome has led to a lot of "hype," our microbes are here to stay and so is research on them.
+Although the human microbiome has gained a lot of attention recently, and in some cases too much attention, our microbes are here to stay and so is research on them.
 Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
+Phylogenetics has contributed significantly to ecology in the past, and indeed has already contributed significantly to microbial ecology; the next generation of methods may require integration of genome-level data.
 

### 2013-09-26 sbreview.tex
blaser material

@@
 Preservation of diversity.
+\cite{blaser2011antibiotic,cho2012human}
 "Overall these studies reveal that human microbiome data has been preserved in some coprolites, and these preserved human microbiomes match more closely to those from the rural communities than to those from cosmopolitan communities.

### 2013-09-27 sbreview.tex
SILVA etc

--- 2013-09-26 sbreview.tex
@@
 
-The parameter regime and focus of microbiome research sits outside of the traditional setting for systematic biology, thus why should our community be interested in what these microbial ecologists and medical researchers have done?
+The parameter regime and focus of microbiome research sits outside of the traditional setting of eukaryotic phylogenetics, thus why should our community be interested in what these microbial ecologists and medical researchers have done?
 This system is data- and question-rich.
@@
 Indeed, just to do ecology you have to do sequencing.
-Many of the issues that are raised by high throughput sequencing of beasties that are more commonly of interest for phylogenetics folk are present, and worse, in microbial studies.
-For instance, the diversity of these organisms is great and organizing them into a taxonomy is a serious challenge, especially in the absence of obvious morphological features.
 
-Because of challenges with species definitions, microbial ecology researchers have developed their own techniques for doing comparing samples which are explicitly phylogenetic.
+Although there appears to be an artificial divide between phylogeny as practiced as part of microbial ecology and that for multicellular organisms, however there are many parallels between the two enterprises.
+Both communities struggle with issues of sequence alignment, large-scale tree reconstruction, species definition.
+However, approaches differ between the microbial ecology community and that of eukaryotic phylogenetics, in part because the former has access to a fantastic diversity of organisms directly through culture-independent methods, leading to additional problems above usual.
+The species concept is problematic for microbes and there is also considerable discussion concerning how to group them into species-like units.
+The diversity of these organisms is great and organizing them into a taxonomy is a serious challenge, especially in the absence of obvious morphological features.
+
+Because of challenges with species definitions, microbial ecology researchers have developed their own techniques for doing comparing samples which are explicitly phylogenetic rather than requiring species concepts.
 Although there is some overlap with previous literature, these techniques are novel and could be used in a wider setting and deserve wider consideration by the phylogenetics community.
@@
 
-In this review I will describe human microbiome research from the perspective of a phylogenetics researcher.
+In this review I intend to inform the \textit{Systematic Biology} readership of phylogenetics-related research happening in microbial ecology and contrast approaches between the two community.
+I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
+I realize that there is substantial overlap-- for instance the microbial community is very interested in unicellular fungi-- but we will use this terminology with the associated caveats.
+There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
 I will first briefly review the recent literature on the human microbiome, then describe novel ways in which human microbiome researchers have used trees and the challenges that human microbiome research poses for the phylogenetics community.
 I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
+I firmly believe that both fields would benefit from more interactions.
+I have made explicit effort to make a neutral comparison between two areas; indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes orders of magnitude larger than typical for eukaryotic phylogenetics.
 
@@
 Considerable attention has been given to how the microbiome influences host phenotype.
+IBD, etc, NNN.
+kwashiorkor \cite{smith2013gut}.
 There is strong evidence of interaction between the gut microbiome and obesity, but the story is yet clear.
@@
 
-Bacteria have been the primary focus of human microbiome research, however other domains have been investigated.
-Archaea and fungi interact with bacterial residents \citep{hoffmann2013archaea}.
-Virus populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
+Bacteria have been the primary focus of human microbiome research, and other domains have been investigated though to a lesser extent.
+Changes in archaeal and fungi have been shown to covary with bacterial residents \citep{hoffmann2013archaea}.
+Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
 
@@
 The second is to randomly shear input DNA and/or RNA and then perform sequencing directly.
-Although there is some vagueness in the usage of these words, we will consistently refer to the former as a \textit{survey} and the second a \textit{metagenome}.
+Although these words are not consistently used in the literature, we will consistently refer to the former as a \textit{survey} and the second a \textit{metagenome}.
 
 Lots of survey, metagenome, and whole-genome sequencing data has been generated by the Human Microbiome Project \citep{methe2012framework} and is available on its website\footnote{\url{http://www.hmpdacc.org/}}.
+The METAHIT study also generated lots of data but it is not freely available. NNN
 
@@
 ``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
-Worse, multiplex PCR is known to create chimeric sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
+Worse, multiplex PCR is known to create chimeric sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}, and chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
 Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb};
@@
 
+XXX Data through time.
+
 \subsection{Metagenomes}
-
-As described above ``metagenome" means that data is sheared randomly.
-This type of data has a complication above that taken from 16S survey data, which is that the genetic region sequenced is unknown in addition to the organism it was taken from.
-However, because it does proceed through an amplification step, it does not have the same primer biases as a marker gene survey, although sequencing is known to have biases.
+As described above ``metagenome" means that data is sheared randomly across the genome, thus the genetic region of a read is unknown in addition to the organism it was derived from.
+Because it does proceed through an amplification step, it does not have the same primer biases as a marker gene survey, although sequencing is known to have biases.
 
@@
 The authors of the metAMOS pipeline \citep{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
-The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \cite{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.
 
@@
 This data will become common for unculturable organisms as single-cell sequencing \citep[reviewed in][]{kalisky2011single} becomes common.
+The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \cite{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.
 
@@
 Interestingly, this latter tool allows for polyphyletic taxonomic groups as allowance for either lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis.
-Our group \citep{matsen2011reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a previously described coloring problem.
+Our group \citep{matsen2012reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a previously described coloring problem.
 
@@
 There is also SAP.
+
 However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple na\"ive Bayes classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
-
+% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
 Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
-
-Tree-based classification.
 
@@
 
+\subsection{The role of Operational Taxonomic Units (OTUs)}
+There continues to be a lively debate on if there is a meaningful concept of species for microbes NNN, however most microbial research has replaced any traditional species concept with the notion of an Operational Taxonomic Units (OTUs).
+An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly a 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
+It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.
+
+There is an entire industry of OTU inference software with various strategies of various speeds; we will not attempt a systematic review here.
+Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch}.
+These methods are very fast and heuristic, in that they are described as an algorithm rather than a routine attempting to optimize some notion of goodness of clustering.
+It is well known that different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
+There are also methods that try to come up with sequence groupings that more reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
+The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type NNN, or that methods have been devised to find OTUs from non-overlapping sequences NNN.
+With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
+
 \subsection{Diversity estimates using phylogenetics}
 Because 16s surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
-For the most part, this literature adapts methods from the classical ecological literature.
+For the most part, this literature adapts methods from the classical ecological literature replacing taxonomic groupings with OTUs.
+%The most common diversity indices are Simpson and Shannon.
+% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
+However, phylogenetic diversity metrics are also used and here we will focus on their applications and methods.
 
-One approach is to apply diversity metrics to classified reads.
-
-Here we will focus on variants of phylogenetic diversity and their use.
-
-\citep{hill1973diversity}
-Diversity and evenness: a unifying notation and its consequences
-
-On expanded marker tree \citep{kembel2011phylogenetic}.
-
-hill numbers
-\citep{chao2010phylogenetic}
-
-\citep{odwyer2012phylogenetic}
-Phylogenetic Diversity Theory Sheds Light on the Structure of Microbial Communities
-
-Despite this work, and the enthusiasm with which microbial ecologists have accepted between-community comparison, phylogenetic alpha diversity seems under-developed.
-We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}.
-We have also looked at PD under rarefaction \citep{nipperess2013mean}.
+Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
+Phylogenetic diversity notions have been applied on metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
+Although abundance-weighted non-phylogenetic diversity measures such as NNN simpson NNN shannon are among the most common, abundance-weighted phylogenetic diversity measures are generally not used.
+Such measures do exist \citep{chao2010phylogenetic}
+We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}, in particular that they do a better job than the traditionally-used OTU-based measures.
+We have also looked at the mean and variance of PD under random subsampling \citep{nipperess2013mean}.
 
 \subsection{Community comparison using phylogenetics}
-
-The level of similarity between samples or groups thereof are often called \emph{beta diversity};
-Traditional way: \citep{jaccard1908nouvelles} Nouvelles recherches sur la distribution florale
-
-UniFrac distance and variants has been very successful.
-\citep{LozuponeKnightUniFrac05}
-\citep{LozuponeEaWeightedUnifrac07}
+The level of similarity between samples or groups thereof called \emph{beta diversity}.
+As with alpha diversity, it is common to use classical measures \citep[e.g.][]{jaccard1908nouvelles} applied to OTU counts, however in this case phylogenetics-based methods are widely used.
+They generally go under variants of ``UniFrac'' as named by Lozupone and Knight \citep{LozuponeKnightUniFrac05;LozuponeEaWeightedUnifrac07}.
+As evidence of their superiority
 \citep{kuczynski2010microbial} Microbial community resemblance methods differ in their ability to detect biologically relevant patterns
 
+The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample or another rather than between samples.
+Weighted UniFrac is an abundance weighted version.
+We have shown that weighted UniFrac is in fact a special case of the earth-movers distance, and is special in that it can be calculated in linear time \citep{evans2012phylogenetic}.
+We also showed that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
+Recent work \citep{chen2012associating} has shown that a partially abundance-weighted variant of UniFrac has greater power to resolve community differences than either unweighted or weighted UniFrac.
+
+The most common way to use a pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
 It recovers gradients.
@@
 ?? Global patterns in the biogeography of bacterial taxa
+Indeed, the observation of differences between two communities is often used as prima fasce evidence of a difference between them, while the lack of such a difference indicates that the communities are not different (amish?)
 
@@
 
-This has precedents.
-\citep{evans2012phylogenetic}
-The phylogenetic Kantorovich-Rubinstein metric for environmental sequence samples
-
-\citep{BikEaMicrobiotaStomach06}
-\citep{PurdomAnalyzingDataGraphs08}
-
-\citep{chen2012associating}
-Associating microbiome composition with environment-- generalized unifrac
-
-\citep{matsen2013edge}
-Edge principal components and squash clustering: using the special structure of phylogenetic placement data for sample comparison
+There have been efforts to augment the ordination visualizations with additional information giving more structure to the visualizations.
+Biplot.
+Elizabeth Purdom \citep{BikEaMicrobiotaStomach06,PurdomAnalyzingDataGraphs08} has incorporated.
+We have developed an alternative ordination process that explicitly labels the axes with weightings on phylogenetic trees that indicate their meanings \citep{matsen2013edge}.
 
@@
 
-As we will see below, 16S is frequently used as a way of measuring diversity.
-Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
-However, \citep{zaneveld2010ribosomal} have shown that there is a correlation between pairwise distances of 16S genes and genetic repertoire.
-This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} using either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
+16S distance is frequently used as a proxy for a functional comparison.
+Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes is commonly acquired horizontally as well as vertically, and horizontal transmission may leave no trace in the 16S ancestry.
+
+However, \citep{zaneveld2010ribosomal} have shown that organisms that are more distant in terms of 16S are also more divergent in terms of gene repertoire.
+Such observations generally follow a nonlinear curve, and the extent to which they lie on the curve appears to be phylum-dependent.
+This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} by either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
+
+Similar logic has been applied to prioritize microbes for sequencing.
+\citet{wu2009phylogeny} have derived a ``phylogeny-driven genomic encyclopaedia of Bacteria and Archaea'' by selecting organisms for sequencing that are divergent from organisms that have a sequenced genome.
+By doing so they have recovered more novel open reading frames than they would have using other methods NNN.
+However, for the equivalent effort for the human microbiome \cite{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
+NNN wasn't there a previous effort that used ppacer?
+% http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294
 
 \subsection{Genome-scale inquiries using phylogenetics}
+With some notable exceptions, mainstream applications of phylogenetics have typically been with the idea of finding ``the'' tree of a collection of human-associated microbes rather than explicitly exploring divergence between various gene trees.
+As described above, whole-genome data is typically used to directly infer functional information rather than information concerning ancestry.
+The debate concerning whether a microbial tree of life is a useful concept does not seem to have dampened human microbiome researchers' enthusiasm for using such a tree.
 
-A phylogeny-driven genomic encyclopaedia of Bacteria and Archaea \citep{wu2009phylogeny}.
-However, for the most wanted \cite{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
-% http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294
+Nevertheless, the work that has been done to infer horizontal gene transfer has revealed interesting results.
+\citet{hehemann2010transfer} found that a gene found in seaweed has been transferred into the microbiome of Japanese such that individuals with this resulting microbiome are better able to digest algae.
+Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
 
-Sushi paper \cite{hehemann2010transfer}
-network of gene transfer \cite{smillie2011ecology}
-
-Reconciliation, team phyldog.
+NNN look for gene transfer work in microbiome.
 
@@
 \subsection{Alignment and tree inference}
+In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.
+This can be defended by saying for many applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
+Furthermore, given that the underlying data is typically 16S only we can expect some tree problems even with the best methods.
+Additionally, as specified below, these data sets can be very large.
+There does not seem to be contentious discussion of specific features of the inferred trees equivalent to, say, the current discussion around the rooting of the placental mammal tree.
 
-Very large data sets.
+Alignment is focused on creation of large curated alignments, and alignment of additional sequences to those established alignments.
+There are two primary such projects: one is the SILVA project \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes \citep{desantis2006greengenes,mcdonald2011improved} database.
+Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
+Taking the length of 16S to be 1543 nt,
+% SILVA
+% 1543 / 42284
+% .96353230536373096207
+SILVA alignment version 115 has XXX sequences and is over 96\% gap,
+% GreenGenes
+% 1543 / 7682
+% .79914084873730799272
+while the GreenGenes 13\_5 alignment has 1,262,986 sequences and  is almost 80\% gap when aligned by SSU-Align \cite{nawrocki2009structural}
+The SILVA alignment is... hand aligned?
 
-Some groups are not interested in accuracy per se.
+Such large data
 
-Alignment.
+MSA tool specializing in 16S \cite{pruesse2012sina}.
+
+PYNAST
+ARB alignment.
+
 \citep{eddy1998profile}
@@
 
-PYNAST
+\cite{peplies2008standard,yarza2008all}
 
@@
 
-Species delineation problem not explicitly considered. Clustering.
-\citep{li2006cdhit}
-Li, W.  and Godzik, A.
-\citep{edgar2010usearch}
-\citep{navlakha2009finding} Finding biologically accurate clusterings in hierarchical tree decompositions using the variation of information
-\citep{white2010alignment} Alignment and clustering of phylogenetic markers-implications for microbial diversity studies
+Living tree project.
+
+Bayesian phylogenetic inference is absent.
+
+However, it is surprising that so much focus is on complete de novo tree inference, and that methods such as DCM NNN have not gained traction.
 
@@
 
-Chimeric sequences are a real problem, and similar to recombination.
+The scale of the data has motivated strategies other than complete phylogenetic inference.
+Although such insertion has a long history as a means to sequentially build a phylogenetic tree NNN farris
+, the first software with insertion specifically as a goal was by \citep{ludwig2004arb} which works by parsimony.
+ARB is used to reconstruct a full tree.
 
-ARB software.
-Phylogenetic placement.
+There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a tree.
+Other analogous methods have been used \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} with various speeds and features.
+Such questions have also found the interest of the alignment community as well.
+\citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
+\citet{mirarabsepp} does an approximately.
 
-\citep{wu2008simple}
-A simple, fast, and accurate method of phylogenomic inference
-
-\citep{matsen2010pplacer}
-
-\citep{ludwig2004arb}
-{{ARB}: a software environment for sequence data}
-\citep{berger2011performance}
-
-\citep{berger2011aligning}
-Aligning short reads to reference alignments and trees
-
-\citep{mirarabsepp}
-{SEPP: SAT{\'e}-Enabled Phylogenetic Placement}
-
-\citep{monierEaLargeViruses08}
-{Taxonomic distribution of large DNA viruses in the sea}
-
-\citep{stark2010mltreemap}
-{{MLTreeMap}-accurate Maximum Likelihood placement of environmental DNA sequences into taxonomic and functional reference phylogenies.}
-
-\citep{vonMeringEaQuantitative08}
-{Quantitative phylogenetic assessment of microbial communities in diverse environments}
-
+There are other strategies still that do something related to phylogenetics.
 \citep{huson2007megan}
-
-\citep{wang2007naive}
-{Naive Bayesian classifier for rapid assignment of rRNA sequences into the new bacterial taxonomy}
-
-\citep{werner2011impact}
-Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
+LCA strategy
 
@@
 QIIME wraps clearcut and \citep{price2010fasttree}.
-
-Bayesian inference is very rare.
 
@@
 
-Usual challenges.
-Non-overlapping reads.
-Assemblies.
+In addition to the usual challenges surrounding large-scale phylogenetic inference, there are some questions that are specific to the microbial setting.
+I have collected some of these questions here.
 
@@
 
-Indeed, in the microbial phylogenetics community there continues to be discussion about whether there is a tree of life.
-Examples.
-
-However, the apparent success of 16s tree based comparisons begs the question of if a better representation of the evolutionary history of the microbes would yield better predictions.
-That is, if the 16S tree is incomplete and useful, what could we replace it with that would also be incomplete but even more useful?
-This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of an organism that allows us to best predict?
+The apparent success of 16s tree based comparisons begs the question of if a more complete representation of the evolutionary history of the microbes would yield better predictions.
+This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of an organism that allows us to best predict features of underlying genomes?
 This representation could be simple.
@@
 It is quite possible that there is nothing better given the inherent noise of the data, but if we allow an arbitrary ``hidden'' object, what such object would perform best?
+For example, collections of reconciled gene trees in the presence of gene deletion, transfer, and loss, and could be used \citep[see][for interesting recent results]{szollHosi2013lateral,szollHosi2013efficient}.
 
-What can we do with improved genome-level phylo reconstruction methods like Bastien's.
+In addition to using phylogenies to infer gene repertoire, it appears that neutral models involving phylogenetics could be more fully developed.
+Methods explicitly invoking trait evolution seem notably absent, with one recent exception \cite{langille2013predictive}.
+The results of this simple method are reasonable.
+It is possible that improved methods, perhaps involving whole-genome evolutionary modeling or models of metabolic network evolution, could shed light on the problem.
+How could we formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on a collection of reconciled gene trees?
 
-Trait evolution models are notably absent for the most part, except for PICRUSt.
-The results of simple methods are reasonable.
-Possible that improved methods, perhaps involving whole-genome evolutionary modeling, could shed light on the problem.
-In any case, what about trait evolution models that take HGT into account?
-What about trait evolution models that are on KEGG?
-Continued research on the correlation between phylogeny and function.
+Given that microbes are cosmopolitan and have been evolving for a long time, another important project is to model community assembly.
+One direction is to apply Hubbell's neutral theory to the human microbiome \citep{fierer2012animalcules,costello2012application}.
+A recent paper does so with an explicitly phylogenetic perspective \citep{o2012phylogenetic}, and includes some comparison of models to data.
+Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \cite{mooers1997inferring,aldous2011five}.
+In this case various (alpha and beta) diversity statistics would take the role of tree shape statistics.
+It will be challenging to bring together macroevolutionary modeling with species abundance modeling, but some very first steps have been made in another setting \citep{lambert2013predicting}.
 
-Trait evolution is used for macroorganisms to describe, say, environmental effects on recent evolution.
-However, evolution of microbes is all ancient?
-Or we can assume that all the genetic diversity needed is already out there, and it's just a question of who gets what?
+Diversity preservation is of interest for microbiome researchers, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms.
+Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer and has made passionate appeals to preserve microbiome diversity \cite{blaser2011antibiotic,cho2012human}.
+Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
+Because microbiomes are seeded from mother to child, there is a somewhat equivalent notion of extinction of microbiome when the chain is interrupted.
+Consequently, a recent study has explicitly contrasted microbiome development in urban, forest-dwelling, and rural African populations \cite{yatsunenko2012human}, and one study has endeavored to characterize the microbiome from ancient feces \cite{tito2012insights}.
 
-Hubbell's neutral theory applied to human microbiome \citep{fierer2012animalcules,costello2012application}.
-What about macroevolutionary modeling in general.
+Perhaps in part because of the importance of mother-to-child transmission, there are indications of coevolution between microbiomes and their hosts.
+\citet{ochman2010evolutionary} found identical trees for primate evolution and microbiome,
+They used maximum parsimony such that each column was a microbe and in each such entry took discrete states according to how much of that microbe was present.
+Although parsimony gave an interesting answer here, the presence of such coevolution raises the question of what sort of forward-time models are appropriate for microbiome change?
+Would methods using these models do better than clustering or distance methods applied to the above distances?
+Other studies \citep[e.g.][]{phillips2012microbiome,delsuc2013convergence} see a combination of historical and dietary influences.
+This seems like a setting in which convergent evolution could be quantified.
 
-Preservation of diversity.
-"Overall these studies reveal that human microbiome data has been preserved in some coprolites, and these preserved human microbiomes match more closely to those from the rural communities than to those from cosmopolitan communities.
-These results suggest that the modern cosmopolitan lifestyle resulted in a dramatic change to the human gut microbiome."
-
-% note there is also http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0013963
-\citet{ochman2010evolutionary} found identical trees for primate evolution and microbiome.
-They used maximum parsimony such that each column was a bug and each entry took discrete states according to how much of that bug was present.
-What sort of models could be used for microbiome evolution?
-Questions of convergent evolution from nutrition versus history.
-
-Coevolutionary studies between hosts and microbes.
-
-What about diversity of organisms within a population.
-ILS?
-
-How could we do independent contrasts on a collection of reconciled gene trees?
-
-Gut microbiomes of Malawian twin pairs discordant for kwashiorkor \cite{smith2013gut}
+Finally, the conventional wisdom from the microbiome community that conclusions of a UniFrac analysis are robust to tree reconstruction methodology begs further exploration.
+Would it be possible to infer an ``equivalence class'' of phylogenetic trees, where two trees are deemed equivalent if they induce the same principal coordinates projection with the same underlying data?
+Given that a tree is an integral part of a UniFrac analysis, it would also be interesting to be able to infer the features of a tree that determine the primary trends in a projection.
+% How bad of tree inference can we do and still get the same trends?
 
@@
 
+% Companies? http://www.ribocon.com/ second genome, etc
+
 Comparative studies of microbiomes.
-\citep{phillips2012microbiome}
 Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography
@@
 
-The field has experienced a frenetic rate of expansion over the past xxx years, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
+Although I have described some areas for future research for phylogeneticists, researchers should be warned that their work may not be widely accepted by the microbial ecology community.
+That is, methodological changes that would be viewed as improvements in, substitution modeling for example, may not readily be taken up by the microbial ecology community.
+My sense is that this is in part because the bulk of microbial ecology researchers are doctors, immunologists, and oceanographers that have an applied focus and for which computational and modeling issues are a nuisance rather than an interesting aspect.
+For this reason, computational analysis is primarily done in advance or as part of established pipelines and is hardened into standard operating procedures \citep{peplies2008standard}.
+
+Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
 Although the human microbiome has gained a lot of attention recently, and in some cases too much attention, our microbes are here to stay and so is research on them.

### 2013-09-27 sbreview.tex
Thoughts about big trees.

--- 2013-09-26 sbreview.tex
@@
 
-Very large data sets.
+In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.
+This is not unreasonable, as for many applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
+Furthermore, if the underlying data it 16S we can expect some tree problems even with the best methods.
 
-Some groups are not interested in accuracy per se.
+Data sets can be very large.
+Size of GreenGenes tree.
 
-Alignment.
+The alignment culture seems to be mostly around large, established alignments.
+One such alignment is the SILVA alignment \citep{pruesse2007silva,quast2013silva}, which is available as part of a web tool.
+
+Because of the high rate of insertion and deletion
+OMG the SILVA alignment is mostly gap.
+
+Fixed alignment inferred with heuristic algorithms.
+
+Such large data
+
+MSA tool specializing in 16S \cite{pruesse2012sina}.
+
+PYNAST
+ARB alignment.
+
 \citep{eddy1998profile}
@@
 Infernal 1.0: inference of RNA alignments
-
-PYNAST
 
@@
 {FastTree 2--approximately maximum-likelihood trees for large alignments}
+
+Living tree project.
 
@@
 ARB software.
+parsimony insertion.
 Phylogenetic placement.
@@
 
+How bad of tree inference can we do and still get the same trends?
+This is kind of a similar question to what is driving UniFrac?
+Obviously everything can matter, but what does matter?
+
 \section{Discussion}

### 2013-09-27 sbreview.tex
Commits of the morning.

--- 2013-09-26 sbreview.tex
@@
 However, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.
+kwashiorkor \cite{smith2013gut}.
 
@@
 
-Very large data sets.
+In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.
+This can be defended by saying for many applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
+Furthermore, given that the underlying data is typically 16S only we can expect some tree problems even with the best methods.
+Additionally, as specified below, these data sets can be very large.
+There does not seem to be contentious discussion of specific features of the inferred trees equivalent to, say, the current discussion around the rooting of the placental mammal tree.
 
-Some groups are not interested in accuracy per se.
+Alignment is focused on creation of large curated alignments, and alignment of additional sequences to those established alignments.
+There are two primary such projects: one is the SILVA project \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes \citep{desantis2006greengenes,mcdonald2011improved} database.
+Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
+Taking the length of 16S to be 1543 nt,
+% SILVA
+% 1543 / 42284
+% .96353230536373096207
+SILVA alignment version 115 has XXX sequences and is over 96\% gap,
+% GreenGenes
+% 1543 / 7682
+% .79914084873730799272
+while the GreenGenes 13\_5 alignment has 1,262,986 sequences and  is almost 80\% gap when aligned by SSU-Align \cite{nawrocki2009structural}
+The SILVA alignment is... hand aligned?
 
-Alignment.
+Such large data
+
+MSA tool specializing in 16S \cite{pruesse2012sina}.
+
+PYNAST
+ARB alignment.
+
 \citep{eddy1998profile}
@@
 {FastTree 2--approximately maximum-likelihood trees for large alignments}
+
+Living tree project.
+
+Given the scale of the problems, it is not surprising Bayesian phylogenetic inference is absent.
 
@@
 
-Bayesian inference is very rare.
-
 Databases:
@@
 However, the apparent success of 16s tree based comparisons begs the question of if a better representation of the evolutionary history of the microbes would yield better predictions.
-That is, if the 16S tree is incomplete and useful, what could we replace it with that would also be incomplete but even more useful?
-This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of an organism that allows us to best predict?
+That is, if the 16S tree is incomplete and useful, what could we replace it with that would also be necessarily incomplete but even more useful?
+This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of an organism that allows us to best predict features of underlying genomes?
 This representation could be simple.
@@
 
-Preservation of diversity.
-\cite{blaser2011antibiotic,cho2012human}
-"Overall these studies reveal that human microbiome data has been preserved in some coprolites, and these preserved human microbiomes match more closely to those from the rural communities than to those from cosmopolitan communities.
-These results suggest that the modern cosmopolitan lifestyle resulted in a dramatic change to the human gut microbiome."
+Diversity preservation is also of interest for microbiome researchers, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity.
+Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer and has made passionate appeals to preserve microbiome diversity \cite{blaser2011antibiotic,cho2012human}.
+Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
+\cite{tito2012insights} also cite knight group paper and global microbiome.
+Because microbiomes are seeded from mother to child, there is a somewhat equivalent notion of extinction of microbiome when the chain is interrupted.
 
@@
 Citizen microbiome projects.
+
+% Companies? http://www.ribocon.com/ second genome, etc
 

### 2013-09-27 sbreview.tex
Alpha and beta diversity, check.

@@
 However, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.
-kwashiorkor \cite{smith2013gut}.
 
-Bacteria have been the primary focus of human microbiome research, however other domains have been investigated.
-Archaea and fungi interact with bacterial residents \citep{hoffmann2013archaea}.
-Virus populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
+Bacteria have been the primary focus of human microbiome research, and other domains have been investigated though to a lesser extent.
+Changes in archaeal and fungi have been shown to covary with bacterial residents \citep{hoffmann2013archaea}.
+Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
 
@@
 
-As described above ``metagenome" means that data is sheared randomly.
-This type of data has a complication above that taken from 16S survey data, which is that the genetic region sequenced is unknown in addition to the organism it was taken from.
-However, because it does proceed through an amplification step, it does not have the same primer biases as a marker gene survey, although sequencing is known to have biases.
+As described above ``metagenome" means that data is sheared randomly across the genome, thus the genetic region of a read is unknown in addition to the organism it was derived from.
+Because it does proceed through an amplification step, it does not have the same primer biases as a marker gene survey, although sequencing is known to have biases.
 
@@
 
+\subsection{The role of Operational Taxonomic Units (OTUs)}
+There continues to be a lively debate on if there is a meaningful concept of species for microbes NNN, however most microbial research has replaced any traditional species concept with the notion of an Operational Taxonomic Units (OTUs).
+An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly a 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
+It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.
+
+There is an entire industry of OTU inference software with various strategies of various speeds; we will not attempt a systematic review here.
+Of the fixed-cutoff methods, the traditional choice has been CD-HIT NNN, which supplanted by UCLUST.
+These methods are very fast and heuristic, in that they are described as an algorithm rather than a routine attempting to optimize some notion of goodness of clustering.
+It is well known that different ways of doing this heuristic clustering can result in very different results NNN.
+There are also methods that try to come up with sequence groupings that more reflect the type of species divisions found in taxonomies NNN VI-CUT.
+The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type NNN, or that methods have been devised to find OTUs from non-overlapping sequences NNN.
+With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
+
 \subsection{Diversity estimates using phylogenetics}
 Because 16s surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
-For the most part, this literature adapts methods from the classical ecological literature.
+For the most part, this literature adapts methods from the classical ecological literature replacing taxonomic groupings with OTUs.
+%The most common diversity indices are Simpson and Shannon.
+% \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
+However, phylogenetic diversity metrics are also used and here we will focus on their applications and methods.
 
-One approach is to apply diversity metrics to classified reads.
+Phylogenetic diversity notions have been applied on metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
+Phylogenetic Diversity Theory Sheds Light on the Structure of Microbial Communities NNN \citep{odwyer2012phylogenetic}
 
-Here we will focus on variants of phylogenetic diversity and their use.
-
-\citep{hill1973diversity}
-Diversity and evenness: a unifying notation and its consequences
-
-On expanded marker tree \citep{kembel2011phylogenetic}.
-
-hill numbers
-\citep{chao2010phylogenetic}
-
-\citep{odwyer2012phylogenetic}
-Phylogenetic Diversity Theory Sheds Light on the Structure of Microbial Communities
-
-Despite this work, and the enthusiasm with which microbial ecologists have accepted between-community comparison, phylogenetic alpha diversity seems under-developed.
-We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}.
-We have also looked at PD under rarefaction \citep{nipperess2013mean}.
+Despite this work, and the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
+Although abundance-weighted non-phylogenetic diversity measures such as NNN simpson NNN shannon are among the most common, abundance-weighted phylogenetic diversity measures are generally not used.
+Such measures do exist \citep{chao2010phylogenetic}
+We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}, in particular that they do a better job than the traditionally-used OTU-based measures.
+We have also looked at the mean and variance of PD under random subsampling \citep{nipperess2013mean}.
 
 \subsection{Community comparison using phylogenetics}
-
-The level of similarity between samples or groups thereof are often called \emph{beta diversity};
-Traditional way: \citep{jaccard1908nouvelles} Nouvelles recherches sur la distribution florale
-
-UniFrac distance and variants has been very successful.
-\citep{LozuponeKnightUniFrac05}
-\citep{LozuponeEaWeightedUnifrac07}
+The level of similarity between samples or groups thereof called \emph{beta diversity}.
+As with alpha diversity, it is common to use classical measures \citep[e.g.][]{jaccard1908nouvelles} applied to OTU counts, however in this case phylogenetics-based methods are widely used.
+They generally go under variants of ``UniFrac'' as named by Lozupone and Knight \citep{LozuponeKnightUniFrac05;LozuponeEaWeightedUnifrac07}.
+As evidence of their superiority
 \citep{kuczynski2010microbial} Microbial community resemblance methods differ in their ability to detect biologically relevant patterns
 
+The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample or another rather than between samples.
+Weighted UniFrac is an abundance weighted version.
+We have shown that weighted UniFrac is in fact a special case of the earth-movers distance, and is special in that it can be calculated in linear time \citep{evans2012phylogenetic}.
+We also showed that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
+Recent work \citep{chen2012associating} has shown that a partially abundance-weighted variant of UniFrac has greater power to resolve community differences than either unweighted or weighted UniFrac.
+
+The most common way to use a pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
 It recovers gradients.

### 2013-09-27 sbreview.tex
Done with applications of phylogenetics.

@@
 
-As we will see below, 16S is frequently used as a way of measuring diversity.
-Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
-However, \citep{zaneveld2010ribosomal} have shown that there is a correlation between pairwise distances of 16S genes and genetic repertoire.
-This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} using either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
+16S distance is frequently used as a proxy for a functional comparison.
+Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes is commonly acquired horizontally as well as vertically, and horizontal transmission may leave no trace in the 16S ancestry.
+
+However, \citep{zaneveld2010ribosomal} have shown that organisms that are more distant in terms of 16S are also more divergent in terms of gene repertoire.
+Such observations generally follow a nonlinear curve, and the extent to which they lie on the curve appears to be phylum-dependent.
+This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} by either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
+
+Similar logic has been applied to prioritize microbes for sequencing.
+\citet{wu2009phylogeny} have derived a ``phylogeny-driven genomic encyclopaedia of Bacteria and Archaea'' by selecting organisms for sequencing that are divergent from organisms that have a sequenced genome.
+By doing so they have recovered more novel open reading frames than they would have using other methods NNN.
+However, for the equivalent effort for the human microbiome \cite{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
+NNN wasn't there a previous effort that used ppacer?
+% http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294
 
 \subsection{Genome-scale inquiries using phylogenetics}
+With some notable exceptions, mainstream applications of phylogenetics have typically been with the idea of finding ``the'' tree of a collection of human-associated microbes rather than explicitly exploring divergence between various gene trees.
+As described above, whole-genome data is typically used to directly infer functional information rather than information concerning ancestry.
+The debate concerning whether a microbial tree of life is a useful concept does not seem to have dampened human microbiome researchers' enthusiasm for using such a tree.
 
-A phylogeny-driven genomic encyclopaedia of Bacteria and Archaea \citep{wu2009phylogeny}.
-However, for the most wanted \cite{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
-% http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294
+Nevertheless, the work that has been done to infer horizontal gene transfer has revealed interesting results.
+\citet{hehemann2010transfer} found that a gene found in seaweed has been transferred into the microbiome of Japanese such that individuals with this resulting microbiome are better able to digest algae.
+Following on this work, \citet{smillie2011ecology} found that the human gut is in fact a common location for gene transfer.
 
-Sushi paper \cite{hehemann2010transfer}
-network of gene transfer \cite{smillie2011ecology}
-
-Reconciliation, team phyldog.
+NNN look for gene transfer work in microbiome.
 
@@
 \subsection{Alignment and tree inference}
-
 In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.

### 2013-09-27 sbreview.tex
Time to shift locations.

@@
 There is an entire industry of OTU inference software with various strategies of various speeds; we will not attempt a systematic review here.
-Of the fixed-cutoff methods, the traditional choice has been CD-HIT NNN, which supplanted by UCLUST.
+Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch}.
 These methods are very fast and heuristic, in that they are described as an algorithm rather than a routine attempting to optimize some notion of goodness of clustering.
-It is well known that different ways of doing this heuristic clustering can result in very different results NNN.
-There are also methods that try to come up with sequence groupings that more reflect the type of species divisions found in taxonomies NNN VI-CUT.
+It is well known that different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
+There are also methods that try to come up with sequence groupings that more reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
 The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type NNN, or that methods have been devised to find OTUs from non-overlapping sequences NNN.
@@
 
-Given the scale of the problems, it is not surprising Bayesian phylogenetic inference is absent.
+Bayesian phylogenetic inference is absent.
 
-Species delineation problem not explicitly considered. Clustering.
-\citep{li2006cdhit}
-Li, W.  and Godzik, A.
-\citep{edgar2010usearch}
-\citep{navlakha2009finding} Finding biologically accurate clusterings in hierarchical tree decompositions using the variation of information
-\citep{white2010alignment} Alignment and clustering of phylogenetic markers-implications for microbial diversity studies
+However, it is surprising that so much focus is on complete de novo tree inference, and that methods such as DCM NNN have not gained traction.
 
@@
 
-\citep{wu2008simple}
+The scale of the data has motivated strategies to insert sequences into a tree.
+Although such insertion has a long history as a means to sequentially build a phylogenetic tree NNN farris
+, the first software with insertion specifically as a goal was by \citep{ludwig2004arb}.
+Other analogous methods have been used \citep{wu2008simple}.
 A simple, fast, and accurate method of phylogenomic inference
@@
 
-\citep{ludwig2004arb}
-{{ARB}: a software environment for sequence data}
 \citep{berger2011performance}

### 2013-09-27 sbreview.tex
Challenges and opportunities roughed out.

@@
 
+XXX Data through time.
+
 \subsection{Metagenomes}
-
 As described above ``metagenome" means that data is sheared randomly across the genome, thus the genetic region of a read is unknown in addition to the organism it was derived from.
@@
 
+Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
 Phylogenetic diversity notions have been applied on metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
-Phylogenetic Diversity Theory Sheds Light on the Structure of Microbial Communities NNN \citep{odwyer2012phylogenetic}
-
-Despite this work, and the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
 Although abundance-weighted non-phylogenetic diversity measures such as NNN simpson NNN shannon are among the most common, abundance-weighted phylogenetic diversity measures are generally not used.
@@
 \citet{hehemann2010transfer} found that a gene found in seaweed has been transferred into the microbiome of Japanese such that individuals with this resulting microbiome are better able to digest algae.
-Following on this work, \citet{smillie2011ecology} found that the human gut is in fact a common location for gene transfer.
+Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
 
@@
 
-Chimeric sequences are a real problem, and similar to recombination.
+The scale of the data has motivated strategies other than complete phylogenetic inference.
+Although such insertion has a long history as a means to sequentially build a phylogenetic tree NNN farris
+, the first software with insertion specifically as a goal was by \citep{ludwig2004arb} which works by parsimony.
+ARB is used to reconstruct a full tree.
 
-ARB software.
-parsimony insertion.
-Phylogenetic placement.
+There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a tree.
+Other analogous methods have been used \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} with various speeds and features.
+Such questions have also found the interest of the alignment community as well.
+\citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
+\citet{mirarabsepp} does an approximately.
 
-The scale of the data has motivated strategies to insert sequences into a tree.
-Although such insertion has a long history as a means to sequentially build a phylogenetic tree NNN farris
-, the first software with insertion specifically as a goal was by \citep{ludwig2004arb}.
-Other analogous methods have been used \citep{wu2008simple}.
-A simple, fast, and accurate method of phylogenomic inference
-
-\citep{matsen2010pplacer}
-
-\citep{berger2011performance}
-
-\citep{berger2011aligning}
-Aligning short reads to reference alignments and trees
-
-\citep{mirarabsepp}
-{SEPP: SAT{\'e}-Enabled Phylogenetic Placement}
-
-\citep{monierEaLargeViruses08}
-{Taxonomic distribution of large DNA viruses in the sea}
-
-\citep{stark2010mltreemap}
-{{MLTreeMap}-accurate Maximum Likelihood placement of environmental DNA sequences into taxonomic and functional reference phylogenies.}
-
-\citep{vonMeringEaQuantitative08}
-{Quantitative phylogenetic assessment of microbial communities in diverse environments}
-
+There are other strategies still that do something related to phylogenetics.
 \citep{huson2007megan}
-
-\citep{wang2007naive}
-{Naive Bayesian classifier for rapid assignment of rRNA sequences into the new bacterial taxonomy}
-
-\citep{werner2011impact}
-Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
+LCA strategy
 
@@
 
-Indeed, in the microbial phylogenetics community there continues to be discussion about whether there is a tree of life.
-Examples.
-
-However, the apparent success of 16s tree based comparisons begs the question of if a better representation of the evolutionary history of the microbes would yield better predictions.
-That is, if the 16S tree is incomplete and useful, what could we replace it with that would also be necessarily incomplete but even more useful?
+The apparent success of 16s tree based comparisons begs the question of if a more complete representation of the evolutionary history of the microbes would yield better predictions.
 This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of an organism that allows us to best predict features of underlying genomes?
@@
 It is quite possible that there is nothing better given the inherent noise of the data, but if we allow an arbitrary ``hidden'' object, what such object would perform best?
+For example, collections of reconciled gene trees in the presence of gene deletion, transfer, and loss, and could be used \citep[see][for interesting recent results]{szollHosi2013lateral,szollHosi2013efficient}.
 
-What can we do with improved genome-level phylo reconstruction methods like Bastien's.
+In addition to using phylogenies to infer gene repertoire, it appears that neutral models involving phylogenetics could be more fully developed.
+Methods explicitly invoking trait evolution seem notably absent, with one recent exception \cite{langille2013predictive}.
+The results of this simple method are reasonable.
+It is possible that improved methods, perhaps involving whole-genome evolutionary modeling or models of metabolic network evolution, could shed light on the problem.
+How could we formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on a collection of reconciled gene trees?
 
-Trait evolution models are notably absent for the most part, except for PICRUSt.
-The results of simple methods are reasonable.
-Possible that improved methods, perhaps involving whole-genome evolutionary modeling, could shed light on the problem.
-In any case, what about trait evolution models that take HGT into account?
-What about trait evolution models that are on KEGG?
-Continued research on the correlation between phylogeny and function.
+Given that microbes are cosmopolitan and have been evolving for a long time, another important project is to model community assembly.
+One direction is to apply Hubbell's neutral theory to the human microbiome \citep{fierer2012animalcules,costello2012application}.
+A recent paper does so with an explicitly phylogenetic perspective \citep{o2012phylogenetic}, and includes some comparison of models to data.
+Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \cite{mooers1997inferring,aldous2011five}.
+In this case various (alpha and beta) diversity statistics would take the role of tree shape statistics.
+It will be challenging to bring together macroevolutionary modeling with species abundance modeling, but some very first steps have been made in another setting \citep{lambert2013predicting}.
 
-Trait evolution is used for macroorganisms to describe, say, environmental effects on recent evolution.
-However, evolution of microbes is all ancient?
-Or we can assume that all the genetic diversity needed is already out there, and it's just a question of who gets what?
-
-Hubbell's neutral theory applied to human microbiome \citep{fierer2012animalcules,costello2012application}.
-What about macroevolutionary modeling in general.
-
-Diversity preservation is also of interest for microbiome researchers, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity.
+Diversity preservation is of interest for microbiome researchers, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms.
 Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer and has made passionate appeals to preserve microbiome diversity \cite{blaser2011antibiotic,cho2012human}.
 Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
-\cite{tito2012insights} also cite knight group paper and global microbiome.
 Because microbiomes are seeded from mother to child, there is a somewhat equivalent notion of extinction of microbiome when the chain is interrupted.
+Consequently, a recent study has explicitly contrasted microbiome development in urban, forest-dwelling, and rural African populations \cite{yatsunenko2012human}, and one study has endeavored to characterize the microbiome from ancient feces \cite{tito2012insights}.
 
-% note there is also http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0013963
-\citet{ochman2010evolutionary} found identical trees for primate evolution and microbiome.
-They used maximum parsimony such that each column was a bug and each entry took discrete states according to how much of that bug was present.
-What sort of models could be used for microbiome evolution?
-Questions of convergent evolution from nutrition versus history.
+Perhaps in part because of the importance of mother-to-child transmission, there are indications of coevolution between microbiomes and their hosts.
+\citet{ochman2010evolutionary} found identical trees for primate evolution and microbiome,
+They used maximum parsimony such that each column was a microbe and in each such entry took discrete states according to how much of that microbe was present.
+Although parsimony gave an interesting answer here, the presence of such coevolution raises the question of what sort of forward-time models are appropriate for microbiome change?
+Would methods using these models do better than clustering or distance methods applied to the above distances?
+Other studies \citep[e.g.][]{phillips2012microbiome,delsuc2013convergence} see a combination of historical and dietary influences.
+This seems like a setting in which convergent evolution could be quantified.
 
-Coevolutionary studies between hosts and microbes.
-
-What about diversity of organisms within a population.
-ILS?
-
-How could we do independent contrasts on a collection of reconciled gene trees?
-
-Gut microbiomes of Malawian twin pairs discordant for kwashiorkor \cite{smith2013gut}
-
-How bad of tree inference can we do and still get the same trends?
-This is kind of a similar question to what is driving UniFrac?
-Obviously everything can matter, but what does matter?
+Finally, the conventional wisdom from the microbiome community that conclusions of a UniFrac analysis are robust to tree reconstruction methodology begs further exploration.
+Would it be possible to infer an ``equivalence class'' of phylogenetic trees, where two trees are deemed equivalent if they induce the same principal coordinates projection with the same underlying data?
+Given that a tree is an integral part of a UniFrac analysis, it would also be interesting to be able to infer the features of a tree that determine the primary trends in a projection.
+% How bad of tree inference can we do and still get the same trends?
 
@@
 Comparative studies of microbiomes.
-\citep{phillips2012microbiome}
 Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography

### 2013-09-27 sbreview.tex
Phylo methods section roughed.

@@
 
+The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy to turn a taxonomy and a collection of BLAST results of a unknown sequence on a database of sequences with those taxonomic labels into taxonomic assignments for the unknown sequence.
+
 A fascinating approach is that of \citet{segata2012metagenomic}, who start by compiling a database of clade-specific genes \citep{segata2011metagenomic}.
@@
 
-Alignment is focused on creation of large curated alignments, and alignment of additional sequences to those established alignments.
+Alignment methods are primarily focused on extending a curated ``seed alignment'' with additional sequences; tools have been created with exactly this application for 16S in mind \citep{desantis2006nast,caporaso2010pynast,pruesse2012sina}.
+The community also uses profile hidden Markov models \citep{eddy1998profile} and CM models \citep{nawrocki2009infernal,nawrocki2009structural}.
+
+The large data sets associated with human microbiome analysis require highly efficient algorithms for \emph{de novo} tree inference.
+Historically this has meant relaxed neighbor joining \cite{evans2006relaxed,sheneman2006clearcut}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
+People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in clearcut \citep{evans2006relaxed,sheneman2006clearcut} program, and QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree 2.
+% AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
+% Bayesian phylogenetic inference is absent.
+
+The scale of the data has motivated strategies other than complete phylogenetic inference.
+Although such insertion has a long history as a means to sequentially build a phylogenetic tree \citep{kluge1969quantitative}, the first software with insertion specifically as a goal was by \citep{ludwig2004arb} which works by parsimony.
+ARB is commonly used to reconstruct a full tree by direct insertion.
+
+There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a tree, sometimes with uncertainty estimates.
+Other analogous methods have been used \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} with various speeds and features.
+Such questions have also found the interest of the alignment community as well.
+\citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
+\citet{mirarabsepp} uses data set partitioning to improve alignments on subsets of taxa.
+
+Considerable effort goes to the creation of large curated alignments and phylogenetic trees on 16S.
 There are two primary such projects: one is the SILVA project \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes \citep{desantis2006greengenes,mcdonald2011improved} database.
@@
 % .96353230536373096207
-SILVA alignment version 115 has XXX sequences and is over 96\% gap,
+SILVA reference alignment version 115 has 479,726 sequences and is over 96\% gap,
 % GreenGenes
@@
 % .79914084873730799272
-while the GreenGenes 13\_5 alignment has 1,262,986 sequences and  is almost 80\% gap when aligned by SSU-Align \cite{nawrocki2009structural}
-The SILVA alignment is... hand aligned?
+while the GreenGenes 13\_5 alignment has 1,262,986 sequences and is almost 80\% gap.
+The SILVA-associated 'all-species living tree' project \cite{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
+The GreenGenes tree is updated by running FastTree from scratch.
+In addition to these 16S-based resources, the MicrobesOnline resource \citep{dehal2010microbesonline} offers a tree-based genome browser.
+On a much smaller scale, there curated reference sets that are microbiome body-site specific \citep{chen2010human,griffen2011core,srinivasan2012bacterial}
 
-Such large data
-
-MSA tool specializing in 16S \cite{pruesse2012sina}.
-
-PYNAST
-ARB alignment.
-
-\citep{eddy1998profile}
-Profile hidden Markov models.
-
-\citep{nawrocki2009infernal}
-Infernal 1.0: inference of RNA alignments
-
-\citep{felsenstein1981evolutionary}
-{Evolutionary trees from DNA sequences: a maximum likelihood approach}
-
-NJ
-\citep{price2010fasttree}
-{FastTree 2--approximately maximum-likelihood trees for large alignments}
-
-Living tree project.
-
-Bayesian phylogenetic inference is absent.
-
-However, it is surprising that so much focus is on complete de novo tree inference, and that methods such as DCM NNN have not gained traction.
-
-Networks not considered.
-
-The scale of the data has motivated strategies other than complete phylogenetic inference.
-Although such insertion has a long history as a means to sequentially build a phylogenetic tree NNN farris
-, the first software with insertion specifically as a goal was by \citep{ludwig2004arb} which works by parsimony.
-ARB is used to reconstruct a full tree.
-
-There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a tree.
-Other analogous methods have been used \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} with various speeds and features.
-Such questions have also found the interest of the alignment community as well.
-\citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
-\citet{mirarabsepp} does an approximately.
-
-There are other strategies still that do something related to phylogenetics.
-\citep{huson2007megan}
-LCA strategy
-
-\subsection{Software}
-
-People do most phylogenetic inferences as part of a pipeline.
-The most popular ones are QIIME \citep{caporaso2010qiime} and mothur \citep{schloss2009introducing}.
-AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
-
-mothur has ported in the clearcut \citep{evans2006relaxed,sheneman2006clearcut} program.
-QIIME wraps clearcut and \citep{price2010fasttree}.
-
-Databases:
-\citep{chen2010human}
-The Human Oral Microbiome Database: a web accessible resource for investigating oral microbe taxonomic and genomic information
-
-\citep{griffen2011core}
-CORE: a phylogenetically-curated 16S rDNA database of the core oral microbiome
-
-\citep{srinivasan2012bacterial}
+% Networks are not considered.
 
@@
 
-In addition to the usual challenges surrounding large-scale phylogenetic inference, there are some questions that are specific to the microbial setting.
-I have collected some of these questions here.
+I have collected some of questions motivated by human microbiome research.
 
-In this paper I have devoted considerable space to the ways in which microbial ecologists have used the 16s tree as a proxy structure for the evolutionary history of their favorite organisms.
+The first and most general challenge is to address the gap between on one hand complete \emph{de novo} tree inference, and sequence insertion or placement that leaves the reference tree fixed.
+Sequence data is continually being added to large databases, and so it would be nice to have methods that could continually update trees.
+% Surprising that methods such as DCM \cite{huson1999disk} have not gained traction.
+
+In this review I have devoted considerable space to the ways in which microbial ecologists have used the 16s tree as a proxy structure for the evolutionary history of their favorite organisms.
 They have even shown that 16S distance recapitulates gene content divergence and used this correlation to predict gene functions.
@@
 
-Finally, the conventional wisdom from the microbiome community that conclusions of a UniFrac analysis are robust to tree reconstruction methodology begs further exploration.
+Finally, the conventional wisdom from the microbiome community that UniFrac analysis is robust to tree reconstruction methodology begs further exploration.
 Would it be possible to infer an ``equivalence class'' of phylogenetic trees, where two trees are deemed equivalent if they induce the same principal coordinates projection with the same underlying data?

### 2013-09-28 sbreview.tex
First fleshed-out draft done.

@@
 
-Considerable attention has been given to how the microbiome influences host phenotype.
-IBD, etc, NNN.
-kwashiorkor \cite{smith2013gut}.
-There is strong evidence of interaction between the gut microbiome and obesity, but the story is yet clear.
+Considerable attention has been given to how the microbiome influences host phenotype \cite[reviewed in][]{cho2012human}.
+% claesson2012gut
+% kwashiorkor \cite{smith2013gut}.
+Considerable attention has been given to the interaction between the gut microbiome and obesity, but the story is yet clear.
 An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
@@
 Lots of survey, metagenome, and whole-genome sequencing data has been generated by the Human Microbiome Project \citep{methe2012framework} and is available on its website\footnote{\url{http://www.hmpdacc.org/}}.
-The METAHIT study also generated lots of data but it is not freely available. NNN
+The MetaHIT study also generated lots of data but it is not available to outside researchers.
 
@@
 For simply separating two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.
+In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \cite{caporaso2011moving}.
 
@@
 Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
-
-XXX Data through time.
 
@@
 The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved} which uses a heuristic algorithm to optimize $F$-measure of precision and recall.
-Interestingly, this latter tool allows for polyphyletic taxonomic groups as allowance for either lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis.
+Interestingly, this latter tool allows for polyphyletic taxonomic groups as allowance for either lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the authors note that this can occur from either phylogenetic error or the taxonomy deviating from the phylogeny.
 Our group \citep{matsen2012reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a previously described coloring problem.
@@
 An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
-There is also SAP.
-
 However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple na\"ive Bayes classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
@@
 The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy to turn a taxonomy and a collection of BLAST results of a unknown sequence on a database of sequences with those taxonomic labels into taxonomic assignments for the unknown sequence.
+Another program, called SAP \cite{munch2008fast,munch2008statistical} infers taxonomic assignment by automatically retrieving  and building a tree on retrieved sequences equipped with taxonomic information.
 
@@
 \subsection{The role of Operational Taxonomic Units (OTUs)}
-There continues to be a lively debate on if there is a meaningful concept of species for microbes NNN, however most microbial research has replaced any traditional species concept with the notion of an Operational Taxonomic Units (OTUs).
+There continues to be a lively debate on if there is a meaningful concept of species for microbes \cite{bapteste2009prokaryotic,caro2012bacterial}, and most human microbiome research has replaced any traditional species concept with the notion of an Operational Taxonomic Units (OTUs).
 An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly a 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
@@
 There are also methods that try to come up with sequence groupings that more reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
-The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type NNN, or that methods have been devised to find OTUs from non-overlapping sequences NNN.
+The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
 With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
@@
 Phylogenetic diversity notions have been applied on metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
-Although abundance-weighted non-phylogenetic diversity measures such as NNN simpson NNN shannon are among the most common, abundance-weighted phylogenetic diversity measures are generally not used.
-Such measures do exist \citep{chao2010phylogenetic}
+Although abundance-weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance-weighted phylogenetic diversity measures are generally not used.
+Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring} but have not seen application in human microbiome studies.
 We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}, in particular that they do a better job than the traditionally-used OTU-based measures.
@@
 \citet{wu2009phylogeny} have derived a ``phylogeny-driven genomic encyclopaedia of Bacteria and Archaea'' by selecting organisms for sequencing that are divergent from organisms that have a sequenced genome.
-By doing so they have recovered more novel open reading frames than they would have using other methods NNN.
+By doing so they have recovered more novel protein families than they would have using methods organized by taxonomy.
 However, for the equivalent effort for the human microbiome \cite{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
-NNN wasn't there a previous effort that used ppacer?
 % http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294
@@
 Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
-
-NNN look for gene transfer work in microbiome.
+\citet{stecher2012gut} find in a mouse model that horizontal transfer between pathogenic bacteria is blocked by commensal bacteria outside of periods of gut inflammation.
 
@@
 
+The approach of considering a collection of genes and their metabolic network as a meta-organism has yielded some interesting results \cite{borenstein2008large,greenblum2012metagenomic}.
+A clear limitation to this approach is that cellular boundaries are ignored: populations are not a freely diffusing soup.
+Could these approaches be improved by using phylogenetic methods to reconstruct the compartmentalization of genomes and processes into cells?
+% As an extension microbes have to be close to one another-- requires population modeling.
+
 Finally, the conventional wisdom from the microbiome community that UniFrac analysis is robust to tree reconstruction methodology begs further exploration.
@@
 \section{Discussion}
-What we can expect next.
-
-Future will give deeper sequencing, which can mean massive parallelization.
-Citizen microbiome projects.
-
-% Companies? http://www.ribocon.com/ second genome, etc
-
-Comparative studies of microbiomes.
-Microbiome analysis among bats describes influences of host phylogeny, life history, physiology and geography
-
-Clinical applications,
-\citep{clarridge2004}
-{{I}mpact of 16{S} r{RNA} gene sequence analysis for identification of bacteria on clinical microbiology and infectious diseases}
-
-For routine identification of a single microbe in culture, MALDI-TOF kicks ass and is cheap.
-Diagnostics involving mixed communities.
-
-Bioinformatics will get hardened into fixed workflows.
-
-Understand cause and effect of diseases.
-Take an ecological view on atopic dermititis.
-
-Although we have raised the possibility of using organism-level modeling via phylogeny
-There is the possibility that we don't have to think about individual organisms at all, but rather can think about the collection of genes and their metabolic network as a meta-organism (cite Elbo).
-Has limitation that you forget about cellular boundaries, which matter-- not a freely diffusing soup.
-Question: to what extent is modeling cellular-level evolutionary history interesting?
-As an extension microbes have to be close to one another-- requires population modeling.
+What can we expect next at the intersection of phylogenetics and the human microbiome?
+At least for the next several years we can expect more related work.
+Future will continue to bring deeper sequencing, which can mean massive parallelization.
+The uBiome project (http://ubiome.com/) promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
+Comparative studies will continue to study what shapes and is shaped by the microbiome.
+However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.
 
 There are limitations to what we can learn using genetics because more intricate processes such as gene regulation may be at play, limiting what sequence-level phylogenetics can do.
-However, evolution still acts and there is nothing saying that we couldn't model evolution of phenotype in some other way.
-
-Future work will need to include specific interactions.
-Host side is sometimes missing, but experiments are good.
-As described above, we can expect much more development in terms of specifics.
-\citep{hooper2012interactions}
+Future work may move from more general ecological models to models that include specific interactions between microbes and the host \citep[e.g.][]{hooper2012interactions}.
+MALDI-TOF is inexpensive, fast, and becoming accurate \citep{clark2013matrix}; it is thus likely to supplant 16S sequencing for routine identification of a single known microbe in culture.
+However, for complex commmunity-level diagnoses sequencing and consequent analysis methods will still be required \citep[see review][]{Rogers2013271}.
 
@@
 Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
-Although the human microbiome has gained a lot of attention recently, and in some cases too much attention, our microbes are here to stay and so is research on them.
+However, microbes are here to stay and so is research on them.
 Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.

### 2013-09-28 sbreview.tex
One... more... reference...

@@
 MALDI-TOF is inexpensive, fast, and becoming accurate \citep{clark2013matrix}; it is thus likely to supplant 16S sequencing for routine identification of a single known microbe in culture.
-However, for complex commmunity-level diagnoses sequencing and consequent analysis methods will still be required \citep[see review][]{Rogers2013271}.
+However, for complex commmunity-level diagnoses sequencing and consequent analysis methods will still be required \citep[see review][]{Rogers2013271}, and inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice \citep{didelot2012transforming}.
 

### 2013-09-28 sbreview.tex
Revised through section 3.

@@
 
-The human microbiome is the collection of microbes that live inside and on the outside of humans.
-Recent work including the Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortia has advanced human microboime research by generating thousands of 16s surveys and terabases of metagenomic data on the human microbiome.
-In this paper I review the impact of phylogenetics and tree-thinking on the methods used in the analysis of human microbiome data, as well as describing current challenges for phylogenetics coming from this type of work.
+The human microbiome is the collection of microbes that live inside and on the surface of humans.
+Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information of microbiome samples.
+Much of the analytical work for these sequences involves phylogenetics, at least indirectly, and methodology has developed in a somewhat different direction than for eukaryotic phylogenetics.
+In this paper I review the field and its methods from the perspective of a phylogeneticist, in particular the impact of tree-thinking on the methods used in the analysis of human microbiome data, as well as describing current challenges for phylogenetics coming from this type of work.
 
@@
 
-The parameter regime and focus of microbiome research sits outside of the traditional setting of eukaryotic phylogenetics, thus why should our community be interested in what these microbial ecologists and medical researchers have done?
-This system is data- and question-rich.
-It absolutely requires molecular methods, as microbes can be identified by their molecular sequence, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
-Indeed, just to do ecology you have to do sequencing.
+The parameter regime and focus of microbiome research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what these microbial ecologists and medical researchers have done?
+The answer is simple: this system is data- and question-rich.
+It absolutely requires molecular methods, as microbes are now primarily identified by their molecular sequence, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
+Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment.
 
-Although there appears to be an artificial divide between phylogeny as practiced as part of microbial ecology and that for multicellular organisms, however there are many parallels between the two enterprises.
+Although there is something of a divide between phylogeny as practiced as part of microbial ecology and that for multicellular organisms, there are many parallels between the two enterprises.
 Both communities struggle with issues of sequence alignment, large-scale tree reconstruction, species definition.
-However, approaches differ between the microbial ecology community and that of eukaryotic phylogenetics, in part because the former has access to a fantastic diversity of organisms directly through culture-independent methods, leading to additional problems above usual.
-The species concept is problematic for microbes and there is also considerable discussion concerning how to group them into species-like units.
-The diversity of these organisms is great and organizing them into a taxonomy is a serious challenge, especially in the absence of obvious morphological features.
+However, approaches differ between the microbial ecology community and that of eukaryotic phylogenetics, in part because the scope of the former contains an almost unlimited diversity of organisms, leading to additional problems above the usual.
+The species concept is even more problematic for microbes than for larger eukaryotes and there is also considerable discussion concerning how to group them into species-like units.
+The organizing microbes into a sensible taxonomy is a serious challenge, especially in the absence of obvious morphological features.
 
-Because of challenges with species definitions, microbial ecology researchers have developed their own techniques for doing comparing samples which are explicitly phylogenetic rather than requiring species concepts.
+Because of this high level of diversity and challenges with species definitions, microbial ecology researchers have developed their own explicitly phylogenetic techniques for doing comparing samples rather than comparing on the level of species abundances.
 Although there is some overlap with previous literature, these techniques are novel and could be used in a wider setting and deserve wider consideration by the phylogenetics community.
 
-The human microbiome part of microbial ecology is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
-Additionally, due to more than a century of hospital lab medicine, our knowledge about human-associated microbes is much richer than about microbes in general.
-It is relatively routine to manipulate the human microbiome or models thereof via intervention studies and germ-free animals.
+The human microbiome is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
+Additionally, due to more than a century of hospital lab medicine, our knowledge about human-associated microbes relatively rich.
+It is also common to manipulate the human microbiome or models thereof via intervention studies and germ-free animals, and is easier to do than for many other microbial communities.
 
-In this review I intend to inform the \textit{Systematic Biology} readership of phylogenetics-related research happening in microbial ecology and contrast approaches between the two community.
-I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
-I realize that there is substantial overlap-- for instance the microbial community is very interested in unicellular fungi-- but we will use this terminology with the associated caveats.
+In this review I will describe phylogenetics-related research happening in microbial ecology and contrast approaches between what I think of as the typical \emph{Systematic Biology} audience and microbial researchers.
+Despite an obvious oversimplification, I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
+I realize that there is substantial overlap-- for instance the microbial community is very interested in unicellular fungi, and many in the SB community do work on microbes-- but this terminology will be useful for concreteness.
 There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
@@
 I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
-I firmly believe that both fields would benefit from more interactions.
-I have made explicit effort to make a neutral comparison between two areas; indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes orders of magnitude larger than typical for eukaryotic phylogenetics.
+I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics; indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes orders of magnitude larger than typical for eukaryotic phylogenetics.
 
@@
 These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells.
-The microbiome has remarkable metabolic potential, with a set of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
-Much of our metabolic interaction with the outside world is mediated by our microbiome, as it has important roles in both vitamin
-% http://www.sciencedirect.com/science/article/pii/S095816691200119X
-and drug \citep{maurice2013xenobiotics} metabolism; our food and drug intake also impacts the diversity of microbes present.
+The microbiome has remarkable metabolic potential, with an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
+Much of our metabolic interaction with the outside world is mediated by our microbiome, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
 In this section I will briefly review what is known about the human microbiome and its effect on our health.
@@
 The human microbiome is an ecosystem.
-The microbiome is dynamic in terms of representation but apparently constant in terms of function \citep{hmp2012structure}.
-The ``core'' microbiome, or the microbiome shared between all humans, has been described \citep{turnbaugh2008core}.
-Human microbiome is spatially organized, as can be seen on skin \citep{grice2009topographical}, with substantial variation in human body habitats across space and time \citep{costello2009bacterial}.
+It is dynamic in terms of taxonomic representation but apparently constant in terms of function \citep{hmp2012structure}.
+There is a ``core'' microbiome which is shared between all humans \citep{turnbaugh2008core}.
+The human microbiome is spatially organized, as can be seen on skin \citep{grice2009topographical}, with substantial variation in human body habitats across space and time \citep{costello2009bacterial}.
 There is a substantial range of inter-individual versus intra-individual variation \citep{hmp2012structure}.
@@
 Our actions can shift the composition of our microbiome.
-Changes in diet very quickly shift its composition, and there is a strong correlation between long-term diet and microbiome \citep{li2009human,wu2011linking}.
+Changes in diet can very quickly shift its composition, but there is also a strong correlation between long-term diet and microbiome \citep{li2009human,wu2011linking}.
 Antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts for years \citep{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}.
 
-Considerable attention has been given to how the microbiome influences host phenotype \cite[reviewed in][]{cho2012human}.
-% claesson2012gut
-% kwashiorkor \cite{smith2013gut}.
-Considerable attention has been given to the interaction between the gut microbiome and obesity, but the story is yet clear.
+The microbiome interacts on many levels with host phenotype \citep[reviewed in][]{cho2012human}.
+For example, the gut microbiome correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to acute metabolic syndromes in rural Africa \citep{smith2013gut}.
+Considerable attention has also been given to the interaction between the gut microbiome and obesity, although the story is yet clear.
 An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
-An obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}.
-Pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}.
-However, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.
+A causal role for microbiome leading to phenotype is clear for mice: an obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}, pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
+However, these promising leads have not yet been confirmed causally or in population studies of humans \citep{zhao2013gut}.
+For example, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.
 
@@
 Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
+We will focus on bacteria here.
 
-In this paper we will be primarily be describing the human microbiome from the perspective of ecological dynamics.
-We will be ignoring recent research on fine-scale immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
-Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and with the human body, ecological methods will continue to have an important role.
+In this paper we will be primarily be describing the human microbiome from a large-scale phylogenetic perspective rather than from the fine-scale perpsective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
+Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and with the human body, a wide perspective will continue to be important.
 
 \section{Investigating the human microbiome via sequencing}
-It is now possible to assay microbial communities in high throughput using via sequencing.
+It is now possible to assay microbial communities in high throughput using sequencing.
 There are two ways of doing so.
-The first is to amplify a specific ``marker'' gene in the genome, and perform sequencing.
+The first is to amplify a specific gene in the genome for sequencing using polymerase chain reaction (PCR).
+Scientists typically pick a ``marker'' gene in that case suitable for phylogenetic analysis.
 The second is to randomly shear input DNA and/or RNA and then perform sequencing directly.
-Although these words are not consistently used in the literature, we will consistently refer to the former as a \textit{survey} and the second a \textit{metagenome}.
+We will consistently refer to the former as a \textit{survey} and the second a \textit{metagenome}, although these words have not always been consistently used in the literature.
 
-Lots of survey, metagenome, and whole-genome sequencing data has been generated by the Human Microbiome Project \citep{methe2012framework} and is available on its website\footnote{\url{http://www.hmpdacc.org/}}.
-The MetaHIT study also generated lots of data but it is not available to outside researchers.
+The Human Microbiome Project \citep{methe2012framework} generated lots of survey, metagenome, and whole-genome sequencing data and this data is available on a dedicated website\footnote{\url{http://www.hmpdacc.org/}}.
+The MetaHIT study \citep{qin2010human} also generated lots of data but it is not available to outside researchers.
 
@@
 Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
-Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
+Their work, and the scientists who followed them, focused on the 16S ribosomal gene as a genetic marker.
 This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
-Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which initiated surveys of all organisms regardless of whether they can be cultured.
+Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled surveys of all organisms regardless of whether they can be cultured.
 
@@
 This can be done with a high level of multiplexing, making an explicit trade-off between depth of sequencing for each specimen and the number of specimens able to be put on the sequencer at the same time.
-This has led to extensive parallelization, most recently by sequencing dozens of samples on the Illumina instrument \citep{degnan2011illumina,caporaso2012ultra}.
+This has led to extensive parallelization, most recently by sequencing dozens of samples at a time on the Illumina instrument \citep{degnan2011illumina,caporaso2012ultra}.
 This leads to the question of how many sequences are needed to characterize the microbial diversity of a given environment.
-For simply separating two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.
-In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \cite{caporaso2011moving}.
+To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.
+In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \citep{caporaso2011moving}.
 
@@
 ``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
-Worse, multiplex PCR is known to create chimeric sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}, and chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
+Worse, multiplex PCR is known to create chimeric sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
+Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
 Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb};
-recent work \citep{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been successful despite a moderate evolutionary signal \citep{klappenbach2000rrna}.
+recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal \citep{klappenbach2000rrna}.
 Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
@@
 \subsection{Metagenomes}
-As described above ``metagenome" means that data is sheared randomly across the genome, thus the genetic region of a read is unknown in addition to the organism it was derived from.
-Because it does proceed through an amplification step, it does not have the same primer biases as a marker gene survey, although sequencing is known to have biases.
+As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific region, thus the genetic region of a read is unknown in addition to the organism it was derived from.
+Because it does proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although sequencing is known to have biases.
 
 It is possible to use metagenomic data as an expanded set of marker genes.
-That is, one can use 16S reads that appear in the metagenome as well as other ``core" genes present in a large proportion of micro-organisms that are expected to follow the same evolutionary path \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}.
-Because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another, and even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
+That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes present in a large proportion of micro-organisms that are expected to follow the same evolutionary path \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}.
+However, because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another, and even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
+The rest of the data can be taxonomically classified
+\citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
 
-Metagenomic data can be used to infer information about metabolic capacity \cite{greenblum2012metagenomic,abubucker2012metabolic}.
-The authors of the metAMOS pipeline \citep{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
+Metagenomic data is often used to infer information of a metabolic rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
+Such inference is beyond the scope of this paper.
 
@@
 The Food and Drug Administration maintains GenomeTrakr, an openly accessible database of whole genome sequences from culture\footnote{\url{http://www.fda.gov/Food/FoodScienceResearch/WholeGenomeSequencingProgramWGS/}}.
-This data will become common for unculturable organisms as single-cell sequencing \citep[reviewed in][]{kalisky2011single} becomes common.
-The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \cite{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.
+This data may become common for unculturable organisms as single-cell sequencing methods \citep[reviewed in][]{kalisky2011single} improve.
+The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \citep{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.
 
@@
 The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy to turn a taxonomy and a collection of BLAST results of a unknown sequence on a database of sequences with those taxonomic labels into taxonomic assignments for the unknown sequence.
-Another program, called SAP \cite{munch2008fast,munch2008statistical} infers taxonomic assignment by automatically retrieving  and building a tree on retrieved sequences equipped with taxonomic information.
+Another program, called SAP \citep{munch2008fast,munch2008statistical} infers taxonomic assignment by automatically retrieving  and building a tree on retrieved sequences equipped with taxonomic information.
 
@@
 \subsection{The role of Operational Taxonomic Units (OTUs)}
-There continues to be a lively debate on if there is a meaningful concept of species for microbes \cite{bapteste2009prokaryotic,caro2012bacterial}, and most human microbiome research has replaced any traditional species concept with the notion of an Operational Taxonomic Units (OTUs).
+There continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, and most human microbiome research has replaced any traditional species concept with the notion of an Operational Taxonomic Units (OTUs).
 An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly a 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
@@
 By doing so they have recovered more novel protein families than they would have using methods organized by taxonomy.
-However, for the equivalent effort for the human microbiome \cite{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
+However, for the equivalent effort for the human microbiome \citep{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
 % http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294
@@
 The large data sets associated with human microbiome analysis require highly efficient algorithms for \emph{de novo} tree inference.
-Historically this has meant relaxed neighbor joining \cite{evans2006relaxed,sheneman2006clearcut}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
+Historically this has meant relaxed neighbor joining \citep{evans2006relaxed,sheneman2006clearcut}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
 People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in clearcut \citep{evans2006relaxed,sheneman2006clearcut} program, and QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree 2.
@@
 while the GreenGenes 13\_5 alignment has 1,262,986 sequences and is almost 80\% gap.
-The SILVA-associated 'all-species living tree' project \cite{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
+The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
 The GreenGenes tree is updated by running FastTree from scratch.
@@
 Sequence data is continually being added to large databases, and so it would be nice to have methods that could continually update trees.
-% Surprising that methods such as DCM \cite{huson1999disk} have not gained traction.
+% Surprising that methods such as DCM \citep{huson1999disk} have not gained traction.
 
@@
 In addition to using phylogenies to infer gene repertoire, it appears that neutral models involving phylogenetics could be more fully developed.
-Methods explicitly invoking trait evolution seem notably absent, with one recent exception \cite{langille2013predictive}.
+Methods explicitly invoking trait evolution seem notably absent, with one recent exception \citep{langille2013predictive}.
 The results of this simple method are reasonable.
@@
 A recent paper does so with an explicitly phylogenetic perspective \citep{o2012phylogenetic}, and includes some comparison of models to data.
-Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \cite{mooers1997inferring,aldous2011five}.
+Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \citep{mooers1997inferring,aldous2011five}.
 In this case various (alpha and beta) diversity statistics would take the role of tree shape statistics.
@@
 Diversity preservation is of interest for microbiome researchers, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms.
-Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer and has made passionate appeals to preserve microbiome diversity \cite{blaser2011antibiotic,cho2012human}.
+Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer and has made passionate appeals to preserve microbiome diversity \citep{blaser2011antibiotic,cho2012human}.
 Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
 Because microbiomes are seeded from mother to child, there is a somewhat equivalent notion of extinction of microbiome when the chain is interrupted.
-Consequently, a recent study has explicitly contrasted microbiome development in urban, forest-dwelling, and rural African populations \cite{yatsunenko2012human}, and one study has endeavored to characterize the microbiome from ancient feces \cite{tito2012insights}.
+Consequently, a recent study has explicitly contrasted microbiome development in urban, forest-dwelling, and rural African populations \citep{yatsunenko2012human}, and one study has endeavored to characterize the microbiome from ancient feces \citep{tito2012insights}.
 
@@
 
-The approach of considering a collection of genes and their metabolic network as a meta-organism has yielded some interesting results \cite{borenstein2008large,greenblum2012metagenomic}.
+The approach of considering a collection of genes and their metabolic network as a meta-organism has yielded some interesting results \citep{borenstein2008large,greenblum2012metagenomic}.
 A clear limitation to this approach is that cellular boundaries are ignored: populations are not a freely diffusing soup.

### 2013-09-28 sbreview.tex
Revision pass.

@@
 \usepackage{palatino}
+\usepackage[hidelinks]{hyperref}
 \usepackage[round]{natbib}
@@
 Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information of microbiome samples.
-Much of the analytical work for these sequences involves phylogenetics, at least indirectly, and methodology has developed in a somewhat different direction than for eukaryotic phylogenetics.
+Much of the analytical work for these sequences involves phylogenetics, at least indirectly, and methodology has developed in a somewhat different direction than for other applications of phylogenetics.
 In this paper I review the field and its methods from the perspective of a phylogeneticist, in particular the impact of tree-thinking on the methods used in the analysis of human microbiome data, as well as describing current challenges for phylogenetics coming from this type of work.
@@
 
-In this section I consider the ways in which phylogenetic thinking has impacted human microbiome research.
-What may be most interesting for this audience is the way in which phylogenetic trees are being used to actively revise taxonomy and used as a structure on which to perform sample comparison.
+In this section I consider the ways in which phylogenetics has impacted human microbiome research.
+What may be most interesting for this audience is the way in which phylogenetic trees are being used to actively revise taxonomy as well as being used as a structure on which to perform sample comparison.
 
@@
 Phylogenetic inference has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
-The most stunning example of this is the discovery that archaea, although morphologically similar to bacteria, form their own separate lineage \citep{woese1977phylogenetic}.
+The clearest such example is the discovery that archaea, although morphologically similar to bacteria, form their own separate lineage \citep{woese1977phylogenetic}.
 
-In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
-These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed to preserve information associated with taxonomic classifications.
+Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
+These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed to preserve information associated with taxonomic classifications.
 Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process is somewhat opaque \citep{kreig1984bergey}.
-The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved} which uses a heuristic algorithm to optimize $F$-measure of precision and recall.
-Interestingly, this latter tool allows for polyphyletic taxonomic groups as allowance for either lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the authors note that this can occur from either phylogenetic error or the taxonomy deviating from the phylogeny.
-Our group \citep{matsen2012reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a previously described coloring problem.
+The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
+Tax2tree uses a heuristic algorithm to optimize $F$-measure of precision and recall for its taxonomic assignments.
+Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
+Our group \citep{matsen2012reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
+Although it is wonderful that several groups are actively working on taxonomic revision, the existence of multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database can be frustrating.
 
 An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
-However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple na\"ive Bayes classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
+However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
 % \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
 Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
-
-The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy to turn a taxonomy and a collection of BLAST results of a unknown sequence on a database of sequences with those taxonomic labels into taxonomic assignments for the unknown sequence.
-Another program, called SAP \citep{munch2008fast,munch2008statistical} infers taxonomic assignment by automatically retrieving  and building a tree on retrieved sequences equipped with taxonomic information.
-
-A fascinating approach is that of \citet{segata2012metagenomic}, who start by compiling a database of clade-specific genes \citep{segata2011metagenomic}.
-Then they classify the origin of a given read as being the only clade that has this gene.
-They show that this has good specificity, however, sensitivity is limited to genomes that have been sequenced.
+The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy to turn a taxonomy and a collection of BLAST results for an unknown sequence on a database of sequences with those taxonomic labels into taxonomic assignments for the unknown sequence.
+Another program, called SAP, infers taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence \citep{munch2008statistical,munch2008fast}.
+\citet{segata2012metagenomic} propose a different approach to inferring organisms present in a metagenomic sample from a metagenome by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
+They show that this has good sensitivity and specificity, however, this method can only be used to classify to organisms that have complete genome sequences.
 
 \subsection{The role of Operational Taxonomic Units (OTUs)}
-There continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, and most human microbiome research has replaced any traditional species concept with the notion of an Operational Taxonomic Units (OTUs).
-An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly a 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
+There continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, and most human microbiome research has replaced any traditional species concept with the notion of Operational Taxonomic Units (OTUs).
+An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly at 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
 It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.
+I briefly describe the mini-industry of OTU clustering techniques to contrast with the phylogenetic literature on species delimitation \citep{pons2006sequence,yang2010bayesian}.
 
-There is an entire industry of OTU inference software with various strategies of various speeds; we will not attempt a systematic review here.
-Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch}.
-These methods are very fast and heuristic, in that they are described as an algorithm rather than a routine attempting to optimize some notion of goodness of clustering.
+\textbf{[Figure: \url{http://i.imgur.com/3rs1bND.png}]?}
+
+There are many OTU clustering strategies of various speeds.
+Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
+These methods are very fast and are heuristic in that they are described as an algorithm rather than as global optimization of some notion of goodness of clustering.
 It is well known that different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
-There are also methods that try to come up with sequence groupings that more reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
+There are also methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
 The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
@@
 \subsection{Diversity estimates using phylogenetics}
-Because 16s surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
-For the most part, this literature adapts methods from the classical ecological literature replacing taxonomic groupings with OTUs.
+Because 16S surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
+For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
 %The most common diversity indices are Simpson and Shannon.
@@
 
-Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
-Phylogenetic diversity notions have been applied on metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
-Although abundance-weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance-weighted phylogenetic diversity measures are generally not used.
-Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring} but have not seen application in human microbiome studies.
-We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \citep{mccoy2013abundance}, in particular that they do a better job than the traditionally-used OTU-based measures.
-We have also looked at the mean and variance of PD under random subsampling \citep{nipperess2013mean}.
+Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (see next section), phylogenetic alpha diversity seems under-developed: whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve phylogenetic diversity (PD) measures.
+Faith's PD \citep{faith1992conservation} has been applied to some 16S survey data \cite{lozupone2007global,costello2009bacterial} and to metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
+
+Although abundance-weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance-weighted phylogenetic diversity measures are not used in human microbiome studies.
+Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
+\citet{mccoy2013abundance} have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome, in particular that they do a better job than the traditionally-used OTU-based measures.
+\citet{nipperess2013mean} have also determined the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.
+
+\textbf{[Figure: \url{http://cl.ly/image/3k3n3W3U443G/ref_tree_bwpd.png} to show a partially abundance weighted PD measure?]}
 
 \subsection{Community comparison using phylogenetics}
-The level of similarity between samples or groups thereof called \emph{beta diversity}.
-As with alpha diversity, it is common to use classical measures \citep[e.g.][]{jaccard1908nouvelles} applied to OTU counts, however in this case phylogenetics-based methods are widely used.
-They generally go under variants of ``UniFrac'' as named by Lozupone and Knight \citep{LozuponeKnightUniFrac05;LozuponeEaWeightedUnifrac07}.
-As evidence of their superiority
-\citep{kuczynski2010microbial} Microbial community resemblance methods differ in their ability to detect biologically relevant patterns
+The level of similarity between samples or groups thereof is called \emph{beta diversity}.
+As with alpha diversity, it is not uncommon to use classical measures \citep[e.g.][]{jaccard1908nouvelles} applied to OTU counts, however phylogenetics-based methods are the most popular.
+They are generally called variants of the ``UniFrac'' dissimilarity as so named by \citet{LozuponeKnightUniFrac05}.
+\citet{kuczynski2010microbial} claim that they are superior to other methods for community comparison using real data and simulations \citep[for a contrary viewpoint using simulations see][]{schloss2008evaluating}.
+% It recovers gradients. \citep{nemergut2011global}
 
-The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample or another rather than between samples.
-Weighted UniFrac is an abundance weighted version.
-We have shown that weighted UniFrac is in fact a special case of the earth-movers distance, and is special in that it can be calculated in linear time \citep{evans2012phylogenetic}.
-We also showed that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
+\textbf{[Figure: canonical unifrac explanation picture?]}
+
+The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample or another via the shortest path rather than edges that connect between samples.
+Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
+\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a special case of the earth-movers distance, and is special in that it can be calculated in linear time, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
 Recent work \citep{chen2012associating} has shown that a partially abundance-weighted variant of UniFrac has greater power to resolve community differences than either unweighted or weighted UniFrac.
 
+\textbf{[Figure: \url{http://cl.ly/image/3K472T0K3S3S/tree_dirtpiles.png}]?}.
+
 The most common way to use a pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
-It recovers gradients.
-\citep{nemergut2011global}
-?? Global patterns in the biogeography of bacterial taxa
-Indeed, the observation of differences between two communities is often used as prima fasce evidence of a difference between them, while the lack of such a difference indicates that the communities are not different (amish?)
+Indeed, the separation of two communities in a principal components plot is often used as \emph{prima facie} evidence of a difference between them
+\citep[e.g.][]{lozupone2007global,costello2009bacterial,yatsunenko2012human}, while the lack of such a difference is interpreted as showing that the communities are not different overall.
 
-Figure: example principal coordinates plot.
+\textbf{[Figure: example principal coordinates plot?]}
 
 There have been efforts to augment the ordination visualizations with additional information giving more structure to the visualizations.
-Biplot.
-Elizabeth Purdom \citep{BikEaMicrobiotaStomach06,PurdomAnalyzingDataGraphs08} has incorporated.
-We have developed an alternative ordination process that explicitly labels the axes with weightings on phylogenetic trees that indicate their meanings \citep{matsen2013edge}.
+Biplots display variables (in the microbial case summarized by taxonomic labels) as points in addition to samples \cite[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
+Elizabeth Purdom \citep{BikEaMicrobiotaStomach06,PurdomAnalyzingDataGraphs08} describes how generalized principal component eigenvectors can be interpreted via weightings on the leaves of a phylogenetic tree.
+\citet{matsen2013edge} have developed a variant of principal components analysis that explicitly labels the axes with weightings on phylogenetic trees that indicate their meanings.
 
@@
 
-16S distance is frequently used as a proxy for a functional comparison.
-Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes is commonly acquired horizontally as well as vertically, and horizontal transmission may leave no trace in the 16S ancestry.
+16S distance is frequently used as a proxy for a functional comparison between human microbiome samples.
+Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes is commonly acquired horizontally as well as vertically, and horizontal transmission leaves no trace in the 16S ancestry.
 
-However, \citep{zaneveld2010ribosomal} have shown that organisms that are more distant in terms of 16S are also more divergent in terms of gene repertoire.
-Such observations generally follow a nonlinear curve, and the extent to which they lie on the curve appears to be phylum-dependent.
-This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} by either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
+However, \citet{zaneveld2010ribosomal} have shown that organisms that are more distant in terms of 16S are also more divergent in terms of gene repertoire.
+Such observations surround a fit nonlinear curve, and the extent to which they lay on the curve appears to be phylum-dependent.
+This ``proxy'' approach has recently been taken to its logical conclusion by methods to infer functional characteristics from a 16S sample using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} by either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
 
@@
 By doing so they have recovered more novel protein families than they would have using methods organized by taxonomy.
-However, for the equivalent effort for the human microbiome \citep{fodor2012most} list phylogenetic methods were not used-- they just wanted to get microbes that were prevalent in the microbiome but not sequenced.
+In a similar effort for the human microbiome \citep{fodor2012most} phylogenetic results were not shown although the authors state that phylogenetic methods did give similar results to their analysis.
 % http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294
@@
 \subsection{Genome-scale inquiries using phylogenetics}
-With some notable exceptions, mainstream applications of phylogenetics have typically been with the idea of finding ``the'' tree of a collection of human-associated microbes rather than explicitly exploring divergence between various gene trees.
+With some notable exceptions, mainstream applications of phylogenetics to a collection of human-associated microbes have typically been with the idea of finding ``the'' tree of such a collection rather than explicitly exploring divergence between various gene trees.
 As described above, whole-genome data is typically used to directly infer functional information rather than information concerning ancestry.
-The debate concerning whether a microbial tree of life is a useful concept does not seem to have dampened human microbiome researchers' enthusiasm for using such a tree.
+The debate concerning whether a microbial tree of life is a useful concept \citep{bapteste2009prokaryotic,caro2012bacterial} does not seem to have dampened human microbiome researchers' enthusiasm for using a single such tree.
 
 Nevertheless, the work that has been done to infer horizontal gene transfer has revealed interesting results.
-\citet{hehemann2010transfer} found that a gene found in seaweed has been transferred into the microbiome of Japanese such that individuals with this resulting microbiome are better able to digest algae.
+\citet{hehemann2010transfer} found that a gene found in seaweed has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
 Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
-\citet{stecher2012gut} find in a mouse model that horizontal transfer between pathogenic bacteria is blocked by commensal bacteria outside of periods of gut inflammation.
+\citet{stecher2012gut} find that in a mouse model, horizontal transfer between pathogenic bacteria is blocked by commensal bacteria except for periods of gut inflammation.
 
@@
 In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.
-This can be defended by saying for many applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
-Furthermore, given that the underlying data is typically 16S only we can expect some tree problems even with the best methods.
+This is defended by saying for applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
+Furthermore, given that the underlying data is typically 16S only we can expect some topological inaccuracy even with the best methods.
 Additionally, as specified below, these data sets can be very large.
@@
 
-Alignment methods are primarily focused on extending a curated ``seed alignment'' with additional sequences; tools have been created with exactly this application for 16S in mind \citep{desantis2006nast,caporaso2010pynast,pruesse2012sina}.
-The community also uses profile hidden Markov models \citep{eddy1998profile} and CM models \citep{nawrocki2009infernal,nawrocki2009structural}.
+Alignment methods are primarily focused on extending a curated ``seed alignment'' by adding additional sequences; several tools have been created with exactly this application for 16S in mind \citep{desantis2006nast,caporaso2010pynast,pruesse2012sina}.
+The community also uses profile hidden Markov models \citep{eddy1998profile} and CM models \citep{nawrocki2009infernal,nawrocki2009structural} to achieve the same result.
 
 The large data sets associated with human microbiome analysis require highly efficient algorithms for \emph{de novo} tree inference.
-Historically this has meant relaxed neighbor joining \citep{evans2006relaxed,sheneman2006clearcut}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
-People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in clearcut \citep{evans2006relaxed,sheneman2006clearcut} program, and QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree 2.
+Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
+People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut \citep{sheneman2006clearcut} program, and QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree 2.
 % AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
@@
 The scale of the data has motivated strategies other than complete phylogenetic inference.
-Although such insertion has a long history as a means to sequentially build a phylogenetic tree \citep{kluge1969quantitative}, the first software with insertion specifically as a goal was by \citep{ludwig2004arb} which works by parsimony.
+Although such insertion has a long history as a means to sequentially build a phylogenetic tree \citep{kluge1969quantitative}, the first software with insertion specifically as a goal was ARB by \citet{ludwig2004arb} which works by parsimony.
 ARB is commonly used to reconstruct a full tree by direct insertion.
 
-There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a tree, sometimes with uncertainty estimates.
-Other analogous methods have been used \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} with various speeds and features.
-Such questions have also found the interest of the alignment community as well.
+There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a so-called fixed \textit{reference tree}, sometimes with uncertainty estimates.
+These programs \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} have various speeds and features.
+This work has also spurred development of specialized alignment tools for this mapping process.
 \citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
-\citet{mirarabsepp} uses data set partitioning to improve alignments on subsets of taxa.
+\citet{mirarabsepp} use data set partitioning to improve alignments on subsets of taxa.
 
 Considerable effort goes to the creation of large curated alignments and phylogenetic trees on 16S.
-There are two primary such projects: one is the SILVA project \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes \citep{desantis2006greengenes,mcdonald2011improved} database.
+There are two primary such projects: one is the SILVA database \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes database \citep{desantis2006greengenes,mcdonald2011improved}.
 Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
@@
 % .96353230536373096207
-SILVA reference alignment version 115 has 479,726 sequences and is over 96\% gap,
+the 479,726 sequence SILVA reference alignment version 115 is over 96\% gap,
 % GreenGenes
@@
 % .79914084873730799272
-while the GreenGenes 13\_5 alignment has 1,262,986 sequences and is almost 80\% gap.
+while the 1,262,986 sequence GreenGenes 13\_5 alignment has is almost 80\% gap.
 The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
 The GreenGenes tree is updated by running FastTree from scratch.
-In addition to these 16S-based resources, the MicrobesOnline resource \citep{dehal2010microbesonline} offers a tree-based genome browser.
+For some reason, some believe that FastTree in particular works well even with such gappy alignments \citep[e.g.][]{sharpton2011phylotu}.
+
+In addition to these 16S-based resources, the MicrobesOnline resource \citep{dehal2010microbesonline} offers a very nice interactive tree-based genome browser.
 On a much smaller scale, there curated reference sets that are microbiome body-site specific \citep{chen2010human,griffen2011core,srinivasan2012bacterial}
@@
 
-I have collected some of questions motivated by human microbiome research.
-
-The first and most general challenge is to address the gap between on one hand complete \emph{de novo} tree inference, and sequence insertion or placement that leaves the reference tree fixed.
-Sequence data is continually being added to large databases, and so it would be nice to have methods that could continually update trees.
+One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference, and sequence insertion or placement that leaves the ``reference tree'' fixed.
+These two extremes are each each useful in different parameter regimes, but something between the two would be very helpful as well.
+For example, sequence data is continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
 % Surprising that methods such as DCM \citep{huson1999disk} have not gained traction.
 
-In this review I have devoted considerable space to the ways in which microbial ecologists have used the 16s tree as a proxy structure for the evolutionary history of their favorite organisms.
+In this review I have devoted considerable space to the ways in which microbial ecologists have used the 16S tree as a proxy structure for the complete evolutionary history of their favorite organisms.
 They have even shown that 16S distance recapitulates gene content divergence and used this correlation to predict gene functions.
-16s tree has taken us remarkably far, but it is not a complete representation of the evolutionary history of these microbes.
+It is well known, however, that any single tree will not give a complete representation of the evolutionary history of a collection of microbes.
 
-The apparent success of 16s tree based comparisons begs the question of if a more complete representation of the evolutionary history of the microbes would yield better predictions.
-This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of an organism that allows us to best predict features of underlying genomes?
+The apparent success of 16S-tree-based comparisons begs the question of if a more complete representation of the evolutionary history of the microbes would yield better comparisons.
+This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of a set of organisms that allows us to best predict features of underlying genomes?
 This representation could be simple.
 For example, one of the results of \citet{zaneveld2010ribosomal} is that 16S correlates better with gene repertoire in some taxonomic groups than others.
-If we were to equip the 16S tree with some measure of the strength of that correlation, would that allow for more accurate prediction?
-Would a species tree reconciled with a collection of gene trees allow for better prediction?
-It is quite possible that there is nothing better given the inherent noise of the data, but if we allow an arbitrary ``hidden'' object, what such object would perform best?
-For example, collections of reconciled gene trees in the presence of gene deletion, transfer, and loss, and could be used \citep[see][for interesting recent results]{szollHosi2013lateral,szollHosi2013efficient}.
+If we were to equip the 16S tree with some measure of the strength of that correlation, would that allow for more precise comparison?
+It is quite possible that these objects would not be able to overcome the inherent noise of the data, but if we allow an arbitrary ``hidden'' object, what such object would perform best?
+For example, collections of reconciled gene trees in the presence of gene deletion, transfer, and loss \citep[see][for interesting recent results]{szollHosi2013efficient,szollHosi2013lateral} could be used.
 
@@
 Methods explicitly invoking trait evolution seem notably absent, with one recent exception \citep{langille2013predictive}.
-The results of this simple method are reasonable.
-It is possible that improved methods, perhaps involving whole-genome evolutionary modeling or models of metabolic network evolution, could shed light on the problem.
-How could we formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on a collection of reconciled gene trees?
+The results of this simple method are reasonable, but would a species tree reconciled with a collection of gene trees allow for better prediction?
+Perhaps improved methods, say involving whole-genome evolutionary modeling or models of metabolic network evolution, could shed light on the problem.
+How would we formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on a collection of reconciled gene trees?
 
-Given that microbes are cosmopolitan and have been evolving for a long time, another important project is to model community assembly.
-One direction is to apply Hubbell's neutral theory to the human microbiome \citep{fierer2012animalcules,costello2012application}.
-A recent paper does so with an explicitly phylogenetic perspective \citep{o2012phylogenetic}, and includes some comparison of models to data.
+An important project is to model community assembly, and perhaps further phylogeny-aware methods could be used.
+One way to model community assembly is to apply Hubbell's neutral theory to the human microbiome \citep{fierer2012animalcules,costello2012application}.
+\citet{o2012phylogenetic} model community assembly with an explicitly phylogenetic perspective, and include some comparison of models to data.
 Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \citep{mooers1997inferring,aldous2011five}.
-In this case various (alpha and beta) diversity statistics would take the role of tree shape statistics.
-It will be challenging to bring together macroevolutionary modeling with species abundance modeling, but some very first steps have been made in another setting \citep{lambert2013predicting}.
+In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics.
+It will be challenging to bring together macroevolutionary modeling with species abundance modeling, but some initial steps have been made in another setting \citep{lambert2013predicting}.
 
-Diversity preservation is of interest for microbiome researchers, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms.
-Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer and has made passionate appeals to preserve microbiome diversity \citep{blaser2011antibiotic,cho2012human}.
+Diversity preservation is of interest for microbiome researchers, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms \citep{hartmann2006maximizing,pardi2007resource}.
+Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer \citep[reviewed in][]{cho2012human} and has made passionate appeals to preserve microbiome diversity \citep{blaser2011antibiotic}.
 Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
 Because microbiomes are seeded from mother to child, there is a somewhat equivalent notion of extinction of microbiome when the chain is interrupted.
-Consequently, a recent study has explicitly contrasted microbiome development in urban, forest-dwelling, and rural African populations \citep{yatsunenko2012human}, and one study has endeavored to characterize the microbiome from ancient feces \citep{tito2012insights}.
+Consequently, \citet{yatsunenko2012human} have explicitly contrasted microbiome development in urban, forest-dwelling, and rural populations, while \citet{tito2012insights} have endeavored to characterize the microbiome from ancient feces.
 
 Perhaps in part because of the importance of mother-to-child transmission, there are indications of coevolution between microbiomes and their hosts.
-\citet{ochman2010evolutionary} found identical trees for primate evolution and microbiome,
+\citet{ochman2010evolutionary} found identical trees for primate and microbiome evolution.
 They used maximum parsimony such that each column was a microbe and in each such entry took discrete states according to how much of that microbe was present.
@@
 Other studies \citep[e.g.][]{phillips2012microbiome,delsuc2013convergence} see a combination of historical and dietary influences.
-This seems like a setting in which convergent evolution could be quantified.
+How can these forces be compared in this setting?
 
@@
 
-Finally, the conventional wisdom from the microbiome community that UniFrac analysis is robust to tree reconstruction methodology begs further exploration.
-Would it be possible to infer an ``equivalence class'' of phylogenetic trees, where two trees are deemed equivalent if they induce the same principal coordinates projection with the same underlying data?
-Given that a tree is an integral part of a UniFrac analysis, it would also be interesting to be able to infer the features of a tree that determine the primary trends in a projection.
-% How bad of tree inference can we do and still get the same trends?
+Finally, the conventional wisdom that UniFrac analysis is robust to tree reconstruction methodology begs further exploration.
+Would it be possible to infer an ``equivalence class'' of phylogenetic trees, where two trees are deemed equivalent if they induce the same principal coordinates projection given the same underlying presence/absence or count data?
+Given that a tree is an integral part of a UniFrac analysis, it would be interesting to be able to infer the features of a tree that determine the primary trends in a projection.
 
@@
 At least for the next several years we can expect more related work.
-Future will continue to bring deeper sequencing, which can mean massive parallelization.
-The uBiome project (http://ubiome.com/) promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
+Future will continue to bring deeper sequencing on more samples.
+The uBiome project (\url{http://ubiome.com/}) promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
 Comparative studies will continue to study what shapes and is shaped by the microbiome.
@@
 There are limitations to what we can learn using genetics because more intricate processes such as gene regulation may be at play, limiting what sequence-level phylogenetics can do.
-Future work may move from more general ecological models to models that include specific interactions between microbes and the host \citep[e.g.][]{hooper2012interactions}.
-MALDI-TOF is inexpensive, fast, and becoming accurate \citep{clark2013matrix}; it is thus likely to supplant 16S sequencing for routine identification of a single known microbe in culture.
-However, for complex commmunity-level diagnoses sequencing and consequent analysis methods will still be required \citep[see review][]{Rogers2013271}, and inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice \citep{didelot2012transforming}.
+Future work may move from more general ecological models to models that include specific interactions between microbes and the host \citep[reviewed in][]{hooper2012interactions}.
 
-Although I have described some areas for future research for phylogeneticists, researchers should be warned that their work may not be widely accepted by the microbial ecology community.
-That is, methodological changes that would be viewed as improvements in, substitution modeling for example, may not readily be taken up by the microbial ecology community.
-My sense is that this is in part because the bulk of microbial ecology researchers are doctors, immunologists, and oceanographers that have an applied focus and for which computational and modeling issues are a nuisance rather than an interesting aspect.
-For this reason, computational analysis is primarily done in advance or as part of established pipelines and is hardened into standard operating procedures \citep{peplies2008standard}.
+Opportunities for clinical applications may present themselves, but probably only some of these will be sequencing based.
+For example, 16S sequencing is likely soon to be completely replaced by Matrix-Assisted Laser Desorption Ionization--Time of Flight (MALDI-TOF) mass spectrometry for routine identification of a single known microbe grown in culture \citep{clark2013matrix}.
+However, for complex commmunity-level diagnoses sequencing and consequent analysis methods will still be required \citep[see review in][]{Rogers2013271}, and inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}.
 
-Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes it can feel like researchers in this area have forgotten some of the basic principles of study design, such as power calculations, that are well recognized in other areas of medical research.
+Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes it can feel like the hype has outmatched the reality.
 However, microbes are here to stay and so is research on them.
 Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
-Phylogenetics has contributed significantly to ecology in the past, and indeed has already contributed significantly to microbial ecology; the next generation of methods may require integration of genome-level data.
+Phylogenetics has already contributed significantly to human microbiome research and will continue to do so.
 
 \section{Acknowledgements}
-Aaron Darling, Connor McCoy,
-FHCRC folk.
-ATD grant.
+The author would like to thank Aaron Darling, David Fredricks, Noah Hoffman, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review.
+FAM is supported in part by NIH R01 HG005966-01 and NSF Award 1223057.
 

### 2013-09-28 post_build.sh
Build script, etc.

@@ -1,0 +1,7 @@
+date=`date +%m-%d-%Hh%M`
+name=sbreview-$date.pdf
+
+cp _build/sbreview.pdf $name
+url=`cloudapp $name | tail -n 1`
+echo "* [$date]($url)" >> README.md
+rm $name

### 2013-09-28 sbreview.tex
Thank Olivier.

@@
 \section{Acknowledgements}
-The author would like to thank Aaron Darling, David Fredricks, Noah Hoffman, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review.
-FAM is supported in part by NIH R01 HG005966-01 and NSF Award 1223057.
+The author would like to thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
+He is also grateful to Aaron Darling, David Fredricks, Noah Hoffman, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, and to support from the NIH (R01 HG005966-01) and NSF (1223057).
 

### 2013-09-30 sbreview.tex
Pass through the introduction.

@@
 It absolutely requires molecular methods, as microbes are now primarily identified by their molecular sequence, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
-Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment.
+Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sequences derived from the sample.
 
-Although there is something of a divide between phylogeny as practiced as part of microbial ecology and that for multicellular organisms, there are many parallels between the two enterprises.
-Both communities struggle with issues of sequence alignment, large-scale tree reconstruction, species definition.
+Although there is something of a divide between phylogeny as practiced as part of microbial ecology on one hand and that for multicellular organisms on the other, there are many parallels between the two enterprises.
+Both communities struggle with issues of sequence alignment, large-scale tree reconstruction, and species definition.
 However, approaches differ between the microbial ecology community and that of eukaryotic phylogenetics, in part because the scope of the former contains an almost unlimited diversity of organisms, leading to additional problems above the usual.
 The species concept is even more problematic for microbes than for larger eukaryotes and there is also considerable discussion concerning how to group them into species-like units.
-The organizing microbes into a sensible taxonomy is a serious challenge, especially in the absence of obvious morphological features.
+Organizing microbes into a sensible taxonomy is a serious challenge, especially in the absence of obvious morphological features.
 
 Because of this high level of diversity and challenges with species definitions, microbial ecology researchers have developed their own explicitly phylogenetic techniques for doing comparing samples rather than comparing on the level of species abundances.
-Although there is some overlap with previous literature, these techniques are novel and could be used in a wider setting and deserve wider consideration by the phylogenetics community.
+Although there is some overlap with previous literature, these techniques and could be used in a wider setting and may deserve broader consideration by the phylogenetics community.
 
 The human microbiome is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
-Additionally, due to more than a century of hospital lab medicine, our knowledge about human-associated microbes relatively rich.
+Additionally, due to more than a century of hospital laboratory work, our knowledge about human-associated microbes relatively rich.
 It is also common to manipulate the human microbiome or models thereof via intervention studies and germ-free animals, and is easier to do than for many other microbial communities.
 
-In this review I will describe phylogenetics-related research happening in microbial ecology and contrast approaches between what I think of as the typical \emph{Systematic Biology} audience and microbial researchers.
+In this review I will describe phylogenetics-related research happening in microbial ecology and contrast approaches between microbial researchers and what I think of as the typical \emph{Systematic Biology} audience.
 Despite an obvious oversimplification, I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
-I realize that there is substantial overlap-- for instance the microbial community is very interested in unicellular fungi, and many in the SB community do work on microbes-- but this terminology will be useful for concreteness.
+I realize that there is substantial overlap-- for instance the microbial community is very interested in unicellular fungi, and additionally many in the SB community do work on microbes-- but this terminology will be useful for concreteness.
 There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
-I will first briefly review the recent literature on the human microbiome, then describe novel ways in which human microbiome researchers have used trees and the challenges that human microbiome research poses for the phylogenetics community.
+I will first briefly review the recent literature on the human microbiome, then describe novel ways in which human microbiome researchers have used trees.
 I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
@@
 \subsection{The role of Operational Taxonomic Units (OTUs)}
-There continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, and most human microbiome research has replaced any traditional species concept with the notion of Operational Taxonomic Units (OTUs).
+Although there continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, most human microbiome research has replaced any traditional species concept with the notion of Operational Taxonomic Units (OTUs).
 An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly at 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
@@
 There are also methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
+\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values.
 The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
@@
 Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (see next section), phylogenetic alpha diversity seems under-developed: whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve phylogenetic diversity (PD) measures.
-Faith's PD \citep{faith1992conservation} has been applied to some 16S survey data \cite{lozupone2007global,costello2009bacterial} and to metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
+Faith's PD \citep{faith1992conservation} has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
 
@@
 There have been efforts to augment the ordination visualizations with additional information giving more structure to the visualizations.
-Biplots display variables (in the microbial case summarized by taxonomic labels) as points in addition to samples \cite[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
+Biplots display variables (in the microbial case summarized by taxonomic labels) as points in addition to samples \citep[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
 Elizabeth Purdom \citep{BikEaMicrobiotaStomach06,PurdomAnalyzingDataGraphs08} describes how generalized principal component eigenvectors can be interpreted via weightings on the leaves of a phylogenetic tree.

### 2013-09-30 sbreview.tex
An editorial pass.

@@
 The microbiome interacts on many levels with host phenotype \citep[reviewed in][]{cho2012human}.
-For example, the gut microbiome correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to acute metabolic syndromes in rural Africa \citep{smith2013gut}.
+For example, the gut microbiome correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to children with acute metabolic syndromes in rural Africa \citep{smith2013gut}.
 Considerable attention has also been given to the interaction between the gut microbiome and obesity, although the story is yet clear.
 An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
-A causal role for microbiome leading to phenotype is clear for mice: an obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}, pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
+A causal role for microbiome leading to phenotype is established for mice: an obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}, the pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
 However, these promising leads have not yet been confirmed causally or in population studies of humans \citep{zhao2013gut}.
@@
 Bacteria have been the primary focus of human microbiome research, and other domains have been investigated though to a lesser extent.
-Changes in archaeal and fungi have been shown to covary with bacterial residents \citep{hoffmann2013archaea}.
+Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea}.
 Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
@@
 
-In this paper we will be primarily be describing the human microbiome from a large-scale phylogenetic perspective rather than from the fine-scale perpsective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
-Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and with the human body, a wide perspective will continue to be important.
+In this paper we will be primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perpsective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
+Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and with the human body, a broad perspective will continue to be important.
 
@@
 To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.
-In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \citep{caporaso2011moving}.
+In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \citep[e.g.][]{caporaso2011moving}.
 
-Despite the high-throughput and low cost of modern sequencing, inherent challenges remain for the use of population census by marker gene sequencing.
-Most fundamentally, various microbes have different DNA extraction efficiencies, even with hard core protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
+Despite the high throughput and low cost of modern sequencing, inherent challenges remain for the use of population census by marker gene sequencing.
+Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
 Current sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
 ``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
-Worse, multiplex PCR is known to create chimeric sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
+Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
 Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
-Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb};
-recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal \citep{klappenbach2000rrna}.
+Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
+Recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
 Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
@@
 \subsection{Metagenomes}
-As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific region, thus the genetic region of a read is unknown in addition to the organism it was derived from.
+As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific region, and thus the genetic region of a read is unknown in addition to the organism it was derived from.
 Because it does proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although sequencing is known to have biases.
@@
 It is possible to use metagenomic data as an expanded set of marker genes.
-That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes present in a large proportion of micro-organisms that are expected to follow the same evolutionary path \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}.
-However, because of the diversity of gene repertoire in microbes, the gene sets may have only limited overlap with one another, and even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
-The rest of the data can be taxonomically classified
-\citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
+That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes present in a large proportion of micro-organisms that are expected to follow the same evolutionary path.
+This is proven to be a useful strategy \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}, however, because of the diversity of gene repertoire in microbes, this core gene set may be relatively small.
+Indeed, even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
+The rest of the data can be taxonomically classified \citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
 
@@
 Whole-genome sequencing from culture is currently being used for microbial outbreak tracking \citep{koser2012rapid,snitkin2012tracking}.
-The Food and Drug Administration maintains GenomeTrakr, an openly accessible database of whole genome sequences from culture\footnote{\url{http://www.fda.gov/Food/FoodScienceResearch/WholeGenomeSequencingProgramWGS/}}.
-This data may become common for unculturable organisms as single-cell sequencing methods \citep[reviewed in][]{kalisky2011single} improve.
+The Food and Drug Administration maintains GenomeTrakr, an openly accessible database of whole genomes sampled from the environment and grown in culture\footnote{\url{http://www.fda.gov/Food/FoodScienceResearch/WholeGenomeSequencingProgramWGS/}}.
+This data may become common for unculturable organisms as single-cell sequencing methods improve \citep[reviewed in][]{kalisky2011single}.
 The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \citep{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.
@@
 Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
-These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed to preserve information associated with taxonomic classifications.
+These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
+% to preserve information associated with taxonomic classifications.
 Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process is somewhat opaque \citep{kreig1984bergey}.
 The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
-Tax2tree uses a heuristic algorithm to optimize $F$-measure of precision and recall for its taxonomic assignments.
+Tax2tree uses a heuristic algorithm to optimize the $F$-measure of precision and recall for its taxonomic assignments.
 Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
-Our group \citep{matsen2012reconciling} has developed ways of quantifying discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
+\citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
 Although it is wonderful that several groups are actively working on taxonomic revision, the existence of multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database can be frustrating.
@@
 Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
-The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy to turn a taxonomy and a collection of BLAST results for an unknown sequence on a database of sequences with those taxonomic labels into taxonomic assignments for the unknown sequence.
-Another program, called SAP, infers taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence \citep{munch2008statistical,munch2008fast}.
-\citet{segata2012metagenomic} propose a different approach to inferring organisms present in a metagenomic sample from a metagenome by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
-They show that this has good sensitivity and specificity, however, this method can only be used to classify to organisms that have complete genome sequences.
+The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy on taxonomies to derive taxonomic assignments for an unknown sequence from BLAST results of an unknown sequence onto a database of sequences with those taxonomic labels.
+\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
+\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample from a metagenome by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
+They show that this has good sensitivity and specificity, however, this method can only be used to classify to organisms whose genome has been sequenced.
 
 \subsection{The role of Operational Taxonomic Units (OTUs)}
-Although there continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, most human microbiome research has replaced any traditional species concept with the notion of Operational Taxonomic Units (OTUs).
+Although there continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, a substantial part of human microbiome research has replaced any traditional species concept with the notion of Operational Taxonomic Units (OTUs).
 An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly at 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
@@
 
-There are many OTU clustering strategies of various speeds.
+There are many OTU clustering methods of various speeds and strategies.
 Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
 These methods are very fast and are heuristic in that they are described as an algorithm rather than as global optimization of some notion of goodness of clustering.
-It is well known that different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
+Different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
 There are also methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
 \citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values.
+
 The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
@@
 
-Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (see next section), phylogenetic alpha diversity seems under-developed: whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve phylogenetic diversity (PD) measures.
-Faith's PD \citep{faith1992conservation} has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads expanded marker tree \citep{kembel2011phylogenetic}.
+Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
+Whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve phylogenetic diversity (PD) measures, and only in the sense of \citet{faith1992conservation}.
+Faith's PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.
 
@@
 Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
-\citet{mccoy2013abundance} have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome, in particular that they do a better job than the traditionally-used OTU-based measures.
-\citet{nipperess2013mean} have also determined the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.
+\citet{mccoy2013abundance} have recently shown that partially abundance-weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the traditionally-used OTU-based measures.
+\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.
 
@@
 \subsection{Community comparison using phylogenetics}
-The level of similarity between samples or groups thereof is called \emph{beta diversity}.
+The level of similarity between samples or groups of samples is called \emph{beta diversity}.
 As with alpha diversity, it is not uncommon to use classical measures \citep[e.g.][]{jaccard1908nouvelles} applied to OTU counts, however phylogenetics-based methods are the most popular.
-They are generally called variants of the ``UniFrac'' dissimilarity as so named by \citet{LozuponeKnightUniFrac05}.
-\citet{kuczynski2010microbial} claim that they are superior to other methods for community comparison using real data and simulations \citep[for a contrary viewpoint using simulations see][]{schloss2008evaluating}.
+They are generally variants of the ``UniFrac'' phylogenetic dissimilarity measure \citep[as described and named by][]{LozuponeKnightUniFrac05}.
+\citet{kuczynski2010microbial} claim that the UniFrac framework is superior to other methods for community comparison via real data and simulations \citep[for a contrary viewpoint using simulations see][]{schloss2008evaluating}.
 % It recovers gradients. \citep{nemergut2011global}
@@
 
-The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample or another via the shortest path rather than edges that connect between samples.
+The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample or another via the shortest path rather than edges that connect between samples \citep{LozuponeKnightUniFrac05}.
 Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
-\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a special case of the earth-movers distance, and is special in that it can be calculated in linear time, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
-Recent work \citep{chen2012associating} has shown that a partially abundance-weighted variant of UniFrac has greater power to resolve community differences than either unweighted or weighted UniFrac.
+\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a special case of the earth-movers distance, is special in that it can be calculated in linear time, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
+Recent work \citep{chen2012associating} has shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
 
@@
 Such observations surround a fit nonlinear curve, and the extent to which they lay on the curve appears to be phylum-dependent.
-This ``proxy'' approach has recently been taken to its logical conclusion by methods to infer functional characteristics from a 16S sample using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} by either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
+This ``proxy'' approach has recently been taken to its logical conclusion by \citet{langille2013predictive}, who develop methods to infer functional characteristics from a 16S sample using discrete trait evolution models on 16S gene trees by either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.
 
@@
 
-Nevertheless, the work that has been done to infer horizontal gene transfer has revealed interesting results.
+Nevertheless, the work that has been done to infer horizontal gene transfer in the human microbiome has revealed interesting results.
 \citet{hehemann2010transfer} found that a gene found in seaweed has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
@@
 In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.
-This is defended by saying for applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
-Furthermore, given that the underlying data is typically 16S only we can expect some topological inaccuracy even with the best methods.
+This is defended by saying that for applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
+Furthermore, given that the underlying data is typically 16S only we can expect some topological inaccuracy in reconstructing the ``tree of cells'' even with the best methods.
 Additionally, as specified below, these data sets can be very large.
 There does not seem to be contentious discussion of specific features of the inferred trees equivalent to, say, the current discussion around the rooting of the placental mammal tree.
+Given this perspective, it is not surprising that Bayesian phylogenetic methods and methods that incorporate alignment uncertainty are absent.
 
@@
 Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
-People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut \citep{sheneman2006clearcut} program, and QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree 2.
+People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut \citep{sheneman2006clearcut} program, or QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree 2.
 % AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
@@
 
-The scale of the data has motivated strategies other than complete phylogenetic inference.
+The scale of the data has motivated strategies other than complete phylogenetic inference, such as the insertion of sequences into an existing phylogenetic tree.
 Although such insertion has a long history as a means to sequentially build a phylogenetic tree \citep{kluge1969quantitative}, the first software with insertion specifically as a goal was ARB by \citet{ludwig2004arb} which works by parsimony.
@@
 \citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
-\citet{mirarabsepp} use data set partitioning to improve alignments on subsets of taxa.
+\citet{mirarabsepp} use data set partitioning to improve alignments on subsets of taxa specifically for this application.
 
@@
 The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
-The GreenGenes tree is updated by running FastTree from scratch.
+The GreenGenes tree is updated by running FastTree from scratch for every release.
 For some reason, some believe that FastTree in particular works well even with such gappy alignments \citep[e.g.][]{sharpton2011phylotu}.
@@
 \section{Phylogenetic challenges and opportunities in human microbiome research}
+
+Many phylogenetic challenges remain in human microbiome research.
 
@@
 The apparent success of 16S-tree-based comparisons begs the question of if a more complete representation of the evolutionary history of the microbes would yield better comparisons.
-This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of a set of organisms that allows us to best predict features of underlying genomes?
+This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of a set of organisms that allows us to best perform proxy whole-genome comparison?
 This representation could be simple.
@@
 If we were to equip the 16S tree with some measure of the strength of that correlation, would that allow for more precise comparison?
-It is quite possible that these objects would not be able to overcome the inherent noise of the data, but if we allow an arbitrary ``hidden'' object, what such object would perform best?
+If we allow an arbitrary ``hidden'' object, what such object would perform best?
 For example, collections of reconciled gene trees in the presence of gene deletion, transfer, and loss \citep[see][for interesting recent results]{szollHosi2013efficient,szollHosi2013lateral} could be used.
 
-In addition to using phylogenies to infer gene repertoire, it appears that neutral models involving phylogenetics could be more fully developed.
-Methods explicitly invoking trait evolution seem notably absent, with one recent exception \citep{langille2013predictive}.
+It appears that neutral models involving phylogenetics could be more fully developed.
+Methods explicitly invoking trait evolution are notably absent, with the recent exception of \citep{langille2013predictive}.
 The results of this simple method are reasonable, but would a species tree reconciled with a collection of gene trees allow for better prediction?
 Perhaps improved methods, say involving whole-genome evolutionary modeling or models of metabolic network evolution, could shed light on the problem.
-How would we formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on a collection of reconciled gene trees?
+Here again the ``tree of life'' problem can be formulated in a practical light: what representation allows for the best prediction of features of underlying genomes?
+How to formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on such an object?
+It is quite possible that these objects would not be able to overcome the inherent noise of the data, but further exploration seems warranted as simple methods give reasonable results.
 
-An important project is to model community assembly, and perhaps further phylogeny-aware methods could be used.
-One way to model community assembly is to apply Hubbell's neutral theory to the human microbiome \citep{fierer2012animalcules,costello2012application}.
+An important project for microbial ecology is to model community assembly, and perhaps further phylogeny-aware methods could be used.
+One way to model community assembly is to apply Hubbell's neutral theory \citep{fierer2012animalcules,costello2012application}.
 \citet{o2012phylogenetic} model community assembly with an explicitly phylogenetic perspective, and include some comparison of models to data.
@@
 In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics.
-It will be challenging to bring together macroevolutionary modeling with species abundance modeling, but some initial steps have been made in another setting \citep{lambert2013predicting}.
+Another challenge is to bring together macroevolutionary modeling with species abundance modeling, but some initial steps have been made in another setting \citep{lambert2013predicting}.
 
-Diversity preservation is of interest for microbiome researchers, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms \citep{hartmann2006maximizing,pardi2007resource}.
+Diversity preservation is of interest for microbiome researchers like it is for eukaryotic organisms, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms \citep{hartmann2006maximizing,pardi2007resource}.
 Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer \citep[reviewed in][]{cho2012human} and has made passionate appeals to preserve microbiome diversity \citep{blaser2011antibiotic}.
-Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
+% Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
 Because microbiomes are seeded from mother to child, there is a somewhat equivalent notion of extinction of microbiome when the chain is interrupted.
 Consequently, \citet{yatsunenko2012human} have explicitly contrasted microbiome development in urban, forest-dwelling, and rural populations, while \citet{tito2012insights} have endeavored to characterize the microbiome from ancient feces.
+How might phylogenetic methods be used in these preservation efforts?
 
@@
 Although parsimony gave an interesting answer here, the presence of such coevolution raises the question of what sort of forward-time models are appropriate for microbiome change?
-Would methods using these models do better than clustering or distance methods applied to the above distances?
-Other studies \citep[e.g.][]{phillips2012microbiome,delsuc2013convergence} see a combination of historical and dietary influences.
+Would methods using these models do better than commonly-applied clustering or distance methods applied to the above distances?
+Some studies \citep[e.g.][]{phillips2012microbiome,delsuc2013convergence} see a combination of historical and dietary influences.
 How can these forces be compared in this setting?
@@
 
-Opportunities for clinical applications may present themselves, but probably only some of these will be sequencing based.
-For example, 16S sequencing is likely soon to be completely replaced by Matrix-Assisted Laser Desorption Ionization--Time of Flight (MALDI-TOF) mass spectrometry for routine identification of a single known microbe grown in culture \citep{clark2013matrix}.
-However, for complex commmunity-level diagnoses sequencing and consequent analysis methods will still be required \citep[see review in][]{Rogers2013271}, and inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}.
+Opportunities for clinical applications may present themselves, but only some of these will be sequencing based.
+For example, routine 16S sequencing is likely soon to be completely replaced by Matrix-Assisted Laser Desorption Ionization--Time of Flight (MALDI-TOF) mass spectrometry for identification of a single known microbe grown in culture \citep{clark2013matrix}.
+However, for diagnoses that are on the level of microbial communities, sequencing and consequent analysis methods will still be required \citep[see review in][]{Rogers2013271}.
+Inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}.
 
 Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes it can feel like the hype has outmatched the reality.
-However, microbes are here to stay and so is research on them.
+However, our microbes are here to stay and so is research on them.
 Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
@@
 \section{Acknowledgements}
-The author would like to thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
-He is also grateful to Aaron Darling, David Fredricks, Noah Hoffman, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, and to support from the NIH (R01 HG005966-01) and NSF (1223057).
+I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
+I am also grateful to Aaron Darling, David Fredricks, Noah Hoffman, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, and to the NIH (R01 HG005966-01) and NSF (1223057) for financial support.
 

### 2013-09-30 sbreview.tex
Edits from excellent feedback by @nhoffman.

@@
 Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
-Their work, and the scientists who followed them, focused on the 16S ribosomal gene as a genetic marker.
+Their work, and the scientists who followed them, focused on the 16S ribosomal gene (henceforth simply ``16s'') as a genetic marker.
 This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
-Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled surveys of all organisms regardless of whether they can be cultured.
+Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled surveys of almost all bacteria regardless of whether they can be cultured.
 
@@
 Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
+A final cause of noise is next-generation sequencing error: this is certainly a problem for surveys and metagenomes, but is becoming less of a problem as technology improves.
+I will not address it specifically except for in the inference of operational taxonomic units as described below.
 
@@
 As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific region, and thus the genetic region of a read is unknown in addition to the organism it was derived from.
-Because it does proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although sequencing is known to have biases.
+Because it does not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although sequencing is known to have biases.
 
@@
 However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
+\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
 % \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
@@
 I briefly describe the mini-industry of OTU clustering techniques to contrast with the phylogenetic literature on species delimitation \citep{pons2006sequence,yang2010bayesian}.
+I will use the term \emph{OTU inference} despite the fact that there is no clear definition of the OTU concept.
 
@@
 
-There are many OTU clustering methods of various speeds and strategies.
+There are many OTU inference methods of various speeds and strategies.
 Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
@@
 The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
+A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as it such errors are especially problematic there.
 With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
@@
 
+As described above, \citet{morgan2010metagenomic} showed that various microbes have different DNA extraction efficiencies, meaning that the representation of marker gene sequences is not representative of the actual communities.
+Furthermore, there was no clear taxonomic signal in their observations of the variability of extraction efficiency, which seems to preclude a correction strategy based on ``species-tree'' phylogenetic modeling.
+However, presumably \textit{something} about their genome is determining extraction efficiency; it would be interesting and useful to search for the genetic determinants.
+As described above, abundances are commonly used as part of community comparison, thus a better quantification of error in those observations of abundance would be a great help.
+
+Noah-- AFAIK nobody has done phylogenetic modeling of primer bias, which would seem to be very useful!
+Couldn't we check for phylogenetic signal and if there is some, do modeling?
+I mean, there has got to be some phylogenetic signal, right?
+
 Finally, the conventional wisdom that UniFrac analysis is robust to tree reconstruction methodology begs further exploration.
@@
 Inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}.
+For all of these measures, it will be important to have rigorous means of quantifying uncertainty for robust diagnostic applications.
 

### 2013-09-30 sbreview.tex
Sample normalization, etc.
Thanks to @b4winckler for the suggestion.

@@
 \hyphenation{Ge-nome Ge-nomes Me-ta-ge-nome Me-ta-ge-nomes Ma-cro-ev-o-lu-tion-ary}
-
-% Noah suggests making a table for
-%alpha diversity
-%beta diversity
-%PCA
-%taxonomic assignment/classification
-%clustering/binning
 
@@
 
+In another vein, \citet{la2012statistical} consider the induced taxonomic tree of a sample as a statistical object and, using a framework where a sampling probability is defined in terms of a distance between such induced trees, define and investigate maximum likelihood estimation of and likelihood ratio tests for these trees.
+They focus on distances between trees induced by matrix metrics on the corresponding adjacency matrices.
+A similar framework was used by \citet{steel2008maximum} to construct maximum likelihood supertrees.
+
 \subsection{Phylogeny and function}
@@
 
-Noah-- AFAIK nobody has done phylogenetic modeling of primer bias, which would seem to be very useful!
-Couldn't we check for phylogenetic signal and if there is some, do modeling?
-I mean, there has got to be some phylogenetic signal, right?
+In a similar vein, assessing the significance level an observed difference between communities is not straightforward and poses difficult problems.
+The randomization test that is very commonly used in combination with UniFrac does not have appropriate properties when in the regime of incomplete sampling with non-independent observations, which is certainly the correct regime for surveys and metagenomes.
+Such non-independence can lead to incorrect rejection of the null hypothesis.
+Imagine, for example, that we have a random observation process on the tree equipped with some collection of ``base observations.''
+Each process takes a random subset of those base observations and then throws down some number of reads for each observation in that set, the number of which has mean significantly greater than 1.
+If the set of base observations is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
+Even basic definitions pose a challenge here: the question of whether two communities are the ``same'' and ``different'' probably needs to be approached from the perspective of ecosystem modeling.
+
+In general, sample normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
+The present sample normalization techniques are simply in terms of \textit{rarefaction}, which means uniform sub-sampling, typically without replacement \citep{caporaso2010qiime,schloss2009introducing}.
+Such normalization implicitly assumes a model whereby reads are sequenced independently of one another, which as explained is almost certainly not the case.
+One might hope to overcome these problems through clustering reads from the ``same'' organism and then performing rarefaction, but as described in the section on OTUs this is a challenge on its own.
+\citet{holmes2012dirichlet,la2012hypothesis} use models where read counts are modeled as overdispersed samples of the true distributions.
+\citet{o2012phylogenetic} provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly, but more work could be done exploring results under deviation from that model.
 
 Finally, the conventional wisdom that UniFrac analysis is robust to tree reconstruction methodology begs further exploration.
-Would it be possible to infer an ``equivalence class'' of phylogenetic trees, where two trees are deemed equivalent if they induce the same principal coordinates projection given the same underlying presence/absence or count data?
+Would it be possible to infer an equivalence class of phylogenetic trees, where two trees are deemed equivalent if they induce the same principal coordinates projection given the same underlying presence/absence or count data?
 Given that a tree is an integral part of a UniFrac analysis, it would be interesting to be able to infer the features of a tree that determine the primary trends in a projection.
@@
 Future will continue to bring deeper sequencing on more samples.
-The uBiome project (\url{http://ubiome.com/}) promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
+The uBiome project\footnote{(\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
 Comparative studies will continue to study what shapes and is shaped by the microbiome.
@@
 I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
-I am also grateful to Aaron Darling, David Fredricks, Noah Hoffman, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, and to the NIH (R01 HG005966-01) and NSF (1223057) for financial support.
+I am also grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, to Noah Hoffman and Bj\"orn Winckler for providing feedback on the manuscript, and to the NIH (R01 HG005966-01) and NSF (1223057) for financial support.
 

### 2013-09-30 sbreview.tex
Adding funkhouser2013mom reference.
H/T @phylogenomics.

@@
 % Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
-Because microbiomes are seeded from mother to child, there is a somewhat equivalent notion of extinction of microbiome when the chain is interrupted.
+Because a child's initial microbiome is transmitted from the mother \citep[reviewed in][]{funkhouser2013mom}, there is a somewhat equivalent notion of extinction of microbiome when the chain is interrupted.
 Consequently, \citet{yatsunenko2012human} have explicitly contrasted microbiome development in urban, forest-dwelling, and rural populations, while \citet{tito2012insights} have endeavored to characterize the microbiome from ancient feces.

### 2013-10-01 sbreview.tex
Bibliography normalization
* replaced full lists of authors where Google Scholar had truncated to "and others", per author guidelines
* added capitalization protection where needed
* switched the few articles that were using quotes to curly braces
* removed a few duplicate articles
* fixed journal name capitalization where needed
Adding paulson2013differential.

@@
 
-In general, sample normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
-The present sample normalization techniques are simply in terms of \textit{rarefaction}, which means uniform sub-sampling, typically without replacement \citep{caporaso2010qiime,schloss2009introducing}.
-Such normalization implicitly assumes a model whereby reads are sequenced independently of one another, which as explained is almost certainly not the case.
-One might hope to overcome these problems through clustering reads from the ``same'' organism and then performing rarefaction, but as described in the section on OTUs this is a challenge on its own.
-\citet{holmes2012dirichlet,la2012hypothesis} use models where read counts are modeled as overdispersed samples of the true distributions.
-\citet{o2012phylogenetic} provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly, but more work could be done exploring results under deviation from that model.
+In general, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
+One type of normalization handles differential depths of sequencing across samples.
+The presently used approach is \textit{rarefaction}, which means uniform sub-sampling, typically without replacement \citep{caporaso2010qiime,schloss2009introducing} to the number of reads in the lowest abundance sample.
+In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another, which as explained is almost certainly not the case.
+An alternative is provided by \citet{o2012phylogenetic}, who provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly.
+This is a good start, but more work could be done exploring results under deviation from that model.
+
+Another type of normalization seeks to infer the true abundances from noisy observations of the various taxonomic groups or OTUs.
+\citet{holmes2012dirichlet,la2012hypothesis} use models where read counts are modeled as overdispersed samples of the true abundance and provide methods for statistical testing.
+\citet{paulson2013differential} estimate true abundances using a zero-inflated Gaussian model for read counts.
+This work could be extended to a phylogenetic context by making use of the relationship between OTUs and modeling the way in which the abundance of one OTU may increase the abundance of a related OTU because of sequencing error or a change of condition that changes the abundance of both.
 
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{plainnat_abbrv}
 \bibliography{sbreview}

### 2013-10-01 sbreview.tex
Including figures into the MS.

@@
 \usepackage{palatino}
+\usepackage{wrapfig}
 \usepackage[hidelinks]{hyperref}
 \usepackage[round]{natbib}
+\usepackage[labelfont=bf,labelsep=period,font=small]{caption} % bold the 'Figure #' in the caption and separate it with a period
 \newcommand{\forarxiv}[1]{#1}
@@
 
-% code
-\newcommand{\program}{\textsf{program}}
-
 \newcommand{\eat}[1]{}
-
-\newcommand{\FIGmassTransport}{\
-\begin{figure}[ht]
-\begin{center}
-  \forarxiv{\includegraphics[width=13cm]{mass_transport.pdf}}
-\end{center}
-\caption{\
-  Caption goes here.
-}
-\label{FIGmassTransport}
-\end{figure}
-}
-\newcommand{\refFIGmassTransport}{1}
 
@@
 Much of the analytical work for these sequences involves phylogenetics, at least indirectly, and methodology has developed in a somewhat different direction than for other applications of phylogenetics.
-In this paper I review the field and its methods from the perspective of a phylogeneticist, in particular the impact of tree-thinking on the methods used in the analysis of human microbiome data, as well as describing current challenges for phylogenetics coming from this type of work.
+In this paper I review the field and its methods from the perspective of a phylogeneticist, as well as describing current challenges for phylogenetics coming from this type of work.
 
@@
 
-\textbf{[Figure: \url{http://i.imgur.com/3rs1bND.png}]?}
-
 There are many OTU inference methods of various speeds and strategies.
@@
 
-\textbf{[Figure: \url{http://cl.ly/image/3k3n3W3U443G/ref_tree_bwpd.png} to show a partially abundance weighted PD measure?]}
+\subsection{Community comparison using phylogenetics}
 
-\subsection{Community comparison using phylogenetics}
+\newcommand{\unifracLegend}{
+    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
+    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
+}
+\forarxiv{
+\begin{wrapfigure}[23]{r}{2.5in}
+  \capstart
+  \vspace{-19pt}
+  \begin{center}
+    \includegraphics[width=2.5in]{figures/unifrac.pdf}
+  \end{center}
+  \vspace{-10pt}
+  \caption{\unifracLegend}
+  \label{fig:unifrac}
+\end{wrapfigure}
+}
+
 The level of similarity between samples or groups of samples is called \emph{beta diversity}.
@@
 
-\textbf{[Figure: canonical unifrac explanation picture?]}
-
 The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample or another via the shortest path rather than edges that connect between samples \citep{LozuponeKnightUniFrac05}.
@@
 
-\textbf{[Figure: \url{http://cl.ly/image/3K472T0K3S3S/tree_dirtpiles.png}]?}.
-
 The most common way to use a pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
@@
 
-\textbf{[Figure: example principal coordinates plot?]}
+\newcommand{\dirtpilesLegend}{
+    Part of a minimal mass movement to calculate the earth-mover's distance on a phylogenetic tree.
+}
+\forarxiv{
+\begin{wrapfigure}[19]{l}{1.6in}
+  \capstart
+% \vspace{-24pt}
+  \begin{center}
+    \includegraphics[width=1.6in]{figures/tree_dirtpiles.pdf}
+  \end{center}
+  \vspace{-5pt}
+  \caption{\dirtpilesLegend}
+  \label{fig:dirtpiles}
+\end{wrapfigure}
+}
 
@@
 Future will continue to bring deeper sequencing on more samples.
-The uBiome project\footnote{(\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
+The uBiome project\footnote{\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
 Comparative studies will continue to study what shapes and is shaped by the microbiome.
@@
 I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
-I am also grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, to Noah Hoffman and Bj\"orn Winckler for providing feedback on the manuscript, and to the NIH (R01 HG005966-01) and NSF (1223057) for financial support.
+I am also grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, to Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and to the NIH (R01 HG005966-01) and NSF (1223057) for financial support.
 
@@
 %\FIGmassTransport
+
+\noindent
+Figure 1: \unifracLegend
+
+\noindent
+Figure 2: \dirtpilesLegend
+
 \clearpage

### 2013-10-02 sbreview.tex
Revision pass through.

@@
 The human microbiome is the collection of microbes that live inside and on the surface of humans.
-Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information of microbiome samples.
-Much of the analytical work for these sequences involves phylogenetics, at least indirectly, and methodology has developed in a somewhat different direction than for other applications of phylogenetics.
+Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information from microbiome samples.
+Much of the analytical work for these sequences involves phylogenetics, at least indirectly, but methodology has developed in a somewhat different direction than for other applications of phylogenetics.
 In this paper I review the field and its methods from the perspective of a phylogeneticist, as well as describing current challenges for phylogenetics coming from this type of work.
@@
 
-The parameter regime and focus of microbiome research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what these microbial ecologists and medical researchers have done?
+The parameter regime and focus of microbiome research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what microbial ecologists and medical researchers have done?
 The answer is simple: this system is data- and question-rich.
-It absolutely requires molecular methods, as microbes are now primarily identified by their molecular sequence, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
-Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sequences derived from the sample.
+It absolutely requires molecular methods, because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
+Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.
 
@@
 However, approaches differ between the microbial ecology community and that of eukaryotic phylogenetics, in part because the scope of the former contains an almost unlimited diversity of organisms, leading to additional problems above the usual.
-The species concept is even more problematic for microbes than for larger eukaryotes and there is also considerable discussion concerning how to group them into species-like units.
+The species concept is even more problematic for microbes than for multicellular organisms, and hence there is also considerable discussion concerning how to group them into species-like units.
 Organizing microbes into a sensible taxonomy is a serious challenge, especially in the absence of obvious morphological features.
 
-Because of this high level of diversity and challenges with species definitions, microbial ecology researchers have developed their own explicitly phylogenetic techniques for doing comparing samples rather than comparing on the level of species abundances.
+Because of this high level of diversity and challenges with species definitions, microbial ecology researchers have developed their own explicitly phylogenetic techniques for comparing samples rather than comparing on the level of species abundances.
 Although there is some overlap with previous literature, these techniques and could be used in a wider setting and may deserve broader consideration by the phylogenetics community.
@@
 The human microbiome is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
-Additionally, due to more than a century of hospital laboratory work, our knowledge about human-associated microbes relatively rich.
-It is also common to manipulate the human microbiome or models thereof via intervention studies and germ-free animals, and is easier to do than for many other microbial communities.
+Additionally, due to more than a century of hospital laboratory work, our knowledge about human-associated microbes is relatively rich.
+It is also common to manipulate the human microbiome or animal models thereof via intervention studies and germ-free animals, which is easier to do than for many other microbial communities.
 
@@
 Despite an obvious oversimplification, I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
-I realize that there is substantial overlap-- for instance the microbial community is very interested in unicellular fungi, and additionally many in the SB community do work on microbes-- but this terminology will be useful for concreteness.
+I realize that there is substantial overlap--- for instance the microbial community is very interested in unicellular fungi, and additionally many in the SB community do work on microbes--- but this terminology will be useful for concreteness.
 There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
@@
 I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
-I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics; indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes orders of magnitude larger than typical for eukaryotic phylogenetics.
+I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics, indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes an order of magnitude larger than typical for eukaryotic phylogenetics.
 
@@
 The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
-These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells.
+These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells of that individual.
 The microbiome has remarkable metabolic potential, with an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
@@
 The microbiome interacts on many levels with host phenotype \citep[reviewed in][]{cho2012human}.
-For example, the gut microbiome correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to children with acute metabolic syndromes in rural Africa \citep{smith2013gut}.
-Considerable attention has also been given to the interaction between the gut microbiome and obesity, although the story is yet clear.
+The gut microbiome in particular correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to children with acute metabolic dysfunction in rural Africa \citep{smith2013gut}.
+Considerable attention has also been given to the interaction between the gut microbiome and obesity, although the story is not yet clear.
 An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
-A causal role for microbiome leading to phenotype is established for mice: an obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}, the pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
+A causal role for the microbiome leading to obesity is established for mice: an obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}, the pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
 However, these promising leads have not yet been confirmed causally or in population studies of humans \citep{zhao2013gut}.
@@
 In this paper we will be primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perpsective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
-Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and with the human body, a broad perspective will continue to be important.
+Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.
 
@@
 It is now possible to assay microbial communities in high throughput using sequencing.
-There are two ways of doing so.
-The first is to amplify a specific gene in the genome for sequencing using polymerase chain reaction (PCR).
-Scientists typically pick a ``marker'' gene in that case suitable for phylogenetic analysis.
-The second is to randomly shear input DNA and/or RNA and then perform sequencing directly.
+One way is to amplify a specific gene in the genome for sequencing using polymerase chain reaction (PCR).
+Scientists typically pick a ``marker'' gene in that case that is meant to recapitulate the ``overall'' evolutionary history of the microbes.
+Another way is to randomly shear input DNA and/or RNA and then perform sequencing directly.
 We will consistently refer to the former as a \textit{survey} and the second a \textit{metagenome}, although these words have not always been consistently used in the literature.
@@
 \subsection{Microbial community estimation using marker gene surveys}
-Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
-Their work, and the scientists who followed them, focused on the 16S ribosomal gene (henceforth simply ``16s'') as a genetic marker.
+Our modern knowledge of the microbial world is in a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
+Their work, and the scientists who followed them, focused on the 16S ribosomal gene (henceforth simply ``16S'') as a genetic marker.
 This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
@@
 This has led to extensive parallelization, most recently by sequencing dozens of samples at a time on the Illumina instrument \citep{degnan2011illumina,caporaso2012ultra}.
-This leads to the question of how many sequences are needed to characterize the microbial diversity of a given environment.
+This brings up the question of how many sequences are needed to characterize the microbial diversity of a given environment.
 To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.
@@
 
-Despite the high throughput and low cost of modern sequencing, inherent challenges remain for the use of population census by marker gene sequencing.
+Despite the high throughput and low cost of modern sequencing, inherent challenges remain for applications of marker gene sequencing to take a census of microbes.
 Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
-Current sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
+Current high throughput sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
 ``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
@@
 Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
-A final cause of noise is next-generation sequencing error: this is certainly a problem for surveys and metagenomes, but is becoming less of a problem as technology improves.
+A final cause of noise is next-generation sequencing error: this is certainly a problem for both surveys and metagenomes, but is becoming less of a problem as technology improves.
 I will not address it specifically except for in the inference of operational taxonomic units as described below.
@@
 \subsection{Metagenomes}
-As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific region, and thus the genetic region of a read is unknown in addition to the organism it was derived from.
-Because it does not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although sequencing is known to have biases.
+As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
+Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although multiplex sequencing is known to have biases of its own.
 
 It is possible to use metagenomic data as an expanded set of marker genes.
-That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes present in a large proportion of micro-organisms that are expected to follow the same evolutionary path.
+That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
 This is proven to be a useful strategy \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}, however, because of the diversity of gene repertoire in microbes, this core gene set may be relatively small.
 Indeed, even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
-The rest of the data can be taxonomically classified \citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
+At least some portion of the rest of the data can be taxonomically classified \citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
 
-Metagenomic data is often used to infer information of a metabolic rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
-Such inference is beyond the scope of this paper.
+Metagenomic data is often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
+Discussing these methods is beyond the scope of this paper.
 
@@
 
-In this section I consider the ways in which phylogenetics has impacted human microbiome research.
-What may be most interesting for this audience is the way in which phylogenetic trees are being used to actively revise taxonomy as well as being used as a structure on which to perform sample comparison.
+In this section I consider the ways in which phylogenetic methodology has impacted human microbiome research.
+What may be most interesting for the \textit{Systematic Biology} audience is the way in which phylogenetic trees are being used to actively revise taxonomy as well as being used as a structure on which to perform sample comparison.
 
@@
 % to preserve information associated with taxonomic classifications.
-Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process is somewhat opaque \citep{kreig1984bergey}.
-The GreenGenes taxonomy \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
+Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process appears opaque \citep{kreig1984bergey}.
+The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
 Tax2tree uses a heuristic algorithm to optimize the $F$-measure of precision and recall for its taxonomic assignments.
@@
 \citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
-Although it is wonderful that several groups are actively working on taxonomic revision, the existence of multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database can be frustrating.
+Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database.
 
@@
 Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
-The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy on taxonomies to derive taxonomic assignments for an unknown sequence from BLAST results of an unknown sequence onto a database of sequences with those taxonomic labels.
+The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy on taxonomies to derive taxonomic assignments for that sequence from BLAST results of an unknown sequence onto a database of sequences with taxonomic labels.
 \citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
-\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample from a metagenome by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
-They show that this has good sensitivity and specificity, however, this method can only be used to classify to organisms whose genome has been sequenced.
+\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
+They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.
 
@@
 Although there continues to be a lively debate on if there is a meaningful concept of species for microbes \citep{bapteste2009prokaryotic,caro2012bacterial}, a substantial part of human microbiome research has replaced any traditional species concept with the notion of Operational Taxonomic Units (OTUs).
-An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly at 97\% sequence identity, such that each OTU consists of sequences that are closer to each other than that cutoff.
+An OTU is a proxy species concept that is typically defined with a fixed divergence cutoff, most commonly at 97\% sequence identity, such that each OTU is a cluster of sequences that are closer to each other than that cutoff.
 It is common for trees to be built on sequence representatives from these OTUs, and the abundance of an OTU to be given by the number of sequences that sit within that cluster.
@@
 
-There are many OTU inference methods of various speeds and strategies.
+There are many OTU inference methods with various speeds and strategies.
 Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
 These methods are very fast and are heuristic in that they are described as an algorithm rather than as global optimization of some notion of goodness of clustering.
-Different ways of doing this heuristic clustering can result in very different results \citep{white2010alignment}.
-There are also methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies \citep{navlakha2009finding}.
+\citet{white2010alignment} show that different ways of doing this heuristic clustering can result in very different results.
+\citet{navlakha2009finding} develop methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies.
 \citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values.
@@
 A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as it such errors are especially problematic there.
-With the exception of this last reference, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
+With the exception of \citet{sharpton2011phylotu}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
 
@@
 Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
-\citet{mccoy2013abundance} have recently shown that partially abundance-weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the traditionally-used OTU-based measures.
+\citet{mccoy2013abundance} have recently shown that partially abundance-weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
 \citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.
@@
 \begin{wrapfigure}[23]{r}{2.5in}
-  \capstart
+  % \capstart
   \vspace{-19pt}
@@
 
-The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample or another via the shortest path rather than edges that connect between samples \citep{LozuponeKnightUniFrac05}.
+The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
 Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
+These dissimilarity measures have hundreds of citations.
 \citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a special case of the earth-movers distance, is special in that it can be calculated in linear time, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
-Recent work \citep{chen2012associating} has shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
+The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree.
+\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
 
-The most common way to use a pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
+The most common way to use a between-sample pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
 Indeed, the separation of two communities in a principal components plot is often used as \emph{prima facie} evidence of a difference between them
@@
 \begin{wrapfigure}[19]{l}{1.6in}
-  \capstart
+  % \capstart
 % \vspace{-24pt}
@@
 
-There have been efforts to augment the ordination visualizations with additional information giving more structure to the visualizations.
-Biplots display variables (in the microbial case summarized by taxonomic labels) as points in addition to samples \citep[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
-Elizabeth Purdom \citep{BikEaMicrobiotaStomach06,PurdomAnalyzingDataGraphs08} describes how generalized principal component eigenvectors can be interpreted via weightings on the leaves of a phylogenetic tree.
-\citet{matsen2013edge} have developed a variant of principal components analysis that explicitly labels the axes with weightings on phylogenetic trees that indicate their meanings.
+There have been efforts to augment these ordination visualizations with additional information giving more structure to the visualizations.
+Biplots display variables (in the microbial case summarized by taxonomic labels) as points along with points representing samples \citep[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
+\citet{PurdomAnalyzingDataGraphs08} describes how generalized principal component eigenvectors can be interpreted via weightings on the leaves of a phylogenetic tree.
+\citet{matsen2013edge} have developed a variant of principal components analysis that explicitly labels the axes with weightings on phylogenetic trees that indicate their influence.
 
@@
 16S distance is frequently used as a proxy for a functional comparison between human microbiome samples.
-Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes is commonly acquired horizontally as well as vertically, and horizontal transmission leaves no trace in the 16S ancestry.
+Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes is commonly acquired horizontally as well as vertically, and horizontal transmission leaves no trace in 16S ancestry.
 
@@
 Similar logic has been applied to prioritize microbes for sequencing.
-\citet{wu2009phylogeny} have derived a ``phylogeny-driven genomic encyclopaedia of Bacteria and Archaea'' by selecting organisms for sequencing that are divergent from organisms that have a sequenced genome.
+\citet{wu2009phylogeny} have derived a ``phylogeny-driven genomic encyclopaedia of Bacteria and Archaea'' by selecting organisms for sequencing that are divergent from sequenced organisms.
 By doing so they have recovered more novel protein families than they would have using methods organized by taxonomy.
-In a similar effort for the human microbiome \citep{fodor2012most} phylogenetic results were not shown although the authors state that phylogenetic methods did give similar results to their analysis.
+In a similar effort for the human microbiome \citep{fodor2012most}, phylogenetic results were not shown although the authors state that phylogenetic methods did give similar results to their analysis.
 % http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041294
@@
 As described above, whole-genome data is typically used to directly infer functional information rather than information concerning ancestry.
-The debate concerning whether a microbial tree of life is a useful concept \citep{bapteste2009prokaryotic,caro2012bacterial} does not seem to have dampened human microbiome researchers' enthusiasm for using a single such tree.
+The continuing debate concerning whether a microbial tree of life is a useful concept \citep{bapteste2009prokaryotic,caro2012bacterial} does not seem to have dampened human microbiome researchers' enthusiasm for using a single such tree.
 
@@
 This is defended by saying that for applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
-Furthermore, given that the underlying data is typically 16S only we can expect some topological inaccuracy in reconstructing the ``tree of cells'' even with the best methods.
+Furthermore, given that the underlying data is typically 16S alone we can expect some topological inaccuracy in reconstructing the ``tree of cells'' even with the best methods.
 Additionally, as specified below, these data sets can be very large.
@@
 Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
-People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut \citep{sheneman2006clearcut} program, or QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree 2.
+People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut \citep{sheneman2006clearcut} program, or QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree.
 % AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
@@
 The scale of the data has motivated strategies other than complete phylogenetic inference, such as the insertion of sequences into an existing phylogenetic tree.
-Although such insertion has a long history as a means to sequentially build a phylogenetic tree \citep{kluge1969quantitative}, the first software with insertion specifically as a goal was ARB by \citet{ludwig2004arb} which works by parsimony.
+Although such insertion has a long history as a means to sequentially build a phylogenetic tree \citep{kluge1969quantitative}, the first software with insertion specifically as a goal was the parsimony insertion tool in the ARB program by \citet{ludwig2004arb}.
 ARB is commonly used to reconstruct a full tree by direct insertion.
@@
 The GreenGenes tree is updated by running FastTree from scratch for every release.
-For some reason, some believe that FastTree in particular works well even with such gappy alignments \citep[e.g.][]{sharpton2011phylotu}.
+There appears to be a commonly held belief that FastTree in particular works well even with such gappy alignments \citep[e.g.][]{sharpton2011phylotu}.
 
@@
 Many phylogenetic challenges remain in human microbiome research.
+Some of them are familiar, such as how to build phylogenies on the scale needed here on data that has many insertions and deletions.
+I review some others here.
 
-One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference, and sequence insertion or placement that leaves the ``reference tree'' fixed.
-These two extremes are each each useful in different parameter regimes, but something between the two would be very helpful as well.
+One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference versus sequence insertion or placement that leaves the ``reference tree'' fixed.
+These two extremes are each each useful in different parameter regimes, but something between the two would be helpful as well.
 For example, sequence data is continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
@@
 The apparent success of 16S-tree-based comparisons begs the question of if a more complete representation of the evolutionary history of the microbes would yield better comparisons.
-This yields a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of a set of organisms that allows us to best perform proxy whole-genome comparison?
+This suggests a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of a set of microbes that allows us to best perform proxy whole-genome comparison?
 This representation could be simple.
@@
 How to formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on such an object?
-It is quite possible that these objects would not be able to overcome the inherent noise of the data, but further exploration seems warranted as simple methods give reasonable results.
+It is quite possible that inference using a more complete representation would not be able to overcome the inherent noise of the data, but further exploration seems warranted as simple methods give reasonable results.
 
@@
 Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \citep{mooers1997inferring,aldous2011five}.
-In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics.
+In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics by reducing a distribution on the tips of a tree down to a real number.
 Another challenge is to bring together macroevolutionary modeling with species abundance modeling, but some initial steps have been made in another setting \citep{lambert2013predicting}.
@@
 % Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
-Because a child's initial microbiome is transmitted from the mother \citep[reviewed in][]{funkhouser2013mom}, there is a somewhat equivalent notion of extinction of microbiome when the chain is interrupted.
+Because a child's initial microbiome is transmitted from the mother \citep[reviewed in][]{funkhouser2013mom}, there is a somewhat equivalent notion of microbiome extinction when the chain is interrupted.
 Consequently, \citet{yatsunenko2012human} have explicitly contrasted microbiome development in urban, forest-dwelling, and rural populations, while \citet{tito2012insights} have endeavored to characterize the microbiome from ancient feces.
@@
 Perhaps in part because of the importance of mother-to-child transmission, there are indications of coevolution between microbiomes and their hosts.
-\citet{ochman2010evolutionary} found identical trees for primate and microbiome evolution.
-They used maximum parsimony such that each column was a microbe and in each such entry took discrete states according to how much of that microbe was present.
+\citet{ochman2010evolutionary} found identical tree topologies for primate and microbiome evolution.
+For the microbiome, they used maximum parsimony such that each column was a microbe and each such entry took discrete states according to how much of that microbe was present.
 Although parsimony gave an interesting answer here, the presence of such coevolution raises the question of what sort of forward-time models are appropriate for microbiome change?
-Would methods using these models do better than commonly-applied clustering or distance methods applied to the above distances?
+Would methods using these models do better than parsimony or commonly-applied phenetic methods applied to the distances described above?
 Some studies \citep[e.g.][]{phillips2012microbiome,delsuc2013convergence} see a combination of historical and dietary influences.
-How can these forces be compared in this setting?
+How can such forces be compared in this setting?
 
@@
 In a similar vein, assessing the significance level an observed difference between communities is not straightforward and poses difficult problems.
-The randomization test that is very commonly used in combination with UniFrac does not have appropriate properties when in the regime of incomplete sampling with non-independent observations, which is certainly the correct regime for surveys and metagenomes.
+The randomization of group membership commonly used in combination with UniFrac to determine significance does not have appropriate properties when in the regime of incomplete sampling with non-independent observations, which is certainly the correct regime for surveys and metagenomes.
 Such non-independence can lead to incorrect rejection of the null hypothesis.
-Imagine, for example, that we have a random observation process on the tree equipped with some collection of ``base observations.''
-Each process takes a random subset of those base observations and then throws down some number of reads for each observation in that set, the number of which has mean significantly greater than 1.
+Imagine, for example, that we have a random observation process on the tree equipped with some collection of ``base observations'' at the leaves.
+Each sample from the process takes a random subset of those base observations and then throws down some number of reads for each observation in that set, the number of which has mean significantly greater than 1.
 If the set of base observations is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
+This would be a false positive.
 Even basic definitions pose a challenge here: the question of whether two communities are the ``same'' and ``different'' probably needs to be approached from the perspective of ecosystem modeling.
@@
 One type of normalization handles differential depths of sequencing across samples.
-The presently used approach is \textit{rarefaction}, which means uniform sub-sampling, typically without replacement \citep{caporaso2010qiime,schloss2009introducing} to the number of reads in the lowest abundance sample.
-In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another, which as explained is almost certainly not the case.
+The presently used approach is \textit{rarefaction}, which means uniform sub-sampling, typically without replacement to the number of reads in the lowest abundance sample \citep{caporaso2010qiime,schloss2009introducing}.
+In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another, which is not the case.
 An alternative is provided by \citet{o2012phylogenetic}, who provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly.
-This is a good start, but more work could be done exploring results under deviation from that model.
+This is a good start, but more work should be done exploring results under deviation from that model.
 
 Another type of normalization seeks to infer the true abundances from noisy observations of the various taxonomic groups or OTUs.
-\citet{holmes2012dirichlet,la2012hypothesis} use models where read counts are modeled as overdispersed samples of the true abundance and provide methods for statistical testing.
+\citet{holmes2012dirichlet} and \citet{la2012hypothesis} use models where read counts are modeled as overdispersed samples of the true abundance and provide methods for statistical testing.
 \citet{paulson2013differential} estimate true abundances using a zero-inflated Gaussian model for read counts.
-This work could be extended to a phylogenetic context by making use of the relationship between OTUs and modeling the way in which the abundance of one OTU may increase the abundance of a related OTU because of sequencing error or a change of condition that changes the abundance of both.
+This work could be extended to a phylogenetic context by making use of the relationship between OTUs, and modeling the way in which the abundance of one OTU may increase the abundance of a related OTU because of sequencing error or a change of condition that changes the abundance of both.
 
@@
 The uBiome project\footnote{\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
-Comparative studies will continue to study what shapes and is shaped by the microbiome.
+Comparative studies will continue to investigate what shapes and is shaped by the microbiome.
 However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.
@@
 There are limitations to what we can learn using genetics because more intricate processes such as gene regulation may be at play, limiting what sequence-level phylogenetics can do.
-Future work may move from more general ecological models to models that include specific interactions between microbes and the host \citep[reviewed in][]{hooper2012interactions}.
+Future work may move from general ecological models to models that include specific interactions between microbes and the host \citep[reviewed in][]{hooper2012interactions}.
 
 Opportunities for clinical applications may present themselves, but only some of these will be sequencing based.
-For example, routine 16S sequencing is likely soon to be completely replaced by Matrix-Assisted Laser Desorption Ionization--Time of Flight (MALDI-TOF) mass spectrometry for identification of a single known microbe grown in culture \citep{clark2013matrix}.
-However, for diagnoses that are on the level of microbial communities, sequencing and consequent analysis methods will still be required \citep[see review in][]{Rogers2013271}.
+For example, routine 16S sequencing is likely to be replaced soon by Matrix-Assisted Laser Desorption Ionization--Time of Flight (MALDI-TOF) mass spectrometry for assignment of a single microbe grown in culture to a database entry \citep{clark2013matrix}.
+However, for diagnoses that are on the level of microbial communities, sequencing and consequent analysis methods will still be required \citep[reviewed in][]{Rogers2013271}.
 Inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}.
@@
 
-Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes it can feel like the hype has outmatched the reality.
+Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes the hype has outmatched the science.
 However, our microbes are here to stay and so is research on them.
@@
 I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
-I am also grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, to Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and to the NIH (R01 HG005966-01) and NSF (1223057) for financial support.
+I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank to Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and thank the NIH (R01 HG005966-01) and NSF (1223057) for financial support.
 

### 2013-10-02 sbreview.tex
A few small fixes from Bastien.

@@
 
-In this paper we will be primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perpsective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
+In this paper we will be primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perspective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
 Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.
@@
 Despite the high throughput and low cost of modern sequencing, inherent challenges remain for applications of marker gene sequencing to take a census of microbes.
-Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
+Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that a collection of sequences need not be representative of the communities from which they were derived \citep{morgan2010metagenomic}.
 Current high throughput sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
@@
 A final cause of noise is next-generation sequencing error: this is certainly a problem for both surveys and metagenomes, but is becoming less of a problem as technology improves.
-I will not address it specifically except for in the inference of operational taxonomic units as described below.
+I will not address it specifically except in the inference of operational taxonomic units as described below.
 
@@
 As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
-Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although multiplex sequencing is known to have biases of its own.
+Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, however extraction efficiency concerns remain and multiplex sequencing is known to have biases of its own.
 
@@
 Metagenomic data is often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
-Discussing these methods is beyond the scope of this paper.
+Discussing these methods is beyond the scope of this paper, as is ``metatranscriptomics": the sequencing of mRNA in bulk.
 
@@
 The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
-A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as it such errors are especially problematic there.
+A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as such errors are especially problematic there.
 With the exception of \citet{sharpton2011phylotu}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
@@
 Nevertheless, the work that has been done to infer horizontal gene transfer in the human microbiome has revealed interesting results.
-\citet{hehemann2010transfer} found that a gene found in seaweed has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
+\citet{hehemann2010transfer} found that a seaweed gene has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
 Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
@@
 Considerable effort goes to the creation of large curated alignments and phylogenetic trees on 16S.
-There are two primary such projects: one is the SILVA database \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes database \citep{desantis2006greengenes,mcdonald2011improved}.
+There are two major projects to do so: one is the SILVA database \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes database \citep{desantis2006greengenes,mcdonald2011improved}.
 Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
@@
 In addition to these 16S-based resources, the MicrobesOnline resource \citep{dehal2010microbesonline} offers a very nice interactive tree-based genome browser.
-On a much smaller scale, there curated reference sets that are microbiome body-site specific \citep{chen2010human,griffen2011core,srinivasan2012bacterial}
+On a much smaller scale, there are microbiome body-site specific reference sets \citep{chen2010human,griffen2011core,srinivasan2012bacterial}
 
@@
 One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference versus sequence insertion or placement that leaves the ``reference tree'' fixed.
-These two extremes are each each useful in different parameter regimes, but something between the two would be helpful as well.
+These two extremes are each useful in different parameter regimes, but something between the two would be helpful as well.
 For example, sequence data is continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
@@
 
-In a similar vein, assessing the significance level an observed difference between communities is not straightforward and poses difficult problems.
+In a similar vein, assessing the significance level of an observed difference between communities is not straightforward and poses difficult problems.
 The randomization of group membership commonly used in combination with UniFrac to determine significance does not have appropriate properties when in the regime of incomplete sampling with non-independent observations, which is certainly the correct regime for surveys and metagenomes.

### 2013-10-03 sbreview.tex
Looking pretty good.

@@
 Several groups are continually revising taxonomy using the results of phylogenetic tree inference.
-These attempts are less ambitious than the development of the PhyloCode \citep{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
+These attempts are less ambitious than the PhyloCode project \citep[to develop a taxonomic scheme expressed directly in terms of a phylogeny; see][]{forey2001phylocode}, and simply work to revise the hierarchical structure of the taxonomy while for the most part leaving taxonomic names fixed.
 % to preserve information associated with taxonomic classifications.
@@
 Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
-The MEGAN program \citep{huson2007megan,huson2011integrative} uses a least common ancestor (LCA) strategy on taxonomies to derive taxonomic assignments for that sequence from BLAST results of an unknown sequence onto a database of sequences with taxonomic labels.
+The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits.
 \citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
@@
 
+\newcommand{\pdLegend}{
+    Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as blue circles and abundances are shown as the size of the circles.
+    Unweighted PD takes the sum of the branch lengths connecting samples.
+    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the areas of the edges.
+}
+\forarxiv{
+\begin{wrapfigure}[23]{r}{2.5in}
+  % \capstart
+  \vspace{-27pt}
+  \begin{center}
+    \includegraphics[width=2.5in]{figures/pd.pdf}
+  \end{center}
+  \vspace{-10pt}
+  \caption{\pdLegend}
+  \label{fig:pd}
+\end{wrapfigure}
+}
+
 The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
@@
 \subsection{Diversity estimates using phylogenetics}
+
 Because 16S surveys are inherently complex and noisy data, summary statistics are often used; summaries of the diversity of a single sample are often called \emph{alpha diversity}.
@@
 Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
-Whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve phylogenetic diversity (PD) measures, and only in the sense of \citet{faith1992conservation}.
-Faith's PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.
-
-Although abundance-weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance-weighted phylogenetic diversity measures are not used in human microbiome studies.
-Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
-\citet{mccoy2013abundance} have recently shown that partially abundance-weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
-\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.
-
-\subsection{Community comparison using phylogenetics}
+Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~\ref{fig:pd}).
+The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
+It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
+Whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve PD: unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.
 
@@
   % \capstart
-  \vspace{-19pt}
+  \vspace{-16pt}
   \begin{center}
@@
 }
+
+Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance weighted phylogenetic diversity measures are not used in human microbiome studies.
+Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not.
+Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
+Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
+\citet{mccoy2013abundance} have recently shown that partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
+\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.
+
+\subsection{Community comparison using phylogenetics}
 
@@
 
-Alignment methods are primarily focused on extending a curated ``seed alignment'' by adding additional sequences; several tools have been created with exactly this application for 16S in mind \citep{desantis2006nast,caporaso2010pynast,pruesse2012sina}.
+Alignment methods are primarily focused on developing automated methods to extend a relatively small hand-curated ``seed alignment'' with additional sequences; several tools have been created with exactly this application for 16S in mind \citep{desantis2006nast,caporaso2010pynast,pruesse2012sina}.
 The community also uses profile hidden Markov models \citep{eddy1998profile} and CM models \citep{nawrocki2009infernal,nawrocki2009structural} to achieve the same result.
@@
 Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
-People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut \citep{sheneman2006clearcut} program, or QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree.
+People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut program \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree.
 % AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
@@
 Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
-Taking the length of 16S to be 1543 nt,
+Taking the length of 16S to be 1543 nucleotides,
 % SILVA
@@
 Methods explicitly invoking trait evolution are notably absent, with the recent exception of \citep{langille2013predictive}.
-The results of this simple method are reasonable, but would a species tree reconciled with a collection of gene trees allow for better prediction?
+The results of this simple method are reasonable, but would a collection of gene trees reconciled with a species tree allow for better prediction?
 Perhaps improved methods, say involving whole-genome evolutionary modeling or models of metabolic network evolution, could shed light on the problem.
@@
 I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
-I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank to Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and thank the NIH (R01 HG005966-01) and NSF (1223057) for financial support.
+I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank Bastien Boussau, Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and thank the NIH (R01 HG005966-01) and NSF (1223057) for financial support.
 
@@
 \section{Figure Legends}
-%\FIGmassTransport
 
 \noindent
-Figure 1: \unifracLegend
+Figure \ref{fig:pd}: \pdLegend
 
 \noindent
-Figure 2: \dirtpilesLegend
+Figure \ref{fig:unifrac}: \unifracLegend
+
+\noindent
+Figure \ref{fig:dirtpiles}: \dirtpilesLegend
 

### 2013-10-03 sbreview.tex
SB submission formatting nonsense mostly.

@@
 \newcommand{\notforarxiv}[1]{}
+
+% % for SB submission
+% \renewcommand{\section}[1]{%
+% \bigskip
+% \begin{center}
+% \begin{Large}
+% \normalfont\scshape #1
+% \medskip
+% \end{Large}
+% \end{center}}
+% \renewcommand{\subsection}[1]{%
+% \bigskip
+% \begin{center}
+% \begin{large}
+% \normalfont\itshape #1
+% \end{large}
+% \end{center}}
+% \renewcommand{\subsubsection}[1]{%
+% \vspace{2ex}
+% \noindent
+% \textit{#1.}---}
+% \renewcommand{\tableofcontents}{}
+% \bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
+% \linespread{1.6}
 
@@
 \notforarxiv{
-\subsubsection{Abstract}
+\newpage
+\section{Abstract}
 }
@@
 \notforarxiv{
-\noindent (Keywords: human microbiome; microbial ecology; phylogenetic methods; review)\\
-\vspace{1.5in}
+\vspace{4.5in}
+\noindent (Keywords: human microbiome; microbial ecology; phylogenetic methods; 16S; metagenome)\\
+\newpage
 }
@@
 Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
-Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~\ref{fig:pd}).
+Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
 The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
@@
 Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance weighted phylogenetic diversity measures are not used in human microbiome studies.
-Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not.
+Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
 Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
@@
 \citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a special case of the earth-movers distance, is special in that it can be calculated in linear time, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
-The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree.
+The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
 \citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
@@
   \begin{center}
-    \includegraphics[width=1.6in]{figures/tree_dirtpiles.pdf}
   \end{center}
@@
 \noindent
-Figure \ref{fig:pd}: \pdLegend
+Figure 1: \pdLegend
 
 \noindent
-Figure \ref{fig:unifrac}: \unifracLegend
+Figure 2: \unifracLegend
 
 \noindent
-Figure \ref{fig:dirtpiles}: \dirtpilesLegend
-
-\clearpage
+Figure 3: \dirtpilesLegend
 

### 2013-10-07 sbreview.tex
Expanded OTU section H/T @b4winckler. Closes #4.
Björn-- in my reading, KS10 was about clustering "true" metagenomic
reads (rather than surveys) and so didn't quite fit. LMK if I missed
something. Thanks for the suggestions!

@@
 \citet{white2010alignment} show that different ways of doing this heuristic clustering can result in very different results.
-\citet{navlakha2009finding} develop methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies.
-\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values.
-
-\newcommand{\pdLegend}{
-    Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as blue circles and abundances are shown as the size of the circles.
-    Unweighted PD takes the sum of the branch lengths connecting samples.
-    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the areas of the edges.
-}
-\forarxiv{
-\begin{wrapfigure}[23]{r}{2.5in}
-  % \capstart
-  \vspace{-27pt}
-  \begin{center}
-    \includegraphics[width=2.5in]{figures/pd.pdf}
-  \end{center}
-  \vspace{-10pt}
-  \caption{\pdLegend}
-  \label{fig:pd}
-\end{wrapfigure}
-}
+\citet{navlakha2009finding} and \citet{WYS13} develop methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies,
+\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values,
+and \cite{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.
 
@@
 
-\newcommand{\unifracLegend}{
-    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
-    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
+\newcommand{\pdLegend}{
+    Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as blue circles and abundances are shown as the size of the circles.
+    Unweighted PD takes the sum of the branch lengths connecting samples.
+    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the areas of the edges.
 }
 \forarxiv{
-\begin{wrapfigure}[23]{r}{2.5in}
+\begin{wrapfigure}[25]{r}{2.2in}
   % \capstart
-  \vspace{-16pt}
+  \vspace{-14pt}
   \begin{center}
-    \includegraphics[width=2.5in]{figures/unifrac.pdf}
+    \includegraphics[width=2.2in]{figures/pd.pdf}
   \end{center}
   \vspace{-10pt}
-  \caption{\unifracLegend}
-  \label{fig:unifrac}
+  \caption{\pdLegend}
+  \label{fig:pd}
 \end{wrapfigure}
@@
 
+\newcommand{\unifracLegend}{
+    The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
+    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
+}
+\forarxiv{
+\begin{wrapfigure}[23]{r}{2.2in}
+  % \capstart
+  \vspace{-26pt}
+  \begin{center}
+    \includegraphics[width=2.2in]{figures/unifrac.pdf}
+  \end{center}
+  \vspace{-10pt}
+  \caption{\unifracLegend}
+  \label{fig:unifrac}
+\end{wrapfigure}
+}
+
 The most common way to use a between-sample pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
@@
 \forarxiv{
-\begin{wrapfigure}[19]{l}{1.6in}
+\begin{wrapfigure}[14]{l}{1.2in}
   % \capstart
-% \vspace{-24pt}
+  \vspace{-17pt}
   \begin{center}
+    \includegraphics[width=1.2in]{figures/tree_dirtpiles.pdf}
   \end{center}
@@
 
-There have been efforts to augment these ordination visualizations with additional information giving more structure to the visualizations.
+There have been several efforts to augment these ordination visualizations with additional information giving more structure to the visualizations.
 Biplots display variables (in the microbial case summarized by taxonomic labels) as points along with points representing samples \citep[e.g.][]{hewitt2013bacterial,lozupone2013meta}.

### 2014-04-04 reviews/response.txt
Adding decision and start of response.

> Reviews from Prof. Frank (Andy) Anderson
> ----------------------------------------

> Thank you for your Systematic Biology submission. It has been reviewed
> by Associate Editor Prof. Olivier Gascuel and three reviewers. Their
> comments are listed at the end of this letter. The reviewers and the
> AE provide some excellent constructive suggestions that I am sure you
> will appreciate.

> I have little to add that wasn't covered by one or more reviewers or
> Olivier in their fine reviews. As an instructor who has almost given
> up on trying to convince microbiology graduate students that they can
> do better analyses than neighbor joining, I was already sensitive to
> some of these differences, but this review helped me appreciate the
> interface between the two disciplines. I agree with Mooers that the
> level of detail varies across the manuscript, but I mainly noticed
> when a topic was summarized quickly and I didn't really understand it
> (I note that Mooers and I both had a question about how tax2tree
> works).

> I have a few comments and questions that may help clarify some of the
> writing. Altogether, I think the manuscript really requires only
> several minor revisions.

> Like Olivier, I note that some of my comments may overlap those of the
> reviewers.

> 1) Pg. 9: "As described above "metagenome" means that data is sheared
> randomly across the genome rather than amplified from a specific
> location," - I assume you mean DNA is sheared randomly?

> 2) Remember, "data" still is a plural noun, at least under my watch!
> Check for "this data", "data is", "data was", etc

> 3) Pg. 10: I don't really understand what "tax2tree" does, based on
> the description here. Does that matter?

> 4) Pg. 14-15: I also don't really understand what UniFrac is or how it
> works, even after looking at Figure 2 (Olivier made a similar
> comment). Could we get some help here?

> 5) The figures need a bit of work. There seems to be some sort of
> equation at the bottom of Figure 1, but I can't decipher it. What do
> the colors mean in Figure 2? I don't find this to be very clear.

> 6) Pg. 20: "How to formulate a useful notion of independent contrasts
> (Felsenstein 1985) on such an object?" - Awkward wording.

> 7) Pg. 23: I don't understand the distinction between "base
> observations" and sample observations here. Can this be clarified?

> 8) Pg. 24: "Future will continue to bring deeper sequencing on more
> samples" - I guess this should be "The future will continue"?

> My recommendation is acceptance pending minor to major changes.
> Actually, no change asked by the reviewers and me (see below) requires
> major efforts, but we all come from different origins, with different
> points of view and expectations, and thus summing all these demands
> will require a bit of time and attention. Basically, we all ask for
> more details and comments here and there, to make the paper more
> precise, comprehensive and self-contained. I am confident that you'll
> find our suggestions useful. I don't believe that the paper should be
> shortened in some places (or very locally), as suggested by Arne
> Mooers. When submitting your revision, please prepare a cover letter
> containing all the comments by the referees and me, with your
> changes/responses.

> Specific comments, possibly redundant with those from the reviewers,
> using Systematic Biology page numbering:

> P4	these techniques AND could

> P8	(Kuczynski et al. 2010). However, for more subtle information

> P8	"for more subtle information deeper sequencing is required" :
> what do you have in mind here?
> Indeed "distinguish between two rather different samples" is a modest
> goal Actually, it's not clear in this subsection what is meant by
> "Microbial community estimation" (title of the subsection). What is
> the goal here? Identify and classify microbes? Measure their
> abundance? Etc.? A few lines and explanations should suffice.

> P11	What is the level of uncertainty in current bacterial
> phylogenies? For example, one commonly considers that only a few
> unknowns remain in mammals phylogeny; what about bacteria? Are people
> discussing the details or the broad lines? Provide a few illustrative
> examples.

> P13 	"There are many OTU inference methods with various speeds and
> strategies" and the two following paragraphs. Describe in a simple way
> the basic principles of these methods, before discussing their
> relative merits (this remark may apply in other places).

> P14 	You describe PD and suggests that it should be used more often.
> But what are the standard alpha diversity measures, and their
> limitations?

> P15 	Are "partially abundance weighted diversity measures" using some
> phylogenetic approaches? Be a bit more specific when describing and
> discussing methods. Here and in many other places (e.g. UniFrac,
> weighted PD, weighted UniFrac, etc.).

> P15 	"Evans and Matsen (2012) showed that weighted UniFrac" check
> sentence.

> P16 	explanations and Fig. 3 on earth-movers distance are unclear to
> me.

> P17 	"By doing so they have recovered more novel protein families
> than they would have using methods organized by taxonomy". I'm lost
> here, as it is said just above that they have derived a
> "phylogeny-driven genomic encyclopedia of Bacteria and Archaea".

> P23-24 Paragraph "In the same vein This would be a false positive." is
> unclear to me.

> P24 Check "The presently used approach is rarefaction, which means
> uniform sub-sampling, typically without replacement to the number of
> reads in the lowest abundance sample"

> P25 Check "Comparative studies will continue to investigate what
> shapes and is shaped by the microbiome"

> P26 "during the 2013 LIRMM Mathematical and Computational Evolutionary
> Biology workshop", change LIRMM into MCEB

> Fig 2. It took time to me to understand and check on the Web that the
> tree must be rooted, to compute the unifrac distance. This is one
> example, among others, that more details are required when describing
> methods, criteria, measures, etc. Some more figures could help, or one
> appendix with formal definitions of the main concepts and tools. At
> least, details and explanations have to be added in the main text.

> Your reference list is extremely long, mostly due to the fact that you
> are citing papers with a lot of authors. Check Syst Biol
> recommendation to authors (I assume that our editorial office will
> answer and help).

> Reviewer 1, Alexis Stamatakis
> -----------------------------

> Comments:
> This is a well written and in depth review on the usage of
> phylogenetics in human microbiome research, which I enjoyed reading.

> I only have a few minor remarks:

> It might be a consideration to cite our recent paper:

> J. Zhang, P. Kapli, P. Pavlidis, A. Stamatakis: "A General Species
> Delimitation Method with Applications to Phylogenetic Placements". In
> Bioinformatics (2013), 29 (22): 2869-2876.

> It would fit will into section "The role of OTUs".

> Apart from that, I believe that a more critical and slightly extended
> discussion of taxonomic classifiers in the very last part of the paper
> should be included. From my point of view, this is still a mostly
> unresolved problem and the currently best classifier, which you may
> also want to cite,  still shows pretty horrible accuracy
> (http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0049334)
> at the genus and family level.

> I would also appreciate if you could explicitly mention the Gartner
> hype curve in the discussion part and provide an estimate as to where
> we are on this curve at present.

> Some typos:

> page 5 lines 7-8 "these techniques and could be used" there seems to
> be a verb missing.

> page 14 lines 12 Here you may add our recent work on PTP, since it
> strives to conduct OTU clustering and phylogenetic inference in one
> single step/pipeline.

> page 15 l 50: in "distance, is special" seems that it should read
> "distance, it is special"

> page 20: lines 51-56: In fact we are currently developing a tool for
> automatic tree extension as described here. While not published yet,
> it is already available as open-source code at:
> http://sco.h-its.org/exelixis/web/software/put/index.html

> Reviewer 2, Arne Mooers
> -----------------------

> This was an interesting manuscript to read, and I compliment whoever
> thought to review Human Microbiome Research (HMR) for the SB
> readership. Overall, there is a lot of very interesting work reported
> on here, though I found the level of detail to be a bit uneven:
> sometimes we are told perhaps too much, and often, what feels like too
> little, in order to make the connections between HMR and SB. There is
> also seems to be some elision between HMR research and metagenomics
> more generally. That isn't a bad thing, but one does sometimes forget
> that we are focussed on HMR.

> Numbers refer to the SB manuscript pagination, rather than to the
> original latex.

> 4-45 additional to
> 5-8 delete and
> 7-23 delete though

> The section "The Human Microbiome" is quite detailed (discussing mouse
> to human transplants wrt obesity, etc.). Given the intense media
> reporting of the work coming from this area, SB readers may not need
> to be primed of the importance of this field. If this were shortened,
> there could be room to expand some of the technical bits that SB
> readers will want that follow. The last part of the section  - about
> sticking to eubacteria and a non-molecular biology approach do seem
> germane.

> 8-12. these data (I guess we will eventually decide that data is
> singular - up to the editors)

> 8-37++ Whereas Woese ++ This is a bit wordy and unnecessary - we know
> that Woese didn't have PCR

> 10-30 Which "databases" ? - this has been written so far as if it were
> being applied to samples, not databases. I am not clear what the unit
> of observation is here.

> 10-32 We quickly switch here from easy background (nonPCR random
> sequencing) to something brand new (at least to me). "At least some
> portion of the rest of the data can be taxonomically classified
> (methods reviewed by Mande et al. 2012); Treangen et al. (2013) report
> speedups and much higher accuracy when reads are assembled before they
> are classified."
> "Rest of data" refers to what - reads found in a database, or reads
> from a sample? "Accuracy" refers to what? "Speed-ups" refer to what?

> 11-53. Here is the first real point of contact between phylogenetics
> and HMR. But what is an "F-measure of precision and recall" - there is
> no citation.

> 12-28 and 12-32. What is the measure of performance that this refers
> to? How do we decide if a method performs well?

> 12-29. Do readers of SB now what "k-mer composition" is? Is it k-means
> clustering?

> 12-37++. Is the MEGAN approach compositional or phylogenetic? I
> believe it is meant to be a hybrid approach (BLAST being
> compositional, but the taxonomic labeling considered phylogenetic),
> but this could be clarified.

> 13-29. It is not clear what heuristics are required for a cut-off
> method. The author says that he is going to describe the OTU inference
> industry, but the paragraph is really just a list of acronyms, and
> then refers obliquely to a "contingency table" of OTU observations.
> So, for example, what is a notion of goodness of clustering that the
> heuristics are trying to achieve? What are the sorts of species
> divisions found in taxonomies? This is an area where more would be
> helpful - and where a shorter introduction might offer space.

> 14-40. Whereas just about every 16S survey investigation involves an
> OTU-based alpha diversity estimate, only a few involve PD: unweighted
> PD has been applied to some 16S survey data (Lozupone and Knight 2007;
> Costello et al. 2009) and to metagenomic reads in a set of marker
> genes (Kembel et al. 2011) - This is an odd pair of sentences: is the
> L&K the only study that used unweighted PD? If so, in what context?

> 14-46. Check this sentence - perhaps start with "While are commonly
> used, abundance weighted PD measures are not.

> 15-30. Perhaps "In contrast to measures of alpha diversity, measures
> of beta diversity do make use of phylogenetics"  This is actually an
> interesting contrast. Is this just historical contingency, or is there
> a logical reason why the two areas disagree on the use of phylogenetic
> information?

> 15-xx. This seems a strong point of contact between metazoan
> phylogenetics and the microbiome folks. UniFrac is clearly described
> on its own terms, except perhaps where I would have written "with the
> fraction of edges connecting samples that are common to two samples"
> rather than "edges that connect between samples" (which doesn't
> strictly make sense). I believe there is the same measure of
> beta-diversity in the community-phylogenetics literature? What is the
> closest analogue?

> 15-4955. The description of Evans and Matsen seems a bit detailed in
> the context -

> 16-16. This requires another sentence of explanation. From this
> reading, the Unifrac method offers a single distance between a pair of
> samples (a ratio of two numbers of edges), rather than a matrix.

> 16-42..51. This is a short paragraph that would seem to explicitly
> link HMR and SB. Why is it so telegraphic?

> 17-45. This seems an odd heading for this section, which seems to be
> about horizontal gene transfer. Given the amount of technical detail
> elsewhere, why are we not told how horizontal gene transfer is
> inferred in these HMR studies?

> Phylogenetic inference. I found this survey very useful. Is there a
> reason this is not an earlier section of the paper, given these are
> the trees upon which many of the comparisons referred to in previous
> sections are based?

> 19-24. "which has ported in the" Is this the correct structure?

> 20-50. "different parameter regimes" - this phrase is used several
> times in this review. What does it mean here? When would we prefer to
> add sequences to a fixed tree rather than re-infer the totality? Is
> there an analogous situation in SB - i.e. is there something SB can
> learn from HMR here?

> I found the challenges section to be rich and well-written, with
> questions that will resonate with the SB community - e.g. the
> discussion of the proxy problem (how edge lengths predict feature
> diversity). The fact that 16S distance does predict whole genome
> distance is highlighted here - it is presented in previous sections,
> but perhaps it could be given more prominence there, given its
> prominence here. This is a general problem.

> 22-10. "Microbial ecology" Perhaps "and explicitly for human microbial
> ecology" to remind us that HMR is the focus here?

> 23-23. "Whole organism approach"  Will this approach be known to SB
> readers? I cannot imagine what is being done here.

> 25-31. "At least for the next several years we can expect more related
> work." What does this mean? Related to what has been reported above?

> 25-32. The future

> I found the final sentence to leave us dangling. After having reviewed
> this manuscript, I could not say exactly where the intersection of SB
> and HMR will be most fruitful for HMR, nor what in HMR might be
> fruitful for SB. The conclusion seems to suggest that phylogenetics
> would not be used in diagnostics, for example. Will it be important in
> primary research then? In constructing the taxonomies the diagnostics
> may require? Finally, the abstract says the review ends by describing
> "the current challenges to phylogenetics coming from this type of
> work."  My reading of the relevant sections is that there is more
> discussion of challenges to phylogenetic HMR research (sampling
> biases, robustness of UniFrac to incorrect tree reconstruction) than
> to phylogenetics per se. The discussion of how to update known trees
> with new sequences is an exception.

> Reviewer 3, Frederic Delsuc
> ---------------------------

> This manuscript presents a review of how phylogenetic tree-thinking
> and methodology is currently used in the study of the human
> microbiome. The author not only reviews the different ways by which
> phylogenetics is currently percolated into microbial ecology, but also
> gives his personal views on the future phylogenetic challenges
> associated with microbiome studies, and insists on the crucial role
> that molecular phylogeneticists has to play in this quickly expanding
> field. Overall, I enjoyed reading this particularly timely manuscript,
> which I find well written. I only have minor comments/suggestions for
> potential improvement.

> Specific comments:
> 1. I think that it should be made clear, from the very beginning of
> the introduction, the definition of the microbiome in order to avoid
> confusion. Indeed, in the current literature, "microbiota" (i.e. the
> collection of microbes living in association with a given organism)
> and "microbiome" (i.e. the collection of genomes of those microbes)
> are more and more often used interchangeably, even though they are not
> referring to the same organizational level. In that sense, the
> definition of the "human microbiome" given at the top of page 5 ("The
> human microbiome is the collection of microbial organisms that live
> inside of and on the surface of humans.") does not seem to be
> accurate. Please clarify.

> 2. In the section introducing the human microbiome, I think that a
> paragraph on the origin and the way in which the microbiome is
> transmitted vertically between the mother and infant is lacking. The
> recent review on maternal transmission by Funkhouser and Bordenstein
> 2013 is cited later, but it would be nice to mention it on page 5,
> especially in the context of the some examples on the impact of birth
> method (vaginal birth vs. cesarean delivery) on future health of
> newborns.

> 3. On page 6, lines 26-27, it might be worth citing also the recent
> paper by Findley et al. (doi:10.1038/nature12171) on the topography of
> fungal and bacterial on human skin.

> 4. On page 6, line 56, maybe also specify that "survey" is the
> equivalent of barcoding, a term with which the Systematic Biology's
> community is more familiar.

> 5. On page 14, lines 14-19, maybe mention the widespread use of
> "rarefaction curves" to enable the comparison between samples of
> different sequencing depth or methods (e.g. 454 vs. Illumina).

> 6. On page 17, lines 54-55, please add some recent references on the
> current debate around the rooting of the placental tree. I don't think
> that most readers of Syst. Biol. will be familiar with this
> discussion.

> 7. On page 18, lines 26-27, RAxML is also implemented in QIIME.

> 8. On page 20, I think that this might be the place to discuss the
> potential use of phylogenetic networks in microbiome research (see
> Parks and Beiko 2012 Mol Biol Evol).

> 9. On page 24, when citing uBiome, it might also be fair to refer the
> similar crowd-funded American Gut Project.

> Typos:
> P3, L40: I would rather use "species delimitation" instead of "species definition".
> P4, L9: "These techniques and could be used"
> P6, L32: "we will be primarily be describing"
> P9, L29, L30: Maybe replace "diversity" by "variability"
> P18, L32: "a long history as a means to sequentially build"
> P19, L16: "GreenGenes 13_5 alignment has is almost 80% gaps"

### 2014-04-05 reviews/response.txt
First pass, fixing typos.

@@
 
+Yes, thanks.
+
 > 2) Remember, "data" still is a plural noun, at least under my watch!
 > Check for "this data", "data is", "data was", etc
+
+Searched and replaced. Thanks!
 
@@
 
+I have changed this to "How might one formulate a useful notion of independent contrasts (Felsenstein 1985) on such an object?" I hope that's better.
+
 > 7) Pg. 23: I don't understand the distinction between "base
@@
 > samples" - I guess this should be "The future will continue"?
+
+Thanks.
+
+AE, Olivier Gascuel
+-------------------
 
@@
 
+Thanks!
+
 > P8	(Kuczynski et al. 2010). However, for more subtle information
-
 > P8	"for more subtle information deeper sequencing is required" :
@@
 
+I have done so, so that the sentence reads "... 2013 MCEB Mathematical and Computational Evolutionary Biology workshop ..." is that right?
+
 > Fig 2. It took time to me to understand and check on the Web that the
@@
 
+Fixed, thanks.
+
 > page 14 lines 12 Here you may add our recent work on PTP, since it
@@
 > "distance, it is special"
+
+Thanks, I've rearranged here.
 
@@
 > 4-45 additional to
+
 > 5-8 delete and
+
+Fixed.
+
 > 7-23 delete though
+
+Nice catch!
 
@@
 > P4, L9: "These techniques and could be used"
+
+Fixed.
+
 > P6, L32: "we will be primarily be describing"
+
+Whoops!
+
 > P9, L29, L30: Maybe replace "diversity" by "variability"
+
+Thanks.
+
 > P18, L32: "a long history as a means to sequentially build"
+
 > P19, L16: "GreenGenes 13_5 alignment has is almost 80% gaps"
+
+Yikes, thanks.

### 2014-04-05 sbreview.tex
First pass, fixing typos.

@@
 \usepackage[labelfont=bf,labelsep=period,font=small]{caption} % bold the 'Figure #' in the caption and separate it with a period
-\newcommand{\forarxiv}[1]{#1}
-\newcommand{\notforarxiv}[1]{}
+\newcommand{\forarxiv}[1]{}
+\newcommand{\notforarxiv}[1]{#1}
 
-% % for SB submission
-% \renewcommand{\section}[1]{%
-% \bigskip
-% \begin{center}
-% \begin{Large}
-% \normalfont\scshape #1
-% \medskip
-% \end{Large}
-% \end{center}}
-% \renewcommand{\subsection}[1]{%
-% \bigskip
-% \begin{center}
-% \begin{large}
-% \normalfont\itshape #1
-% \end{large}
-% \end{center}}
-% \renewcommand{\subsubsection}[1]{%
-% \vspace{2ex}
-% \noindent
-% \textit{#1.}---}
-% \renewcommand{\tableofcontents}{}
-% \bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
-% \linespread{1.6}
+% for SB submission
+\renewcommand{\section}[1]{%
+\bigskip
+\begin{center}
+\begin{Large}
+\normalfont\scshape #1
+\medskip
+\end{Large}
+\end{center}}
+\renewcommand{\subsection}[1]{%
+\bigskip
+\begin{center}
+\begin{large}
+\normalfont\itshape #1
+\end{large}
+\end{center}}
+\renewcommand{\subsubsection}[1]{%
+\vspace{2ex}
+\noindent
+\textit{#1.}---}
+\renewcommand{\tableofcontents}{}
+\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
+\linespread{1.6}
 
@@
 Although there is something of a divide between phylogeny as practiced as part of microbial ecology on one hand and that for multicellular organisms on the other, there are many parallels between the two enterprises.
-Both communities struggle with issues of sequence alignment, large-scale tree reconstruction, and species definition.
+Both communities struggle with issues of sequence alignment, large-scale tree reconstruction, and species delimitation.
 However, approaches differ between the microbial ecology community and that of eukaryotic phylogenetics, in part because the scope of the former contains an almost unlimited diversity of organisms, leading to additional problems above the usual.
@@
 Because of this high level of diversity and challenges with species definitions, microbial ecology researchers have developed their own explicitly phylogenetic techniques for comparing samples rather than comparing on the level of species abundances.
-Although there is some overlap with previous literature, these techniques and could be used in a wider setting and may deserve broader consideration by the phylogenetics community.
+Although there is some overlap with previous literature, these techniques could be used in a wider setting and may deserve broader consideration by the phylogenetics community.
 
@@
 
-Bacteria have been the primary focus of human microbiome research, and other domains have been investigated though to a lesser extent.
+Bacteria have been the primary focus of human microbiome research, and other domains have been investigated to a lesser extent.
 Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea}.
@@
 
-In this paper we will be primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perspective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
+In this paper we will primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perspective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
 Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.
@@
 
-The Human Microbiome Project \citep{methe2012framework} generated lots of survey, metagenome, and whole-genome sequencing data and this data is available on a dedicated website\footnote{\url{http://www.hmpdacc.org/}}.
+The Human Microbiome Project \citep{methe2012framework} generated lots of survey, metagenome, and whole-genome sequencing data and these data are available on a dedicated website\footnote{\url{http://www.hmpdacc.org/}}.
 The MetaHIT study \citep{qin2010human} also generated lots of data but it is not available to outside researchers.
@@
 \subsection{Metagenomes}
-As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
+As described above ``metagenome" means that DNA is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
 Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, however extraction efficiency concerns remain and multiplex sequencing is known to have biases of its own.
@@
 That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
-This is proven to be a useful strategy \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}, however, because of the diversity of gene repertoire in microbes, this core gene set may be relatively small.
+This is proven to be a useful strategy \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}, however, because of the variability of gene repertoire in microbes, this core gene set may be relatively small.
 Indeed, even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
@@
 
-Metagenomic data is often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
+Metagenomic data sets are often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
 Discussing these methods is beyond the scope of this paper, as is ``metatranscriptomics": the sequencing of mRNA in bulk.
@@
 These dissimilarity measures have hundreds of citations.
-\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a special case of the earth-movers distance, is special in that it can be calculated in linear time, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
+\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, that it can be calculated in linear time in contrast to many applications of the earth-movers distance, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
 The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
@@
 With some notable exceptions, mainstream applications of phylogenetics to a collection of human-associated microbes have typically been with the idea of finding ``the'' tree of such a collection rather than explicitly exploring divergence between various gene trees.
-As described above, whole-genome data is typically used to directly infer functional information rather than information concerning ancestry.
+As described above, whole-genome data sets are typically used to directly infer functional information rather than information concerning ancestry.
 The continuing debate concerning whether a microbial tree of life is a useful concept \citep{bapteste2009prokaryotic,caro2012bacterial} does not seem to have dampened human microbiome researchers' enthusiasm for using a single such tree.
@@
 This is defended by saying that for applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
-Furthermore, given that the underlying data is typically 16S alone we can expect some topological inaccuracy in reconstructing the ``tree of cells'' even with the best methods.
+Furthermore, given that the underlying data sets are typically 16S alone we can expect some topological inaccuracy in reconstructing the ``tree of cells'' even with the best methods.
 Additionally, as specified below, these data sets can be very large.
@@
 % .79914084873730799272
-while the 1,262,986 sequence GreenGenes 13\_5 alignment has is almost 80\% gap.
+while the 1,262,986 sequence GreenGenes 13\_5 alignment is almost 80\% gap.
 The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
@@
 These two extremes are each useful in different parameter regimes, but something between the two would be helpful as well.
-For example, sequence data is continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
+For example, sequence data sets are continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
 % Surprising that methods such as DCM \citep{huson1999disk} have not gained traction.
@@
 Here again the ``tree of life'' problem can be formulated in a practical light: what representation allows for the best prediction of features of underlying genomes?
-How to formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on such an object?
+How might one formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on such an object?
 It is quite possible that inference using a more complete representation would not be able to overcome the inherent noise of the data, but further exploration seems warranted as simple methods give reasonable results.
@@
 At least for the next several years we can expect more related work.
-Future will continue to bring deeper sequencing on more samples.
+Future research projects will continue to bring deeper sequencing on more samples.
 The uBiome project\footnote{\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
@@
 \section{Acknowledgements}
-I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 LIRMM Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
+I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 MCEB Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
 I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank Bastien Boussau, Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and thank the NIH (R01 HG005966-01) and NSF (1223057) for financial support.

### 2014-04-11 reviews/response.txt
WIP

@@
 
+I have rephrased so that it's clear that the matrix comes from applying
+the measure to all pairs of sequences in a sample.
+
 > 16-42..51. This is a short paragraph that would seem to explicitly
 > link HMR and SB. Why is it so telegraphic?
+
+I have added a clarifying phrase, but I'm not sure what more is needed.
+The two communities have both identified a given technique (although
+neither has really gotten much traction).
 
@@
 
+Explained now, thanks.
+
 > Phylogenetic inference. I found this survey very useful. Is there a
@@
 
+I suppose I wanted to talk about all of the interesting things that
+microbiome researchers are doing with trees before
+borderline-criticizing the ways that they make them! The readers of SB
+know what it means to infer a tree, so I don't think that this section
+is a prerequisite for the previous ones.
+
 > 19-24. "which has ported in the" Is this the correct structure?
+
+That was what was intended, but I've updated it to use less
+programmer-speak.
 
@@
 > learn from HMR here?
+
+I've simply removed the sentence. The idea of this paragraph is to
+motivate updating of phylogenetic trees with new sequences, and I've
+more fully explained that and added a reference.
 
@@
 
+I have added a sentence emphasizing that this assumption underlies the
+way researchers interpret UniFrac.
+
 > 22-10. "Microbial ecology" Perhaps "and explicitly for human microbial
 > ecology" to remind us that HMR is the focus here?
+
+Added, thanks.
 
@@
 
+You are right that this was not explained properly. I have simply
+removed the paragraph.
+
 > 25-31. "At least for the next several years we can expect more related
 > work." What does this mean? Related to what has been reported above?
+
+I have made this more specific.
 
@@
 > fruitful for SB. The conclusion seems to suggest that phylogenetics
-> would not be used in diagnostics, for example. Will it be important in
-> primary research then? In constructing the taxonomies the diagnostics
-> may require? Finally, the abstract says the review ends by describing
+> would not be used in diagnostics, for example.
+
+I have clarified this sentence now to clarify that some applications
+will go but many new ones will be needed.
+
+> Will it be important in primary research then? In constructing the
+> taxonomies the diagnostics may require?
+
+The rest of the paper is about how phylogenetics has impacted HMR, and
+these applications will continue into the future. I'm not sure how to
+make this more clear.
+
+> Finally, the abstract says the review ends by describing
 > "the current challenges to phylogenetics coming from this type of
@@
 > with new sequences is an exception.
+
+I suppose I don't differentiate between the two.
 
@@
 
+Added.
+
 > 3. On page 6, lines 26-27, it might be worth citing also the recent
@@
 > fungal and bacterial on human skin.
+
+Thanks.
 
@@
 
+Added, thanks.
+
 > 5. On page 14, lines 14-19, maybe mention the widespread use of
@@
 > different sequencing depth or methods (e.g. 454 vs. Illumina).
+
+I do describe strategies for read count normalization in the next paragraph, but have added the "rarefaction curve" terminology.
 
@@
 
+Good point. I've put in the two 2013 MBE papers. If there is something else I should cite let me know.
+
 > 7. On page 18, lines 26-27, RAxML is also implemented in QIIME.
+
+Added.
 
@@
 
+Thank you for pointing me to this reference.
+
 > 9. On page 24, when citing uBiome, it might also be fair to refer the
@@
 
+Thanks.
+
 > Typos:
 > P3, L40: I would rather use "species delimitation" instead of "species definition".
+
+Thanks.
+
 > P4, L9: "These techniques and could be used"
@@
 
+Reworded, thanks.
+
 > P19, L16: "GreenGenes 13_5 alignment has is almost 80% gaps"

### 2014-04-11 sbreview.tex
WIP

@@
 Much of our metabolic interaction with the outside world is mediated by our microbiome, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
+Our microbiomes have been historically transmitted vertically between mother and infant by seeding in the birth canal and breastfeeding;  immune dysfunction has been associated with the interruption of these practices \citep[reviewed in][]{funkhouser2013mom}.
 In this section I will briefly review what is known about the human microbiome and its effect on our health.
@@
 Bacteria have been the primary focus of human microbiome research, and other domains have been investigated to a lesser extent.
-Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea}.
+Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea} and have a nonuniform distribution across the human skin \cite{Findley2013-rm}.
 Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
@@
 This gene was chosen because it has regions of high and low diversity, which enable resolution on a variety of evolutionary time scales.
-Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled surveys of almost all bacteria regardless of whether they can be cultured.
+Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled detection of almost all bacteria and archaea regardless of whether they can be cultured.
+In microbial ecology, the census of bacteria in a given environment using marker gene amplification and sequencing are generally called ``marker gene surveys,'' which is equivalent to the ``barcoding'' terminology more commonly used for eukaryotes.
 
@@
 Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
-\citet{mccoy2013abundance} have recently shown that partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and "normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
-\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths.
+\citet{mccoy2013abundance} have recently shown that partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
+\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''
 
@@
 These dissimilarity measures have hundreds of citations.
-\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, that it can be calculated in linear time in contrast to many applications of the earth-movers distance, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
+\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
 The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
 \citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
+
+Faith2009-gs
 
@@
 
-The most common way to use a between-sample pairwise distance matrix found from an application of UniFrac is to apply an ordination method such as principal coordinates analysis.
+The most common way to use a distance matrix obtained from applying UniFrac to all pairs of samples is to apply an ordination method such as principal coordinates analysis.
 Indeed, the separation of two communities in a principal components plot is often used as \emph{prima facie} evidence of a difference between them
@@
 They focus on distances between trees induced by matrix metrics on the corresponding adjacency matrices.
-A similar framework was used by \citet{steel2008maximum} to construct maximum likelihood supertrees.
+A similar framework was used by \citet{steel2008maximum} to construct maximum likelihood supertrees, for which they use common distance measures in phylogenetics such as the subtree-prune-regraft metric.
 
@@
 16S distance is frequently used as a proxy for a functional comparison between human microbiome samples.
+Indeed, researchers using UniFrac frequently don't think of their comparisons as being in terms of a single gene, but rather in terms of an abstracted measure of community function.
 Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes is commonly acquired horizontally as well as vertically, and horizontal transmission leaves no trace in 16S ancestry.
@@
 
-\subsection{Genome-scale inquiries using phylogenetics}
+\subsection{Horizontal gene transfer}
 With some notable exceptions, mainstream applications of phylogenetics to a collection of human-associated microbes have typically been with the idea of finding ``the'' tree of such a collection rather than explicitly exploring divergence between various gene trees.
@@
 
-Nevertheless, the work that has been done to infer horizontal gene transfer in the human microbiome has revealed interesting results.
+Nevertheless, the work that has been done concerning horizontal gene transfer in the human microbiome has revealed interesting results.
 \citet{hehemann2010transfer} found that a seaweed gene has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
@@
 \citet{stecher2012gut} find that in a mouse model, horizontal transfer between pathogenic bacteria is blocked by commensal bacteria except for periods of gut inflammation.
+Horizontal transfer of genes is inferred in these studies by finding highly similar subsequences in otherwise less related organisms.
 
@@
 Additionally, as specified below, these data sets can be very large.
-There does not seem to be contentious discussion of specific features of the inferred trees equivalent to, say, the current discussion around the rooting of the placental mammal tree.
+There does not seem to be contentious discussion of specific features of the inferred trees equivalent to, say, the current discussion around the rooting of the placental mammal tree \citep{morgan2013heterogeneous,romiguier2013less}.
 Given this perspective, it is not surprising that Bayesian phylogenetic methods and methods that incorporate alignment uncertainty are absent.
@@
 Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
-People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has ported in the clearcut program \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut and FastTree.
+People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has incorporated the clearcut code \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut, FastTree, and RAxML \cite{Stamatakis2006-yz}.
 % AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
@@
 The scale of the data has motivated strategies other than complete phylogenetic inference, such as the insertion of sequences into an existing phylogenetic tree.
-Although such insertion has a long history as a means to sequentially build a phylogenetic tree \citep{kluge1969quantitative}, the first software with insertion specifically as a goal was the parsimony insertion tool in the ARB program by \citet{ludwig2004arb}.
+Although such insertion has long been used as a means to build a phylogenetic tree sequentially \citep{kluge1969quantitative}, the first software with insertion specifically as a goal was the parsimony insertion tool in the ARB program by \citet{ludwig2004arb}.
 ARB is commonly used to reconstruct a full tree by direct insertion.
@@
 
-% Networks are not considered.
-
 \section{Phylogenetic challenges and opportunities in human microbiome research}
@@
 
-One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference versus sequence insertion or placement that leaves the ``reference tree'' fixed.
-These two extremes are each useful in different parameter regimes, but something between the two would be helpful as well.
+One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference versus sequence insertion or placement that leaves the ``reference tree'' fixed, with the idea that such an algorithm would retain the efficiency characteristics of placement algorithms while allowing the reference tree to change.
 For example, sequence data sets are continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
-% Surprising that methods such as DCM \citep{huson1999disk} have not gained traction.
+\citet{Izquierdo-Carrasco2014-hu} have taken a step in this direction by developing an informatic framework that updates alignments and builds larger trees using previous smaller trees as starting points.
 
@@
 If we allow an arbitrary ``hidden'' object, what such object would perform best?
-For example, collections of reconciled gene trees in the presence of gene deletion, transfer, and loss \citep[see][for interesting recent results]{szollHosi2013efficient,szollHosi2013lateral} could be used.
+An important first step in this direction has been taken by \citet{Parks2012-os} who define community comparison metrics on phylogenetic split systems.
+An alternative would be to use collections of reconciled gene trees in the presence of gene deletion, transfer, and loss \citep[e.g.][]{szollHosi2013efficient,szollHosi2013lateral}.
 
@@
 
-An important project for microbial ecology is to model community assembly, and perhaps further phylogeny-aware methods could be used.
+Developing community assembly models forms an important project for microbial ecology generally and human microbial ecology in particular, phylogeny-aware methods could be further developed.
 One way to model community assembly is to apply Hubbell's neutral theory \citep{fierer2012animalcules,costello2012application}.
@@
 
-The approach of considering a collection of genes and their metabolic network as a meta-organism has yielded some interesting results \citep{borenstein2008large,greenblum2012metagenomic}.
-A clear limitation to this approach is that cellular boundaries are ignored: populations are not a freely diffusing soup.
-Could these approaches be improved by using phylogenetic methods to reconstruct the compartmentalization of genomes and processes into cells?
-% As an extension microbes have to be close to one another-- requires population modeling.
-
 As described above, \citet{morgan2010metagenomic} showed that various microbes have different DNA extraction efficiencies, meaning that the representation of marker gene sequences is not representative of the actual communities.
@@
 
-In general, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
+For both alpha and beta diversity studies, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
 One type of normalization handles differential depths of sequencing across samples.
@@
 \citet{holmes2012dirichlet} and \citet{la2012hypothesis} use models where read counts are modeled as overdispersed samples of the true abundance and provide methods for statistical testing.
-\citet{paulson2013differential} estimate true abundances using a zero-inflated Gaussian model for read counts.
+\citet{paulson2013differential} estimate true abundances using a zero-inflated Gaussian mixture model for read counts, while \citet{McMurdie2014-mn} claim better performance using a Gamma-Poisson mixture.
+
 This work could be extended to a phylogenetic context by making use of the relationship between OTUs, and modeling the way in which the abundance of one OTU may increase the abundance of a related OTU because of sequencing error or a change of condition that changes the abundance of both.
@@
 What can we expect next at the intersection of phylogenetics and the human microbiome?
-At least for the next several years we can expect more related work.
+At least for the next several years we can expect the research questions described above to continue to unfold.
 Future research projects will continue to bring deeper sequencing on more samples.
-The uBiome project\footnote{\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
+The uBiome \footnote{\url{http://ubiome.com/}} and American Gut \footnote{\url{http://americangut.org/}} projects promise to bring gut microbiome sequencing to the average citizen for a low price.
 Comparative studies will continue to investigate what shapes and is shaped by the microbiome.
@@
 
-Opportunities for clinical applications may present themselves, but only some of these will be sequencing based.
+Opportunities for clinical applications will present themselves, although the specifics will change.
 For example, routine 16S sequencing is likely to be replaced soon by Matrix-Assisted Laser Desorption Ionization--Time of Flight (MALDI-TOF) mass spectrometry for assignment of a single microbe grown in culture to a database entry \citep{clark2013matrix}.
-However, for diagnoses that are on the level of microbial communities, sequencing and consequent analysis methods will still be required \citep[reviewed in][]{Rogers2013271}.
-Inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}.
+However, for diagnoses that are on the level of microbial communities, sequencing and consequent analysis methods (possibly including phylogenetics) will still be required \citep[reviewed in][]{Rogers2013271}.
+Inexpensive whole-genome sequencing will certainly have a profound impact on clinical practice and epidemiological studies \citep{didelot2012transforming}, and this genome-scale data requires evolutionary analysis methods to interpret it.
 For all of these measures, it will be important to have rigorous means of quantifying uncertainty for robust diagnostic applications.

### 2014-04-11 reviews/response.txt
Done for the day.

@@
 
+I can't quite understand this suggestion. All three papers listed used
+unweighted PD.
+
 > 14-46. Check this sentence - perhaps start with "While are commonly
 > used, abundance weighted PD measures are not.
+
+I have updated this sentence based on your suggestion, thanks.
 
@@
 > information?
+
+I think that this refers to 14-30, and I've simply removed the sentence.
+With regards to your question, I don't know. It seems surprising to me.
 
@@
 
+I don't follow your "with the fraction of edges connecting samples that are common to two samples" correction.
+There is some overlap with the PD literature as now described and cited.
+
 > 15-4955. The description of Evans and Matsen seems a bit detailed in
 > the context -
+
+I've shortened and simplified, thanks.
 

### 2014-04-11 sbreview.tex
Done for the day.

@@
 
-Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
 Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
@@
 
-Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance weighted phylogenetic diversity measures are not used in human microbiome studies.
+Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are commonly used in human microbiome studies, abundance weighted phylogenetic diversity measures are not.
 Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
@@
 \citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
-
-Faith2009-gs
+There are clear connections between UniFrac and phylogenetic diversity explored by \cite{Faith2009-gs}, who point out related measures in Faith's earlier work.
 

### 2014-04-14 sbreview.tex
Trying to explain what OTU clustering wants to do.

@@
 
+In phylogenetics we would like to compare reconstructions to an object that is generally not knowable---the ``true'' historical phylogenetic tree---OTU inference has a variety of desirable outcomes, only some of which are knowable.
+\cite{Wang2013-ol} divide performance measures into \textit{external} measures, which compare an inferred clustering to some defined outcome, and \textit{internal} measures, which give overall descriptive statistics on inferred clustering.
+External measures applied to observed or simulated data include obtaining the same number of OTUs as taxonomic groups at some level \citep[e.g.][]{edgar2013uparse}, scoring deviation from a decomposition of a phylogenetic tree \citep[e.g.][]{navlakha2009finding}, or a given set of taxonomic classifications \citep[e.g.][]{Cai2011-bi}.
+Internal measures evaluate the clustering in various ways without reference to a ``true'' clustering, with the general idea that within-cluster distances should be small compared to between-cluster distances.
+
 There are many OTU inference methods with various speeds and strategies.
-Of the fixed-cutoff methods, the traditional choice has been CD-HIT \citep{li2006cdhit}, which seems to have been supplanted by clustering features of USEARCH \citep{edgar2010usearch} and perhaps now its descendant UPARSE \citep{edgar2013uparse}.
-These methods are very fast and are heuristic in that they are described as an algorithm rather than as global optimization of some notion of goodness of clustering.
-\citet{white2010alignment} show that different ways of doing this heuristic clustering can result in very different results.
-\citet{navlakha2009finding} and \citet{WYS13} develop methods that try to come up with sequence groupings that more closely reflect the type of species divisions found in taxonomies,
-\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values,
-and \cite{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.
+Some methods proceed through a list of sequences and progressively add each either to an existing cluster or start a new cluster with that sequence, such as CD-HIT \citep{li2006cdhit} and USEARCH \citep{edgar2010usearch}, while UPARSE \citep{edgar2013uparse} uses a similar strategy while also attempting to correct for sequencing error.
+\citet{white2010alignment} show that different ways of doing this genre of heuristic clustering can result in very different results.
+\citet{Cai2011-bi} perform highly efficient clustering using a pseudometric based partition tree, which can be thought of as a hierarchical clustering tree with a fixed set of internal node heights.
+\citet{navlakha2009finding} take a semi-supervised approach in that input includes sequences along with a subset of sequences that are equipped with taxonomic classifications; the algorithm then groups all sequences (many in novel clusters) into clusters that have similar properties as the example taxonomic clusters.
+\citet{Wang2013-ol} optimize a criterion of cluster modularity.
+\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \cite{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.
 
-The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
+The centrality of the OTU concept can be seen by the fact that the by-sample table of OTU observations (i.e.\ the matrix of counts with rows representing samples and columns representing OTUs) is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
 A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as such errors are especially problematic there.

### 2014-04-14 reviews/response.txt
Trying to explain what OTU clustering wants to do.
Done with Mooers review!

@@
 
+I'm afraid I can't find to what you refer.
+
 > 5-8 delete and
@@
 
+The editors have stated that they are not concerned with this becoming
+too long, and I've been surprised that what seems like an absurdly big
+amount of media hype to me doesn't reach everyone. So I'm leaving it in
+unless the editors say otherwise.
+
 > 8-12. these data (I guess we will eventually decide that data is
 > singular - up to the editors)
+
+I have fixed all these.
 
@@
 
+Er, perhaps this is a generational thing? I had no idea how they used to
+infer molecular sequences before Pace's phyloseminar.
+
 > 10-30 Which "databases" ? - this has been written so far as if it were
@@
 > of observation is here.
+
+I have clarified with a slightly expanded explanation, thanks.
 
@@
 
+Thanks for reporting this. I'm now more specific.
+
 > 11-53. Here is the first real point of contact between phylogenetics
@@
 
+I've added a mini-definition.
+
 > 12-28 and 12-32. What is the measure of performance that this refers
@@
 
+Clarified, thanks.
+
 > 12-29. Do readers of SB now what "k-mer composition" is? Is it k-means
 > clustering?
+
+I've added in a definition.
 
@@
 > but this could be clarified.
+
+I've added in a clarification, thanks.
 
@@
 > helpful - and where a shorter introduction might offer space.
+
+These are excellent questions, and the landscape is a little complex. I
+have expanded this section to explain what it is that these methods
+would like to achieve.
 

### 2014-04-14 sbreview.tex
Done with Mooers review!

@@
 That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
-This is proven to be a useful strategy \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}, however, because of the variability of gene repertoire in microbes, this core gene set may be relatively small.
-Indeed, even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
-At least some portion of the rest of the data can be taxonomically classified \citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.
+This is proven to be a useful strategy, and several groups have built databases of core gene families and provide programs and/or web tools to subset a metagenome to that core set of genes \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic,Darling2014-ux}.
+However, because of the variability of gene repertoire in microbes, this core gene set may be relatively small: even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
+At least some portion of the rest of the other approximately 99 percent of the metagenome can be taxonomically classified using methods reviewed by \citep{mande2012classification}.
+\citet{treangen2013metamos} report shorter run times and much higher accuracy (for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.
 
@@
 The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
-Tax2tree uses a heuristic algorithm to optimize the $F$-measure of precision and recall for its taxonomic assignments.
+Tax2tree uses a heuristic algorithm to optimize the $F$-measure (the harmonic mean of precision and recall) for its taxonomic assignments.
 Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
@@
 An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
-However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on $k$-mer composition \citep{wang2007naive,rosen2008metagenome}.
+However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on counts of DNA substrings of a given length, which are often called $k$-mer classifiers \citep{wang2007naive,rosen2008metagenome}.
+For those studies and others, the conventional metrics of classification performance such as precision and recall are applied to data simulated from known and taxonomically classified genomes.
 \citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
@@
 Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
-The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits.
+The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the phylogenetic tree.
 \citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.

### 2014-04-15 sbreview.tex
Mostly done with Alexis' comments.

@@
 A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as such errors are especially problematic there.
-With the exception of \citet{sharpton2011phylotu}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
+\citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a large number of sequencing reads and so can be used as an OTU clustering method.
+With the exception of the work of \citet{sharpton2011phylotu,Zhang2013-wh}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
 

### 2014-04-16 sbreview.tex
Making some progress through Olivier's comments.

@@
 \citet{wu2009phylogeny} have derived a ``phylogeny-driven genomic encyclopaedia of Bacteria and Archaea'' by selecting organisms for sequencing that are divergent from sequenced organisms.
-By doing so they have recovered more novel protein families than they would have using methods organized by taxonomy.
+They have recovered more novel protein families using these phylogeny-based approaches than they would have using methods organized by selecting microbes to sequence based on their taxonomic labels.
 In a similar effort for the human microbiome \citep{fodor2012most}, phylogenetic results were not shown although the authors state that phylogenetic methods did give similar results to their analysis.
@@
 If the set of base observations is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
-This would be a false positive.
+This observed significant difference would thus be a false positive.
 Even basic definitions pose a challenge here: the question of whether two communities are the ``same'' and ``different'' probably needs to be approached from the perspective of ecosystem modeling.
@@
 One type of normalization handles differential depths of sequencing across samples.
-The presently used approach is \textit{rarefaction}, which means uniform sub-sampling, typically without replacement to the number of reads in the lowest abundance sample \citep{caporaso2010qiime,schloss2009introducing}.
+The presently used approach is \textit{rarefaction}, which means uniform sub-sampling to the number of reads in the lowest abundance sample \citep{caporaso2010qiime,schloss2009introducing}.
 In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another, which is not the case.

### 2014-04-16 reviews/response.txt
Mostly done with Alexis' comments.
Making some progress through Olivier's comments.
Responded to Olivier's comments.

@@
 
-I have changed this to "How might one formulate a useful notion of independent contrasts (Felsenstein 1985) on such an object?" I hope that's better.
+I have changed this to "How might one formulate a useful notion of
+independent contrasts (Felsenstein 1985) on such an object?" I hope
+that's better.
 
@@
 > what do you have in mind here?
+
+I made this more specific.
+
 > Indeed "distinguish between two rather different samples" is a modest
-> goal Actually, it's not clear in this subsection what is meant by
+> goal
+
+Indeed, but it's not immediately obvious what level of sequencing depth
+is needed even in that situation.
+
+> Actually, it's not clear in this subsection what is meant by
 > "Microbial community estimation" (title of the subsection). What is
@@
 > abundance? Etc.? A few lines and explanations should suffice.
+
+I have revised the heading as well as separated out a paragraph
+describing terminology and the goal.
 
@@
 
+A great suggestion.
+I have added material at the beginning of the "Tree-thinking in human microbiome research" section.
+I don't think that examples are too productive here because people haven't gotten to that level of detail yet.
+
 > P13 	"There are many OTU inference methods with various speeds and
@@
 
+Given the diversity of OTU methodologies, it's not easy to generalize
+about their operation. However, I have added a paragraph explaining what
+it is that OTU inference methods are trying to do. I hope that helps.
+
 > P14 	You describe PD and suggests that it should be used more often.
@@
 > limitations?
+
+Thank you for this excellent suggestion.
 
@@
 
+You are right. I hope my edits make this more clear.
+
 > P15 	"Evans and Matsen (2012) showed that weighted UniFrac" check
@@
 
+I've shortened and simplified this sentence.
+
 > P16 	explanations and Fig. 3 on earth-movers distance are unclear to
 > me.
+
+I have expanded the (previously insufficient) descriptions.
 
@@
 
+I've made that more specific now.
+
 > P23-24 Paragraph "In the same vein This would be a false positive." is
 > unclear to me.
+
+I've made the false positive part more specific.
 
@@
 
+Simplified, thanks.
+
 > P25 Check "Comparative studies will continue to investigate what
 > shapes and is shaped by the microbiome"
+
+I like this sentence.
 
@@
 > Bioinformatics (2013), 29 (22): 2869-2876.
+> It would fit will into section "The role of OTUs".
 
-> It would fit will into section "The role of OTUs".
+It's cited now.
 
@@
 
+Right, thanks.
+
 > page 15 l 50: in "distance, is special" seems that it should read
@@
 > http://sco.h-its.org/exelixis/web/software/put/index.html
+
+The paper has been published now and I now cite it. Thanks!
 

### 2014-04-16 sbreview.tex
Responded to Olivier's comments.

@@
 
-\subsection{Microbial community estimation using marker gene surveys}
+\subsection{Inferring microbial community composition using marker gene surveys}
 Our modern knowledge of the microbial world is in a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages \citep{fox1977comparative}.
@@
 Regions of low diversity in 16S also enabled the development of the first ``universal'' 16S PCR primers \citep{lane1985rapid} which enabled detection of almost all bacteria and archaea regardless of whether they can be cultured.
-In microbial ecology, the census of bacteria in a given environment using marker gene amplification and sequencing are generally called ``marker gene surveys,'' which is equivalent to the ``barcoding'' terminology more commonly used for eukaryotes.
+
+In microbial ecology, the census of bacteria in a given environment using marker gene amplification and sequencing are generally called ``marker gene surveys.''
+This terminology is equivalent to the ``barcoding'' terminology more commonly used for eukaryotes.
+Such surveys would ideally return a census of all of the microbes in a sample along with their abundances.
 
@@
 This brings up the question of how many sequences are needed to characterize the microbial diversity of a given environment.
-To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however for more subtle information deeper sequencing is required.
+To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however to compare more similar samples deeper sequencing is required.
 In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \citep[e.g.][]{caporaso2011moving}.
@@
 
+Before proceeding, we note that phylogenetics for microbes differs in some important respects from, say, mammal phylogenetics.
+Where the phylogenetics researchers for mammals are converging on one or two possible basic structures for the evolution of mammals, much of the deep history of microbes remains obscure.
+Inference of this history is complicated by the fact that any tree-like signal in the overall evolution of bacteria is restricted to a small set of so-called ``core'' genes \citep{bapteste2009prokaryotic,Leigh2011-zu,Lang2013-zt}.
+
 \subsection{Phylogenetics and taxonomy}
@@
 
+It is not straightforward to define a clear notion of optimality for OTU inference.
 In phylogenetics we would like to compare reconstructions to an object that is generally not knowable---the ``true'' historical phylogenetic tree---OTU inference has a variety of desirable outcomes, only some of which are knowable.
@@
 For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
-%The most common diversity indices are Simpson and Shannon.
+For example, the most commonly used index is the Simpson index \cite{simpson1949measurement}, which is simply the sum of the squared frequencies across OTUs.
+The drawback of applying such a diversity metric to a collection of OTUs is that the overall structure of the diversity is lost, in that two closely related OTUs contribute as much to the diversity measure as two distant ones.
 % \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
-However, phylogenetic diversity metrics are also used and here we will focus on their applications and methods.
+Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used and here we will focus on their applications and methods.
 
@@
 Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
-\citet{mccoy2013abundance} have recently shown that partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
+Abundance-weighted measures commonly weight edges in proportion to abundance, but one can also construct ``partially abundance weighted'' measures by weighting edges by abundances transformed by a less than linear function.
+\citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
 \citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''
@@
 These dissimilarity measures have hundreds of citations.
-\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
-The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
+\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
+The earth-movers distance between two probability distributions in this case can be defined using a physical analogy as the minimum amount of ``work'' required to move piles of dirt (each dirt pile configuration sized in proportion to one probability distribution) in one configuration to another along the tree, where work is defined as mass times distance.
+In this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
 \citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
@@
 \newcommand{\dirtpilesLegend}{
-    Part of a minimal mass movement to calculate the earth-mover's distance on a phylogenetic tree.
+    Part of a minimal mass movement to calculate the earth-mover's distance between two probability distributions on a phylogenetic tree.
+    For this, each probability distribution is considered as a configuration of dirt piles on the tree, and the distance between two such dirt pile configurations is defined to be the minimum amount of physical ``work'' required to move the dirt in one configuration to the other.
 }

### 2014-04-16 reviews/response.txt
Just one comment left.

@@
 
+I have improved the description, with an overview sentence for those who
+want to skip the details.
+
 > 4) Pg. 14-15: I also don't really understand what UniFrac is or how it
@@
 
+I have much improved the caption for Figure 2. I think it should be
+clear now.
+
 > 5) The figures need a bit of work. There seems to be some sort of
@@
 > the colors mean in Figure 2? I don't find this to be very clear.
+
+I have made a connection between the figure and the caption via the
+mathematical notation used. For Figure 2, I have greatly expanded the
+caption to describe colors.
 
@@
 > observations" and sample observations here. Can this be clarified?
+
+I have rewritten this with a little notation, thanks.
 

### 2014-04-16 sbreview.tex
Just one comment left.

@@
 The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
-Tax2tree uses a heuristic algorithm to optimize the $F$-measure (the harmonic mean of precision and recall) for its taxonomic assignments.
+Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree as follows.
+The first step is to label the internal nodes of the tree to optimize the $F$-measure (the harmonic mean of precision and recall) across clades where the taxonomic label of a clade is thought of as a prediction for the descendant sequence taxonomic labels; the second step reassigns sequence taxonomic labels based on these clade taxonomic labels.
 Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
@@
     Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as blue circles and abundances are shown as the size of the circles.
-    Unweighted PD takes the sum of the branch lengths connecting samples.
-    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the areas of the edges.
+    Unweighted PD takes the sum of the branch lengths $\ell$ connecting samples.
+    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the areas $A$ of edges.
 }
@@
     The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
+    Assume that the sequence data to build the phylogenetic tree is divided into two samples, the green sample and the blue sample, which might represent samples from two locations.
     When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
+    Bottom pictorial equation shows the ratios of interest for UniFrac: the branch length unique to one sample (green or blue, but not brown) divided by the total branch length.
+    The ratio is smaller when the samples are interspersed (left) than they are when separate (right tree).
 }
@@
 Such non-independence can lead to incorrect rejection of the null hypothesis.
-Imagine, for example, that we have a random observation process on the tree equipped with some collection of ``base observations'' at the leaves.
-Each sample from the process takes a random subset of those base observations and then throws down some number of reads for each observation in that set, the number of which has mean significantly greater than 1.
-If the set of base observations is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
-This observed significant difference would thus be a false positive.
-Even basic definitions pose a challenge here: the question of whether two communities are the ``same'' and ``different'' probably needs to be approached from the perspective of ecosystem modeling.
+Imagine, for example, that we have a random observation process on the tree equipped with some collection $B$ of observations at the leaves.
+Each sample from the process takes a random subset from $B$ and then throws down some number of reads for each observation in that set, the number of which has mean significantly greater than 1.
+If the size of $B$ is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
+In trying to remedy such false-positive identification of differences, it becomes clear that even basic definitions pose a challenge: the question of whether two communities are the ``same'' and ``different'' probably needs to be approached from the perspective of ecosystem modeling.
 

### 2014-04-16 reviews/response.txt
Responded to all reviews!

@@
 
-A great suggestion.
-I have added material at the beginning of the "Tree-thinking in human microbiome research" section.
-I don't think that examples are too productive here because people haven't gotten to that level of detail yet.
+A great suggestion. I have added material at the beginning of the
+"Tree-thinking in human microbiome research" section. I don't think that
+examples are too productive here because people haven't gotten to that
+level of detail yet.
 
@@
 
-I have done so, so that the sentence reads "... 2013 MCEB Mathematical and Computational Evolutionary Biology workshop ..." is that right?
+I have done so, so that the sentence reads "... 2013 MCEB Mathematical
+and Computational Evolutionary Biology workshop ..." is that right?
 
@@
 
+Given that taxonomic classification is not always a phylogenetic
+problem, I don't think that it merits its own section. I have cited your
+suggestion and other recent progress in this field for readers who want
+to dig deeper.
+
 > I would also appreciate if you could explicitly mention the Gartner
@@
 > we are on this curve at present.
+
+I don't see how this would contribute to the scientific goals of the
+paper.
 
@@
 
-I don't follow your "with the fraction of edges connecting samples that are common to two samples" correction.
-There is some overlap with the PD literature as now described and cited.
+I don't follow your "with the fraction of edges connecting samples that
+are common to two samples" correction. There is some overlap with the PD
+literature as now described and cited.
 
@@
 
-I do describe strategies for read count normalization in the next paragraph, but have added the "rarefaction curve" terminology.
+I do describe strategies for read count normalization in the next
+paragraph, but have added the "rarefaction curve" terminology.
 
@@
 
-Good point. I've put in the two 2013 MBE papers. If there is something else I should cite let me know.
+Good point. I've put in the two 2013 MBE papers. If there is something
+else I should cite let me know.
 
@@
 > Typos:
-> P3, L40: I would rather use "species delimitation" instead of "species definition".
+> P3, L40: I would rather use "species delimitation" instead of "species
+> definition".
 

### 2014-04-16 sbreview.tex
Responded to all reviews!

@@
 However, because of the variability of gene repertoire in microbes, this core gene set may be relatively small: even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
-At least some portion of the rest of the other approximately 99 percent of the metagenome can be taxonomically classified using methods reviewed by \citep{mande2012classification}.
-\citet{treangen2013metamos} report shorter run times and much higher accuracy (for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.
+At least some portion of the rest of the other approximately 99 percent of the metagenome can be taxonomically classified using a diversity of methods described below.
 
@@
 They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.
+Other metagenomic classification techniques have been reviewed by \citet{mande2012classification}; interesting recent progress has been made by \citet{Lanzen2012-ul}, \cite{Koslicki2013-mc}, and \cite{Droge2014-lj}.
+\citet{treangen2013metamos} report shorter run times and much higher accuracy (for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.
 

### 2014-04-17 sbreview.tex
Moving from plainnat_abbrv to sysbio.
Loads of bib fixes.

@@
 \citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
-There are clear connections between UniFrac and phylogenetic diversity explored by \cite{Faith2009-gs}, who point out related measures in Faith's earlier work.
+There are clear connections between UniFrac and phylogenetic diversity explored by \cite{faith2009cladistic}, who point out related measures in Faith's earlier work.
 
@@
 \citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
-\citet{mirarabsepp} use data set partitioning to improve alignments on subsets of taxa specifically for this application.
+\citet{Mirarab2012-pk} use data set partitioning to improve alignments on subsets of taxa specifically for this application.
+\citet{Brown2013-ia} use locality-sensitive hashing to obtain placement more than two orders of magnitude faster than the \textit{pplacer} program of \cite{matsen2010pplacer}.
 
@@
 \noindent
-Figure 1: \pdLegend
+\begin{figure}[ht]
+  \caption{\pdLegend}
+  \label{fig:pd}
+\end{figure}
 
 \noindent
-Figure 2: \unifracLegend
+\begin{figure}[ht]
+  \caption{\unifracLegend}
+  \label{fig:unifrac}
+\end{figure}
 
 \noindent
-Figure 3: \dirtpilesLegend
+\begin{figure}[ht]
+  \caption{\dirtpilesLegend}
+  \label{fig:dirtpiles}
+\end{figure}
 
@@
 
-\bibliographystyle{plainnat_abbrv}
+\bibliographystyle{sysbio}
 \bibliography{sbreview}

### 2014-04-17 sbreview.tex
Capitulating to the microbiota-police.

@@
 
-The human microbiome is the collection of microbes that live inside and on the surface of humans.
+The human microbiome is the ensemble of genes in the microbes that live inside and on the surface of humans.
 Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information from microbiome samples.
@@
 \vspace{4.5in}
-\noindent (Keywords: human microbiome; microbial ecology; phylogenetic methods; 16S; metagenome)\\
+\noindent (Keywords: human microbiome; human microbiota; microbial ecology; phylogenetic methods; 16S; metagenome)\\
 \newpage
@@
 
-The parameter regime and focus of microbiome research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what microbial ecologists and medical researchers have done?
+The parameter regime and focus of human-associated microbial research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what microbial ecologists and medical researchers have done?
 The answer is simple: this system is data- and question-rich.
@@
 
-The human microbiome is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
+The human-associated microbial assemblage is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
 Additionally, due to more than a century of hospital laboratory work, our knowledge about human-associated microbes is relatively rich.
-It is also common to manipulate the human microbiome or animal models thereof via intervention studies and germ-free animals, which is easier to do than for many other microbial communities.
+This collection of microbes living inside and on top of us is called the \textit{human microbiota}, and the ensemble of genes in those microbes is called the \textit{human microbiome}, although usage of these terms varies \citep{Boon2013-mc}.
 
@@
 There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
-I will first briefly review the recent literature on the human microbiome, then describe novel ways in which human microbiome researchers have used trees.
+I will first briefly review the recent literature on the human microbiota, then describe novel ways in which human microbiome researchers have used trees.
 I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
@@
 
-\section{The human microbiome}
-The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
+\section{The human microbiota}
+The human microbiota is the collection of microbial organisms that live inside of and on the surface of humans.
 These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells of that individual.
-The microbiome has remarkable metabolic potential, with an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
-Much of our metabolic interaction with the outside world is mediated by our microbiome, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
-Our microbiomes have been historically transmitted vertically between mother and infant by seeding in the birth canal and breastfeeding;  immune dysfunction has been associated with the interruption of these practices \citep[reviewed in][]{funkhouser2013mom}.
-In this section I will briefly review what is known about the human microbiome and its effect on our health.
+The microbiota have remarkable metabolic potential, being an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
+Much of our metabolic interaction with the outside world is mediated by our microbiota, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
+Traditionally, our microbiota have been transmitted from mother to infant in the birth canal and by breastfeeding; immune dysfunction has been associated with the interruption of these practices \citep[reviewed in][]{funkhouser2013mom}.
+In this section I will briefly review what is known about the human microbiota and its effect on our health.
 
-The human microbiome is an ecosystem.
+The human microbiota form an ecosystem.
 It is dynamic in terms of taxonomic representation but apparently constant in terms of function \citep{hmp2012structure}.
-There is a ``core'' microbiome which is shared between all humans \citep{turnbaugh2008core}.
-The human microbiome is spatially organized, as can be seen on skin \citep{grice2009topographical}, with substantial variation in human body habitats across space and time \citep{costello2009bacterial}.
+There is a ``core'' microbiota which is shared between all humans \citep{turnbaugh2008core}.
+The human microbiota is spatially organized, as can be seen on skin \citep{grice2009topographical}, with substantial variation in human body habitats across space and time \citep{costello2009bacterial}.
 There is a substantial range of inter-individual versus intra-individual variation \citep{hmp2012structure}.
 
-Our actions can shift the composition of our microbiome.
-Changes in diet can very quickly shift its composition, but there is also a strong correlation between long-term diet and microbiome \citep{li2009human,wu2011linking}.
+Our actions can shift the composition of our microbiota.
+Changes in diet can very quickly shift its composition, but there is also a strong correlation between long-term diet and microbiota \citep{li2009human,wu2011linking}.
 Antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts for years \citep{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}.
 
-The microbiome interacts on many levels with host phenotype \citep[reviewed in][]{cho2012human}.
-The gut microbiome in particular correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to children with acute metabolic dysfunction in rural Africa \citep{smith2013gut}.
-Considerable attention has also been given to the interaction between the gut microbiome and obesity, although the story is not yet clear.
+The microbiota interact on many levels with host phenotype \citep[reviewed in][]{cho2012human}.
+The gut microbiota in particular correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to children with acute metabolic dysfunction in rural Africa \citep{smith2013gut}.
+Considerable attention has also been given to the interaction between gut microbiota and obesity, although the story is not yet clear.
 An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
-A causal role for the microbiome leading to obesity is established for mice: an obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}, the pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
+A causal role for the microbiota leading to obesity has been established for mice: an obese phenotype can be transferred from mouse to mouse by gut microbial transplantation \citep{turnbaugh2006obesity}, the pregnant human gut microbiota leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
 However, these promising leads have not yet been confirmed causally or in population studies of humans \citep{zhao2013gut}.
@@
 
-Bacteria have been the primary focus of human microbiome research, and other domains have been investigated to a lesser extent.
-Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea} and have a nonuniform distribution across the human skin \cite{Findley2013-rm}.
+Bacteria have been the primary focus of human microbiota research, and other domains have been investigated to a lesser extent.
+Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea} and have a nonuniform distribution across the human skin \citep{Findley2013-rm}.
 Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
@@
 
-In this paper we will primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perspective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
-Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.
+In this paper we will primarily be describing the human microbiota from a community-level phylogenetic perspective rather than from the fine-scale perspective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
+Our understanding of the true effect of the microbiota will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.
 
@@
 Abundance-weighted measures commonly weight edges in proportion to abundance, but one can also construct ``partially abundance weighted'' measures by weighting edges by abundances transformed by a less than linear function.
-\citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
+\citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiota; in particular that they do a better job than the commonly-used OTU-based measures.
 \citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''
@@
 Nevertheless, the work that has been done concerning horizontal gene transfer in the human microbiome has revealed interesting results.
-\citet{hehemann2010transfer} found that a seaweed gene has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
+\citet{hehemann2010transfer} found that a seaweed gene has been transferred into a bacterium in the gut microbiota of Japanese such that individuals with this resulting microbiota are better able to digest the algae in their diet.
 Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
@@
 
-Diversity preservation is of interest for microbiome researchers like it is for eukaryotic organisms, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms \citep{hartmann2006maximizing,pardi2007resource}.
-Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer \citep[reviewed in][]{cho2012human} and has made passionate appeals to preserve microbiome diversity \citep{blaser2011antibiotic}.
-% Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
-Because a child's initial microbiome is transmitted from the mother \citep[reviewed in][]{funkhouser2013mom}, there is a somewhat equivalent notion of microbiome extinction when the chain is interrupted.
-Consequently, \citet{yatsunenko2012human} have explicitly contrasted microbiome development in urban, forest-dwelling, and rural populations, while \citet{tito2012insights} have endeavored to characterize the microbiome from ancient feces.
+Diversity preservation is of interest for microbiota researchers like it is for eukaryotic organisms, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms \citep{hartmann2006maximizing,pardi2007resource}.
+Martin Blaser in particular has argued that changes in our microbiota are leading to an increase in autoimmune disease and certain types of cancer \citep[reviewed in][]{cho2012human} and has made passionate appeals to preserve microbiota diversity \citep{blaser2011antibiotic}.
+Because a child's initial microbiota is transmitted from the mother \citep[reviewed in][]{funkhouser2013mom}, there is a somewhat equivalent notion of microbiota extinction when the chain is interrupted.
+Consequently, \citet{yatsunenko2012human} have explicitly contrasted microbiota development in urban, forest-dwelling, and rural populations, while \citet{tito2012insights} have endeavored to characterize the microbiota from ancient feces.
 How might phylogenetic methods be used in these preservation efforts?
 
-Perhaps in part because of the importance of mother-to-child transmission, there are indications of coevolution between microbiomes and their hosts.
+Perhaps in part because of the importance of mother-to-child transmission, there are indications of coevolution between microbiota and their hosts.
 \citet{ochman2010evolutionary} found identical tree topologies for primate and microbiome evolution.
-For the microbiome, they used maximum parsimony such that each column was a microbe and each such entry took discrete states according to how much of that microbe was present.
-Although parsimony gave an interesting answer here, the presence of such coevolution raises the question of what sort of forward-time models are appropriate for microbiome change?
+For the microbiota, they used maximum parsimony such that each column was a microbe and each such entry took discrete states according to how much of that microbe was present.
+Although parsimony gave an interesting answer here, the presence of such coevolution raises the question of what sort of forward-time models are appropriate for microbiota change?
 Would methods using these models do better than parsimony or commonly-applied phenetic methods applied to the distances described above?
@@
 \section{Discussion}
-What can we expect next at the intersection of phylogenetics and the human microbiome?
+What can we expect next at the intersection of phylogenetics and the human microbiota?
 At least for the next several years we can expect the research questions described above to continue to unfold.
@@
 The uBiome \footnote{\url{http://ubiome.com/}} and American Gut \footnote{\url{http://americangut.org/}} projects promise to bring gut microbiome sequencing to the average citizen for a low price.
-Comparative studies will continue to investigate what shapes and is shaped by the microbiome.
+Comparative studies will continue to investigate what shapes and is shaped by the microbiota.
 However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.
@@
 
-Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes the hype has outmatched the science.
+Human microbiota research has experienced a frenetic rate of expansion over the past decade, and sometimes the hype has outmatched the science.
 However, our microbes are here to stay and so is research on them.
-Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
-Phylogenetics has already contributed significantly to human microbiome research and will continue to do so.
+Thus we can look forward to the field of human microbiota analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
+Phylogenetics has already contributed significantly to research on the human microbiota and will continue to do so.
 

### 2014-04-22 sbreview.tex
Bit o fiddle.

@@
 Additionally, due to more than a century of hospital laboratory work, our knowledge about human-associated microbes is relatively rich.
-This collection of microbes living inside and on top of us is called the \textit{human microbiota}, and the ensemble of genes in those microbes is called the \textit{human microbiome}, although usage of these terms varies \citep{Boon2013-mc}.
+This collection of microbes living inside and on our surfaces is called the \textit{human microbiota}, and the ensemble of genetic information in those microbes is called the \textit{human microbiome}, although usage of these terms varies \citep{Boon2013-mc}.
 
@@
 Much of our metabolic interaction with the outside world is mediated by our microbiota, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
-Traditionally, our microbiota have been transmitted from mother to infant in the birth canal and by breastfeeding; immune dysfunction has been associated with the interruption of these practices \citep[reviewed in][]{funkhouser2013mom}.
+Traditionally, our microbiota have been transmitted from mother to infant in the birth canal and by breastfeeding; immune dysfunction has been associated with the disruption of these practices \citep[reviewed in][]{funkhouser2013mom}.
 In this section I will briefly review what is known about the human microbiota and its effect on our health.
@@
 However, because of the variability of gene repertoire in microbes, this core gene set may be relatively small: even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
-At least some portion of the rest of the other approximately 99 percent of the metagenome can be taxonomically classified using a diversity of methods described below.
+At least some portion of the rest of the other approximately 99 percent of the metagenome can be taxonomically classified using one of the methods described below.
 
@@
 
-Before proceeding, we note that phylogenetics for microbes differs in some important respects from, say, mammal phylogenetics.
-Where the phylogenetics researchers for mammals are converging on one or two possible basic structures for the evolution of mammals, much of the deep history of microbes remains obscure.
+Before proceeding, we note that phylogenetics for microbes differs in some important respects from, say, mammalian phylogenetics.
+Where phylogenetic research on mammals is converging on one or two possible basic structures for their evolution, much of the deep history of microbes remains obscure.
 Inference of this history is complicated by the fact that any tree-like signal in the overall evolution of bacteria is restricted to a small set of so-called ``core'' genes \citep{bapteste2009prokaryotic,Leigh2011-zu,Lang2013-zt}.
@@
 The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
-Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree as follows.
-The first step is to label the internal nodes of the tree to optimize the $F$-measure (the harmonic mean of precision and recall) across clades where the taxonomic label of a clade is thought of as a prediction for the descendant sequence taxonomic labels; the second step reassigns sequence taxonomic labels based on these clade taxonomic labels.
+Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree which is annotated as follows.
+The first step is to label the internal nodes of the tree to optimize the $F$-measure (the harmonic mean of precision and recall) across clades.
+where the taxonomic label of a clade is thought of as a prediction for the descendant sequence taxonomic labels.
+The second step reassigns sequence taxonomic labels based on these clade taxonomic labels.
 Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.

### 2014-04-22 sbreview.tex
Tweaks before printing out.

@@
 They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.
-Other metagenomic classification techniques have been reviewed by \citet{mande2012classification}; interesting recent progress has been made by \citet{Lanzen2012-ul}, \cite{Koslicki2013-mc}, and \cite{Droge2014-lj}.
+Other metagenomic classification techniques have been reviewed by \citet{mande2012classification}; interesting recent progress has been made by \citet{Lanzen2012-ul}, \citet{Koslicki2013-mc}, and \citet{Droge2014-lj}.
 \citet{treangen2013metamos} report shorter run times and much higher accuracy (for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.
@@
 In phylogenetics we would like to compare reconstructions to an object that is generally not knowable---the ``true'' historical phylogenetic tree---OTU inference has a variety of desirable outcomes, only some of which are knowable.
-\cite{Wang2013-ol} divide performance measures into \textit{external} measures, which compare an inferred clustering to some defined outcome, and \textit{internal} measures, which give overall descriptive statistics on inferred clustering.
+\citet{Wang2013-ol} divide performance measures into \textit{external} measures, which compare an inferred clustering to some defined outcome, and \textit{internal} measures, which give overall descriptive statistics on inferred clustering.
 External measures applied to observed or simulated data include obtaining the same number of OTUs as taxonomic groups at some level \citep[e.g.][]{edgar2013uparse}, scoring deviation from a decomposition of a phylogenetic tree \citep[e.g.][]{navlakha2009finding}, or a given set of taxonomic classifications \citep[e.g.][]{Cai2011-bi}.
@@
 \citet{Wang2013-ol} optimize a criterion of cluster modularity.
-\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \cite{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.
+\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.
 
@@
 \citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a large number of sequencing reads and so can be used as an OTU clustering method.
-With the exception of the work of \citet{sharpton2011phylotu,Zhang2013-wh}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
+With the exception of the work of \citet{sharpton2011phylotu} and \citet{Zhang2013-wh}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
 
@@
 For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
-For example, the most commonly used index is the Simpson index \cite{simpson1949measurement}, which is simply the sum of the squared frequencies across OTUs.
-The drawback of applying such a diversity metric to a collection of OTUs is that the overall structure of the diversity is lost, in that two closely related OTUs contribute as much to the diversity measure as two distant ones.
+For example, the most commonly used index is the Simpson index \citep{simpson1949measurement}, which is simply the sum of the squared frequencies across OTUs.
+The drawback of applying such a diversity metric to a collection of OTUs is that the large-scale structure of the diversity is lost, such that two closely related OTUs contribute as much to the diversity measure as two distant ones.
 % \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
-Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used and here we will focus on their applications and methods.
+Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used and here we describe their application.
 
@@
 It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
-Whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve PD: unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.
+Whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate, only a few involve PD: unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.
 
@@
 \citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
-There are clear connections between UniFrac and phylogenetic diversity explored by \cite{faith2009cladistic}, who point out related measures in Faith's earlier work.
+There are clear connections between UniFrac and phylogenetic diversity explored by \citet{faith2009cladistic}, who point out related measures in Faith's earlier work.
 
@@
 Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
-People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has incorporated the clearcut code \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut, FastTree, and RAxML \cite{Stamatakis2006-yz}.
+People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has incorporated the clearcut code \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut, FastTree, and RAxML \citep{Stamatakis2006-yz}.
 % AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
@@
 \citet{Mirarab2012-pk} use data set partitioning to improve alignments on subsets of taxa specifically for this application.
-\citet{Brown2013-ia} use locality-sensitive hashing to obtain placement more than two orders of magnitude faster than the \textit{pplacer} program of \cite{matsen2010pplacer}.
+\citet{Brown2013-ia} use locality-sensitive hashing to obtain placement more than two orders of magnitude faster than the \textit{pplacer} program of \citet{matsen2010pplacer}.
 
@@
 In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics by reducing a distribution on the tips of a tree down to a real number.
-Another challenge is to bring together macroevolutionary modeling with species abundance modeling, but some initial steps have been made in another setting \citep{lambert2013predicting}.
+Another challenge is to bring together macroevolutionary modeling with species abundance modeling, where some initial steps have been made in another setting \citep{lambert2013predicting}.
 
@@
 Imagine, for example, that we have a random observation process on the tree equipped with some collection $B$ of observations at the leaves.
-Each sample from the process takes a random subset from $B$ and then throws down some number of reads for each observation in that set, the number of which has mean significantly greater than 1.
+Each sample from the process takes a random subset from $B$ and then throws down some number of reads for each observation in that subset, with the number of reads having a mean significantly greater than 1.
 If the size of $B$ is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
@@
 
-For both alpha and beta diversity studies, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
+For both alpha and beta diversity measurement, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
 One type of normalization handles differential depths of sequencing across samples.
@@
 Future research projects will continue to bring deeper sequencing on more samples.
-The uBiome \footnote{\url{http://ubiome.com/}} and American Gut \footnote{\url{http://americangut.org/}} projects promise to bring gut microbiome sequencing to the average citizen for a low price.
+The uBiome\footnote{\url{http://ubiome.com/}} and American Gut\footnote{\url{http://americangut.org/}} projects promise to bring gut microbiome sequencing to the average citizen for a low price.
 Comparative studies will continue to investigate what shapes and is shaped by the microbiota.
@@
 
-Human microbiota research has experienced a frenetic rate of expansion over the past decade, and sometimes the hype has outmatched the science.
+Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes the hype has outmatched the science.
 However, our microbes are here to stay and so is research on them.
-Thus we can look forward to the field of human microbiota analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
-Phylogenetics has already contributed significantly to research on the human microbiota and will continue to do so.
+Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
+Phylogenetics has already contributed significantly to research on the human microbiome and will continue to do so.
 

### 2014-04-23 sbreview.tex
Fixen sie formatten for arxiv.
Edits from read-through.

@@
 The human microbiome is the ensemble of genes in the microbes that live inside and on the surface of humans.
-Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information from microbiome samples.
+Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing and genetic analysis of microbiome samples.
 Much of the analytical work for these sequences involves phylogenetics, at least indirectly, but methodology has developed in a somewhat different direction than for other applications of phylogenetics.
@@
 The answer is simple: this system is data- and question-rich.
-It absolutely requires molecular methods, because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
+It absolutely requires molecular methods because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
 Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.
@@
 I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
-I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics, indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes an order of magnitude larger than typical for eukaryotic phylogenetics.
+I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics; indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes an order of magnitude larger than typical for eukaryotic phylogenetics.
 
@@
 The microbiota have remarkable metabolic potential, being an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
-Much of our metabolic interaction with the outside world is mediated by our microbiota, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
-Traditionally, our microbiota have been transmitted from mother to infant in the birth canal and by breastfeeding; immune dysfunction has been associated with the disruption of these practices \citep[reviewed in][]{funkhouser2013mom}.
+Much of our metabolic interaction with the outside world is mediated by our microbiota, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}.
+Our food and drug intake, in turn, impacts the diversity of microbes present.
+Traditionally, our microbiota have been transmitted from mother to infant in the birth canal and by breastfeeding \citep[reviewed in][]{funkhouser2013mom}.
 In this section I will briefly review what is known about the human microbiota and its effect on our health.
@@
 In microbial ecology, the census of bacteria in a given environment using marker gene amplification and sequencing are generally called ``marker gene surveys.''
-This terminology is equivalent to the ``barcoding'' terminology more commonly used for eukaryotes.
+This terminology is equivalent to the ``barcoding'' terminology more commonly used for eukaryotic surveys using 18S or the fungal internal transcribed spacer (ITS).
 Such surveys would ideally return a census of all of the microbes in a sample along with their abundances.
@@
 Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
-Recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
+This can distort inferences concerning actual microbe abundances based on read copy number.
+Recent work by \citet{kembel2012incorporating} implements the independent contrasts method \citep{felsenstein1985phylogenies} to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
 Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
@@
 \subsection{Metagenomes}
-As described above ``metagenome" means that DNA is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
-Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, however extraction efficiency concerns remain and multiplex sequencing is known to have biases of its own.
+As described above ``metagenome" means that DNA is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to the organism from which it came.
+Because metagenomes do not proceed through an amplification step, they do not have the same PCR primer biases as a marker gene survey, however extraction efficiency concerns remain and multiplex sequencing is known to have biases of its own.
 
-It is possible to use metagenomic data as an expanded set of marker genes.
+It is possible to subset metagenomic data to marker genes.
 That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
-This is proven to be a useful strategy, and several groups have built databases of core gene families and provide programs and/or web tools to subset a metagenome to that core set of genes \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic,Darling2014-ux}.
+This is proven to be a useful strategy, and several groups have built databases of core gene families as well as provided programs and/or web tools to phylogenetically analyze metagenomes subset to those core genes \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic,Darling2014-ux}.
 However, because of the variability of gene repertoire in microbes, this core gene set may be relatively small: even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
@@
 Metagenomic data sets are often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
-Discussing these methods is beyond the scope of this paper, as is ``metatranscriptomics": the sequencing of mRNA in bulk.
+Discussing these methods is beyond the scope of this paper, as is the sequencing of mRNA in bulk which is called ``metatranscriptomics".
 
 \subsection{Whole genomes}
-Whole-genome sequencing from culture is currently being used for microbial outbreak tracking \citep{koser2012rapid,snitkin2012tracking}.
+In addition to conventional genome sequencing for microbial genomes, whole-genome sequencing from culture is currently being used for microbial outbreak tracking \citep{koser2012rapid,snitkin2012tracking}.
 The Food and Drug Administration maintains GenomeTrakr, an openly accessible database of whole genomes sampled from the environment and grown in culture\footnote{\url{http://www.fda.gov/Food/FoodScienceResearch/WholeGenomeSequencingProgramWGS/}}.
-This data may become common for unculturable organisms as single-cell sequencing methods improve \citep[reviewed in][]{kalisky2011single}.
+This data may become more common for unculturable organisms as single-cell sequencing methods improve \citep[reviewed in][]{kalisky2011single}.
 The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \citep{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.
@@
 Phylogenetic inference has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
-The clearest such example is the discovery that archaea, although morphologically similar to bacteria, form their own separate lineage \citep{woese1977phylogenetic}.
+The clearest such example is the discovery that archaea, although similar to bacteria in their gross morphology, form their own separate lineage \citep{woese1977phylogenetic}.
 
@@
 % to preserve information associated with taxonomic classifications.
-Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy, although the actual revision process appears opaque \citep{kreig1984bergey}.
+Bergey's Manual of Systematic Bacteriology has officially adopted 16S as the basis for their taxonomy \citep{kreig1984bergey}, although the actual revision process appears opaque.
 The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
-Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree which is annotated as follows.
-The first step is to label the internal nodes of the tree to optimize the $F$-measure (the harmonic mean of precision and recall) across clades.
-where the taxonomic label of a clade is thought of as a prediction for the descendant sequence taxonomic labels.
+Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree, which is annotated as follows.
+The first step is to label the internal nodes of the tree to optimize the $F$-measure (the harmonic mean of precision and recall) across clades, where the taxonomic label of a clade is thought of as a prediction for the descendant sequence taxonomic labels.
 The second step reassigns sequence taxonomic labels based on these clade taxonomic labels.
-Interestingly, tax2tree allows for polyphyletic taxonomic groups as allowance for either phylogenetic error, lack of resolution of the 16S gene or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
+Interestingly, tax2tree allows for polyphyletic taxonomic groups as an allowance for either phylogenetic error, lack of resolution of the 16S gene, or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
 \citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
-Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI sequence database.
+Although it is wonderful that several groups are actively working on taxonomic revision, it can be frustrating to have multiple different taxonomies with no easy way to translate between them or to the taxonomic names provided in the NCBI or EMBL sequence databases.
 
@@
 For those studies and others, the conventional metrics of classification performance such as precision and recall are applied to data simulated from known and taxonomically classified genomes.
-\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
 % \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
 Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
-The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the phylogenetic tree.
+The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the taxonomic tree.
 \citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
+\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
 \citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
@@
 Other metagenomic classification techniques have been reviewed by \citet{mande2012classification}; interesting recent progress has been made by \citet{Lanzen2012-ul}, \citet{Koslicki2013-mc}, and \citet{Droge2014-lj}.
-\citet{treangen2013metamos} report shorter run times and much higher accuracy (for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.
+\citet{treangen2013metamos} report shorter run times and much higher accuracy (again, for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.
 
@@
 It is not straightforward to define a clear notion of optimality for OTU inference.
-In phylogenetics we would like to compare reconstructions to an object that is generally not knowable---the ``true'' historical phylogenetic tree---OTU inference has a variety of desirable outcomes, only some of which are knowable.
-\citet{Wang2013-ol} divide performance measures into \textit{external} measures, which compare an inferred clustering to some defined outcome, and \textit{internal} measures, which give overall descriptive statistics on inferred clustering.
+While in phylogenetics we would like to compare reconstructions to an object that is generally not knowable---the ``true'' historical phylogenetic tree---OTU inference has a variety of desirable outcomes, only some of which are knowable.
+\citet{Wang2013-ol} divide performance measures into \textit{external} measures, which compare an inferred clustering to some defined outcome, and \textit{internal} measures, which give overall descriptive statistics on an inferred clustering.
 External measures applied to observed or simulated data include obtaining the same number of OTUs as taxonomic groups at some level \citep[e.g.][]{edgar2013uparse}, scoring deviation from a decomposition of a phylogenetic tree \citep[e.g.][]{navlakha2009finding}, or a given set of taxonomic classifications \citep[e.g.][]{Cai2011-bi}.
@@
 \citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.
+\citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a relatively large number of sequencing reads and so can be used as an OTU clustering method.
 
@@
 A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as such errors are especially problematic there.
-\citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a large number of sequencing reads and so can be used as an OTU clustering method.
 With the exception of the work of \citet{sharpton2011phylotu} and \citet{Zhang2013-wh}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.
@@
 For the most part, this literature adapts methods from the classical ecological literature by substituting OTUs for taxonomic groups.
-For example, the most commonly used index is the Simpson index \citep{simpson1949measurement}, which is simply the sum of the squared frequencies across OTUs.
+For example, the most commonly used index is the Simpson index \citep{simpson1949measurement}, which is simply the sum of the squared frequencies of the OTUs.
 The drawback of applying such a diversity metric to a collection of OTUs is that the large-scale structure of the diversity is lost, such that two closely related OTUs contribute as much to the diversity measure as two distant ones.
 % \citep{hill1973diversity} Diversity and evenness: a unifying notation and its consequences
-Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used and here we describe their application.
-
-Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
-The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
-It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
-Whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate, only a few involve PD: unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.
+Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used.
+However, whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate, only a few involve PD.
 
@@
 
+Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
+The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
+It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
+Unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.
+
 Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are commonly used in human microbiome studies, abundance weighted phylogenetic diversity measures are not.
@@
 Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
-Abundance-weighted measures commonly weight edges in proportion to abundance, but one can also construct ``partially abundance weighted'' measures by weighting edges by abundances transformed by a less than linear function.
+Abundance-weighted measures commonly weight edges in proportion to abundance, but one can also construct ``partially abundance weighted'' measures by weighting edges by abundances transformed by a sublinear function.
 \citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiota; in particular that they do a better job than the commonly-used OTU-based measures.
@@
 
-The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
-Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
-These dissimilarity measures have hundreds of citations.
-\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
-The earth-movers distance between two probability distributions in this case can be defined using a physical analogy as the minimum amount of ``work'' required to move piles of dirt (each dirt pile configuration sized in proportion to one probability distribution) in one configuration to another along the tree, where work is defined as mass times distance.
-In this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
-\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
-There are clear connections between UniFrac and phylogenetic diversity explored by \citet{faith2009cladistic}, who point out related measures in Faith's earlier work.
-
 \newcommand{\unifracLegend}{
     The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
-    Assume that the sequence data to build the phylogenetic tree is divided into two samples, the green sample and the blue sample, which might represent samples from two locations.
-    When samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch lengths that can be \textit{uniquely} assigned to one sample or another compared to when they are separate (right tree).
-    Bottom pictorial equation shows the ratios of interest for UniFrac: the branch length unique to one sample (green or blue, but not brown) divided by the total branch length.
+    Assume that the sequence data to build the phylogenetic tree derives from two samples: the green sample and the blue sample.
+    When the samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch length that sits ancestral to clades that are \textit{uniquely} composed of one sample or another, compared to when they are separate (right tree).
+    The bottom pictorial equation shows the ratios of interest for UniFrac: the branch length unique to one sample (green or blue, but not brown) divided by the total branch length.
     The ratio is smaller when the samples are interspersed (left) than they are when separate (right tree).
@@
 \forarxiv{
-\begin{wrapfigure}[23]{r}{2.2in}
+\begin{wrapfigure}[32]{r}{2.2in}
   % \capstart
@@
 
+The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
+Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
+These dissimilarity measures have hundreds of citations.
+\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
+The earth-movers distance between two probability distributions in this case can be defined using a physical analogy as the minimum amount of ``work'', defined as mass times distance, required to move probability mass in one distribution to another along the tree.
+In this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
+\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
+There are clear connections between UniFrac and phylogenetic diversity explored by \citet{faith2009cladistic}, who point out related measures in Faith's earlier work.
+
 The most common way to use a distance matrix obtained from applying UniFrac to all pairs of samples is to apply an ordination method such as principal coordinates analysis.
@@
 \forarxiv{
-\begin{wrapfigure}[14]{l}{1.2in}
+\begin{wrapfigure}[24]{l}{1.5in}
   % \capstart
@@
 There have been several efforts to augment these ordination visualizations with additional information giving more structure to the visualizations.
-Biplots display variables (in the microbial case summarized by taxonomic labels) as points along with points representing samples \citep[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
+Biplots display variables (in the microbial case summarized by taxonomic labels) as points along with the points representing samples \citep[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
 \citet{PurdomAnalyzingDataGraphs08} describes how generalized principal component eigenvectors can be interpreted via weightings on the leaves of a phylogenetic tree.
@@
 16S distance is frequently used as a proxy for a functional comparison between human microbiome samples.
-Indeed, researchers using UniFrac frequently don't think of their comparisons as being in terms of a single gene, but rather in terms of an abstracted measure of community function.
+Indeed, researchers using UniFrac don't always think of their comparisons as being in terms of a single gene, but rather in terms of an abstracted measure of community function.
 Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes is commonly acquired horizontally as well as vertically, and horizontal transmission leaves no trace in 16S ancestry.
@@
 Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
-\citet{stecher2012gut} find that in a mouse model, horizontal transfer between pathogenic bacteria is blocked by commensal bacteria except for periods of gut inflammation.
+\citet{stecher2012gut} found that in a mouse model, horizontal transfer between pathogenic bacteria is blocked by commensal bacteria except for periods of gut inflammation.
 Horizontal transfer of genes is inferred in these studies by finding highly similar subsequences in otherwise less related organisms.
@@
 Historically this has meant relaxed neighbor joining \citep{evans2006relaxed}, but more recently FastTree 2 \citep{price2010fasttree} has emerged as the \textit{de facto} standard.
-People do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has incorporated the clearcut code \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut, FastTree, and RAxML \citep{Stamatakis2006-yz}.
+Researchers do most phylogenetic inferences as part of a pipeline such as mothur \citep{schloss2009introducing} which has incorporated the clearcut code \citep{sheneman2006clearcut}, or QIIME \citep{caporaso2010qiime}, which wraps clearcut, FastTree, and RAxML \citep{Stamatakis2006-yz}.
 % AXIOME streamlines and manages analysis of small subunit (SSU) rRNA marker data in QIIME and mothur \citep{lynch2013axiome}.
@@
 
-There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a so-called fixed \textit{reference tree}, sometimes with uncertainty estimates.
+There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a so-called fixed reference tree, sometimes with uncertainty estimates.
 These programs \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} have various speeds and features.
@@
 while the 1,262,986 sequence GreenGenes 13\_5 alignment is almost 80\% gap.
-The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
+The SILVA-associated `all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
 The GreenGenes tree is updated by running FastTree from scratch for every release.
@@
 In addition to these 16S-based resources, the MicrobesOnline resource \citep{dehal2010microbesonline} offers a very nice interactive tree-based genome browser.
-On a much smaller scale, there are microbiome body-site specific reference sets \citep{chen2010human,griffen2011core,srinivasan2012bacterial}
+On a much smaller scale, there are microbiome body-site specific reference sequence sets \citep{chen2010human,griffen2011core,srinivasan2012bacterial}
 
@@
 Many phylogenetic challenges remain in human microbiome research.
-Some of them are familiar, such as how to build phylogenies on the scale needed here on data that has many insertions and deletions.
+Some of them are familiar, such as how to build large phylogenies on data that has many insertions and deletions.
 I review some others here.
@@
 If we allow an arbitrary ``hidden'' object, what such object would perform best?
-An important first step in this direction has been taken by \citet{Parks2012-os} who define community comparison metrics on phylogenetic split systems.
+\citet{Parks2012-os} have expanded the range of choices by defining community comparison metrics on phylogenetic split systems.
 An alternative would be to use collections of reconciled gene trees in the presence of gene deletion, transfer, and loss \citep[e.g.][]{szollHosi2013efficient,szollHosi2013lateral}.
@@
 It appears that neutral models involving phylogenetics could be more fully developed.
-Methods explicitly invoking trait evolution are notably absent, with the recent exception of \citep{langille2013predictive}.
+Methods explicitly invoking trait evolution are notably absent, with the recent exception of the work by \citet{langille2013predictive}.
 The results of this simple method are reasonable, but would a collection of gene trees reconciled with a species tree allow for better prediction?
@@
 
-Developing community assembly models forms an important project for microbial ecology generally and human microbial ecology in particular, phylogeny-aware methods could be further developed.
-One way to model community assembly is to apply Hubbell's neutral theory \citep{fierer2012animalcules,costello2012application}.
+While developing community assembly models forms an important project for microbial ecology generally and human-associated microbial ecology in particular, phylogeny-aware methods could be further developed.
+One way to model microbial community assembly is to apply Hubbell's neutral theory \citep{fierer2012animalcules,costello2012application}.
 \citet{o2012phylogenetic} model community assembly with an explicitly phylogenetic perspective, and include some comparison of models to data.
@@
 In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics by reducing a distribution on the tips of a tree down to a real number.
-Another challenge is to bring together macroevolutionary modeling with species abundance modeling, where some initial steps have been made in another setting \citep{lambert2013predicting}.
+Another challenge is to bring together macroevolutionary modeling with species abundance modeling, where some initial work has been done by \citet{lambert2013predicting} in another setting.
 
@@
 Martin Blaser in particular has argued that changes in our microbiota are leading to an increase in autoimmune disease and certain types of cancer \citep[reviewed in][]{cho2012human} and has made passionate appeals to preserve microbiota diversity \citep{blaser2011antibiotic}.
-Because a child's initial microbiota is transmitted from the mother \citep[reviewed in][]{funkhouser2013mom}, there is a somewhat equivalent notion of microbiota extinction when the chain is interrupted.
-Consequently, \citet{yatsunenko2012human} have explicitly contrasted microbiota development in urban, forest-dwelling, and rural populations, while \citet{tito2012insights} have endeavored to characterize the microbiota from ancient feces.
+Because a child's initial microbiota is transmitted from the mother \citep[reviewed in][]{funkhouser2013mom}, there is a somewhat equivalent notion of microbiota extinction when the chain is interrupted via cesarean section and infant formula.
+In order to characterize extant diversity, \citet{yatsunenko2012human} have explicitly contrasted microbiota development in urban, forest-dwelling, and rural populations, while \citet{tito2012insights} have endeavored to characterize the microbiota from ancient feces.
 How might phylogenetic methods be used in these preservation efforts?
 
-Perhaps in part because of the importance of mother-to-child transmission, there are indications of coevolution between microbiota and their hosts.
+There are indications of coevolution between microbiota and their hosts.
 \citet{ochman2010evolutionary} found identical tree topologies for primate and microbiome evolution.
-For the microbiota, they used maximum parsimony such that each column was a microbe and each such entry took discrete states according to how much of that microbe was present.
+For the microbiota, they used maximum parsimony such that each column represented a microbe and each such entry took discrete states according to how much of that microbe was present.
 Although parsimony gave an interesting answer here, the presence of such coevolution raises the question of what sort of forward-time models are appropriate for microbiota change?
@@
 
-In a similar vein, assessing the significance level of an observed difference between communities is not straightforward and poses difficult problems.
-The randomization of group membership commonly used in combination with UniFrac to determine significance does not have appropriate properties when in the regime of incomplete sampling with non-independent observations, which is certainly the correct regime for surveys and metagenomes.
+In a similar vein, assessing the significance level of an observed difference between communities poses difficult problems.
+The randomization of group membership commonly used in combination with UniFrac to determine significance does not have appropriate properties in the regime of incomplete sampling with non-independent observations, which is certainly the correct regime for surveys and metagenomes.
 Such non-independence can lead to incorrect rejection of the null hypothesis.
-Imagine, for example, that we have a random observation process on the tree equipped with some collection $B$ of observations at the leaves.
-Each sample from the process takes a random subset from $B$ and then throws down some number of reads for each observation in that subset, with the number of reads having a mean significantly greater than 1.
-If the size of $B$ is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
+Imagine, for example, that we have a random process as follows.
+Each sample from the process takes a random subset of ``observations'' from the leaves and then throws down some number of reads for each observation in that subset, with the number of reads having a mean significantly greater than 1.
+If the number of leaves is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
 In trying to remedy such false-positive identification of differences, it becomes clear that even basic definitions pose a challenge: the question of whether two communities are the ``same'' and ``different'' probably needs to be approached from the perspective of ecosystem modeling.
@@
 The presently used approach is \textit{rarefaction}, which means uniform sub-sampling to the number of reads in the lowest abundance sample \citep{caporaso2010qiime,schloss2009introducing}.
-In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another, which is not the case.
+In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another.
+This is not the case.
 An alternative is provided by \citet{o2012phylogenetic}, who provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly.
@@
 
-There are limitations to what we can learn using genetics because more intricate processes such as gene regulation may be at play, limiting what sequence-level phylogenetics can do.
+There are limitations to what we can learn using genetic sequences because more intricate processes such as gene regulation may be at play, limiting what sequence-level phylogenetics can do.
 Future work may move from general ecological models to models that include specific interactions between microbes and the host \citep[reviewed in][]{hooper2012interactions}.

### 2014-06-03 reviews/response.txt
Getting back decision.

--- 2014-04-16 reviews/response.txt
@@
-> Reviews from Prof. Frank (Andy) Anderson
-> ----------------------------------------
+> 1) Pg.
+3: “It absolutely requires molecular methods because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization” – This suggests that molecular sequences have traditionally been characterized via morphological or phenotypic means.
+I suppose that’s indirectly true, but I don’t think it is the best way to describe it.
 
-> Thank you for your Systematic Biology submission. It has been reviewed
-> by Associate Editor Prof. Olivier Gascuel and three reviewers. Their
-> comments are listed at the end of this letter. The reviewers and the
-> AE provide some excellent constructive suggestions that I am sure you
-> will appreciate.
+> 2) Pg.
+10: “Inference of this history is complicated by the fact that any tree-like signal in the overall evolution of bacteria is restricted to a small set of so-called “core” genes” – Would it be more accurate to write that tree-like signal regarding the deep relationships of bacteria is restricted to a small set of so-called “core” genes?
+I suspect all genes carry some tree-like signal, even in microbes, though some have been subject to much more lateral gene transfer than others.
 
-> I have little to add that wasn't covered by one or more reviewers or
-> Olivier in their fine reviews. As an instructor who has almost given
-> up on trying to convince microbiology graduate students that they can
-> do better analyses than neighbor joining, I was already sensitive to
-> some of these differences, but this review helped me appreciate the
-> interface between the two disciplines. I agree with Mooers that the
-> level of detail varies across the manuscript, but I mainly noticed
-> when a topic was summarized quickly and I didn't really understand it
-> (I note that Mooers and I both had a question about how tax2tree
-> works).
+> 3) Pg. 14: “first with a Dirichlet multinomial mixtures on 3-mer profiles” – Should this just be “mixture”?
 
-> I have a few comments and questions that may help clarify some of the
-> writing. Altogether, I think the manuscript really requires only
-> several minor revisions.
+> 4) Pg.
+15: “However, whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate” – This requires some modification.
+Maybe add “one” after “every”.
 
-> Like Olivier, I note that some of my comments may overlap those of the
-> reviewers.
+> 5) Pg.
+16: “PD under random sub-sampling, which is often applied to enable comparison be- tween samples of different sequencing depths and called “rarefaction curves.”” – Would it be preferable to write that the random subsamples are described (or depicted) with rarefaction curves?
 
-> 1) Pg. 9: "As described above "metagenome" means that data is sheared
-> randomly across the genome rather than amplified from a specific
-> location," - I assume you mean DNA is sheared randomly?
+> 6) Pg.
+23: “The apparent success of 16S-tree-based comparisons begs the question of if a more complete representation of the evolutionary history…” – “Begging the question” means something else in philosophical circles.
+You should probably just use “raises the question” here.
 
-Yes, thanks.
+Your revision must comply with our instructions to authors (in this letter, with additional instructions at http://www.oxfordjournals.org/our_journals/sysbio/for_authors).
+This applies even if your original submission deviated from Systematic Biology style and corrections were not included in the notes from the editors or reviewers.
+Therefore, READ AND APPLY OUR INSTRUCTIONS BEFORE SUBMITTING A REVISION.
+Failure to do so will result in significantly delayed processing of your paper.
+If anything remains unclear after reading the instructions, use recent issues of the journal to find examples.
+If necessary, feel free to contact the Managing Editor at sysbio.editorialoffice@oup.com for help.
 
-> 2) Remember, "data" still is a plural noun, at least under my watch!
-> Check for "this data", "data is", "data was", etc
+After you've uploaded your revision, carefully view the Manuscript Central version to verify that all figures and other files display correctly, and that you’ve followed all of our author instructions.
 
-Searched and replaced. Thanks!
+The cost of printed color figures is $600 each.
+Authors are normally expected to cover this cost, but we do have limited funds available for authors who are SSB members, and cannot pay the full amount.
+If you feel your circumstances create an unusual financial need, please explain in your cover letter.
+The maximum allowance is one color figure per paper.
+All color costs can be avoided if you decide to have the figures printed in black and white, but be shown in color in the online version of the paper.
+If you do this, the captions must be worded such that they are appropriate for both situations (e.g., descriptions should not name colors), because the captions in print and online will be identical even if color is used online only.
+Do not submit two versions of any figure.
+Instead, make sure the color figure is also easily legible in grayscale, then submit only the color version.
+In your cover letter, please make your intentions regarding the use of color clear.
 
-> 3) Pg. 10: I don't really understand what "tax2tree" does, based on
-> the description here. Does that matter?
+Journal Covers: Please consider submitting a suggested cover image.
+They can be illustrations of theory, photos of organisms, a combination of the two, or alternatives.
+These can be uploaded with a revision of the manuscript or may be sent later.
+Only images for which you have copyright permission or that are not under copyright may be submitted.
+The color figure fee does not apply to images chosen by the Editor to be on the cover of the issue.
+If you have questions about possible cover designs, please email sysbio.editorialoffice@oup.com.
+We encourage and greatly appreciate cover image suggestions.
 
-> 4) Pg. 14-15: I also don't really understand what UniFrac is or how it
-> works, even after looking at Figure 2 (Olivier made a similar
-> comment). Could we get some help here?
+Make sure all section headings conform to Systematic Biology style for first, second and third levels.
+Use of incorrect styles is potentially confusing and, in any case, is likely to delay processing of the manuscript.
 
-> 5) The figures need a bit of work. There seems to be some sort of
-> equation at the bottom of Figure 1, but I can't decipher it. What do
-> the colors mean in Figure 2? I don't find this to be very clear.
+Tables should have a single-sentence informative title above the table, with any other descriptive information located below the table in the form of notes and/or specific footnotes.
 
-> 6) Pg. 20: "How to formulate a useful notion of independent contrasts
-> (Felsenstein 1985) on such an object?" - Awkward wording.
+When references are grouped together in parentheses in the text, they should be listed in ascending chronological order. Multiple references in a single year should be alphabetized.
 
-I have changed this to "How might one formulate a useful notion of
-independent contrasts (Felsenstein 1985) on such an object?" I hope
-that's better.
+All funding used for this work should be listed in a "Funding" section preceding the Acknowledgements. Please give the full official name of each funding body.
 
-> 7) Pg. 23: I don't understand the distinction between "base
-> observations" and sample observations here. Can this be clarified?
+AE
+--
 
-> 8) Pg. 24: "Future will continue to bring deeper sequencing on more
-> samples" - I guess this should be "The future will continue"?
+> Thanks for submitting the revised version of your review paper
+> “Phylogenetics and the human microbiome” to Systematic Biology
+> (special online issue, after MCEB2013). It has been read by the same
+> three referees as the original submission: Alexis Stamatakis, an
+> anonymous referee (2), and Frédéric Delsuc. All three recommend
+> acceptance with minor or no change. Actually, both of them (Alexis and
+> Fred) recommend to publish the paper as is, and referee 2 has only
+> minor comments on figure 2. I read the paper and reach the same
+> conclusion. My (minor) comments are provided below. I believe that you
+> have done a great job in putting all of these works and references
+> together, in a very comprehensive way, with highly motivating
+> perspectives for future research in phylogenetics. I’m thus happy to
+> recommend acceptance with quite minor changes (mostly in figures 2 and
+> 3).
 
-Thanks.
+> 9 (Kuczynski et al. 2010). However
 
-AE, Olivier Gascuel
--------------------
+> P12 “Several groups are continually revising taxonomy...” and P13 “An
+> obvious application of phylogenetics is to perform taxonomic
+> classification...” We have here two long and technical paragraphs
+> (e.g. F-measure, precision and recall, but what is classified and
+> how?). You should check this again for clarity and to ensure
+> self-contentedness.
 
-> My recommendation is acceptance pending minor to major changes.
-> Actually, no change asked by the reviewers and me (see below) requires
-> major efforts, but we all come from different origins, with different
-> points of view and expectations, and thus summing all these demands
-> will require a bit of time and attention. Basically, we all ask for
-> more details and comments here and there, to make the paper more
-> precise, comprehensive and self-contained. I am confident that you'll
-> find our suggestions useful. I don't believe that the paper should be
-> shortened in some places (or very locally), as suggested by Arne
-> Mooers. When submitting your revision, please prepare a cover letter
-> containing all the comments by the referees and me, with your
-> changes/responses.
+> P16 connecting taxa in the set (Faith 1992).
 
-> Specific comments, possibly redundant with those from the reviewers,
-> using Systematic Biology page numbering:
+> Fig 1 The "graphical formulae" [l(segments) A(segments)] at the bottom
+> of the figure do not help. This can be better described by
+> checking-improving the figure legend, notably "the abundance of the
+> taxa on either side of the branch" (how is it computed? give an
+> example if there are several options)
 
-> P4	these techniques AND could
+> Fig 2 Your two cases are extreme. Having one (or a few) shared (brown)
+> edges in the right tree, and one unshared (blue or green branch) edge
+> in the left tree would help. See also comments of referee 2.
 
-Thanks!
+> Fig 3 and its legend, as well as the corresponding text P17-18 are
+> quite unclear to me.... It would be interesting to have more
+> explanations, both on the math side and on the application to sequence
+> reads.
 
-> P8	(Kuczynski et al. 2010). However, for more subtle information
-> P8	"for more subtle information deeper sequencing is required" :
-> what do you have in mind here?
+Reviewer: 1
+-----------
 
-I made this more specific.
+> The author has addressed all of my minor suggestions for revision and
+> has also fully addressed the suggestions made by the other reviewers.
 
-> Indeed "distinguish between two rather different samples" is a modest
-> goal
+Reviewer: 2
+-----------
 
-Indeed, but it's not immediately obvious what level of sequencing depth
-is needed even in that situation.
+> As noted in my first review, I like the idea of this manuscript,and
+> much of the content.  The revised version is very close to the first
+> version, so I assume the author is happy with it.  I certainly
+> appreciated the diff.pdf.
 
-> Actually, it's not clear in this subsection what is meant by
-> "Microbial community estimation" (title of the subsection). What is
-> the goal here? Identify and classify microbes? Measure their
-> abundance? Etc.? A few lines and explanations should suffice.
+> I am still not clear how the author chooses what needs technical
+> discussion and what does not, and more technical detail has been added
+> to this version.  However other, perhaps more technically-inclined
+> reviewers, did not really pick this up, so, in the interests of
+> efficiency, I won't belabour the point. The author must have a clear
+> view who the intended audience is.
 
-I have revised the heading as well as separated out a paragraph
-describing terminology and the goal.
+> I will say that, while figure 2 does a fairly good job of illustrating
+> the critical UniFrac method (though the common brown internal edges
+> could be named explicitly in the caption), the definition in the text
+> does not: "The name UniFrac is a contraction of “Unique Fraction,”
+> which refers to the fact that the original UniFrac definition compares
+> the fraction of edges that connect only tips from one sample via the
+> shortest path rather than edges that connect between samples (Fig. 2;
+> Lozupone and Knight 2005)" makes very little sense to me, particularly
+> "edges between samples". Could this be reworded in a more intuitive
+> way that refers explicitly to the tree(s) being inferred and the tips
+> used to infer the trees?  So, for instance, does one infer one tree
+> per sample and then one tree of all samples?
 
-> P11	What is the level of uncertainty in current bacterial
-> phylogenies? For example, one commonly considers that only a few
-> unknowns remain in mammals phylogeny; what about bacteria? Are people
-> discussing the details or the broad lines? Provide a few illustrative
-> examples.
+Reviewer: 3
+-----------
 
-A great suggestion.
-I have added material at the beginning of the "Tree-thinking in human microbiome research" section.
-I don't think that examples are too productive here because people haven't gotten to that level of detail yet.
-
-> P13 	"There are many OTU inference methods with various speeds and
-> strategies" and the two following paragraphs. Describe in a simple way
-> the basic principles of these methods, before discussing their
-> relative merits (this remark may apply in other places).
-
-Given the diversity of OTU methodologies, it's not easy to generalize
-about their operation. However, I have added a paragraph explaining what
-it is that OTU inference methods are trying to do. I hope that helps.
-
-> P14 	You describe PD and suggests that it should be used more often.
-> But what are the standard alpha diversity measures, and their
-> limitations?
-
-Thank you for this excellent suggestion.
-
-> P15 	Are "partially abundance weighted diversity measures" using some
-> phylogenetic approaches? Be a bit more specific when describing and
-> discussing methods. Here and in many other places (e.g. UniFrac,
-> weighted PD, weighted UniFrac, etc.).
-
-You are right. I hope my edits make this more clear.
-
-> P15 	"Evans and Matsen (2012) showed that weighted UniFrac" check
-> sentence.
-
-I've shortened and simplified this sentence.
-
-> P16 	explanations and Fig. 3 on earth-movers distance are unclear to
-> me.
-
-I have expanded the (previously insufficient) descriptions.
-
-> P17 	"By doing so they have recovered more novel protein families
-> than they would have using methods organized by taxonomy". I'm lost
-> here, as it is said just above that they have derived a
-> "phylogeny-driven genomic encyclopedia of Bacteria and Archaea".
-
-I've made that more specific now.
-
-> P23-24 Paragraph "In the same vein This would be a false positive." is
-> unclear to me.
-
-I've made the false positive part more specific.
-
-> P24 Check "The presently used approach is rarefaction, which means
-> uniform sub-sampling, typically without replacement to the number of
-> reads in the lowest abundance sample"
-
-Simplified, thanks.
-
-> P25 Check "Comparative studies will continue to investigate what
-> shapes and is shaped by the microbiome"
-
-I like this sentence.
-
-> P26 "during the 2013 LIRMM Mathematical and Computational Evolutionary
-> Biology workshop", change LIRMM into MCEB
-
-I have done so, so that the sentence reads "... 2013 MCEB Mathematical and Computational Evolutionary Biology workshop ..." is that right?
-
-> Fig 2. It took time to me to understand and check on the Web that the
-> tree must be rooted, to compute the unifrac distance. This is one
-> example, among others, that more details are required when describing
-> methods, criteria, measures, etc. Some more figures could help, or one
-> appendix with formal definitions of the main concepts and tools. At
-> least, details and explanations have to be added in the main text.
-
-> Your reference list is extremely long, mostly due to the fact that you
-> are citing papers with a lot of authors. Check Syst Biol
-> recommendation to authors (I assume that our editorial office will
-> answer and help).
-
-> Reviewer 1, Alexis Stamatakis
-> -----------------------------
-
-> Comments:
-> This is a well written and in depth review on the usage of
-> phylogenetics in human microbiome research, which I enjoyed reading.
-
-> I only have a few minor remarks:
-
-> It might be a consideration to cite our recent paper:
-
-> J. Zhang, P. Kapli, P. Pavlidis, A. Stamatakis: "A General Species
-> Delimitation Method with Applications to Phylogenetic Placements". In
-> Bioinformatics (2013), 29 (22): 2869-2876.
-> It would fit will into section "The role of OTUs".
-
-It's cited now.
-
-> Apart from that, I believe that a more critical and slightly extended
-> discussion of taxonomic classifiers in the very last part of the paper
-> should be included. From my point of view, this is still a mostly
-> unresolved problem and the currently best classifier, which you may
-> also want to cite,  still shows pretty horrible accuracy
-> (http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0049334)
-> at the genus and family level.
-
-> I would also appreciate if you could explicitly mention the Gartner
-> hype curve in the discussion part and provide an estimate as to where
-> we are on this curve at present.
-
-> Some typos:
-
-> page 5 lines 7-8 "these techniques and could be used" there seems to
-> be a verb missing.
-
-Fixed, thanks.
-
-> page 14 lines 12 Here you may add our recent work on PTP, since it
-> strives to conduct OTU clustering and phylogenetic inference in one
-> single step/pipeline.
-
-Right, thanks.
-
-> page 15 l 50: in "distance, is special" seems that it should read
-> "distance, it is special"
-
-Thanks, I've rearranged here.
-
-> page 20: lines 51-56: In fact we are currently developing a tool for
-> automatic tree extension as described here. While not published yet,
-> it is already available as open-source code at:
-> http://sco.h-its.org/exelixis/web/software/put/index.html
-
-The paper has been published now and I now cite it. Thanks!
-
-> Reviewer 2, Arne Mooers
-> -----------------------
-
-> This was an interesting manuscript to read, and I compliment whoever
-> thought to review Human Microbiome Research (HMR) for the SB
-> readership. Overall, there is a lot of very interesting work reported
-> on here, though I found the level of detail to be a bit uneven:
-> sometimes we are told perhaps too much, and often, what feels like too
-> little, in order to make the connections between HMR and SB. There is
-> also seems to be some elision between HMR research and metagenomics
-> more generally. That isn't a bad thing, but one does sometimes forget
-> that we are focussed on HMR.
-
-> Numbers refer to the SB manuscript pagination, rather than to the
-> original latex.
-
-> 4-45 additional to
-
-I'm afraid I can't find to what you refer.
-
-> 5-8 delete and
-
-Fixed.
-
-> 7-23 delete though
-
-Nice catch!
-
-> The section "The Human Microbiome" is quite detailed (discussing mouse
-> to human transplants wrt obesity, etc.). Given the intense media
-> reporting of the work coming from this area, SB readers may not need
-> to be primed of the importance of this field. If this were shortened,
-> there could be room to expand some of the technical bits that SB
-> readers will want that follow. The last part of the section  - about
-> sticking to eubacteria and a non-molecular biology approach do seem
-> germane.
-
-The editors have stated that they are not concerned with this becoming
-too long, and I've been surprised that what seems like an absurdly big
-amount of media hype to me doesn't reach everyone. So I'm leaving it in
-unless the editors say otherwise.
-
-> 8-12. these data (I guess we will eventually decide that data is
-> singular - up to the editors)
-
-I have fixed all these.
-
-> 8-37++ Whereas Woese ++ This is a bit wordy and unnecessary - we know
-> that Woese didn't have PCR
-
-Er, perhaps this is a generational thing? I had no idea how they used to
-infer molecular sequences before Pace's phyloseminar.
-
-> 10-30 Which "databases" ? - this has been written so far as if it were
-> being applied to samples, not databases. I am not clear what the unit
-> of observation is here.
-
-I have clarified with a slightly expanded explanation, thanks.
-
-> 10-32 We quickly switch here from easy background (nonPCR random
-> sequencing) to something brand new (at least to me). "At least some
-> portion of the rest of the data can be taxonomically classified
-> (methods reviewed by Mande et al. 2012); Treangen et al. (2013) report
-> speedups and much higher accuracy when reads are assembled before they
-> are classified."
-> "Rest of data" refers to what - reads found in a database, or reads
-> from a sample? "Accuracy" refers to what? "Speed-ups" refer to what?
-
-Thanks for reporting this. I'm now more specific.
-
-> 11-53. Here is the first real point of contact between phylogenetics
-> and HMR. But what is an "F-measure of precision and recall" - there is
-> no citation.
-
-I've added a mini-definition.
-
-> 12-28 and 12-32. What is the measure of performance that this refers
-> to? How do we decide if a method performs well?
-
-Clarified, thanks.
-
-> 12-29. Do readers of SB now what "k-mer composition" is? Is it k-means
-> clustering?
-
-I've added in a definition.
-
-> 12-37++. Is the MEGAN approach compositional or phylogenetic? I
-> believe it is meant to be a hybrid approach (BLAST being
-> compositional, but the taxonomic labeling considered phylogenetic),
-> but this could be clarified.
-
-I've added in a clarification, thanks.
-
-> 13-29. It is not clear what heuristics are required for a cut-off
-> method. The author says that he is going to describe the OTU inference
-> industry, but the paragraph is really just a list of acronyms, and
-> then refers obliquely to a "contingency table" of OTU observations.
-> So, for example, what is a notion of goodness of clustering that the
-> heuristics are trying to achieve? What are the sorts of species
-> divisions found in taxonomies? This is an area where more would be
-> helpful - and where a shorter introduction might offer space.
-
-These are excellent questions, and the landscape is a little complex. I
-have expanded this section to explain what it is that these methods
-would like to achieve.
-
-> 14-40. Whereas just about every 16S survey investigation involves an
-> OTU-based alpha diversity estimate, only a few involve PD: unweighted
-> PD has been applied to some 16S survey data (Lozupone and Knight 2007;
-> Costello et al. 2009) and to metagenomic reads in a set of marker
-> genes (Kembel et al. 2011) - This is an odd pair of sentences: is the
-> L&K the only study that used unweighted PD? If so, in what context?
-
-I can't quite understand this suggestion. All three papers listed used
-unweighted PD.
-
-> 14-46. Check this sentence - perhaps start with "While are commonly
-> used, abundance weighted PD measures are not.
-
-I have updated this sentence based on your suggestion, thanks.
-
-> 15-30. Perhaps "In contrast to measures of alpha diversity, measures
-> of beta diversity do make use of phylogenetics"  This is actually an
-> interesting contrast. Is this just historical contingency, or is there
-> a logical reason why the two areas disagree on the use of phylogenetic
-> information?
-
-I think that this refers to 14-30, and I've simply removed the sentence.
-With regards to your question, I don't know. It seems surprising to me.
-
-> 15-xx. This seems a strong point of contact between metazoan
-> phylogenetics and the microbiome folks. UniFrac is clearly described
-> on its own terms, except perhaps where I would have written "with the
-> fraction of edges connecting samples that are common to two samples"
-> rather than "edges that connect between samples" (which doesn't
-> strictly make sense). I believe there is the same measure of
-> beta-diversity in the community-phylogenetics literature? What is the
-> closest analogue?
-
-I don't follow your "with the fraction of edges connecting samples that are common to two samples" correction.
-There is some overlap with the PD literature as now described and cited.
-
-> 15-4955. The description of Evans and Matsen seems a bit detailed in
-> the context -
-
-I've shortened and simplified, thanks.
-
-> 16-16. This requires another sentence of explanation. From this
-> reading, the Unifrac method offers a single distance between a pair of
-> samples (a ratio of two numbers of edges), rather than a matrix.
-
-I have rephrased so that it's clear that the matrix comes from applying
-the measure to all pairs of sequences in a sample.
-
-> 16-42..51. This is a short paragraph that would seem to explicitly
-> link HMR and SB. Why is it so telegraphic?
-
-I have added a clarifying phrase, but I'm not sure what more is needed.
-The two communities have both identified a given technique (although
-neither has really gotten much traction).
-
-> 17-45. This seems an odd heading for this section, which seems to be
-> about horizontal gene transfer. Given the amount of technical detail
-> elsewhere, why are we not told how horizontal gene transfer is
-> inferred in these HMR studies?
-
-Explained now, thanks.
-
-> Phylogenetic inference. I found this survey very useful. Is there a
-> reason this is not an earlier section of the paper, given these are
-> the trees upon which many of the comparisons referred to in previous
-> sections are based?
-
-I suppose I wanted to talk about all of the interesting things that
-microbiome researchers are doing with trees before
-borderline-criticizing the ways that they make them! The readers of SB
-know what it means to infer a tree, so I don't think that this section
-is a prerequisite for the previous ones.
-
-> 19-24. "which has ported in the" Is this the correct structure?
-
-That was what was intended, but I've updated it to use less
-programmer-speak.
-
-> 20-50. "different parameter regimes" - this phrase is used several
-> times in this review. What does it mean here? When would we prefer to
-> add sequences to a fixed tree rather than re-infer the totality? Is
-> there an analogous situation in SB - i.e. is there something SB can
-> learn from HMR here?
-
-I've simply removed the sentence. The idea of this paragraph is to
-motivate updating of phylogenetic trees with new sequences, and I've
-more fully explained that and added a reference.
-
-> I found the challenges section to be rich and well-written, with
-> questions that will resonate with the SB community - e.g. the
-> discussion of the proxy problem (how edge lengths predict feature
-> diversity). The fact that 16S distance does predict whole genome
-> distance is highlighted here - it is presented in previous sections,
-> but perhaps it could be given more prominence there, given its
-> prominence here. This is a general problem.
-
-I have added a sentence emphasizing that this assumption underlies the
-way researchers interpret UniFrac.
-
-> 22-10. "Microbial ecology" Perhaps "and explicitly for human microbial
-> ecology" to remind us that HMR is the focus here?
-
-Added, thanks.
-
-> 23-23. "Whole organism approach"  Will this approach be known to SB
-> readers? I cannot imagine what is being done here.
-
-You are right that this was not explained properly. I have simply
-removed the paragraph.
-
-> 25-31. "At least for the next several years we can expect more related
-> work." What does this mean? Related to what has been reported above?
-
-I have made this more specific.
-
-> 25-32. The future
-
-> I found the final sentence to leave us dangling. After having reviewed
-> this manuscript, I could not say exactly where the intersection of SB
-> and HMR will be most fruitful for HMR, nor what in HMR might be
-> fruitful for SB. The conclusion seems to suggest that phylogenetics
-> would not be used in diagnostics, for example.
-
-I have clarified this sentence now to clarify that some applications
-will go but many new ones will be needed.
-
-> Will it be important in primary research then? In constructing the
-> taxonomies the diagnostics may require?
-
-The rest of the paper is about how phylogenetics has impacted HMR, and
-these applications will continue into the future. I'm not sure how to
-make this more clear.
-
-> Finally, the abstract says the review ends by describing
-> "the current challenges to phylogenetics coming from this type of
-> work."  My reading of the relevant sections is that there is more
-> discussion of challenges to phylogenetic HMR research (sampling
-> biases, robustness of UniFrac to incorrect tree reconstruction) than
-> to phylogenetics per se. The discussion of how to update known trees
-> with new sequences is an exception.
-
-I suppose I don't differentiate between the two.
-
-> Reviewer 3, Frederic Delsuc
-> ---------------------------
-
-> This manuscript presents a review of how phylogenetic tree-thinking
-> and methodology is currently used in the study of the human
-> microbiome. The author not only reviews the different ways by which
-> phylogenetics is currently percolated into microbial ecology, but also
-> gives his personal views on the future phylogenetic challenges
-> associated with microbiome studies, and insists on the crucial role
-> that molecular phylogeneticists has to play in this quickly expanding
-> field. Overall, I enjoyed reading this particularly timely manuscript,
-> which I find well written. I only have minor comments/suggestions for
-> potential improvement.
-
-> Specific comments:
-> 1. I think that it should be made clear, from the very beginning of
-> the introduction, the definition of the microbiome in order to avoid
-> confusion. Indeed, in the current literature, "microbiota" (i.e. the
-> collection of microbes living in association with a given organism)
-> and "microbiome" (i.e. the collection of genomes of those microbes)
-> are more and more often used interchangeably, even though they are not
-> referring to the same organizational level. In that sense, the
-> definition of the "human microbiome" given at the top of page 5 ("The
-> human microbiome is the collection of microbial organisms that live
-> inside of and on the surface of humans.") does not seem to be
-> accurate. Please clarify.
-
-> 2. In the section introducing the human microbiome, I think that a
-> paragraph on the origin and the way in which the microbiome is
-> transmitted vertically between the mother and infant is lacking. The
-> recent review on maternal transmission by Funkhouser and Bordenstein
-> 2013 is cited later, but it would be nice to mention it on page 5,
-> especially in the context of the some examples on the impact of birth
-> method (vaginal birth vs. cesarean delivery) on future health of
-> newborns.
-
-Added.
-
-> 3. On page 6, lines 26-27, it might be worth citing also the recent
-> paper by Findley et al. (doi:10.1038/nature12171) on the topography of
-> fungal and bacterial on human skin.
-
-Thanks.
-
-> 4. On page 6, line 56, maybe also specify that "survey" is the
-> equivalent of barcoding, a term with which the Systematic Biology's
-> community is more familiar.
-
-Added, thanks.
-
-> 5. On page 14, lines 14-19, maybe mention the widespread use of
-> "rarefaction curves" to enable the comparison between samples of
-> different sequencing depth or methods (e.g. 454 vs. Illumina).
-
-I do describe strategies for read count normalization in the next paragraph, but have added the "rarefaction curve" terminology.
-
-> 6. On page 17, lines 54-55, please add some recent references on the
-> current debate around the rooting of the placental tree. I don't think
-> that most readers of Syst. Biol. will be familiar with this
-> discussion.
-
-Good point. I've put in the two 2013 MBE papers. If there is something else I should cite let me know.
-
-> 7. On page 18, lines 26-27, RAxML is also implemented in QIIME.
-
-Added.
-
-> 8. On page 20, I think that this might be the place to discuss the
-> potential use of phylogenetic networks in microbiome research (see
-> Parks and Beiko 2012 Mol Biol Evol).
-
-Thank you for pointing me to this reference.
-
-> 9. On page 24, when citing uBiome, it might also be fair to refer the
-> similar crowd-funded American Gut Project.
-
-Thanks.
-
-> Typos:
-> P3, L40: I would rather use "species delimitation" instead of "species definition".
-
-Thanks.
-
-> P4, L9: "These techniques and could be used"
-
-Fixed.
-
-> P6, L32: "we will be primarily be describing"
-
-Whoops!
-
-> P9, L29, L30: Maybe replace "diversity" by "variability"
-
-Thanks.
-
-> P18, L32: "a long history as a means to sequentially build"
-
-Reworded, thanks.
-
-> P19, L16: "GreenGenes 13_5 alignment has is almost 80% gaps"
-
-Yikes, thanks.
+> I went through both the revised manuscript and the author’s response
+> letter. I think that the author has made significant efforts to
+> address all editorial and referee comments, including my own concerns.
+> I would therefore recommend the acceptation of the revised manuscript
+> basically as is.

### 2014-06-03 sbreview.tex
Doing the easy changes.

@@
 This brings up the question of how many sequences are needed to characterize the microbial diversity of a given environment.
-To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial} however to compare more similar samples deeper sequencing is required.
+To distinguish between two rather different samples, relatively few sequences per sample are required \citep{kuczynski2010microbial}, however, to compare more similar samples deeper sequencing is required.
 In addition to sequencing samples across individuals, this parallelization has also enabled sampling through time \citep[e.g.][]{caporaso2011moving}.
@@
 \citet{Wang2013-ol} optimize a criterion of cluster modularity.
-\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixtures on 3-mer profiles, and then a minimum description length criterion.
+\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixture on 4-mer profiles, and then a minimum description length criterion.
 \citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a relatively large number of sequencing reads and so can be used as an OTU clustering method.
@@
 Phylogenetic diversity metrics, which do take this overall diversity structure into account, are also used.
-However, whereas just about every of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate, only a few involve PD.
+However, whereas just about every one of the hundreds or thousands of 16S surveys applies an OTU-based alpha diversity estimate, only a few involve PD.
 
@@
 
-The apparent success of 16S-tree-based comparisons begs the question of if a more complete representation of the evolutionary history of the microbes would yield better comparisons.
+The apparent success of 16S-tree-based comparisons raises the question of if a more complete representation of the evolutionary history of the microbes would yield better comparisons.
 This suggests a practical perspective on the theoretical issue of the tree of life: what is the representation of the genetic ancestry of a set of microbes that allows us to best perform proxy whole-genome comparison?

### 2014-06-24 reviews/response.txt
Doing the easy changes.
Addressing some comments.

@@
+Editor
+------
+
 > 1) Pg.
@@
 I suppose that’s indirectly true, but I don’t think it is the best way to describe it.
+
+I've clarified this, thanks.
 
@@
 
+Absolutely. I've changed the text accordingly.
+
 > 3) Pg. 14: “first with a Dirichlet multinomial mixtures on 3-mer profiles” – Should this just be “mixture”?
+
+Right! Thanks.
 
@@
 
+Done, thanks.
+
 > 5) Pg.
 16: “PD under random sub-sampling, which is often applied to enable comparison be- tween samples of different sequencing depths and called “rarefaction curves.”” – Would it be preferable to write that the random subsamples are described (or depicted) with rarefaction curves?
+
+You caught a strange sentence there. I've just cut the strange bits and linked to a later discussion.
 
@@
 
-Your revision must comply with our instructions to authors (in this letter, with additional instructions at http://www.oxfordjournals.org/our_journals/sysbio/for_authors).
-This applies even if your original submission deviated from Systematic Biology style and corrections were not included in the notes from the editors or reviewers.
-Therefore, READ AND APPLY OUR INSTRUCTIONS BEFORE SUBMITTING A REVISION.
-Failure to do so will result in significantly delayed processing of your paper.
-If anything remains unclear after reading the instructions, use recent issues of the journal to find examples.
-If necessary, feel free to contact the Managing Editor at sysbio.editorialoffice@oup.com for help.
+I didn't know that! Thank you.
 
-After you've uploaded your revision, carefully view the Manuscript Central version to verify that all figures and other files display correctly, and that you’ve followed all of our author instructions.
-
-The cost of printed color figures is $600 each.
-Authors are normally expected to cover this cost, but we do have limited funds available for authors who are SSB members, and cannot pay the full amount.
-If you feel your circumstances create an unusual financial need, please explain in your cover letter.
-The maximum allowance is one color figure per paper.
 All color costs can be avoided if you decide to have the figures printed in black and white, but be shown in color in the online version of the paper.
@@
 
-Journal Covers: Please consider submitting a suggested cover image.
-They can be illustrations of theory, photos of organisms, a combination of the two, or alternatives.
-These can be uploaded with a revision of the manuscript or may be sent later.
-Only images for which you have copyright permission or that are not under copyright may be submitted.
-The color figure fee does not apply to images chosen by the Editor to be on the cover of the issue.
-If you have questions about possible cover designs, please email sysbio.editorialoffice@oup.com.
-We encourage and greatly appreciate cover image suggestions.
-
 Make sure all section headings conform to Systematic Biology style for first, second and third levels.
@@
 
-Tables should have a single-sentence informative title above the table, with any other descriptive information located below the table in the form of notes and/or specific footnotes.
-
 When references are grouped together in parentheses in the text, they should be listed in ascending chronological order. Multiple references in a single year should be alphabetized.
-
-All funding used for this work should be listed in a "Funding" section preceding the Acknowledgements. Please give the full official name of each funding body.
 
@@
 
+Thanks. I've added commas.
+
 > P12 “Several groups are continually revising taxonomy...” and P13 “An
@@
 > self-contentedness.
+
+You are right. I've decided to just eliminate the technical sentences.
 

### 2014-06-24 sbreview.tex
Addressing some comments.

@@
 \usepackage[labelfont=bf,labelsep=period,font=small]{caption} % bold the 'Figure #' in the caption and separate it with a period
-\newcommand{\forarxiv}[1]{}
-\newcommand{\notforarxiv}[1]{#1}
+\newcommand{\forarxiv}[1]{#1}
+\newcommand{\notforarxiv}[1]{}
 
 % for SB submission
-\renewcommand{\section}[1]{%
-\bigskip
-\begin{center}
-\begin{Large}
-\normalfont\scshape #1
-\medskip
-\end{Large}
-\end{center}}
-\renewcommand{\subsection}[1]{%
-\bigskip
-\begin{center}
-\begin{large}
-\normalfont\itshape #1
-\end{large}
-\end{center}}
-\renewcommand{\subsubsection}[1]{%
-\vspace{2ex}
-\noindent
-\textit{#1.}---}
-\renewcommand{\tableofcontents}{}
-\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
-\linespread{1.6}
+% \renewcommand{\section}[1]{%
+% \bigskip
+% \begin{center}
+% \begin{Large}
+% \normalfont\scshape #1
+% \medskip
+% \end{Large}
+% \end{center}}
+% \renewcommand{\subsection}[1]{%
+% \bigskip
+% \begin{center}
+% \begin{large}
+% \normalfont\itshape #1
+% \end{large}
+% \end{center}}
+% \renewcommand{\subsubsection}[1]{%
+% \vspace{2ex}
+% \noindent
+% \textit{#1.}---}
+% \renewcommand{\tableofcontents}{}
+% \bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
+% \linespread{1.6}
 
@@
 The answer is simple: this system is data- and question-rich.
-It absolutely requires molecular methods because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
+It absolutely requires molecular methods because microbes are now primarily identified by their molecular sequences, and identification of microbes by their molecular sequences is much more straightforward to do in high throughput than morphological or phenotypic characterization.
 Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.
@@
 Where phylogenetic research on mammals is converging on one or two possible basic structures for their evolution, much of the deep history of microbes remains obscure.
-Inference of this history is complicated by the fact that any tree-like signal in the overall evolution of bacteria is restricted to a small set of so-called ``core'' genes \citep{bapteste2009prokaryotic,Leigh2011-zu,Lang2013-zt}.
+Inference of this history is complicated by the fact that any tree-like signal in the deep evolutionary relationships between bacteria is restricted to a small set of so-called ``core'' genes \citep{bapteste2009prokaryotic,Leigh2011-zu,Lang2013-zt}.
 
@@
 The GreenGenes group \citep{desantis2006greengenes} has been very active in updating their taxonomy according to 16S, first with their GRUNT tool \citep{dalevi2007automated} and more recently with their tax2tree tool \citep{mcdonald2011improved}.
-Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree, which is annotated as follows.
-The first step is to label the internal nodes of the tree to optimize the $F$-measure (the harmonic mean of precision and recall) across clades, where the taxonomic label of a clade is thought of as a prediction for the descendant sequence taxonomic labels.
-The second step reassigns sequence taxonomic labels based on these clade taxonomic labels.
-Interestingly, tax2tree allows for polyphyletic taxonomic groups as an allowance for either phylogenetic error, lack of resolution of the 16S gene, or taxonomic groups with no evolutionary basis; the method does not attempt to signal the cause of such polyphyletic groups.
+Tax2tree uses a heuristic algorithm to reassign sequence taxonomic labels so that they are concordant with a given rooted phylogenetic tree in a way that allows polyphyletic taxonomic groups.
 \citet{matsen2012reconciling} developed algorithms to quantify discordance between phylogeny and taxonomy based on a coloring problem previously described in the computer science literature \citep{moran2008convex}.
@@
 Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
-The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
+The (unweighted and unrooted) PD of a set of taxa $S$ is simply the total of the lengths of the branches which connect taxa in $S$ \citet{faith1992conservation}.
 It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
@@
 \citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiota; in particular that they do a better job than the commonly-used OTU-based measures.
-\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''
+\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling (see discussion of rarefaction below).
 
@@
 I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 MCEB Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
-I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank Bastien Boussau, Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript, and thank the NIH (R01 HG005966-01) and NSF (1223057) for financial support.
+I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank Bastien Boussau, Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript.
+This work was supported by National Institutes of health grant R01-HG005966-01 and National Science Foundation grant 1223057.
 

### 2014-06-24 sbreview.tex
Sorting references.

@@
 Recent work by \citet{kembel2012incorporating} implements the independent contrasts method \citep{felsenstein1985phylogenies} to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
-Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
+Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{mcnabb2004hsp65,case2007rpob}, however 16S remains the dominant locus used by a large margin.
 A final cause of noise is next-generation sequencing error: this is certainly a problem for both surveys and metagenomes, but is becoming less of a problem as technology improves.
@@
 
-Metagenomic data sets are often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
+Metagenomic data sets are often used to infer information about metabolism rather than phylogenetic nature \citep{abubucker2012metabolic,greenblum2012metagenomic}.
 Discussing these methods is beyond the scope of this paper, as is the sequencing of mRNA in bulk which is called ``metatranscriptomics".
@@
 This data may become more common for unculturable organisms as single-cell sequencing methods improve \citep[reviewed in][]{kalisky2011single}.
-The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \citep{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.
+The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \citep{emerson2012metagenomic,howe2012assembling,iverson2012untangling,pell2012scaling,podell2013assembly}.
 
@@
 The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the taxonomic tree.
-\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
+\citet{munch2008fast,munch2008statistical} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
 \citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
@@
 There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a so-called fixed reference tree, sometimes with uncertainty estimates.
-These programs \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} have various speeds and features.
+These programs \citep{monierEaLargeViruses08,vonMeringEaQuantitative08,wu2008simple,matsen2010pplacer,stark2010mltreemap,berger2011performance} have various speeds and features.
 This work has also spurred development of specialized alignment tools for this mapping process.
@@
 While developing community assembly models forms an important project for microbial ecology generally and human-associated microbial ecology in particular, phylogeny-aware methods could be further developed.
-One way to model microbial community assembly is to apply Hubbell's neutral theory \citep{fierer2012animalcules,costello2012application}.
+One way to model microbial community assembly is to apply Hubbell's neutral theory \citep{costello2012application,fierer2012animalcules}.
 \citet{o2012phylogenetic} model community assembly with an explicitly phylogenetic perspective, and include some comparison of models to data.
@@
 One type of normalization handles differential depths of sequencing across samples.
-The presently used approach is \textit{rarefaction}, which means uniform sub-sampling to the number of reads in the lowest abundance sample \citep{caporaso2010qiime,schloss2009introducing}.
+The presently used approach is \textit{rarefaction}, which means uniform sub-sampling to the number of reads in the lowest abundance sample \citep{schloss2009introducing,caporaso2010qiime}.
 In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another.

### 2014-06-24 response/response.txt
response, not reviews

--- 2014-06-24 reviews/response.txt
@@
 In your cover letter, please make your intentions regarding the use of color clear.
-
-Make sure all section headings conform to Systematic Biology style for first, second and third levels.
-Use of incorrect styles is potentially confusing and, in any case, is likely to delay processing of the manuscript.
-
-When references are grouped together in parentheses in the text, they should be listed in ascending chronological order. Multiple references in a single year should be alphabetized.
 

### 2014-06-24 response/v1/response.txt
response, not reviews

> Reviews from Prof. Frank (Andy) Anderson
> ----------------------------------------

> Thank you for your Systematic Biology submission. It has been reviewed
> by Associate Editor Prof. Olivier Gascuel and three reviewers. Their
> comments are listed at the end of this letter. The reviewers and the
> AE provide some excellent constructive suggestions that I am sure you
> will appreciate.

> I have little to add that wasn't covered by one or more reviewers or
> Olivier in their fine reviews. As an instructor who has almost given
> up on trying to convince microbiology graduate students that they can
> do better analyses than neighbor joining, I was already sensitive to
> some of these differences, but this review helped me appreciate the
> interface between the two disciplines. I agree with Mooers that the
> level of detail varies across the manuscript, but I mainly noticed
> when a topic was summarized quickly and I didn't really understand it
> (I note that Mooers and I both had a question about how tax2tree
> works).

> I have a few comments and questions that may help clarify some of the
> writing. Altogether, I think the manuscript really requires only
> several minor revisions.

> Like Olivier, I note that some of my comments may overlap those of the
> reviewers.

> 1) Pg. 9: "As described above "metagenome" means that data is sheared
> randomly across the genome rather than amplified from a specific
> location," - I assume you mean DNA is sheared randomly?

Yes, thanks.

> 2) Remember, "data" still is a plural noun, at least under my watch!
> Check for "this data", "data is", "data was", etc

Searched and replaced. Thanks!

> 3) Pg. 10: I don't really understand what "tax2tree" does, based on
> the description here. Does that matter?

I have improved the description, with an overview sentence for those who
want to skip the details.

> 4) Pg. 14-15: I also don't really understand what UniFrac is or how it
> works, even after looking at Figure 2 (Olivier made a similar
> comment). Could we get some help here?

I have much improved the caption for Figure 2. I think it should be
clear now.

> 5) The figures need a bit of work. There seems to be some sort of
> equation at the bottom of Figure 1, but I can't decipher it. What do
> the colors mean in Figure 2? I don't find this to be very clear.

I have made a connection between the figure and the caption via the
mathematical notation used. For Figure 2, I have greatly expanded the
caption to describe colors.

> 6) Pg. 20: "How to formulate a useful notion of independent contrasts
> (Felsenstein 1985) on such an object?" - Awkward wording.

I have changed this to "How might one formulate a useful notion of
independent contrasts (Felsenstein 1985) on such an object?" I hope
that's better.

> 7) Pg. 23: I don't understand the distinction between "base
> observations" and sample observations here. Can this be clarified?

I have rewritten this with a little notation, thanks.

> 8) Pg. 24: "Future will continue to bring deeper sequencing on more
> samples" - I guess this should be "The future will continue"?

Thanks.

AE, Olivier Gascuel
-------------------

> My recommendation is acceptance pending minor to major changes.
> Actually, no change asked by the reviewers and me (see below) requires
> major efforts, but we all come from different origins, with different
> points of view and expectations, and thus summing all these demands
> will require a bit of time and attention. Basically, we all ask for
> more details and comments here and there, to make the paper more
> precise, comprehensive and self-contained. I am confident that you'll
> find our suggestions useful. I don't believe that the paper should be
> shortened in some places (or very locally), as suggested by Arne
> Mooers. When submitting your revision, please prepare a cover letter
> containing all the comments by the referees and me, with your
> changes/responses.

> Specific comments, possibly redundant with those from the reviewers,
> using Systematic Biology page numbering:

> P4	these techniques AND could

Thanks!

> P8	(Kuczynski et al. 2010). However, for more subtle information
> P8	"for more subtle information deeper sequencing is required" :
> what do you have in mind here?

I made this more specific.

> Indeed "distinguish between two rather different samples" is a modest
> goal

Agreed, but it's not immediately obvious what level of sequencing depth
is needed even in that situation.

> Actually, it's not clear in this subsection what is meant by
> "Microbial community estimation" (title of the subsection). What is
> the goal here? Identify and classify microbes? Measure their
> abundance? Etc.? A few lines and explanations should suffice.

I have revised the heading as well as separated out a paragraph
describing terminology and the goal.

> P11	What is the level of uncertainty in current bacterial
> phylogenies? For example, one commonly considers that only a few
> unknowns remain in mammals phylogeny; what about bacteria? Are people
> discussing the details or the broad lines? Provide a few illustrative
> examples.

A great suggestion. I have added material at the beginning of the
"Tree-thinking in human microbiome research" section. I don't think that
examples are too productive here because people haven't gotten to that
level of detail yet.

> P13 	"There are many OTU inference methods with various speeds and
> strategies" and the two following paragraphs. Describe in a simple way
> the basic principles of these methods, before discussing their
> relative merits (this remark may apply in other places).

Given the diversity of OTU methodologies, it's not easy to generalize
about their operation. However, I have added a paragraph explaining what
it is that OTU inference methods are trying to do. I hope that helps.

> P14 	You describe PD and suggests that it should be used more often.
> But what are the standard alpha diversity measures, and their
> limitations?

Thank you for this excellent suggestion.

> P15 	Are "partially abundance weighted diversity measures" using some
> phylogenetic approaches? Be a bit more specific when describing and
> discussing methods. Here and in many other places (e.g. UniFrac,
> weighted PD, weighted UniFrac, etc.).

You are right. I hope my edits make this more clear.

> P15 	"Evans and Matsen (2012) showed that weighted UniFrac" check
> sentence.

I've shortened and simplified this sentence.

> P16 	explanations and Fig. 3 on earth-movers distance are unclear to
> me.

I have expanded the (previously insufficient) descriptions.

> P17 	"By doing so they have recovered more novel protein families
> than they would have using methods organized by taxonomy". I'm lost
> here, as it is said just above that they have derived a
> "phylogeny-driven genomic encyclopedia of Bacteria and Archaea".

I've made that more specific now.

> P23-24 Paragraph "In the same vein This would be a false positive." is
> unclear to me.

I've made the false positive part more specific.

> P24 Check "The presently used approach is rarefaction, which means
> uniform sub-sampling, typically without replacement to the number of
> reads in the lowest abundance sample"

Simplified, thanks.

> P25 Check "Comparative studies will continue to investigate what
> shapes and is shaped by the microbiome"

I like this sentence.

> P26 "during the 2013 LIRMM Mathematical and Computational Evolutionary
> Biology workshop", change LIRMM into MCEB

I have done so, so that the sentence reads "... 2013 MCEB Mathematical
and Computational Evolutionary Biology workshop ..." is that right?

> Fig 2. It took time to me to understand and check on the Web that the
> tree must be rooted, to compute the unifrac distance. This is one
> example, among others, that more details are required when describing
> methods, criteria, measures, etc. Some more figures could help, or one
> appendix with formal definitions of the main concepts and tools. At
> least, details and explanations have to be added in the main text.

> Your reference list is extremely long, mostly due to the fact that you
> are citing papers with a lot of authors. Check Syst Biol
> recommendation to authors (I assume that our editorial office will
> answer and help).

> Reviewer 1, Alexis Stamatakis
> -----------------------------

> Comments:
> This is a well written and in depth review on the usage of
> phylogenetics in human microbiome research, which I enjoyed reading.

> I only have a few minor remarks:

> It might be a consideration to cite our recent paper:

> J. Zhang, P. Kapli, P. Pavlidis, A. Stamatakis: "A General Species
> Delimitation Method with Applications to Phylogenetic Placements". In
> Bioinformatics (2013), 29 (22): 2869-2876.
> It would fit will into section "The role of OTUs".

It's cited now.

> Apart from that, I believe that a more critical and slightly extended
> discussion of taxonomic classifiers in the very last part of the paper
> should be included. From my point of view, this is still a mostly
> unresolved problem and the currently best classifier, which you may
> also want to cite,  still shows pretty horrible accuracy
> (http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0049334)
> at the genus and family level.

Given that taxonomic classification is not always a phylogenetic
problem, I don't think that it merits its own section. I have cited your
suggestion and other recent progress in this field for readers who want
to dig deeper.

> I would also appreciate if you could explicitly mention the Gartner
> hype curve in the discussion part and provide an estimate as to where
> we are on this curve at present.

I don't see how this would contribute to the scientific goals of the
paper.

> Some typos:

> page 5 lines 7-8 "these techniques and could be used" there seems to
> be a verb missing.

Fixed, thanks.

> page 14 lines 12 Here you may add our recent work on PTP, since it
> strives to conduct OTU clustering and phylogenetic inference in one
> single step/pipeline.

Right, thanks.

> page 15 l 50: in "distance, is special" seems that it should read
> "distance, it is special"

Thanks, I've rearranged here.

> page 20: lines 51-56: In fact we are currently developing a tool for
> automatic tree extension as described here. While not published yet,
> it is already available as open-source code at:
> http://sco.h-its.org/exelixis/web/software/put/index.html

The paper has been published now and I now cite it. Thanks!

> Reviewer 2, Arne Mooers
> -----------------------

> This was an interesting manuscript to read, and I compliment whoever
> thought to review Human Microbiome Research (HMR) for the SB
> readership. Overall, there is a lot of very interesting work reported
> on here, though I found the level of detail to be a bit uneven:
> sometimes we are told perhaps too much, and often, what feels like too
> little, in order to make the connections between HMR and SB. There is
> also seems to be some elision between HMR research and metagenomics
> more generally. That isn't a bad thing, but one does sometimes forget
> that we are focussed on HMR.

> Numbers refer to the SB manuscript pagination, rather than to the
> original latex.

> 4-45 additional to

I'm afraid I can't find to what you refer.

> 5-8 delete and

Fixed.

> 7-23 delete though

Nice catch!

> The section "The Human Microbiome" is quite detailed (discussing mouse
> to human transplants wrt obesity, etc.). Given the intense media
> reporting of the work coming from this area, SB readers may not need
> to be primed of the importance of this field. If this were shortened,
> there could be room to expand some of the technical bits that SB
> readers will want that follow. The last part of the section  - about
> sticking to eubacteria and a non-molecular biology approach do seem
> germane.

The editors have stated that they are not concerned with this becoming
too long, and I've been surprised that what seems like an absurdly big
amount of media hype to me doesn't reach everyone. So I'm leaving it in
unless the editors say otherwise.

> 8-12. these data (I guess we will eventually decide that data is
> singular - up to the editors)

I have fixed all these.

> 8-37++ Whereas Woese ++ This is a bit wordy and unnecessary - we know
> that Woese didn't have PCR

Er, perhaps this is a generational thing? I had no idea how they used to
infer molecular sequences before Pace's phyloseminar.

> 10-30 Which "databases" ? - this has been written so far as if it were
> being applied to samples, not databases. I am not clear what the unit
> of observation is here.

I have clarified with a slightly expanded explanation, thanks.

> 10-32 We quickly switch here from easy background (nonPCR random
> sequencing) to something brand new (at least to me). "At least some
> portion of the rest of the data can be taxonomically classified
> (methods reviewed by Mande et al. 2012); Treangen et al. (2013) report
> speedups and much higher accuracy when reads are assembled before they
> are classified."
> "Rest of data" refers to what - reads found in a database, or reads
> from a sample? "Accuracy" refers to what? "Speed-ups" refer to what?

Thanks for reporting this. I'm now more specific.

> 11-53. Here is the first real point of contact between phylogenetics
> and HMR. But what is an "F-measure of precision and recall" - there is
> no citation.

I've added a mini-definition.

> 12-28 and 12-32. What is the measure of performance that this refers
> to? How do we decide if a method performs well?

Clarified, thanks.

> 12-29. Do readers of SB now what "k-mer composition" is? Is it k-means
> clustering?

I've added in a definition.

> 12-37++. Is the MEGAN approach compositional or phylogenetic? I
> believe it is meant to be a hybrid approach (BLAST being
> compositional, but the taxonomic labeling considered phylogenetic),
> but this could be clarified.

I've added in a clarification, thanks.

> 13-29. It is not clear what heuristics are required for a cut-off
> method. The author says that he is going to describe the OTU inference
> industry, but the paragraph is really just a list of acronyms, and
> then refers obliquely to a "contingency table" of OTU observations.
> So, for example, what is a notion of goodness of clustering that the
> heuristics are trying to achieve? What are the sorts of species
> divisions found in taxonomies? This is an area where more would be
> helpful - and where a shorter introduction might offer space.

These are excellent questions, and the landscape is a little complex. I
have expanded this section to explain what it is that these methods
would like to achieve.

> 14-40. Whereas just about every 16S survey investigation involves an
> OTU-based alpha diversity estimate, only a few involve PD: unweighted
> PD has been applied to some 16S survey data (Lozupone and Knight 2007;
> Costello et al. 2009) and to metagenomic reads in a set of marker
> genes (Kembel et al. 2011) - This is an odd pair of sentences: is the
> L&K the only study that used unweighted PD? If so, in what context?

I can't quite understand this suggestion. All three papers listed used
unweighted PD.

> 14-46. Check this sentence - perhaps start with "While are commonly
> used, abundance weighted PD measures are not.

I have updated this sentence based on your suggestion, thanks.

> 15-30. Perhaps "In contrast to measures of alpha diversity, measures
> of beta diversity do make use of phylogenetics"  This is actually an
> interesting contrast. Is this just historical contingency, or is there
> a logical reason why the two areas disagree on the use of phylogenetic
> information?

I think that this refers to 14-30, and I've simply removed the sentence.
With regards to your question, I don't know. It seems surprising to me.

> 15-xx. This seems a strong point of contact between metazoan
> phylogenetics and the microbiome folks. UniFrac is clearly described
> on its own terms, except perhaps where I would have written "with the
> fraction of edges connecting samples that are common to two samples"
> rather than "edges that connect between samples" (which doesn't
> strictly make sense). I believe there is the same measure of
> beta-diversity in the community-phylogenetics literature? What is the
> closest analogue?

I don't follow your "with the fraction of edges connecting samples that
are common to two samples" correction. There is some overlap with the PD
literature as now described and cited.

> 15-4955. The description of Evans and Matsen seems a bit detailed in
> the context -

I've shortened and simplified, thanks.

> 16-16. This requires another sentence of explanation. From this
> reading, the Unifrac method offers a single distance between a pair of
> samples (a ratio of two numbers of edges), rather than a matrix.

I have rephrased so that it's clear that the matrix comes from applying
the measure to all pairs of sequences in a sample.

> 16-42..51. This is a short paragraph that would seem to explicitly
> link HMR and SB. Why is it so telegraphic?

I have added a clarifying phrase, but I'm not sure what more is needed.
The two communities have both identified a given technique (although
neither has really gotten much traction).

> 17-45. This seems an odd heading for this section, which seems to be
> about horizontal gene transfer. Given the amount of technical detail
> elsewhere, why are we not told how horizontal gene transfer is
> inferred in these HMR studies?

Explained now, thanks.

> Phylogenetic inference. I found this survey very useful. Is there a
> reason this is not an earlier section of the paper, given these are
> the trees upon which many of the comparisons referred to in previous
> sections are based?

I suppose I wanted to talk about all of the interesting things that
microbiome researchers are doing with trees before
borderline-criticizing the ways that they make them! The readers of SB
know what it means to infer a tree, so I don't think that this section
is a prerequisite for the previous ones.

> 19-24. "which has ported in the" Is this the correct structure?

That was what was intended, but I've updated it to use less
programmer-speak.

> 20-50. "different parameter regimes" - this phrase is used several
> times in this review. What does it mean here? When would we prefer to
> add sequences to a fixed tree rather than re-infer the totality? Is
> there an analogous situation in SB - i.e. is there something SB can
> learn from HMR here?

I've simply removed the sentence. The idea of this paragraph is to
motivate updating of phylogenetic trees with new sequences, and I've
more fully explained that and added a reference.

> I found the challenges section to be rich and well-written, with
> questions that will resonate with the SB community - e.g. the
> discussion of the proxy problem (how edge lengths predict feature
> diversity). The fact that 16S distance does predict whole genome
> distance is highlighted here - it is presented in previous sections,
> but perhaps it could be given more prominence there, given its
> prominence here. This is a general problem.

I have added a sentence emphasizing that this assumption underlies the
way researchers interpret UniFrac.

> 22-10. "Microbial ecology" Perhaps "and explicitly for human microbial
> ecology" to remind us that HMR is the focus here?

Added, thanks.

> 23-23. "Whole organism approach"  Will this approach be known to SB
> readers? I cannot imagine what is being done here.

You are right that this was not explained properly. I have simply
removed the paragraph.

> 25-31. "At least for the next several years we can expect more related
> work." What does this mean? Related to what has been reported above?

I have made this more specific.

> 25-32. The future

> I found the final sentence to leave us dangling. After having reviewed
> this manuscript, I could not say exactly where the intersection of SB
> and HMR will be most fruitful for HMR, nor what in HMR might be
> fruitful for SB. The conclusion seems to suggest that phylogenetics
> would not be used in diagnostics, for example.

I have clarified this sentence now to clarify that some applications
will go but many new ones will be needed.

> Will it be important in primary research then? In constructing the
> taxonomies the diagnostics may require?

The rest of the paper is about how phylogenetics has impacted HMR, and
these applications will continue into the future. I'm not sure how to
make this more clear.

> Finally, the abstract says the review ends by describing
> "the current challenges to phylogenetics coming from this type of
> work."  My reading of the relevant sections is that there is more
> discussion of challenges to phylogenetic HMR research (sampling
> biases, robustness of UniFrac to incorrect tree reconstruction) than
> to phylogenetics per se. The discussion of how to update known trees
> with new sequences is an exception.

I suppose I don't differentiate between the two.

> Reviewer 3, Frederic Delsuc
> ---------------------------

> This manuscript presents a review of how phylogenetic tree-thinking
> and methodology is currently used in the study of the human
> microbiome. The author not only reviews the different ways by which
> phylogenetics is currently percolated into microbial ecology, but also
> gives his personal views on the future phylogenetic challenges
> associated with microbiome studies, and insists on the crucial role
> that molecular phylogeneticists has to play in this quickly expanding
> field. Overall, I enjoyed reading this particularly timely manuscript,
> which I find well written. I only have minor comments/suggestions for
> potential improvement.

> Specific comments:
> 1. I think that it should be made clear, from the very beginning of
> the introduction, the definition of the microbiome in order to avoid
> confusion. Indeed, in the current literature, "microbiota" (i.e. the
> collection of microbes living in association with a given organism)
> and "microbiome" (i.e. the collection of genomes of those microbes)
> are more and more often used interchangeably, even though they are not
> referring to the same organizational level. In that sense, the
> definition of the "human microbiome" given at the top of page 5 ("The
> human microbiome is the collection of microbial organisms that live
> inside of and on the surface of humans.") does not seem to be
> accurate. Please clarify.

I have changed the definitions along the lines you suggest and have
provided an introduction to the terminology.

> 2. In the section introducing the human microbiome, I think that a
> paragraph on the origin and the way in which the microbiome is
> transmitted vertically between the mother and infant is lacking. The
> recent review on maternal transmission by Funkhouser and Bordenstein
> 2013 is cited later, but it would be nice to mention it on page 5,
> especially in the context of the some examples on the impact of birth
> method (vaginal birth vs. cesarean delivery) on future health of
> newborns.

Added.

> 3. On page 6, lines 26-27, it might be worth citing also the recent
> paper by Findley et al. (doi:10.1038/nature12171) on the topography of
> fungal and bacterial on human skin.

Thanks.

> 4. On page 6, line 56, maybe also specify that "survey" is the
> equivalent of barcoding, a term with which the Systematic Biology's
> community is more familiar.

Added, thanks.

> 5. On page 14, lines 14-19, maybe mention the widespread use of
> "rarefaction curves" to enable the comparison between samples of
> different sequencing depth or methods (e.g. 454 vs. Illumina).

I do describe strategies for read count normalization in the next
paragraph, but have added the "rarefaction curve" terminology.

> 6. On page 17, lines 54-55, please add some recent references on the
> current debate around the rooting of the placental tree. I don't think
> that most readers of Syst. Biol. will be familiar with this
> discussion.

Good point. I've put in the two 2013 MBE papers. If there is something
else I should cite let me know.

> 7. On page 18, lines 26-27, RAxML is also implemented in QIIME.

Added.

> 8. On page 20, I think that this might be the place to discuss the
> potential use of phylogenetic networks in microbiome research (see
> Parks and Beiko 2012 Mol Biol Evol).

Thank you for pointing me to this reference.

> 9. On page 24, when citing uBiome, it might also be fair to refer the
> similar crowd-funded American Gut Project.

Thanks.

> Typos:
> P3, L40: I would rather use "species delimitation" instead of "species
> definition".

Thanks.

> P4, L9: "These techniques and could be used"

Fixed.

> P6, L32: "we will be primarily be describing"

Whoops!

> P9, L29, L30: Maybe replace "diversity" by "variability"

Thanks.

> P18, L32: "a long history as a means to sequentially build"

Reworded, thanks.

> P19, L16: "GreenGenes 13_5 alignment has is almost 80% gaps"

Yikes, thanks.

### 2014-06-24 response/response.txt
Avoiding references to color.

@@
 I didn't know that! Thank you.
-
-All color costs can be avoided if you decide to have the figures printed in black and white, but be shown in color in the online version of the paper.
-If you do this, the captions must be worded such that they are appropriate for both situations (e.g., descriptions should not name colors), because the captions in print and online will be identical even if color is used online only.
-Do not submit two versions of any figure.
-Instead, make sure the color figure is also easily legible in grayscale, then submit only the color version.
-In your cover letter, please make your intentions regarding the use of color clear.
 

### 2014-06-24 sbreview.tex
Avoiding references to color.

@@
 \newcommand{\pdLegend}{
-    Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as blue circles and abundances are shown as the size of the circles.
+    Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as circles and abundances are shown as the size of the circles.
     Unweighted PD takes the sum of the branch lengths $\ell$ connecting samples.
@@
     The UniFrac divergence measure \citep[figure adapted from][]{LozuponeKnightUniFrac05}.
-    Assume that the sequence data to build the phylogenetic tree derives from two samples: the green sample and the blue sample.
+    Assume that the sequence data to build the phylogenetic tree derives from two samples: the light-shaded sample and the dark-shaded sample (green and blue in the online version).
     When the samples are interspersed across the tree (left tree), they have a smaller \textit{fraction} of branch length that sits ancestral to clades that are \textit{uniquely} composed of one sample or another, compared to when they are separate (right tree).
-    The bottom pictorial equation shows the ratios of interest for UniFrac: the branch length unique to one sample (green or blue, but not brown) divided by the total branch length.
+    The bottom pictorial equation shows the ratios of interest for UniFrac: the branch length unique to one sample divided by the total branch length.
     The ratio is smaller when the samples are interspersed (left) than they are when separate (right tree).

### 2014-06-24 response/response.txt
Addressing the last of the comments.

@@
 
+I have changed this to:
+
+"The (unweighted and unrooted) PD of a set of taxa $S$ is simply the total of the lengths of the branches which connect taxa in $S$ (Faith 1992)."
+
 > Fig 1 The "graphical formulae" [l(segments) A(segments)] at the bottom
@@
 
+I have removed the graphical formulae, and given an example abundance-weighted measure.
+
 > Fig 2 Your two cases are extreme. Having one (or a few) shared (brown)
@@
 > in the left tree would help. See also comments of referee 2.
+
+I have made it less extreme according to your suggestion, thanks.
 
@@
 > reads.
+
+I feel that by expanding the explanation here I would contribute to the un-evenness of coverage described by the other reviewers.
+The earth-movers distance is a standard way to compare probability distributions on metric spaces and has a Wikipedia page.
+I have added a classical and a recent book reference.
 
@@
 
+Well, I chose to discuss things at depth those subjects I thought would
+interest the phylogenetics community. I have removed discussion of one
+outlier as described above, but am open to hearing about others that
+should be removed.
+
 > I will say that, while figure 2 does a fairly good job of illustrating
@@
 
+Thank you for your comments-- the text has been updated (I discussed
+these changes via email with Arne directly).
+
 Reviewer: 3

### 2014-06-24 sbreview.tex
Addressing the last of the comments.

@@
     Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as circles and abundances are shown as the size of the circles.
-    Unweighted PD takes the sum of the branch lengths $\ell$ connecting samples.
-    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the areas $A$ of edges.
+    Unweighted PD takes the sum of the branch lengths connecting samples.
+    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the total area of the edges.
+    One such abundance-weighted measure simply takes the absolute value of the difference of the total taxon abundance on one side compared to the other.
 }
@@
 
-The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
+To calculate the UniFrac divergence between two samples of reads, one builds a tree on both samples and then performs a calculation on the branch lengths according whether branches have descendants in both samples or one sample only.
+Specifically, the original UniFrac divergence is the fraction of the total branch length of the tree coming from branches with descendant leaves in exactly one of the two samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
 Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
@@
 \citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
-The earth-movers distance between two probability distributions in this case can be defined using a physical analogy as the minimum amount of ``work'', defined as mass times distance, required to move probability mass in one distribution to another along the tree.
+The earth-movers distance \citep{monge1781memoire,Villani2003-wv} between two probability distributions in this case can be defined using a physical analogy as the minimum amount of ``work'', defined as mass times distance, required to move probability mass in one distribution to another along the tree.
 In this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
@@
     Part of a minimal mass movement to calculate the earth-mover's distance between two probability distributions on a phylogenetic tree.
-    For this, each probability distribution is considered as a configuration of dirt piles on the tree, and the distance between two such dirt pile configurations is defined to be the minimum amount of physical ``work'' required to move the dirt in one configuration to the other.
+    For this, each probability distribution is considered as a configuration of dirt piles (round bumps in the figure) on the tree, and the distance between two such dirt pile configurations is defined to be the minimum amount of physical ``work'' required to move the dirt in one configuration to the other.
 }

### 2014-06-24 sbreview.tex
Typo fix.

@@
 \citet{Wang2013-ol} optimize a criterion of cluster modularity.
-\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixture on 4-mer profiles, and then a minimum description length criterion.
+\citet{hao2011clustering} use a Gaussian mixture model formulation for clustering to avoid fixed cutoff values, and \citet{CWC12} use a two step process, first with a Dirichlet multinomial mixture on 3-mer profiles, and then a minimum description length criterion.
 \citet{Zhang2013-wh} have developed a phylogenetic means to do species delimitation that scales to a relatively large number of sequencing reads and so can be used as an OTU clustering method.

### 2014-06-27 sbreview.tex
Repagination for arxiv.
Fixes from reading diff.

@@
 The answer is simple: this system is data- and question-rich.
-It absolutely requires molecular methods because microbes are now primarily identified by their molecular sequences, and identification of microbes by their molecular sequences is much more straightforward to do in high throughput than morphological or phenotypic characterization.
+Microbes are now primarily identified by their molecular sequences because such molecular identification is much more straightforward to do in high throughput than morphological or phenotypic characterization.
+This motivates methods to organize and derive knowledge from those sequences via molecular evolution.
 Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.
@@
 \forarxiv{
-\begin{wrapfigure}[25]{r}{2.2in}
+\begin{wrapfigure}[26]{r}{2.2in}
   % \capstart
-  \vspace{-14pt}
+  \vspace{-25pt}
   \begin{center}
@@
 Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
-The (unweighted and unrooted) PD of a set of taxa $S$ is simply the total of the lengths of the branches which connect taxa in $S$ \citet{faith1992conservation}.
+The (unweighted) PD of a set of taxa $S$ in an unrooted phylogenetic tree is simply the total of the lengths of the branches that sit between taxa in $S$ \citep{faith1992conservation}.
 It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
@@
 \forarxiv{
-\begin{wrapfigure}[32]{r}{2.2in}
+\begin{wrapfigure}[33]{r}{2.2in}
   % \capstart
-  \vspace{-26pt}
+  \vspace{-4pt}
   \begin{center}
@@
 
+There have been several efforts to augment these ordination visualizations with additional information giving more structure to the visualizations.
+Biplots display variables (in the microbial case summarized by taxonomic labels) as points along with the points representing samples \citep[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
+\citet{PurdomAnalyzingDataGraphs08} describes how generalized principal component eigenvectors can be interpreted via weightings on the leaves of a phylogenetic tree.
+\citet{matsen2013edge} have developed a variant of principal components analysis that explicitly labels the axes with weightings on phylogenetic trees that indicate their influence.
+
 \newcommand{\dirtpilesLegend}{
@@
 \forarxiv{
-\begin{wrapfigure}[24]{l}{1.5in}
+\begin{wrapfigure}[25]{l}{2.in}
   % \capstart
@@
   \begin{center}
-    \includegraphics[width=1.2in]{figures/tree_dirtpiles.pdf}
+    \includegraphics[width=1.85in]{figures/tree_dirtpiles.pdf}
   \end{center}
@@
 }
-
-There have been several efforts to augment these ordination visualizations with additional information giving more structure to the visualizations.
-Biplots display variables (in the microbial case summarized by taxonomic labels) as points along with the points representing samples \citep[e.g.][]{hewitt2013bacterial,lozupone2013meta}.
-\citet{PurdomAnalyzingDataGraphs08} describes how generalized principal component eigenvectors can be interpreted via weightings on the leaves of a phylogenetic tree.
-\citet{matsen2013edge} have developed a variant of principal components analysis that explicitly labels the axes with weightings on phylogenetic trees that indicate their influence.
 

### 2014-06-28 sbreview.tex
Few more fixes.

@@
     Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as circles and abundances are shown as the size of the circles.
-    Unweighted PD takes the sum of the branch lengths connecting samples.
+    Unweighted PD takes the total length of branches sitting between tree tips represented in a sample.
     Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the total area of the edges.
-    One such abundance-weighted measure simply takes the absolute value of the difference of the total taxon abundance on one side compared to the other.
+    One such abundance-weighted measure simply takes the absolute value of the difference of the total read abundance on one side compared to the other.
 }
@@
 Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
-The (unweighted) PD of a set of taxa $S$ in an unrooted phylogenetic tree is simply the total of the lengths of the branches that sit between taxa in $S$ \citep{faith1992conservation}.
+The (unweighted) PD of a set of taxa $S$ in an unrooted phylogenetic tree is simply the total length of the branches that sit between taxa in $S$ \citep{faith1992conservation}.
 It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.

### 2014-06-28 sbreview.tex
quantify diversity of a sample
Deleting a problematic sentence.

@@
 Microbes are now primarily identified by their molecular sequences because such molecular identification is much more straightforward to do in high throughput than morphological or phenotypic characterization.
-This motivates methods to organize and derive knowledge from those sequences via molecular evolution.
 Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.
@@
 
-Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
+Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to quantify the diversity of a sample (Fig.~1).
 The (unweighted) PD of a set of taxa $S$ in an unrooted phylogenetic tree is simply the total length of the branches that sit between taxa in $S$ \citep{faith1992conservation}.

### 2014-07-18 sbreview.tex
Formatting.
Thank peer reviewers
Arne's version of the unifrac sentence.

@@
 \forarxiv{
-\begin{wrapfigure}[26]{r}{2.2in}
+\begin{wrapfigure}[27]{r}{2.2in}
   % \capstart
-  \vspace{-25pt}
+  % \vspace{-25pt}
   \begin{center}
@@
 
-To calculate the UniFrac divergence between two samples of reads, one builds a tree on both samples and then performs a calculation on the branch lengths according whether branches have descendants in both samples or one sample only.
+To calculate the UniFrac (``unique fraction'') divergence between two samples of reads, one builds a global tree based on both samples and then calculates the fraction of the total tree length found in only one (i.e.
+the fraction of the tree unique to one) of the two samples.
 Specifically, the original UniFrac divergence is the fraction of the total branch length of the tree coming from branches with descendant leaves in exactly one of the two samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
@@
 I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank Bastien Boussau, Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript.
+The manuscript was greatly improved by thoughtful peer review from Frank (Andy) Anderson, Olivier Gascuel, Alexis Stamatakis, Frédéric Delsuc, and an anonymous referee.
 This work was supported by National Institutes of health grant R01-HG005966-01 and National Science Foundation grant 1223057.

### 2014-07-18 sbreview.tex
Smoothing out the local sentences.

@@
 
-To calculate the UniFrac (``unique fraction'') divergence between two samples of reads, one builds a global tree based on both samples and then calculates the fraction of the total tree length found in only one (i.e.
-the fraction of the tree unique to one) of the two samples.
-Specifically, the original UniFrac divergence is the fraction of the total branch length of the tree coming from branches with descendant leaves in exactly one of the two samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
+To calculate the unweighted UniFrac (``unique fraction'') divergence between two samples of reads, one builds a global tree based on both samples and then calculates the fraction of the total tree length found in only one (i.e.
+the fraction of the tree unique to one) of the two samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
 Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
